---
title: "The Brighton Secondary School System"
subtitle: "A strategic view"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
abstract: |
  Brighton & Hove is deciding what to do about a secondary school system
  that is losing children faster than it is losing places. This document
  assembles what can be established about that system from published
  data alone: where the schools and the children are, how reachable the
  schools actually are by bus, how demand has moved over sixteen years,
  and what a model of the whole system says about the choices in front
  of the council. It is written to be argued with.
---

```{r setup}
#| include: false

source(here::here("R", "00_core.R"))

library(scales)
library(patchwork)

knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

# ---- Everything the document draws on --------------------------------
oi     <- bh_data("open_inputs.rds")
acc    <- bh_data("accessibility.rds")
dep    <- bh_data("deprivation_open.rds")
brt    <- bh_data("brightopia.rds")
env    <- bh_data("sensitivity_envelope.rds")
perf   <- bh_data("performance_panel.rds")
lsoa   <- bh_data("lsoa.geojson")
catch  <- bh_data("catchments_current.geojson") %>% st_transform(4326)

# Loaded here as well as in the sections that use them, so the executive
# summary can quote figures that are computed further down the document.
rc     <- bh_data("reception_cohort.rds")
os     <- bh_data("open_scenarios.rds")
conv   <- bh_data("adjudicator_conversion.rds")
fp     <- bh_data("factsheet_panel.rds")

fmt_n   <- function(x, d = 0) formatC(round(x, d), big.mark = ",", format = "f", digits = d)
fmt_pct <- function(x, d = 0) paste0(formatC(x, format = "f", digits = d), "%")
```

```{r exec-figures}
#| include: false

# Headline figures for the executive summary. Every one of these is
# recomputed in the section it belongs to; they are gathered here only
# so the summary can quote them before those sections run.

ex_now      <- rc$secondary_city %>% slice_max(year, n = 1)
ex_peak     <- rc$secondary_city %>% slice_max(y7_offers, n = 1)
ex_proj     <- rc$projection %>% slice_max(entry_year, n = 1)
ex_fall     <- 100 * (ex_proj$central - ex_now$y7_offers) / ex_now$y7_offers
ex_places   <- sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"])
ex_surplus  <- ex_places - ex_proj$central

ex_zero30   <- acc$lsoa %>% filter(places_30 < 1)
ex_corner   <- acc$worst_corner

ex_rat <- conv$conv %>%
  group_by(school) %>%
  summarise(rate = sum(p1_offered) / sum(p1_named), .groups = "drop")
ex_open <- sum(ex_rat$rate > 0.999)

ex_r2_att8 <- os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"]
ex_r2_va   <- os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"]

# Same specification as section 2.3, which follows the Lever report:
# log outcome, logged rates, KS2 score entered linearly.
ex_nat <- perf$national %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  mutate(ks2_c = KS2ASS - 100)
ex_fsm   <- summary(lm(log(ATT8SCR) ~ log(PTFSM6CLA1A), data = ex_nat))$r.squared
ex_prior <- summary(lm(log(ATT8SCR) ~ ks2_c, data = ex_nat))$r.squared
ex_absn  <- summary(lm(log(ATT8SCR) ~ log(PERCTOT), data = ex_nat))$r.squared

ex_lh <- brt$at_original_beta %>%
  filter(str_detect(name, "Longhill")) %>%
  mutate(fill_2026 = modelled / pan2026, fill_2024 = modelled / pan2024)

# The between-school share of *pupil-level* variance. This cannot be
# computed from school-level aggregates - it needs pupil records - so it
# is quoted from the school-effectiveness literature, not estimated
# here. Defined at the top because both the executive summary and
# section 2.4 use it.
SCHOOL_LO <- 8; SCHOOL_HI <- 15

# Variance decomposition from the Lever model, for the summary. Section
# 2.4 recomputes these from the same object.
ex_sed <- bh_data("school_effect_decomp.rds")
.pick <- function(d, pat) d[["Share of total %"]][grepl(pat, d$Component)][1]
ex_wf_all <- .pick(ex_sed$decomp_all, "^Endogenous")
ex_ex_all <- .pick(ex_sed$decomp_all, "^Exogenous")
ex_wf_dis <- .pick(ex_sed$decomp_dis, "^Endogenous")

# The share a school can actually reach: workforce plus the
# school-controllable half of absence. Section 2.5 derives this.
ex_lev        <- bh_data("school_leverage.rds")$leverage
ex_reach      <- ex_lev$reachable[ex_lev$who == "All pupils"]
ex_abs_school <- ex_lev$absence_school[ex_lev$who == "All pupils"]
# Recomputed from the fitted model rather than the cached table, so that
# workforce + school-reachable absence actually sums to the reachable
# total quoted alongside it. It differs from the cached 3.3% by under a
# percentage point; section 2.5 explains why.
ex_wf_recomp  <- ex_lev$workforce[ex_lev$who == "All pupils"]
```

::: {.callout-note appearance="simple"}
## Everything here is built from published data

No pupil-level record is used anywhere in this document. Every figure
comes from data the council, the DfE or the ONS has already published,
or from a model built on top of it. Where a question can only be
answered with data the council holds but has not released, that is said
plainly rather than glossed over --- and section 9 lists what those
releases would be.
:::

# Executive summary {#sec-exec}

## The situation

Brighton & Hove has a secondary school system built for more children
than it now has, and the shortfall is going to get worse before anything
the council decides can affect it. Year 7 offers peaked at
`r fmt_n(ex_peak$y7_offers)` in `r ex_peak$year` and stood at
`r fmt_n(ex_now$y7_offers)` in `r ex_now$year`. The children who will
enter Year 7 in `r ex_proj$entry_year` are already in the city's primary
schools, and there are about **`r fmt_n(ex_proj$central)`** of them ---
a further fall of `r fmt_pct(abs(ex_fall), 0)`. Against roughly
`r fmt_n(ex_places)` published places, that implies something like
**`r fmt_n(ex_surplus)` surplus places**, or two and a half average
secondary schools' worth.

The council's own catchment forecasts sit above that projection in
almost every year, and the gap is widest in the Longhill catchment ---
the place where the consequences of being wrong are largest.

## What the evidence says

**This is not a school quality problem.** Across English secondaries,
absence explains `r fmt_pct(100 * ex_absn, 0)` of the variation in
Attainment 8 and prior attainment `r fmt_pct(100 * ex_prior, 0)`, while
deprivation explains only `r fmt_pct(100 * ex_fsm, 0)`. Once intake is
accounted for, Brighton's schools mostly perform above what their intake
predicts. The largest lever available to the city is **absence**, not
social mixing --- established in detail in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).

**Moving children between schools is a weak instrument, and this is the
single most important point for admissions policy.** Decomposing the
variance in Attainment 8 shows that a school can reach about
**`r fmt_pct(ex_reach, 0)`** of what separates its results from another's
--- consistent with sixty years of school-effectiveness research. The
rest arrives with the children. Crucially, **almost all of that reachable
share runs through attendance**: the workforce a school hires accounts
for only `r fmt_pct(ex_wf_recomp, 1)`, while the school-controllable half
of absence accounts for `r fmt_pct(ex_abs_school, 0)`. Redistribution touches
neither. It changes which building a child attends without changing their
attendance or their circumstances. This is not an argument that schools
do not matter --- it is an argument that the council's lever is
attendance, and attendance is not an admissions policy.

**Admissions criteria only bind at some schools.**
`r ex_open` of the `r nrow(ex_rat)` schools offered a place to every
first-preference applicant. For those schools the catchment rule, the
sibling rule and the free school meals quota do nothing at all. The city
runs two admissions systems under one set of rules, and public argument
about criteria concerns only one of them.

**Families are choosing on the wrong number.** Headline Attainment 8
explains `r fmt_pct(100 * ex_r2_att8, 0)` of the variation in how
heavily each school is preferred. The value-added measure --- what the
school actually contributes --- explains
`r fmt_pct(100 * ex_r2_va, 0)`. Because the headline score is largely a
description of the existing intake, choice becomes self-fulfilling.

**Physical access is unequal, and unequal in the worst direction.**
`r nrow(ex_zero30)` of `r nrow(acc$lsoa)` neighbourhoods --- about
`r fmt_n(sum(ex_zero30$Oi))` cohort-aged children --- can reach **no**
secondary school place within 30 minutes by walking and bus.
Accessibility and child poverty correlate at
`r sprintf("%.2f", acc$acc_dep_spearman)`: `r nrow(ex_corner)`
neighbourhoods sit in both the least-reachable third and the most
deprived third.

**Longhill has already been shrunk to about what geography supports.**
A model of where children would go if only distance and school size
mattered gives it `r fmt_n(ex_lh$modelled)` children ---
`r fmt_pct(100 * ex_lh$fill_2024, 0)` of its former admission number of
`r fmt_n(ex_lh$pan2024)`, but `r fmt_pct(100 * ex_lh$fill_2026, 0)` of
the reduced `r fmt_n(ex_lh$pan2026)` now in force. The reduction has
already done most of what a reduction can do, and the remaining problem
is a falling cohort against a reserve that supports about
`r sprintf("%.1f", os$years_left)` more years.

## What follows

1. **Decide about Longhill on the financial timetable**, not the
   consultation one. The horizon is about
   `r sprintf("%.1f", os$years_left)` years.
2. **Treat the instruments as a package.** Moving the school, changing
   its admission number and redrawing catchments each fall short alone
   and compose when combined. Picking one and waiting is the worst
   available option.
3. **Publish a contextualised measure alongside Attainment 8.** It costs
   nothing and begins to unwind a loop the authority currently sustains
   through its own admissions guide.
4. **Treat the bus network as an admissions instrument.** It is the only
   lever here that widens choice without taking anything from anyone.
5. **Release six tables** (§9) that would convert most of the bands in
   this document into estimates.

## How confident to be

The demographic findings are the firmest: they are counts of children
already in school, adjusted by a transfer rate that has varied by under
two percentage points in a decade. The accessibility findings are firm
in direction and approximate in magnitude --- modelled journey times,
not measured ones. The model results in sections 7 and 8 are the
softest: the model is swept rather than calibrated, because calibration
needs data the council has not released. Every comparison there is
robust across the swept range; no individual number should be read as a
prediction.

# The secondary school system {#sec-system}

Ten secondary schools serve Brighton & Hove: six community schools run
by the local authority, two academies (Brighton Aldridge and Portslade
Aldridge) and two church schools (King's and Cardinal Newman) that set
their own admissions. An eleventh, Peacehaven Community School, sits
just outside the boundary but inside the travel-to-school geography, and
is included throughout because families on the eastern edge of the city
treat it as a real option.

The city allocates places by **catchment area with a lottery tie-break**
for oversubscribed schools --- an arrangement that is very unusual in
England, where distance from the school gate usually decides priority.
It dates from the closure of CoMArt in the east of the city in 2005, and
the catchment system that replaced distance-based allocation in 2008.
CoMArt is marked on the map below because the shape of the system is
still partly the shape of that closure.

## Where the schools, the catchments and the children are {#sec-orientation}

```{r fig-orientation}
#| fig-cap: "Schools, catchment boundaries and the cohort-aged child population. Use the control at the top right to turn each layer on and off."

# ---- Schools ---------------------------------------------------------
sch <- schools_sf(oi$schools)

sch_popup <- sprintf(
  "<b>%s</b><br>Admission number 2024: %s<br>Planned 2030: %s%s",
  sch$name, sch$pan2024, sch$pan2030,
  ifelse(is.na(sch$catchment),
         "<br><i>Faith school, admits city-wide</i>",
         paste0("<br>Catchment: ", CATCH_LABELS[sch$boundary_catchment])))

# ---- Children per LSOA ----------------------------------------------
# Oi is the cohort-aged child estimate per zone; zones are LSOA x
# catchment, so an LSOA split by a boundary has to be summed back up.
children <- oi$zones %>%
  group_by(lsoa) %>%
  summarise(children = sum(Oi, na.rm = TRUE), .groups = "drop")

lsoa_pts <- lsoa %>%
  left_join(children, by = c("lsoa21cd" = "lsoa")) %>%
  filter(!is.na(children), children > 0) %>%
  st_point_on_surface() %>%
  suppressWarnings()
pt_xy <- st_coordinates(lsoa_pts)
lsoa_pts$lon <- pt_xy[, 1]; lsoa_pts$lat <- pt_xy[, 2]

# Radius on the square root of the count, so circle *area* is
# proportional to the number of children rather than the radius being -
# the latter exaggerates big areas by a factor of the count.
r_scale <- function(n) 3 + 17 * sqrt(n / max(n, na.rm = TRUE))

pal_catch <- colorFactor(unname(CATCH_COLOURS), names(CATCH_COLOURS))

m <- leaflet(width = "100%", height = 620) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -0.10, lat = 50.84, zoom = 12) %>%
  addPolygons(
    data = catch, group = "Catchments",
    fillColor = ~pal_catch(catchment), fillOpacity = 0.13,
    color = ~pal_catch(catchment), weight = 2.5, opacity = 0.85,
    label = ~unname(CATCH_LABELS[catchment]),
    highlightOptions = highlightOptions(weight = 4, fillOpacity = 0.25,
                                        bringToFront = FALSE)) %>%
  addCircleMarkers(
    data = lsoa_pts, lng = ~lon, lat = ~lat, group = "Children",
    radius = ~r_scale(children),
    fillColor = "#1f4e79", fillOpacity = 0.35,
    color = "#1f4e79", weight = 1, opacity = 0.55,
    label = ~sprintf("%s: about %s children in the cohort",
                     lsoa21nm, round(children)))

m <- add_school_layer(m, sch, group = "Schools", popup = sch_popup)

# CoMArt, closed 2005
m <- m %>% addCircleMarkers(
  lng = COMART$lon, lat = COMART$lat, group = "CoMArt (closed 2005)",
  radius = 6, fillColor = "#b2182b", fillOpacity = 0.9,
  color = "white", weight = 2,
  popup = "<b>CoMArt</b><br>Closed 2005. Its closure is why the city moved to catchments with a lottery tie-break.",
  label = COMART$name)
ic <- logo_icon(COMART$logo)
if (!is.null(ic))
  m <- m %>% addMarkers(lng = COMART$lon, lat = COMART$lat, icon = ic,
                        group = "CoMArt (closed 2005)",
                        label = COMART$name)

legend_html <- paste0(
  '<div style="background:white;padding:8px 12px;border-radius:4px;',
  'box-shadow:0 1px 5px rgba(0,0,0,.3);font-family:sans-serif;font-size:12px;line-height:1.6">',
  '<div style="font-weight:bold;margin-bottom:4px">Catchment areas, 2025/26</div>',
  paste0('<div style="display:flex;align-items:center;gap:6px">',
         '<span style="display:inline-block;width:14px;height:14px;background:',
         unname(CATCH_COLOURS), ';border-radius:2px;flex-shrink:0"></span><span>',
         unname(CATCH_LABELS), '</span></div>', collapse = ''),
  '<div style="margin-top:6px;padding-top:6px;border-top:1px solid #ddd">',
  '<span style="display:inline-block;width:14px;height:14px;background:#1f4e79;',
  'opacity:.35;border-radius:50%;margin-right:6px"></span>Children per LSOA',
  ' <span style="color:#777">(area &prop; count)</span></div></div>')

m %>%
  addLayersControl(
    overlayGroups = c("Catchments", "Schools", "Children", "CoMArt (closed 2005)"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("CoMArt (closed 2005)") %>%
  addControl(html = legend_html, position = "bottomright")
```

Three things are worth noticing on that map before any analysis begins.

**The catchments are not equivalent objects.** Four of them contain a
single school; two contain a pair. A child in the Varndean / Dorothy
Stringer or Hove Park / Blatchington Mill catchment has two schools to
name and a realistic chance of one of them. A child in the Longhill,
Patcham, BACA or PACA catchment has one. That asymmetry does more work
in this system than almost anything else, and section 5 returns to it.

**The children are not spread evenly.** The cohort is concentrated in
the centre and west; the east of the city, which contains Longhill's
entire catchment, is thin. Longhill's problem is visible on this map
before any model is run.

**Two schools sit outside the geography entirely.** King's and Cardinal
Newman admit city-wide on faith criteria. Together they hold
`r fmt_n(sum(oi$schools$pan2026[is.na(oi$schools$catchment)]))` places,
about
`r fmt_pct(100 * sum(oi$schools$pan2026[is.na(oi$schools$catchment)]) / sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]))`
of the city total. Any statement about what catchments do has to be read
against that.

## Where the children actually are {#sec-child-density}

The circles on the map above are LSOA totals, and an LSOA is a blunt
unit --- about 1,500 people, drawn for administrative convenience rather
than to describe a neighbourhood. Two finer views of the same population
follow. They are shown side by side because they answer slightly
different questions, and it is worth deciding which one this document
should carry.

```{r pcd-data}
#| include: false

pcd <- bh_data("postcode_children.csv") %>%
  filter(!is.na(easting), !is.na(northing), children > 0)

# Land mask: the union of the city's LSOAs. Without it a smoothed
# surface runs cheerfully out into the Channel.
city <- lsoa %>%
  filter(lsoa21cd %in% unique(pcd$lsoa)) %>%
  st_transform(27700) %>%
  st_union() %>%
  st_make_valid()
```

::: {.panel-tabset}

### Density surface

```{r fig-child-density}
#| fig-cap: "Children aged 0-18 per hectare, smoothed from postcode-level estimates with a 400 m kernel and clipped to the built-up area. Census 2021 household composition apportioned to postcode centroids."

BW <- 400          # kernel bandwidth in metres - a walkable neighbourhood
GRID <- 320        # grid resolution

# A weighted kernel density: each child contributes one point, so the
# surface is the density of children rather than of postcodes. Counts are
# apportioned estimates and so fractional; rounding for replication
# changes the total by well under a percent.
reps <- round(pcd$children)
kx <- rep(pcd$easting,  reps)
ky <- rep(pcd$northing, reps)

bb <- st_bbox(city)
k <- MASS::kde2d(kx, ky, h = c(BW, BW), n = GRID,
                 lims = c(bb["xmin"] - 500, bb["xmax"] + 500,
                          bb["ymin"] - 500, bb["ymax"] + 500))

# kde2d integrates to 1 over square metres. Scale to children per
# hectare so the legend means something to a reader.
dens <- k$z * sum(reps) * 1e4

brks <- c(0, 2, 5, 10, 15, 22, 30, max(45, ceiling(max(dens))))
pal_d <- viridisLite::viridis(length(brks) - 1, option = "magma",
                              direction = -1, begin = 0.08)

# isoband wants z indexed [y, x]; kde2d returns [x, y].
bands <- isoband::isobands(x = k$x, y = k$y, z = t(dens),
                           levels_low = brks[-length(brks)],
                           levels_high = brks[-1])

band_sf <- st_sf(
  lo = brks[-length(brks)], hi = brks[-1],
  geometry = st_sfc(isoband::iso_to_sfg(bands), crs = 27700)) %>%
  st_make_valid() %>%
  st_intersection(city) %>%
  filter(!st_is_empty(geometry)) %>%
  # Simplify while still in metres. Contour bands off a 320 x 320 grid
  # carry far more vertices than a screen can show; 15 m is invisible at
  # any zoom this map offers and cuts the embedded geometry sharply.
  # This has to happen before the transform - in 4326 the tolerance
  # would be read as degrees and flatten the whole surface.
  st_simplify(dTolerance = 15, preserveTopology = TRUE) %>%
  st_transform(4326)

band_sf$col <- pal_d[seq_len(nrow(band_sf))]
band_sf$lab <- sprintf("%g - %g children per hectare", band_sf$lo, band_sf$hi)

leaflet(width = "100%", height = 600,
        options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = band_sf, fillColor = ~col, fillOpacity = 0.72,
              color = NA, weight = 0, label = ~lab, group = "Children") %>%
  addPolygons(data = catch, fill = FALSE, color = "#333333",
              weight = 1.6, opacity = 0.7, group = "Catchments",
              label = ~unname(CATCH_LABELS[catchment])) %>%
  add_school_layer(sch, group = "Schools", popup = sch_popup) %>%
  addLayersControl(overlayGroups = c("Children", "Catchments", "Schools"),
                   options = layersControlOptions(collapsed = TRUE)) %>%
  addLegend(colors = band_sf$col, labels = band_sf$lab,
            title = "Children 0-18<br>per hectare",
            position = "bottomright", opacity = 0.8)
```

### Postcode detail

```{r fig-pcd-dots}
#| fig-cap: "Every residential postcode in the city, sized by the number of children aged 0-18 and coloured by income deprivation affecting children. Decile 1 is the most deprived tenth of neighbourhoods nationally."

pcd_sf <- pcd %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>%
  st_transform(4326) %>%
  filter(!is.na(idaci_decile))

pal_idaci <- colorFactor("RdYlBu", domain = sort(unique(pcd_sf$idaci_decile)))

# Radius on the square root of the count so circle area is proportional
# to children. Scaled well down from the LSOA circles in 2.1: there are
# 4,229 postcodes against 179 LSOAs, so the same per-circle size would
# cover the city in a single wash of ink. The median postcode draws at
# about 3.5 px and the largest at about 9.
#
# preferCanvas matters here rather than being a nicety. Leaflet's
# default renderer gives every marker its own SVG element, and four
# thousand of them is enough to hang or crash a browser tab. On canvas
# they are drawn into a single element and the map stays responsive.
leaflet(width = "100%", height = 600,
        options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = catch, fill = FALSE, color = "#333333",
              weight = 1.6, opacity = 0.7, group = "Catchments",
              label = ~unname(CATCH_LABELS[catchment])) %>%
  addCircleMarkers(
    data = pcd_sf, group = "Postcodes",
    radius = ~0.5 + sqrt(children),
    fillColor = ~pal_idaci(idaci_decile), fillOpacity = 0.6,
    stroke = FALSE,
    label = ~sprintf("%s | %.0f children 0-18 | IDACI decile %s",
                     postcode, children, idaci_decile)) %>%
  add_school_layer(sch, group = "Schools", popup = sch_popup) %>%
  addLayersControl(overlayGroups = c("Postcodes", "Catchments", "Schools"),
                   options = layersControlOptions(collapsed = TRUE)) %>%
  addLegend(pal = pal_idaci, values = sort(unique(pcd_sf$idaci_decile)),
            title = "IDACI decile<br><span style='font-weight:normal'>1 = most deprived</span>",
            position = "bottomright", opacity = 0.9)
```

:::

```{r density-check}
#| include: false
# Sanity check on the surface. The comparison has to be like for like:
# the surface is children per hectare, so testing it against the raw
# count in each postcode mostly measures how much postcodes differ in
# size (that correlation is only ~0.36 and means little). Compare it
# instead with the children within 500 m of each postcode, which is the
# same quantity the surface is estimating. A transposed grid would show
# up here as a collapse towards the swapped-index value.
ix <- findInterval(pcd$easting,  k$x)
iy <- findInterval(pcd$northing, k$y)
ok <- ix > 0 & ix <= length(k$x) & iy > 0 & iy <= length(k$y)
sampled <- dens[cbind(ix[ok], iy[ok])]

xy <- as.matrix(pcd[ok, c("easting", "northing")])
ch <- pcd$children[ok]
local500 <- vapply(seq_len(nrow(xy)), function(i)
  sum(ch[(xy[, 1] - xy[i, 1])^2 + (xy[, 2] - xy[i, 2])^2 <= 500^2]),
  numeric(1))

dens_cor  <- cor(sampled, local500, method = "spearman")
dens_swap <- suppressWarnings(
  cor(dens[cbind(iy[ok], ix[ok])], local500, method = "spearman"))
stopifnot(dens_cor > 0.6, dens_cor > dens_swap + 0.2)

n_pcd  <- nrow(pcd)
```

The surface and the dots are built from the same
`r fmt_n(n_pcd)` postcodes and the same
`r fmt_n(sum(pcd$children))` children, and they show the same city ---
the smoothed surface tracks the number of children within 500 metres of
each postcode at a rank correlation of
`r sprintf("%.2f", dens_cor)`. What differs is what each one makes easy
to see.

The **surface** shows where children are concentrated: a dense central
and western band, thinning sharply east of the marina. It answers "where
is the demand?" and it does not pretend to precision it lacks.

The **postcode map** shows the same concentration but adds the social
gradient, and it exposes something the surface cannot --- that
deprivation in this city is fine-grained. Deprived and comfortable
postcodes sit within a few hundred metres of each other, particularly in
the centre. That texture matters later: section 4 shows that poor
transport access and child poverty coincide, and section 5 that
catchment boundaries drawn at this scale will always cut through mixed
ground rather than separating rich areas from poor ones.

::: {.callout-note appearance="simple"}
## Two limits on both maps

These are **Census 2021 output-area household counts apportioned to
postcode centroids**, not a register of children. They are estimates,
and at postcode level they are noisy estimates --- reliable in aggregate,
not for any individual postcode.

They also cover **Brighton & Hove only**. The Peacehaven and Telscombe
area, which contributes about `r fmt_n(sum(oi$zones$Oi[oi$zones$area == "Expansion area"]))`
cohort-aged children to the study area, is absent from both. The
LSOA-level figures used everywhere else in this document do include it.
:::

## What the attainment data actually says {#sec-lever}

The debate about Brighton's schools is usually conducted in the currency
of headline attainment --- Attainment 8, league table position, "good"
and "bad" schools. That currency is close to worthless for the question
the council actually faces, and it is worth being precise about why,
because the usual explanation is also wrong.

```{r lever-models}
#| include: false

# These follow the specification used in "How to Pull the Right Lever"
# and the RPE paper:
#
#   log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL)
#                  + ks2_c + ...
#
# Rates are logged - deprivation, absence, EAL - so their coefficients
# are elasticities. The prior-attainment term is a mean KS2 score
# centred at 100 and entered linearly, because it is a score rather
# than a rate: logging it is neither interpretable nor better fitting.
# That distinction matters. An earlier draft of this section logged a
# *proportion* with low prior attainment instead, which fitted worse and
# silently dropped the 132 highest-attaining intakes in the country,
# since a log cannot take their zero.
#
# Absence here is PERCTOT, the overall absence rate, again matching the
# house model rather than the persistent-absence measure.

# Every model in this section is fitted on the same rows. lm() drops
# incomplete cases silently, so without this the figures would not be
# comparable with each other or with the decomposition below - and the
# difference is not cosmetic. Deprivation's R2 alone is 0.21 across all
# schools reporting it, and 0.37 across the schools that also report a
# KS2 score, because the ones missing a KS2 score are not a random
# sample of schools.
nat <- perf$national %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  mutate(ks2_c = KS2ASS - 100) %>%
  filter(!is.na(ks2_c), !is.na(PTFSM6CLA1A), !is.na(PERCTOT),
         PTFSM6CLA1A > 0, PERCTOT > 0)

m_fsm   <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A), data = nat)
m_prior <- lm(log(ATT8SCR) ~ ks2_c, data = nat)
m_abs1  <- lm(log(ATT8SCR) ~ log(PERCTOT), data = nat)
m_both  <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT), data = nat)
m_full  <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + ks2_c,
              data = nat)

R2 <- function(m) 100 * summary(m)$r.squared
cf <- coef(m_full)

bh_latest <- perf$bh %>%
  filter(year_label == perf$latest, !is.na(ATT8SCR), !is.na(PTFSM6CLA1A),
         !is.na(PERCTOT), !is.na(KS2ASS)) %>%
  mutate(ks2_c = KS2ASS - 100)
bh_resid <- bh_latest %>%
  mutate(pred = exp(predict(m_full, newdata = .)), resid = ATT8SCR - pred)
above <- sum(bh_resid$resid > 0, na.rm = TRUE)
```

The common account is that attainment tracks deprivation. It does, but
weakly, and stopping there points the city at the wrong lever.

The specification below follows the model in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html):
Attainment 8 and the rate variables are logged, so their coefficients
are elasticities, while prior attainment enters as a mean KS2 score
because it is a score and not a rate.

Across `r fmt_n(nobs(m_fsm))` mainstream secondaries in England, the
proportion of disadvantaged pupils explains **`r fmt_pct(R2(m_fsm))`**
of the variation in Attainment 8. Prior attainment explains
**`r fmt_pct(R2(m_prior))`**. And the single strongest of the three is
neither: **absence alone explains `r fmt_pct(R2(m_abs1))`**. All three
together reach `r fmt_pct(R2(m_full))`.

The elasticities make the point more sharply than the R² values do. In
the combined model a one per cent rise in absence is associated with a
**`r sprintf("%.2f", abs(cf[["log(PERCTOT)"]]))` per cent** fall in
Attainment 8; a one per cent rise in the disadvantaged share with a fall
of just **`r sprintf("%.2f", abs(cf[["log(PTFSM6CLA1A)"]]))` per cent**
--- roughly
`r sprintf("%.0f", abs(cf[["log(PERCTOT)"]]) / abs(cf[["log(PTFSM6CLA1A)"]]))`
times smaller.

That ordering is the finding. Who is poor matters least of the three.
What children could already do at eleven matters more. Whether they are
in the room matters most.

```{r fig-att8-prior}
#| fig-cap: "Attainment 8 against two features of intake, every mainstream secondary in England in the latest year, with Brighton & Hove schools highlighted. Prior attainment and absence both order schools far more sharply than deprivation does."
#| fig-height: 5

# Attainment 8 is on a log axis in every panel, matching the model. The
# x axis is logged for the two rates and left linear for the KS2 score,
# so each panel is drawn on the scale its term is actually fitted on.
panel <- function(xvar, xlab, ttl, logx) {
  d <- nat %>% filter(!is.na(.data[[xvar]]))
  b <- bh_latest %>% filter(!is.na(.data[[xvar]]))
  if (logx) { d <- d %>% filter(.data[[xvar]] > 0); b <- b %>% filter(.data[[xvar]] > 0) }
  p <- ggplot(d, aes(.data[[xvar]], ATT8SCR)) +
    geom_point(colour = "grey80", size = 0.9, alpha = 0.5) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
                colour = "grey30", linewidth = 0.7, linetype = "22") +
    geom_point(data = b, colour = "#b2182b", size = 2.4) +
    ggrepel::geom_text_repel(
      data = b %>% filter(SCHNAME %in% c("Longhill High School", "King's School")),
      aes(label = SCHNAME), size = 2.9, colour = "#7f0f22",
      min.segment.length = 0, segment.size = 0.3, seed = 1) +
    scale_y_log10(breaks = c(10, 20, 30, 40, 50, 60, 70)) +
    labs(x = xlab, y = "Attainment 8", subtitle = ttl) +
    theme_bh(11)
  if (logx)
    p <- p + scale_x_log10(labels = label_percent(scale = 1),
                           breaks = c(1, 2, 5, 10, 20, 50))
  p
}

(panel("PTFSM6CLA1A", "Disadvantaged pupils",
       sprintf("Deprivation, logged (R² = %.2f)", summary(m_fsm)$r.squared), TRUE) |
 panel("ks2_c", "Mean KS2 prior attainment, centred",
       sprintf("Prior attainment, linear (R² = %.2f)", summary(m_prior)$r.squared), FALSE) |
 panel("PERCTOT", "Overall absence rate",
       sprintf("Absence, logged (R² = %.2f)", summary(m_abs1)$r.squared), TRUE)) +
  plot_annotation(
    title = "Attainment is a description of intake, and of who turns up",
    subtitle = sprintf("Mainstream secondaries in England, %s. Attainment 8 log scaled throughout; rates logged, the KS2 score linear. Red points are Brighton & Hove.",
                       perf$latest),
    caption = "Source: DfE performance tables. Specification follows How to Pull the Right Lever.",
    theme = theme_bh())
```

Two consequences follow, and they point in the same direction.

**A headline score is mostly a description of who walks through the
door.** Once prior attainment and deprivation are accounted for,
**`r above` of the `r nrow(bh_resid)`** Brighton & Hove schools score
above what the national relationship predicts for their intake. On the
measure that isolates the school's own contribution, this is not a city
with a school quality problem.

**The lever that moves is attendance, not social mix.** Absence and
deprivation travel together, and absence is the mechanism through which
much of the deprivation effect actually operates --- which is why
redistributing disadvantaged children between schools is a far weaker
instrument than it is usually assumed to be.

::: {.callout-important appearance="simple"}
## Two cautions

**Absence is partly an outcome as well as a cause.** A child who is
disengaged attends less *and* attains less, so a single-equation fit
overstates how much of that `r fmt_pct(R2(m_full))` is absence doing the
causal work. The Lever report models this properly --- with school, year,
region and inspection-rating effects, and a first stage that models
absence itself --- rather than reading it off a scatterplot. These three
panels are an illustration of that model's premise, not a substitute for
it.

**These are single-year, single-equation fits.** The full model is
fitted on a multi-year panel with random effects and a wider set of
school characteristics, including English as an additional language,
teacher retention and admissions policy. The ordering of the three terms
here matches it; the individual R² values will not, and none of them
should be quoted as a causal effect.
:::

::: {.callout-tip}
## The detail behind this is published separately

The contextualised analysis --- value-added modelling across every local
authority in England, and what it says about where Brighton's schools
really sit --- is set out in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).
Its central findings are that Brighton's schools perform well once
intake is accounted for, and that **absence, not social mixing, is the
largest single lever available to the city**. This document takes those
findings as given and turns to the spatial and demographic questions the
Lever report does not address.
:::

```{r fig-ranking-flip}
#| fig-cap: "The same ten schools ranked two ways: by headline Attainment 8, and by the value-added measure from the Lever report. Lines crossing means the two measures disagree about which schools are doing well."
#| fig-height: 5.5

flip <- oi$attract %>%
  filter(!is.na(att8), !is.na(va_lever), name != "Peacehaven Community School") %>%
  mutate(rank_att8 = rank(-att8), rank_va = rank(-va_lever)) %>%
  select(name, rank_att8, rank_va) %>%
  pivot_longer(starts_with("rank_"), names_to = "measure", values_to = "rank") %>%
  mutate(measure = factor(measure, c("rank_att8", "rank_va"),
                          c("Attainment 8\n(headline)", "Value added\n(Lever report)")))

moved <- flip %>%
  pivot_wider(names_from = measure, values_from = rank) %>%
  rename(att8 = 2, va = 3) %>%
  mutate(shift = att8 - va)

ggplot(flip, aes(measure, rank, group = name)) +
  geom_line(aes(colour = name), linewidth = 1.1, alpha = 0.85, show.legend = FALSE) +
  geom_point(aes(colour = name), size = 3, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = flip %>% filter(measure == levels(flip$measure)[1]),
    aes(label = name), hjust = 1, nudge_x = -0.08, size = 3,
    direction = "y", segment.size = 0.2, seed = 1) +
  ggrepel::geom_text_repel(
    data = flip %>% filter(measure == levels(flip$measure)[2]),
    aes(label = name), hjust = 0, nudge_x = 0.08, size = 3,
    direction = "y", segment.size = 0.2, seed = 1) +
  scale_y_reverse(breaks = 1:10) +
  scale_x_discrete(expand = expansion(add = c(0.85, 0.85))) +
  labs(x = NULL, y = "Rank (1 = highest)",
       title = "Which schools look good depends on which measure you use",
       subtitle = "Ranked by headline attainment, and by contribution to progress once intake is accounted for",
       caption = "Sources: DfE performance tables; How to Pull the Right Lever.") +
  theme_bh() +
  theme(panel.grid.major.x = element_blank())
```

```{r flip-stats}
#| include: false
# Ties are real here - two schools fall by the same four places - so
# collapse to a list rather than letting slice_min pick one silently.
and_list <- function(x) if (length(x) < 2) x else
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))

up   <- moved %>% filter(shift == max(shift))
down <- moved %>% filter(shift == min(shift))
n_moved3 <- sum(abs(moved$shift) >= 3)
```

The lines cross a great deal. **`r and_list(up$name)`** rises
`r abs(up$shift[1])` places when you move from the headline measure to
the value-added one; **`r and_list(down$name)`** falls
`r abs(down$shift[1])`. `r n_moved3` of the ten schools move by three
places or more.

The schools that fall are not bad schools, and the schools that rise are
not suddenly good ones. The two measures are answering different
questions: one asks what pupils at this school achieved, the other asks
how much of that the school is responsible for. Families reading the
published headline are answering the first question while believing they
are answering the second --- and section 5 shows that this has a
measurable effect on where demand actually goes.

## How much can a school change at all? {#sec-school-effects}

There is a prior question underneath all of this, and it is the one that
bears most directly on what the council can achieve by moving children
between schools.

```{r variance-decomp}
#| include: false

# Shapley (LMG) decomposition, following RPE_Paper.qmd. A t-value in a
# multiple regression describes each predictor's *unique* contribution,
# so where predictors are as correlated as these are it cannot rank
# them. The Shapley share instead averages each variable's incremental
# R2 across every order in which the three could enter the model, which
# is the standard way of apportioning variance that predictors share.

r2_of <- function(vs) {
  rhs <- if (!length(vs)) "1" else paste(vs, collapse = " + ")
  summary(lm(as.formula(paste("log(ATT8SCR) ~", rhs)), data = nat))$r.squared
}
three <- c("log(PTFSM6CLA1A)", "log(PERCTOT)", "ks2_c")
perms <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))

shap <- setNames(numeric(3), three)
for (p in perms) {
  sofar <- character(0)
  for (i in p) {
    v <- three[i]
    shap[v] <- shap[v] + (r2_of(c(sofar, v)) - r2_of(sofar))
    sofar <- c(sofar, v)
  }
}
shap <- shap / length(perms)
shap_share <- 100 * shap / sum(shap)

# How much of the explained variance is shared rather than unique: the
# gap between the full model and the sum of each term's last-in
# contribution.
uniq   <- vapply(three, function(v) r2_of(three) - r2_of(setdiff(three, v)),
                 numeric(1))
r2_all <- r2_of(three)
shared_pct <- 100 * (r2_all - sum(uniq)) / r2_all

LAB <- c("log(PTFSM6CLA1A)" = "Disadvantage",
         "log(PERCTOT)"     = "Absence",
         "ks2_c"            = "Prior attainment")

```

Since @Coleman1966, study after study has found that the share of
variation in pupils' outcomes attributable to differences *between
schools* --- as opposed to differences in the pupils they admit and the
circumstances those pupils bring with them --- is somewhere in the order
of **`r SCHOOL_LO` to `r SCHOOL_HI` per cent**. The landmark English
study settled on about a tenth [@SmithTomlinson1989], and the finding
has held up across countries and phases [@TeddlieReynolds2000].

The multilevel model behind *How to Pull the Right Lever* decomposes
that variance directly, and lands in the same place.

```{r fig-school-share}
#| fig-cap: "Where the variance in school-level Attainment 8 sits, decomposed from the multilevel model, for all pupils and for the two pupil groups separately. Absence is shown as its own row rather than buried in the inherited block, split into the structural share a first-stage model predicts from intake and area, and the school-level residual it does not. The persistent school effect is an upper bound: it bundles genuine school ethos and management with anything unmeasured about the catchment, such as cultural capital or parental motivation."
#| fig-height: 9.4

sed <- bh_data("school_effect_decomp.rds")
lev <- bh_data("school_leverage.rds")
LV  <- lev$leverage

GRP_COL <- c("Inherited (exogenous)"      = "#49a0c4",
             "Absence: structural"        = "#7FB3D5",
             "Absence: school-reachable"  = "#F2A03D",
             "Persistent school effect"   = "#6A4C93",
             "Workforce"                  = "#ED357D",
             "Place / noise"              = "#BBBBBB")

ABS_LAB <- "Absence (part inherited, part school-reachable)"
EXO_LAB <- "Inherited: cohort, prior attainment, neighbourhood"

# Absence is published inside the exogenous block, but it is the one
# term that is partly within a school's control. Split it out as its own
# row. The published totals are kept; the share of the exogenous block
# that is absence is taken from the fitted model, and the split of that
# absence into structural and school-reachable from the stage-1 model.
prep <- function(d, who_lab) {
  r        <- LV[LV$who == who_lab, ]
  abs_frac <- r$absence / r$exogenous
  exo_tot  <- d[["Share of total %"]][grepl("^Exogenous", d$Component)]
  abs_tot  <- exo_tot * abs_frac

  bind_rows(
    d %>%
      filter(!grepl("^Exogenous", Component)) %>%
      transmute(Component, share = `Share of total %`,
                grp_type = case_when(
                  grepl("^Endogenous", Component) ~ "Workforce",
                  Component == "School (persistent, unexplained)" ~ "Persistent school effect",
                  TRUE ~ "Place / noise")),
    tibble(Component = EXO_LAB, share = exo_tot - abs_tot,
           grp_type = "Inherited (exogenous)"),
    tibble(Component = ABS_LAB,
           share = c(abs_tot * lev$struct_share, abs_tot * lev$school_share),
           grp_type = c("Absence: structural", "Absence: school-reachable"))
  ) %>%
    mutate(who = who_lab,
           Component = sub("^Endogenous, within-school \\(workforce\\)$",
                           "Workforce (the part a school hires)", Component))
}

decomp <- bind_rows(prep(sed$decomp_all, "All pupils"),
                    prep(sed$decomp_dis, "Disadvantaged pupils"),
                    prep(sed$decomp_non, "Non-disadvantaged pupils")) %>%
  mutate(who = factor(who, c("All pupils", "Disadvantaged pupils",
                             "Non-disadvantaged pupils")),
         grp_type = factor(grp_type, levels = names(GRP_COL)))

# Order rows by the all-pupil total for each component, so the three
# panels read in the same order and can be compared row by row.
ord <- decomp %>%
  filter(who == "All pupils") %>%
  group_by(Component) %>% summarise(t = sum(share), .groups = "drop") %>%
  arrange(t) %>% pull(Component)
decomp <- decomp %>% mutate(Component = factor(Component, ord))

# One label per row, placed at the row total rather than per segment, so
# the split absence bar reads as a single quantity.
tot_lab <- decomp %>%
  group_by(who, Component) %>% summarise(share = sum(share), .groups = "drop")

ggplot(decomp, aes(share, Component, fill = grp_type)) +
  geom_col(width = 0.72) +
  geom_text(data = tot_lab, inherit.aes = FALSE, hjust = -0.15, size = 3,
            mapping = aes(x = share, y = Component,
                          label = sprintf("%.1f%%", share))) +
  facet_wrap(~ who, ncol = 1) +
  scale_fill_manual(values = GRP_COL, name = NULL) +
  scale_x_continuous(limits = c(0, 52), expand = expansion(mult = c(0, 0.02))) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(x = "Share of total outcome variance (%)", y = NULL,
       title = "Most of what separates schools' results arrives with the children",
       subtitle = "Absence shown separately: it is the one term that is partly inherited and partly within a school's reach",
       caption = "Sources: How to Pull the Right Lever, school effect decomposition; RPE paper two-stage absence model.") +
  theme_bh(11) +
  theme(legend.position = "bottom", panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8),
        strip.text = element_text(face = "bold", hjust = 0))
```

```{r decomp-stats}
#| include: false
pick <- function(d, pat) d[["Share of total %"]][grepl(pat, d$Component)][1]
wf_all  <- pick(sed$decomp_all, "^Endogenous")
ex_all  <- pick(sed$decomp_all, "^Exogenous")
sch_all <- pick(sed$decomp_all, "^School \\(persistent")
wf_dis  <- pick(sed$decomp_dis, "^Endogenous")
ex_dis  <- pick(sed$decomp_dis, "^Exogenous")
sch_dis <- pick(sed$decomp_dis, "^School \\(persistent")
wf_non  <- pick(sed$decomp_non, "^Endogenous")
ex_non  <- pick(sed$decomp_non, "^Exogenous")
```

For all pupils, the part a school **directly controls** through the
staff it hires accounts for about **`r fmt_pct(LV$workforce[LV$who == "All pupils"], 1)`**
of the variance. What it inherits with the intake --- cohort composition,
prior attainment, neighbourhood --- accounts for roughly
**`r fmt_pct(LV$exogenous[LV$who == "All pupils"] - LV$absence[LV$who == "All pupils"], 0)`**.
Even the persistent school effect, at `r sch_all`%, is an upper bound: it
is what is left once everything measured is accounted for, so it bundles
genuine school ethos and leadership together with anything unmeasured
about the catchment --- cultural capital, parental motivation, and the
rest.

Absence has its own row because it is the one term that will not sit in
either camp. The published decomposition files it with the inherited
block, but a school's attendance is partly a fact about its intake and
partly a fact about what it does. The next section takes that apart, and
it turns out to matter more than anything else on the chart.

**For disadvantaged pupils --- the group any mixing policy is aimed at
--- the workforce term falls to `r wf_dis`%**, against `r ex_dis`%
inherited, with a persistent school effect of `r sch_dis`%. The lever
gets *smaller*, not larger, precisely where the policy is pointed. For
non-disadvantaged pupils it is `r wf_non`% against `r ex_non`% --- so
the pattern is not an artefact of one group, and the school-controllable
share is smallest for the children the policy is meant to help.

## Absence is the term that moves {#sec-absence-split}

The published decomposition files absence inside the exogenous block ---
"cohort, prior attainment, absence, neighbourhood" --- as though a school
simply receives its attendance rate. The two-stage work in the RPE paper
shows that it does not. A first-stage model of school absence on intake,
area and segregation explains only about half of its variation; the rest
is a residual that pastoral systems, attendance officers, family liaison
and persistent-absentee follow-up act on directly. That is why it is
drawn as its own split row above, and it is worth being precise about
the size of each part.

```{r leverage-stats}
#| include: false
lev <- bh_data("school_leverage.rds")
LV  <- lev$leverage
lv  <- function(who, col) LV[[col]][LV$who == who]
```

Across `r fmt_n(lev$stage1_n)` school-years, that first stage accounts
for **`r fmt_pct(100 * lev$struct_share)`** of the variation in absence.
The other **`r fmt_pct(100 * lev$school_share)`** is school-level
residual. Absence is not one thing: roughly half of it is structural and
outside a school's reach, and roughly half is not.

That changes the arithmetic, because absence is the largest single term
in the exogenous block --- `r fmt_pct(lv("All pupils", "absence"), 1)` of
total variance for all pupils, and
`r fmt_pct(lv("Disadvantaged pupils", "absence"), 1)` for disadvantaged
pupils, where it is the biggest component of all.

```{r fig-leverage}
#| fig-cap: "The same variance, re-cut by what a school can actually act on. The absence term is split into the structural share the first-stage model predicts and the school-level residual it does not."
#| fig-height: 4

LV %>%
  transmute(who = factor(who, LV$who),
            `Workforce` = workforce,
            `Absence, school-reachable` = absence_school,
            `Absence, structural` = absence_structural,
            `Other inherited` = exogenous - absence) %>%
  pivot_longer(-who, names_to = "part", values_to = "pct") %>%
  mutate(part = factor(part, c("Workforce", "Absence, school-reachable",
                               "Absence, structural", "Other inherited"))) %>%
  ggplot(aes(pct, fct_rev(who), fill = part)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = ifelse(pct >= 3, sprintf("%.0f%%", pct), "")),
            position = position_stack(vjust = 0.5), size = 3.1,
            colour = "white", fontface = "bold") +
  scale_fill_manual(values = c("Workforce" = "#ED357D",
                               "Absence, school-reachable" = "#F2A03D",
                               "Absence, structural" = "#7FB3D5",
                               "Other inherited" = "#49a0c4"), name = NULL) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(x = "Share of total outcome variance (%)", y = NULL,
       title = "What a school can actually act on is mostly attendance",
       subtitle = "Fixed-effect variance only; the random-effect components are omitted here",
       caption = "Sources: How to Pull the Right Lever; RPE paper two-stage absence model.") +
  theme_bh(11) +
  theme(legend.position = "bottom", panel.grid.major.y = element_blank())
```

Adding the school-controllable half of absence to the workforce term
gives a school-reachable share of **`r fmt_pct(lv("All pupils", "reachable"), 1)`**
for all pupils, **`r fmt_pct(lv("Disadvantaged pupils", "reachable"), 1)`**
for disadvantaged pupils and
**`r fmt_pct(lv("Non-disadvantaged pupils", "reachable"), 1)`** for
non-disadvantaged --- strikingly consistent, and sitting squarely inside
the `r SCHOOL_LO`--`r SCHOOL_HI` per cent that sixty years of
school-effectiveness research has settled on.

::: {.callout-note appearance="simple"}
## A small discrepancy worth naming

The workforce figures in this section are recomputed from the fitted
models rather than read off the published table, so that the parts sum
to the totals quoted beside them. They come out slightly lower than the
published version --- `r fmt_pct(lv("All pupils", "workforce"), 1)`
against `r wf_all`% for all pupils --- because the cached table and the
model objects available here are not from quite the same fit. The
difference is under a percentage point and changes nothing in the
argument, but the two numbers are not identical and it would be worse to
present them as if they were.
:::

So the earlier conclusion needs restating, more precisely and more
usefully:

::: {.callout-important}
## The corrected version

A school can reach about **`r fmt_pct(lv("All pupils", "reachable"), 0)`**
of the variance in its results --- not the `r wf_all`% the workforce term
alone suggests. But **almost all of that leverage runs through
attendance**, not through who is sitting in the classroom.

For disadvantaged pupils the point is sharper still. The workforce term
is `r fmt_pct(lv("Disadvantaged pupils", "workforce"), 1)`; the
school-reachable absence term is
`r fmt_pct(lv("Disadvantaged pupils", "absence_school"), 1)` ---
`r sprintf("%.0f", lv("Disadvantaged pupils", "absence_school") / lv("Disadvantaged pupils", "workforce"))`
times larger. The lever that works for the children the policy is aimed
at is attendance, and it is not an admissions lever.

Moving children between schools does not touch either term. It changes
which building a child attends without changing their attendance, and
without changing the intake characteristics that account for the rest.
That is the case against expecting much from redistribution --- and,
equally, the case for expecting a great deal from attendance work.
:::

::: {.callout-note appearance="simple"}
## Does the shared variance change this?

It is a fair challenge. The two fixed blocks are not independent ---
schools with difficult intakes also tend to have less stable workforces
--- so the variance they share has to be allocated somehow, and the
answer depends on how.

The decomposition above credits each block with its covariance against
the total. An alternative is to split the shared part evenly, as a
Shapley decomposition does. On the fitted models the two blocks
correlate at about **+0.3 to +0.4**, and the choice is not immaterial:
the workforce share moves from 2.5% to **7.5%** for all pupils, from
0.8% to **3.1%** for disadvantaged pupils, and from 3.2% to **8.1%** for
non-disadvantaged pupils --- roughly a threefold difference.

It does not, however, change the conclusion. Under the more generous
allocation the workforce share is still about eight times smaller than
the inherited share for all pupils, and about seventeen times smaller
for disadvantaged pupils. The random-effect components --- the
persistent school effect among them --- are untouched either way,
because variance components in a multilevel model are orthogonal by
construction and have no shared part to argue over.

The figures quoted here are the published ones, on the covariance split.
Read the workforce term as *at most* a few per cent rather than as a
precise quantity.
:::

```{r fig-shapley}
#| fig-cap: "How the explained variance in school-level Attainment 8 divides between the three main intake measures, by Shapley decomposition."
#| fig-height: 3

tibble(term = LAB[three], share = shap_share) %>%
  ggplot(aes(reorder(term, share), share)) +
  geom_col(fill = "#2166ac", width = 0.6) +
  geom_text(aes(label = fmt_pct(share, 0)), hjust = -0.18, size = 3.4) +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(shap_share) * 1.22)) +
  labs(x = NULL, y = "Share of explained variance (%)",
       title = "And within that inherited share, how it divides",
       subtitle = sprintf("Shapley decomposition of school-level Attainment 8 (R² = %.2f)", r2_all),
       caption = "Source: DfE performance tables.") +
  theme_bh(11) +
  theme(panel.grid.major.y = element_blank())
```

Two things follow, and together they are the most important finding in
this section for the decisions in front of the council.

**Moving children between schools operates on the small term.**
Redistributing pupils changes *which* school a child attends. It does
not change the intake characteristics and social circumstances that
account for `r ex_all`% of the variance, and it does not touch the
workforce that accounts for `r wf_all`%. A policy of social mixing
through admissions is therefore working on the smaller term by
construction --- not because mixing is undesirable, but because the
arithmetic limits what it can deliver.

The next section qualifies this in an important way. One large item
inside that `r ex_all`% is not inherited at all.

**And the measured drivers cannot be cleanly separated anyway.** Of the
`r fmt_pct(100 * r2_all, 0)` of school-level variation the three
measures jointly explain, **`r fmt_pct(shared_pct, 0)` is variance they
*share*** rather than variance any one of them owns. Disadvantage,
absence and prior attainment correlate at
`r sprintf("%.2f", cor(log(nat$PTFSM6CLA1A), nat$ks2_c))`,
`r sprintf("%.2f", cor(log(nat$PERCTOT), nat$ks2_c))` and
`r sprintf("%.2f", cor(log(nat$PTFSM6CLA1A), log(nat$PERCTOT)))`
because they are substantially the same phenomenon seen three ways. Any
policy that targets one of them in isolation --- and an admissions
policy targets disadvantage in isolation --- is pulling on a rope
attached to the other two.

::: {.callout-important}
## What this does and does not say

It does **not** say schools do not matter. A persistent school effect of
`r sch_all`% across three and a half thousand secondary schools is a
great deal, and section 2.3 showed Brighton's schools mostly performing
above what their intakes predict. Nor does it say that segregation
between schools is acceptable; there are fairness arguments for mixed
intakes that have nothing to do with attainment.

What it says is narrower and harder to argue with: **the council should
not expect a redistribution policy to move attainment much**, because
the mechanism it operates through is the smallest of the terms
available. The instruments that reach the larger terms --- attendance
above all --- are not admissions instruments at all.
:::

## A school's disadvantaged results are a noisy signal {#sec-noise}

One row on that chart is easy to skip past and should not be. For
disadvantaged pupils, **transient cohort-year noise is the
second-largest component of all** --- larger than the persistent school
effect, larger than absence, larger than anything except what the
children bring with them.

```{r noise-stats}
#| include: false
pt <- function(d, pat) d[["Att8 points"]][grepl(pat, d$Component)][1]
sc <- function(d, pat) d[["Share of total %"]][grepl(pat, d$Component)][1]

noise <- tibble(
  who = factor(c("All pupils", "Disadvantaged pupils", "Non-disadvantaged pupils"),
               c("All pupils", "Disadvantaged pupils", "Non-disadvantaged pupils")),
  persistent = c(pt(sed$decomp_all, "^School \\(persistent"),
                 pt(sed$decomp_dis, "^School \\(persistent"),
                 pt(sed$decomp_non, "^School \\(persistent")),
  transient  = c(pt(sed$decomp_all, "^Transient"),
                 pt(sed$decomp_dis, "^Transient"),
                 pt(sed$decomp_non, "^Transient")))

n_dis_tr  <- pt(sed$decomp_dis, "^Transient")
n_dis_sch <- pt(sed$decomp_dis, "^School \\(persistent")
n_all_tr  <- pt(sed$decomp_all, "^Transient")
n_all_sch <- pt(sed$decomp_all, "^School \\(persistent")
# Two schools with identical true performance still differ by this much
# in a given year, on average, purely from cohort-year variation.
gap_dis <- n_dis_tr * sqrt(2)
```

```{r fig-noise}
#| fig-cap: "The persistent difference between schools against the year-to-year wobble within them, in Attainment 8 points. For disadvantaged pupils the wobble is the larger of the two."
#| fig-height: 3.4

noise %>%
  pivot_longer(-who, names_to = "part", values_to = "pts") %>%
  mutate(part = factor(part, c("persistent", "transient"),
                       c("Persistent difference between schools",
                         "Year-to-year noise within a school"))) %>%
  ggplot(aes(pts, fct_rev(who), fill = part)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.66) +
  geom_text(aes(label = sprintf("%.1f", pts)),
            position = position_dodge(width = 0.72), hjust = -0.25, size = 3.2) +
  scale_fill_manual(values = c("Persistent difference between schools" = "#6A4C93",
                               "Year-to-year noise within a school" = "#BBBBBB"),
                    name = NULL) +
  scale_x_continuous(limits = c(0, max(noise$transient) * 1.25),
                     expand = expansion(mult = c(0, 0.02))) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(x = "Attainment 8 points (standard deviation)", y = NULL,
       title = "For disadvantaged pupils, the noise is bigger than the signal",
       caption = "Source: How to Pull the Right Lever, school effect decomposition.") +
  theme_bh(11) +
  theme(legend.position = "bottom", panel.grid.major.y = element_blank())
```

For **all** pupils the school signal wins: the persistent difference
between schools is `r sprintf("%.1f", n_all_sch)` Attainment 8 points
against `r sprintf("%.1f", n_all_tr)` points of year-to-year noise. For
**disadvantaged** pupils it reverses --- `r sprintf("%.1f", n_dis_sch)`
points of persistent difference against **`r sprintf("%.1f", n_dis_tr)`
points of noise**.

The reason is arithmetic rather than mysterious. A secondary school's
disadvantaged cohort is small --- often forty or fifty pupils --- and
small numbers move about. A handful of children having a bad year, or a
good one, shifts the published figure by more than any real difference
in what the school does.

The consequence is uncomfortable and worth stating plainly. **Two
schools doing genuinely equally well by their disadvantaged pupils will
differ by around `r sprintf("%.0f", gap_dis)` Attainment 8 points in a
given year for no reason at all.** That is larger than most of the gaps
that get argued about. A single year of disadvantaged results is not
evidence about a school; it is one draw from a distribution.

::: {.callout-important}
## What follows for how the city reads its own data

**Never judge a school on one year of disadvantaged results.** Three or
more years, averaged, or the comparison is mostly noise. This applies to
league tables, to consultation documents, and to the case made for or
against any individual school.

**Do not expect to detect an admissions policy's effect this way
either.** If redistribution moves disadvantaged attainment by a point or
two --- which section 2.4 suggests is optimistic --- that signal sits
well inside `r sprintf("%.0f", gap_dis)` points of year-to-year variation.
The policy could work exactly as intended and be invisible for years;
it could also fail and appear to succeed. Evaluating it will need
attendance and intake measures, not headline outcomes.

**This bears directly on the free school meals review.** The FSM
admissions priority took effect for entry in September 2023. Those
children entered Year 7 that autumn and will not sit GCSEs until
**summer 2028**. So a review conducted now is not working with thin
outcome evidence --- it is working with *none*: not one cohort admitted
under the policy has yet reached an examination. Even in 2028 there will
be a single cohort, and a single cohort of disadvantaged pupils in one
catchment carries the `r sprintf("%.1f", n_dis_tr)`-point year-to-year
variation described above.

That is not an argument against reviewing the policy. Admission
patterns, intake composition by catchment, and the operation of the
criteria itself can all be examined now, and section 5 does some of
that. It is an argument against reviewing it **on attainment**, or
against any conclusion of the form "results have or have not improved
since the policy came in". No honest analysis can support such a claim
from the data that exists, in either direction.

**This compounds the problem in section 2.3.** Families are choosing on
a measure that is largely a description of intake, *and* that measure is
noisiest precisely for the pupils whose schools are most often
criticised on it.
:::


# Demographic futures {#sec-demography}

The single most important fact about this system is that it is shrinking,
and that the shrinkage is already locked in. The children who will enter
Year 7 in 2033 have already started primary school. Their number is not
a forecast in any meaningful sense --- it is a count, adjusted for a
transfer rate that has been remarkably stable for a decade.

## What has already happened {#sec-cohort-history}

```{r cohort-data}
#| include: false
rc <- bh_data("reception_cohort.rds")

y7    <- rc$secondary_city
rec   <- rc$reception_city
peak  <- y7 %>% slice_max(y7_offers, n = 1)
now   <- y7 %>% slice_max(year, n = 1)
proj  <- rc$projection
last_proj <- proj %>% slice_max(entry_year, n = 1)
drop_pct <- 100 * (last_proj$central - now$y7_offers) / now$y7_offers
```

Year 7 offers across the city peaked at **`r fmt_n(peak$y7_offers)`** in
`r peak$year`. In `r now$year` the figure was
**`r fmt_n(now$y7_offers)`** --- a fall of
`r fmt_pct(100 * (peak$y7_offers - now$y7_offers) / peak$y7_offers, 1)`
from the peak. That decline is not a projection. It has happened.

```{r fig-cohort-history}
#| fig-cap: "Reception and Year 7 offers made across Brighton & Hove. The reception line leads the Year 7 line by seven years, so it is a preview of secondary demand rather than a separate story."
#| fig-height: 5

bind_rows(
  y7  %>% transmute(year, offers = y7_offers, phase = "Year 7 offers"),
  rec %>% transmute(year, offers = rec_offers, phase = "Reception offers")) %>%
  ggplot(aes(year, offers, colour = phase)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Reception offers" = "#2c7fb8",
                                 "Year 7 offers" = "#b2182b"), name = NULL) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Offers made",
       title = "Both phases are past their peak",
       subtitle = "Reception offers turn down first, and Year 7 follows seven years later",
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh()
```

## The cohort already in school {#sec-projection}

The transfer rate from a reception cohort to the Year 7 cohort seven
years later has been strikingly stable. Across the six cohorts where
both ends are observable, **`r fmt_pct(100 * rc$survival_recent, 1)`** of
reception offers reappear as Year 7 offers, with a standard deviation of
just `r fmt_pct(100 * rc$survival_sd, 1)`. The city loses about one child
in nine between ages four and eleven --- to independent schools, to
moves out of the area, and to schools across the boundary.

Applying that rate to reception cohorts already in school gives a
projection that depends on almost no assumptions:

```{r fig-projection}
#| fig-cap: "Year 7 demand projected from reception cohorts already in school. The band is the range implied by the observed year-to-year variation in the transfer rate."
#| fig-height: 5

ggplot() +
  geom_ribbon(data = proj, aes(entry_year, ymin = lo, ymax = hi),
              fill = "#b2182b", alpha = 0.18) +
  geom_line(data = proj, aes(entry_year, central),
            colour = "#b2182b", linewidth = 1.1, linetype = "22") +
  geom_point(data = proj, aes(entry_year, central), colour = "#b2182b", size = 2) +
  geom_line(data = y7, aes(year, y7_offers), colour = "grey25", linewidth = 1.1) +
  geom_point(data = y7, aes(year, y7_offers), colour = "grey25", size = 2) +
  annotate("text", x = 2016, y = 2480, label = "Observed", colour = "grey25",
           hjust = 0, fontface = "bold", size = 3.6) +
  annotate("text", x = 2029.5, y = 2180, label = "Projected from\nchildren already in school",
           colour = "#b2182b", hjust = 0, size = 3.4, lineheight = 0.95) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Year 7 offers",
       title = sprintf("A further %s fall is already in the primary schools",
                       fmt_pct(abs(drop_pct), 0)),
       subtitle = sprintf("From %s in %s to about %s by %s",
                          fmt_n(now$y7_offers), now$year,
                          fmt_n(last_proj$central), last_proj$entry_year),
       caption = "Sources: BHCC allocation factsheets; transfer rate estimated from six observed cohorts.") +
  theme_bh()
```

By **`r last_proj$entry_year`** the city should expect about
**`r fmt_n(last_proj$central)`** Year 7 children, against
`r fmt_n(now$y7_offers)` today --- a further fall of
`r fmt_pct(abs(drop_pct), 0)`. Against a current published admission
capacity of about
`r fmt_n(sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]))`
places, that is roughly
`r fmt_n(sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]) - last_proj$central)`
surplus places --- the equivalent of two and a half average-sized
secondary schools standing empty.

## Where the council's forecast differs {#sec-council-forecast}

The council publishes its own catchment-level forecasts. They are
consistently **higher** than what the primary cohorts support, and the
gap widens with distance.

```{r fig-forecast-gap}
#| fig-cap: "The council's own demand forecast against a projection built from reception cohorts already in school. Positive values mean the council expects more children than the primary registers contain."
#| fig-height: 4.6

rc$comparison %>%
  mutate(gap = ons_demand - reception_based) %>%
  ggplot(aes(entry_year, gap)) +
  geom_col(fill = "#7b3294", alpha = 0.85, width = 0.65) +
  geom_hline(yintercept = 0, colour = "grey30") +
  geom_text(aes(label = sprintf("%+.0f", gap),
                vjust = ifelse(gap >= 0, -0.4, 1.3)), size = 3.2) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.18))) +
  labs(x = "Year of Year 7 entry", y = "Council forecast minus cohort projection",
       title = "The council expects more children than the primary registers hold",
       subtitle = "Difference in Year 7 places, city-wide",
       caption = "Sources: BHCC published forecasts; reception-based projection as above.") +
  theme_bh()
```

The city-wide figure hides a good deal. Broken down by catchment, and
set against the council's own two successive forecasts, the picture is
less tidy.

```{r fig-catchment-projections}
#| fig-cap: "Year 7 offers for each catchment: what has actually happened, our projection from primary cohorts already in school, and the council's two most recent forecasts of the same years. Dotted line is the catchment's admission number."
#| fig-height: 8
#| fig-width: 10

# Four series on one set of axes. The council labels its catchments
# differently from the model tables, so everything is converted to model
# keys first - see as_model_catchment() in R/00_core.R.
CATCH_DISPLAY <- c(CATCH_LABELS[unname(CATCH_FROM_MODEL)], "Religious schools")
names(CATCH_DISPLAY) <- c(names(CATCH_FROM_MODEL), "Religious schools")

council25 <- rc$council %>%
  transmute(catchment = as_model_catchment(CatchmentGroup),
            entry_year, offers = council, PAN,
            series = "Council forecast (Oct-25)")

council24 <- bh_data("council_forecast_oct24.csv") %>%
  transmute(catchment = as_model_catchment(CatchmentGroup),
            entry_year, offers = council_oct24,
            series = "Council forecast (Oct-24, superseded)")

series <- bind_rows(
  rc$observed_by_catchment %>%
    transmute(catchment, entry_year, offers = y7_offers,
              series = "Actual offers"),
  rc$by_catchment %>%
    transmute(catchment, entry_year, offers = projected,
              series = "Our projection (from primary cohorts)"),
  council25 %>% select(-PAN),
  council24) %>%
  filter(!is.na(catchment), entry_year >= 2016) %>%
  mutate(panel = factor(unname(CATCH_DISPLAY[catchment]),
                        unname(CATCH_DISPLAY)),
         series = factor(series, c(
           "Actual offers", "Our projection (from primary cohorts)",
           "Council forecast (Oct-25)",
           "Council forecast (Oct-24, superseded)")))

stopifnot(!any(is.na(series$panel)))

pan_lines <- council25 %>%
  distinct(catchment, PAN) %>%
  mutate(panel = factor(unname(CATCH_DISPLAY[catchment]), unname(CATCH_DISPLAY)))

ggplot(series, aes(entry_year, offers, colour = series)) +
  geom_hline(data = pan_lines, aes(yintercept = PAN),
             linetype = "dotted", colour = "grey30", linewidth = 0.45) +
  geom_line(linewidth = 0.75) +
  geom_point(size = 1.2) +
  facet_wrap(~ panel, ncol = 2, scales = "free_y") +
  scale_colour_manual(values = c(
    "Actual offers"                          = "#1f78b4",
    "Our projection (from primary cohorts)"  = "#33a02c",
    "Council forecast (Oct-25)"              = "#e31a1c",
    "Council forecast (Oct-24, superseded)"  = "#fb9a99"), name = NULL) +
  scale_y_continuous(limits = c(0, NA)) +
  guides(colour = guide_legend(nrow = 2)) +
  labs(x = NULL, y = "Year 7 offers",
       title = "Every catchment, four views of the same future",
       subtitle = "Dotted line is the catchment admission number",
       caption = "Sources: BHCC allocation factsheets; BHCC forecasts, Appendices 6 and 7.") +
  theme_bh(11) +
  theme(legend.position = "top", strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r proj-stats}
#| include: false
# How far the council moved between its own two forecasts, on the years
# both cover - a measure of how settled the council's own view is.
rev <- inner_join(
  council25 %>% select(catchment, entry_year, oct25 = offers),
  council24 %>% select(catchment, entry_year, oct24 = offers),
  by = c("catchment", "entry_year")) %>%
  mutate(shift = oct25 - oct24)
rev_by_catch <- rev %>%
  group_by(catchment) %>%
  summarise(mean_shift = mean(shift), max_abs = max(abs(shift)), .groups = "drop") %>%
  arrange(desc(abs(mean_shift)))
biggest_rev <- rev_by_catch %>% slice(1)
n_down <- sum(rev_by_catch$mean_shift < 0)
n_catch <- nrow(rev_by_catch)

# Where our projection sits relative to the council's current forecast,
# on the years both cover.
gap_by_catch <- inner_join(
  council25 %>% select(catchment, entry_year, council = offers),
  rc$by_catchment %>% select(catchment, entry_year, ours = projected),
  by = c("catchment", "entry_year")) %>%
  group_by(catchment) %>%
  summarise(mean_gap = mean(ours - council), .groups = "drop") %>%
  arrange(mean_gap)
n_below <- sum(gap_by_catch$mean_gap < 0)
closest <- gap_by_catch %>% slice_min(abs(mean_gap), n = 1)
closest_txt <- if (abs(closest$mean_gap) < 1) "under a place" else
  sprintf("%.0f places", abs(closest$mean_gap))
spell <- function(n) c("one","two","three","four","five","six","seven",
                       "eight","nine","ten")[n]
down_txt <- if (n_down == n_catch)
  sprintf("Every one of the %s catchments was", spell(n_catch)) else
  sprintf("%s of the %s catchments were", spell(n_down), spell(n_catch))
```

Two things stand out.

**Our projection sits below the council's in
`r n_below` of the `r nrow(gap_by_catch)` catchments** --- the city-wide
gap of the previous chart, distributed. The closest agreement is in
`r CATCH_DISPLAY[[as.character(closest$catchment)]]`, where the two
differ by `r closest_txt` a year on average.

**And the council's own view has moved a long way in a single year.**
Between the Oct-24 and Oct-25 forecasts, for the years both cover, the
figure for
**`r CATCH_DISPLAY[[as.character(biggest_rev$catchment)]]`** moved by an
average of `r sprintf("%+.0f", biggest_rev$mean_shift)` places a year,
with a largest single-year revision of `r biggest_rev$max_abs`.

The direction is the part worth noting. **`r down_txt` revised downward**, none upward. That is not
forecasting noise; it is a consistent correction in one direction, and
it points the same way as the projection from primary cohorts. The
council's own figures are moving toward the lower number --- they have
not arrived at it yet, and decisions with twenty-year consequences are
being taken against a quantity still in motion.

## The forecast is not wrong in total. It is wrong by place {#sec-forecast-errors}

The comparison above sets one projection against another, and a reader
is entitled to ask why either should be believed. For 2026 that question
can be settled, because the offers have been made. Both the council's
forecast and our projection can be checked against what happened.

```{r tbl-2026-errors}
#| tbl-cap: "Year 7 offers in 2026: what the council forecast, what our projection gave, and what actually happened."

err26 <- rc$observed_by_catchment %>%
  filter(entry_year == 2026) %>%
  select(catchment, actual = y7_offers) %>%
  inner_join(council25 %>% filter(entry_year == 2026) %>%
               select(catchment, PAN, council = offers), by = "catchment") %>%
  left_join(rc$by_catchment %>% filter(entry_year == 2026) %>%
              select(catchment, ours = projected), by = "catchment") %>%
  mutate(council_err = council - actual, ours_err = round(ours - actual)) %>%
  arrange(desc(council_err))

err26 %>%
  transmute(Catchment = unname(CATCH_DISPLAY[catchment]),
            PAN, Actual = actual,
            `Council forecast` = council,
            `Council error` = sprintf("%+d", council_err),
            `Our projection` = round(ours),
            `Our error` = sprintf("%+d", ours_err)) %>%
  knitr::kable(align = "lrrrrrr")
```

```{r err-stats}
#| include: false
net_err <- sum(err26$council_err)
abs_err <- sum(abs(err26$council_err))
lh26  <- err26 %>% filter(catchment == "Longhill")
hb26  <- err26 %>% filter(catchment == "Hove_Blatch")
paca26 <- err26 %>% filter(catchment == "PACA")
lh_ratio <- lh26$council / lh26$actual
```

::: {.callout-note appearance="simple"}
## Read our column with more suspicion than the council's

The council's forecast is a genuine prediction: published in advance,
checked here against what happened. Ours is not, quite. The retention
ratio it uses is fitted on cohorts that include the 2019 reception
group, which *is* the 2026 Year 7 group --- so 2026 is partly inside the
window our method was calibrated on, and its apparent accuracy is
flattered. The honest comparison is between the council's forecast and
the outturn; our column is there for orientation, not as a claim to have
done better.
:::

City-wide the council's 2026 forecast was close: it was
`r sprintf("%+d", net_err)` places out in total, on a cohort of over two
thousand. By catchment it was not close at all. The individual errors
sum to `r abs_err` places in absolute terms --- roughly
`r sprintf("%.0f", abs_err / abs(net_err))` times the net error. The
council is forecasting the right number of children and putting a great
many of them in the wrong place.

Two catchments carry most of that, and they miss in opposite directions.

**Longhill was forecast `r lh26$council` and received `r lh26$actual`
--- `r sprintf("%.1f", lh_ratio)` times too high.** This is a catchment
whose admission number is `r lh26$PAN` and which filled
`r fmt_pct(100 * lh26$actual / lh26$PAN, 0)` of it. The council's own
method already applies the largest leakage deduction in the city to this
catchment, at over 22 per cent, and it is still not close.

**Hove Park / Blatchington Mill was forecast `r hb26$council` and
received `r hb26$actual`** --- `r abs(hb26$council_err)` places too low,
the largest under-estimate of any catchment.

::: {.callout-warning}
## Question for the council 1: why is the forecast wrong in these two places, and in opposite directions?

The Hove Park / Blatchington Mill under-estimate has a candidate
explanation. The neighbouring PACA catchment was over-forecast by
`r paca26$council_err` places --- close to the
`r abs(hb26$council_err)` that Hove Park / Blatchington Mill was under
by. Children the model expected to appear in Portslade appearing in Hove
instead would account for much of it. That is testable, and the council
holds the data to test it.

**Longhill has no such explanation.** Its neighbours were not
over-forecast in a way that could absorb an
`r sprintf("%+d", lh26$council_err)`-place error --- BACA was
`r sprintf("%+d", err26$council_err[err26$catchment == "BACA"])` and
Patcham `r sprintf("%+d", err26$council_err[err26$catchment == "Patcham"])`.
The children the council expected to enrol at Longhill did not turn up
in an adjacent catchment. They did not turn up at all.

So the questions are:

1. **What accounts for the Longhill over-forecast, if not
   cross-catchment movement?** Out-of-city flow to Lewes district and
   the independent sector are the obvious candidates, and the council
   holds allocation records that would separate them.
2. **Is the leakage rate for the Longhill catchment being estimated on
   data old enough to predate the change it is meant to capture?** A
   rate calibrated on historical patterns will lag a shift that is
   still happening.
3. **Is the Hove Park / Blatchington Mill under-estimate the
   counterpart of the PACA over-estimate?** If so, the model's
   catchment-of-residence assumption is diverging from where families
   actually apply, and the two errors should be corrected together
   rather than separately.

These matter because the Longhill decision is being taken against the
forecast, and the forecast has been wrong about Longhill by a factor of
`r sprintf("%.1f", lh_ratio)` in the most recent year it can be checked
against.
:::

```{r lh-gap}
#| include: false
lh_council <- rc$council %>%
  mutate(catchment = as_model_catchment(CatchmentGroup)) %>%
  filter(catchment == "Longhill")
lh_cohort <- rc$by_catchment %>% filter(catchment == "Longhill")
lh_join <- lh_council %>%
  inner_join(lh_cohort %>% select(entry_year, projected), by = "entry_year") %>%
  mutate(gap = council - projected)
stopifnot(nrow(lh_join) > 0)
```

The city-wide gap is modest in most years. Broken down by catchment it
is not, and Longhill is where it opens widest.

```{r fig-longhill-gap}
#| fig-cap: "Longhill catchment: the council's forecast, the projection from its own primary cohorts, and the school's admission number."
#| fig-height: 4.6

lh_join %>%
  select(entry_year, Council = council, `From primary cohorts` = projected) %>%
  pivot_longer(-entry_year, names_to = "series", values_to = "children") %>%
  ggplot(aes(entry_year, children, colour = series)) +
  geom_hline(yintercept = rc$council_lh_pan, linetype = "31", colour = "grey45") +
  annotate("text", x = min(lh_join$entry_year), y = rc$council_lh_pan + 7,
           label = sprintf("Admission number (%s)", rc$council_lh_pan),
           hjust = 0, size = 3.1, colour = "grey35") +
  geom_line(linewidth = 1.1) + geom_point(size = 2.4) +
  scale_colour_manual(values = c("Council" = "#7b3294",
                                 "From primary cohorts" = "#008837"), name = NULL) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year of Year 7 entry", y = "Children",
       title = "Longhill: the two forecasts disagree, and both sit below the admission number",
       subtitle = sprintf("Mean gap of %.0f children a year",
                          mean(lh_join$gap, na.rm = TRUE)),
       caption = "Sources: BHCC catchment forecasts; reception-based projection.") +
  theme_bh()
```

The council's forecast for the Longhill catchment runs an average of
**`r fmt_n(mean(lh_join$gap, na.rm = TRUE))` children a year above** what
that catchment's own primary cohorts support. Both series sit below the
school's admission number of `r rc$council_lh_pan` in every year. This
matters because decisions about Longhill's future are being taken
against the higher of two numbers, and the higher one is the one the
primary registers do not support.

## The Peacehaven accession {#sec-peacehaven}

```{r peacehaven}
#| include: false
area_children <- oi$zones %>%
  group_by(area) %>%
  summarise(lsoas = n_distinct(lsoa), children = sum(Oi, na.rm = TRUE),
            .groups = "drop")
exp_area <- area_children %>% filter(area == "Expansion area")
tot_children <- sum(area_children$children)
ph_pan <- oi$schools$pan2026[oi$schools$name == "Peacehaven Community School"]
```

The catchment system reaches beyond the city boundary. The Peacehaven
and Telscombe area contributes **`r exp_area$lsoas` LSOAs** and about
**`r fmt_n(exp_area$children)` cohort-aged children**, roughly
`r fmt_pct(100 * exp_area$children / tot_children, 0)`
of the total the model covers.

In headcount terms this is close to self-supporting: Peacehaven
Community School has an admission number of `r ph_pan`, against about
`r fmt_n(exp_area$children)` children in its area. The difficulty is not
the arithmetic but the direction of travel. These families sit at the
eastern end of a transport network that runs east--west along the coast,
and section 4 shows that they are among the least well served in the
whole study area --- not because they are far from *a* school, but
because they are far from *most* schools, and therefore have the fewest
alternatives if their first choice is full.

# How reachable are the schools? {#sec-access}

Almost all of the public argument about Brighton's schools is conducted
in terms of admissions rules --- catchments, priorities, tie-breaks.
Underneath the rules sits a physical constraint that no rule can undo:
how long it actually takes a eleven-year-old to get to each school on a
bus. This section establishes that constraint before section 5 returns
to the rules.

All journey times here are routed with `r5r` over the merged street
network and bus timetable for the area, for a weekday morning arrival.
They include walking to the stop, waiting, the ride itself, and any
transfer.

## What counts as a long journey to school? {#sec-benchmarks}

Before mapping anything it is worth fixing what these minutes mean,
because a number like "twenty-eight minutes" is only interpretable
against something.

```{r benchmark-stats}
#| include: false
bm  <- acc$benchmarks
sta <- acc$statutory
```

| Benchmark | Figure | Source |
|:--|:--|:--|
| Average one-way school trip, England | **`r bm$nts_min` minutes** | National Travel Survey, five-year average to 2019 |
| Average one-way trip distance | **`r bm$nts_miles` miles** | as above |
| Statutory maximum, secondary age | **`r bm$dfe_max_sec` minutes** each way | DfE home-to-school travel guidance |
| Statutory maximum, primary age | **`r bm$dfe_max_pri` minutes** each way | as above |

Two cautions about the `r bm$nts_min`-minute average. It covers **ages 5
to 16 together**, and secondary journeys are longer than primary ones,
so it understates the secondary figure. And it counts **all modes**,
including the car --- which carries
`r 30`% of secondary school trips nationally against
`r 18`% by local bus. The journeys modelled here are walking and bus
only, so they describe the trip facing a family without a car. That is
the right comparison for this document, but it is not a like-for-like
one with the national number.

The `r bm$dfe_max_sec`-minute figure is firmer. It is the Department for
Education's guidance on how long a home-to-school journey should
reasonably take for a child of secondary age, each way, including
walking to a pick-up point.

::: {.callout-important}
## A correction to the routed times, and why it matters

The routed matrix returns a walk-and-bus itinerary for every
neighbourhood-school pair. Where the bus network is awkward it returns a
poor one: **`r fmt_pct(100 * acc$walk_fallback$share, 0)` of the
`r fmt_n(acc$walk_fallback$total)` neighbourhood-school pairs were
modelled as taking longer than simply walking the distance would**. In
the worst case a journey of about four kilometres came out at 99
minutes.

That is a property of the router, not of the city. A family facing a
99-minute bus ride to a school four kilometres away walks instead. So
every journey is now the **lesser of the routed bus time and the time it
would take to walk**, and the bus is used only where it is genuinely
faster.

The walking pace is not invented. Kent et al. [-@Kent2026], a systematic
review of active school travel, finds children's mean walking trip to
school is **`r fmt_n(acc$walk_fallback$review$dist_m)` metres in
`r sprintf("%.1f", acc$walk_fallback$review$time_min)` minutes** ---
`r sprintf("%.2f", acc$walk_fallback$kmh)` km/h --- and that is the speed
used, with a circuity factor of `r acc$walk_fallback$circuity` to turn
straight-line distance into street distance.

::: {.callout-note appearance="simple"}
## This is a ceiling, not a claim about how children travel

The same review finds children walk about a kilometre on average and
seldom beyond `r fmt_n(acc$walk_fallback$review$dist_range[2])` metres.
Capping a four-kilometre journey at its walking time does **not** assert
that a child walks four kilometres. It asserts that no journey should be
*modelled* as costing more than walking would, because a family with
that option would take it. The cap is an upper bound on modelled cost,
not a mode assumption.
:::

The correction is not cosmetic, and it is applied **upstream, in the
open model's own inputs** --- so every section of this document and of
the technical companion now uses the corrected costs. Sections 7 and 8
were re-run against them. It is worth being open that an earlier version
of this analysis overstated the headline figures:

| | Uncorrected | With the walk fallback |
|:--|--:|--:|
| Journey to the nearest school, child-weighted | 22.8 min | **`r sprintf("%.1f", acc$city$near_now)` min** |
| Neighbourhoods reaching no place within 30 min | 37 | **`r acc$city$zero30_now`** |
| Children in those neighbourhoods | 474 | **`r sprintf("%.0f", acc$city$zero30_children_now)`** |
:::

The corrected figure is a useful check in itself. A child-weighted mean
of **`r sprintf("%.1f", acc$city$near_now)` minutes** to the nearest
school sits almost exactly on the national average of `r bm$nts_min`
minutes --- which is reassuring for a model built from timetables and
street networks rather than from observed journeys.

### Against the statutory guidance {#sec-statutory}

The DfE limit applies to the journey a child actually has to make, so
the test is the time to the school their catchment entitles them to ---
not to the nearest school of any kind.

```{r tbl-statutory}
#| tbl-cap: "Journey to the nearest school in the child's own catchment, by catchment, against the DfE guidance."

acc$by_catch_stat %>%
  transmute(Catchment = unname(CATCH_DISPLAY[catchment]),
            `Mean minutes` = sprintf("%.1f", mean_min),
            `Worst zone` = sprintf("%.0f", max_min),
            `Zones over 45 min` = over_45,
            `Children affected` = sprintf("%.0f", children_over_45)) %>%
  knitr::kable(align = "lrrrr")
```

**No neighbourhood in the city exceeds the
`r bm$dfe_max_sec`-minute secondary limit** to its own catchment school.
The worst is `r sprintf("%.0f", sta$worst)` minutes. On that test the
city passes, and it is worth saying so plainly given how much of this
document is critical.

`r sta$over_45` zones --- about `r sprintf("%.0f", sta$over_45_children)`
children --- exceed the `r bm$dfe_max_pri`-minute figure that applies to
primary-age children. That is not a breach for secondary pupils, but it
is a reasonable marker of a long journey, and it identifies where the
pressure sits.

## Journey time to a single school {#sec-one-school}

```{r r5-meta}
#| include: false
bmeta <- readRDS(file.path(DATA, "travel", "build_metadata.rds"))
rp <- bmeta$params
```

::: {.callout-note collapse="true"}
## How these journey times are calculated, and what they are not

Every figure in this section comes from routing each origin to each
school over a street network and a bus timetable. The details matter,
because they set the limits of what the numbers can support.

**What was routed.** `r5r` over a merged OpenStreetMap extract and a
GTFS bus timetable, from **`r fmt_n(bmeta$n_origins)` postcode origins**
to `r nrow(bmeta$destinations)` destinations --- the eleven schools in
the expanded authority plus three East Sussex schools that Peacehaven
families genuinely choose between, and one hypothetical site. Postcode
results are then aggregated to the LSOA-by-catchment zones used
everywhere else, weighted by child population.

**The parameters.** Walking and public transport combined; up to
`r rp$max_walk_time` minutes of walking; trips capped at
`r rp$max_trip_duration` minutes; the median of a
`r rp$time_window`-minute departure window. Arrival is set for
**`r sub(":00$", "", rp$default_time)`** at every school except Longhill,
which uses **`r sub(":00$", "", rp$longhill_time)`** because its buses
run earlier. That asymmetry is deliberate.

**Two extracts, merged.** Geofabrik splits its Sussex data along the
county boundary, and the split runs straight through this study area:
Brighton & Hove sits in the west extract, the Peacehaven expansion area
in the east. r5r accepts exactly one `.pbf` file, so the two were merged
into `r bmeta$pbf`. Supplying both separately does not error --- it
silently returns nothing.

**The timetable is a timetable.** This is the single most important
caveat. The GTFS feed (`r bmeta$gtfs`, from the Department for
Transport's Open Bus Data Service) describes the service that was
*scheduled*, not the service that ran. The routing date is chosen
automatically as the weekday with the most services active in the feed,
because it is a multi-period export and only one window carries a full
timetable. So these are **timetabled journey times, not measured ones**.
They do not know about a bus that failed to arrive, one that was already
full, or one running ten minutes late.

**What is missing entirely.** No school buses, no parental lifts, no
cycling, no walking routes children actually prefer to the shortest one.
Car journeys are excluded by design --- the question here is what a
family without a car faces --- but nationally the car carries about 30%
of secondary school trips, so these times describe a subset of families
rather than the average one.

**Where the school gate is.** `r5r` snaps each school to the nearest
point on the street network, and which point that is can matter more
than it should. Dorothy Stringer and Varndean are 470 metres apart, yet
the matrix makes Stringer slower from most of the city by a median of
about five and a half minutes. That may be a real difference in bus
access or an artefact of the snapping; it has not been established
which, and it is large enough to affect anything said about Stringer
specifically. Longhill's position is untouched by it, because the
anomaly sits between two central schools.

**Bus stop coverage** was checked at five points across the study area:

```{r tbl-gtfs-coverage}
#| tbl-cap: "Bus stops within 2 km of each check point, in the feed used."
bmeta$gtfs_coverage %>%
  transmute(Place = place, `Stops within 2 km` = stops_within_2km) %>%
  knitr::kable(align = "lr")
```

The feed thins eastwards but does not stop at the city boundary, which
is what matters: the expansion area is genuinely served rather than
appearing unreachable because the timetable runs out.
:::

```{r journey-surfaces}
#| include: false

nm <- bh_data("network_maps.rds")

# One surface per school, from the zone x school cost table. Zones are
# LSOA x catchment, so collapse to LSOA weighting by the child
# population before joining to the polygons.
zone_w <- oi$zones %>% select(zone, lsoa, Oi)
ELM_LABEL <- "Longhill AT ELM GROVE (alternative site)"

to_lsoa <- function(costs, label) {
  costs %>%
    inner_join(zone_w, by = "zone") %>%
    group_by(name = label(name), lsoa) %>%
    summarise(mins = weighted.mean(cij, pmax(Oi, 1e-6), na.rm = TRUE),
              .groups = "drop")
}

surf_all <- bind_rows(
  to_lsoa(oi$costs_now, identity),
  # The relocation variant: only Longhill's column differs, so take just
  # that one and label it as the alternative site.
  to_lsoa(oi$costs_elm %>% filter(name == "Longhill High School"),
          function(x) rep(ELM_LABEL, length(x))))

# One colour scale for every school, so the maps are comparable with
# each other rather than each being stretched to its own range.
pal_time <- colorNumeric("magma", domain = range(surf_all$mins, na.rm = TRUE),
                         reverse = TRUE)

# Polygon detail is the binding constraint here: twelve layers means the
# geometry is embedded twelve times. Simplified at 40 m it is
# indistinguishable at any zoom this map offers and roughly halves the
# payload. Simplify in metres, before the transform.
lsoa_lite <- lsoa %>%
  st_transform(27700) %>%
  st_simplify(dTolerance = 40, preserveTopology = TRUE) %>%
  st_transform(4326)

sch_pts <- schools_sf(oi$schools)
elm_pt <- st_as_sf(tibble(easting = oi$elm_grove$easting,
                          northing = oi$elm_grove$northing),
                   coords = c("easting", "northing"), crs = 27700) %>%
  st_transform(4326)
elm_xy <- st_coordinates(elm_pt)
```

```{r fig-journey-picker}
#| fig-cap: "Walk-and-bus journey time from every neighbourhood, for a weekday morning arrival. Choose a school from the control at the top right. The colour scale is shared across all schools, so the maps are directly comparable. The last option places Longhill at the alternative Elm Grove site."
#| fig-height: 6.4

m_pick <- leaflet(width = "100%", height = 600,
                  options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles(providers$CartoDB.Positron)

layer_names <- c(sort(unique(oi$costs_now$name)), ELM_LABEL)

for (s in layer_names) {
  d <- lsoa_lite %>%
    inner_join(surf_all %>% filter(name == s), by = c("lsoa21cd" = "lsoa")) %>%
    filter(is.finite(mins))

  # The school's own marker, at Elm Grove for the relocation layer.
  if (identical(s, ELM_LABEL)) {
    px <- elm_xy[, 1]; py <- elm_xy[, 2]; plab <- oi$elm_grove$label
  } else {
    r <- sch_pts[sch_pts$name == s, ]
    px <- r$lon; py <- r$lat; plab <- s
  }

  m_pick <- m_pick %>%
    addPolygons(data = d, group = s,
                fillColor = ~pal_time(mins), fillOpacity = 0.8,
                color = "white", weight = 0.3,
                label = ~sprintf("%s: %.0f minutes", lsoa21nm, mins)) %>%
    addCircleMarkers(lng = px, lat = py, group = s,
                     radius = 7, fillColor = "#1a9850", fillOpacity = 1,
                     color = "white", weight = 2, label = plab)
}

m_pick %>%
  addPolygons(data = catch, group = "Catchment boundaries",
              fill = FALSE, color = "#111111", weight = 2.2, opacity = 0.85,
              label = ~unname(CATCH_LABELS[catchment])) %>%
  addLayersControl(baseGroups = layer_names,
                   overlayGroups = "Catchment boundaries",
                   options = layersControlOptions(collapsed = TRUE)) %>%
  hideGroup("Catchment boundaries") %>%
  addLegend(pal = pal_time, values = surf_all$mins, title = "Minutes",
            position = "bottomright", opacity = 0.85)
```

```{r tbl-journey-times}
#| tbl-cap: "Average walk-and-bus journey time to each school. The unweighted column treats every neighbourhood equally; the weighted column weights each by the number of cohort-aged children living there, which is the figure that describes what families actually face."

jt <- bind_rows(
  oi$costs_now %>% inner_join(zone_w, by = "zone") %>%
    group_by(name) %>%
    summarise(unw = mean(cij), wtd = weighted.mean(cij, Oi), .groups = "drop"),
  oi$costs_elm %>% filter(name == "Longhill High School") %>%
    inner_join(zone_w, by = "zone") %>%
    summarise(name = ELM_LABEL, unw = mean(cij), wtd = weighted.mean(cij, Oi))
) %>%
  arrange(wtd) %>%
  mutate(gap = wtd - unw)

jt %>%
  transmute(School = sub(" AT ELM GROVE \\(alternative site\\)",
                         " (at Elm Grove)", name),
            `Unweighted mean` = sprintf("%.1f", unw),
            `Child-weighted mean` = sprintf("%.1f", wtd),
            `Difference` = sprintf("%+.1f", gap)) %>%
  knitr::kable(align = "lrrr")
```

```{r journey-stats}
#| include: false
lh_now <- jt %>% filter(name == "Longhill High School")
lh_elm <- jt %>% filter(name == ELM_LABEL)
city_only <- jt %>% filter(!name %in% c(ELM_LABEL, "Peacehaven Community School"))
worst <- city_only %>% slice_max(wtd, n = 1)
best  <- city_only %>% slice_min(wtd, n = 1)
elm_rank <- sum(city_only$wtd < lh_elm$wtd) + 1
```

Three things come out of that table.

**Longhill is the least accessible school in the city**, at
`r sprintf("%.1f", lh_now$wtd)` minutes on the child-weighted measure
against `r sprintf("%.1f", best$wtd)` for
`r best$name`. Only Peacehaven, outside the authority, is further from
the city's children.

**Weighting by where children live makes almost every school look
worse.** The weighted mean exceeds the unweighted one for
`r sum(jt$gap > 0)` of the `r nrow(jt)` rows. Children are not
distributed evenly across the map: they are concentrated in places
slightly further from the schools than the average neighbourhood is.
The two largest gaps belong to
**`r and_list(city_only %>% slice_max(gap, n = 2) %>% pull(name))`**
(`r paste(sprintf("%+.1f", city_only %>% slice_max(gap, n = 2) %>% pull(gap)), collapse = " and ")`
minutes) --- which are also the two least full schools in the city. The
children in those catchments live further from their own school than an
average neighbourhood does, which is a different problem from the school
simply being far away, and a harder one to fix with a bus route.

**Moving Longhill to Elm Grove changes its position completely.** The
child-weighted journey time falls from
`r sprintf("%.1f", lh_now$wtd)` minutes to
**`r sprintf("%.1f", lh_elm$wtd)`** --- a saving of
`r sprintf("%.1f", lh_now$wtd - lh_elm$wtd)` minutes, which would take
it from the least accessible school in the city to
`r if (elm_rank == 1) "the most accessible" else paste0("the ", c("first","second","third","fourth","fifth")[elm_rank], " most accessible")`.
That is the single largest effect of any intervention considered
anywhere in this document, and section 8 returns to what it would and
would not solve.

## What one of those journeys actually looks like {#sec-legs}

A single number in minutes hides what the journey involves. These are
seven real routed journeys, drawn leg by leg: walking legs follow the
street network, bus legs follow the service's own shape.

```{r fig-legs}
#| fig-cap: "Seven worked journeys, routed leg by leg. Walking legs in orange, bus legs in blue. Each can be checked against local knowledge."
#| fig-height: 6

rg <- bh_data("route_geometries.rds")
legs <- rg$legs %>% st_transform(4326)

pal_mode <- colorFactor(c("BUS" = "#2166ac", "WALK" = "#e08214"),
                        domain = c("BUS", "WALK"))

mleg <- leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron)

for (j in sort(unique(legs$journey))) {
  d <- legs %>% filter(journey == j)
  tot <- rg$summary$total_minutes[rg$summary$journey == j]
  mleg <- mleg %>%
    addPolylines(data = d, group = j,
                 color = ~pal_mode(leg_mode), weight = 5, opacity = 0.85,
                 label = ~sprintf("%s - %s leg, %.0f min%s", j, leg_mode, minutes,
                                  ifelse(nzchar(route), paste0(" (service ", route, ")"), "")),
                 popup = ~sprintf("<b>%s</b><br>%s minutes in total", j, tot))
}

mleg %>%
  addLayersControl(overlayGroups = sort(unique(legs$journey)),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  addLegend(pal = pal_mode, values = c("BUS", "WALK"), title = "Leg",
            position = "bottomright", opacity = 0.9)
```

```{r tbl-journeys}
#| tbl-cap: "The same seven journeys, summarised."
rg$summary %>%
  transmute(Journey = journey,
            `Total minutes` = total_minutes,
            Legs = legs,
            `Bus services used` = ifelse(nzchar(services), services, "walk only")) %>%
  arrange(`Total minutes`) %>%
  knitr::kable()
```

Woodingdean to Longhill takes 11 minutes. Whitehawk to Dorothy Stringer
takes 52, for a journey of about the same straight-line distance. The
network is not symmetric, and neither is opportunity.

## Which school is actually nearest {#sec-nearest}

```{r fig-nearest}
#| fig-cap: "The nearest secondary school to each neighbourhood by routed walk-and-bus time, regardless of catchment. Turn on the catchment boundaries to see where the two disagree."
#| fig-height: 6

near <- lsoa %>%
  left_join(acc$lsoa %>% select(lsoa, nearest_school, nearest_min),
            by = c("lsoa21cd" = "lsoa")) %>%
  filter(!is.na(nearest_school))

pal_near <- colorFactor("Set3", domain = sort(unique(near$nearest_school)))

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = near, group = "Nearest school",
              fillColor = ~pal_near(nearest_school), fillOpacity = 0.72,
              color = "white", weight = 0.4,
              label = ~sprintf("%s: %s, %.0f min", lsoa21nm, nearest_school, nearest_min)) %>%
  # This is the map where the overlay earns its place: the colours are
  # the nearest school and the outlines are the catchment a child is
  # actually assigned to, so any mismatch is visible directly.
  addPolygons(data = catch, group = "Catchment boundaries",
              fill = FALSE, color = "#111111", weight = 2.2, opacity = 0.85,
              label = ~unname(CATCH_LABELS[catchment])) %>%
  addLayersControl(overlayGroups = c("Nearest school", "Catchment boundaries"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("Catchment boundaries") %>%
  addLegend(pal = pal_near, values = near$nearest_school, title = "Nearest school",
            position = "bottomright", opacity = 0.85)
```

```{r nearest-stats}
#| include: false
near_tab <- acc$lsoa %>%
  count(nearest_school, wt = Oi, name = "children") %>%
  arrange(desc(children)) %>%
  mutate(pct = 100 * children / sum(children))
```

This map is worth comparing with the catchment map in section 2. They
are not the same shape. The nearest school by bus is not always the
catchment school, and for
`r fmt_pct(100 * mean(acc$lsoa$nearest_min > 25), 0)`
of neighbourhoods even the nearest school is more than 25 minutes away.

## Potential accessibility: how much school is within reach {#sec-potential}

Nearest-school distance answers a narrow question. It says nothing about
whether a neighbourhood has *one* school within reach or *five* --- and
that difference is the whole of what a family's realistic options
amount to when their first preference is full.

Two measures are used here, deliberately, because they fail in opposite
directions.

```{r acc-map-data}
#| include: false

acc_map <- lsoa_lite %>%
  inner_join(acc$lsoa, by = c("lsoa21cd" = "lsoa"))

NOW_LAB <- "Today (Longhill at Ovingdean)"
ELM_SCN <- "With Longhill at Elm Grove"

# Both measures and both scenarios share one label, so whichever layer a
# reader is on they can see all four numbers for the neighbourhood under
# the cursor.
acc_lab <- sprintf(
  paste0("%s<br>gravity index %.0f today, %.0f at Elm Grove",
         "<br>%.0f places within 30 min today, %.0f at Elm Grove",
         "<br>nearest school %s, %.0f min"),
  acc_map$lsoa21nm, acc_map$A_index, acc_map$A_index_elm,
  acc_map$places_30, acc_map$places_30_elm,
  acc_map$nearest_school, acc_map$nearest_min) %>%
  lapply(htmltools::HTML)

# Each pair of scenario layers shares a colour scale, so switching
# between them shows a real change rather than a rescaling.
pal_grav <- colorNumeric("viridis", reverse = TRUE,
                         domain = c(acc_map$A_index, acc_map$A_index_elm))
pal_cum  <- colorNumeric("viridis", reverse = TRUE,
                         domain = c(acc_map$places_30, acc_map$places_30_elm))

acc_base <- function() {
  leaflet(width = "100%", height = 560,
          options = leafletOptions(preferCanvas = TRUE)) %>%
    addProviderTiles(providers$CartoDB.Positron)
}

#' Two-scenario choropleth with a radio control to switch between them,
#' and the catchment boundaries as an optional overlay.
acc_scenario_map <- function(col_now, col_elm, pal, legend_title) {
  acc_base() %>%
    addPolygons(data = acc_map, group = NOW_LAB,
                fillColor = pal(acc_map[[col_now]]),
                fillOpacity = 0.8, color = "white", weight = 0.3,
                label = acc_lab) %>%
    addPolygons(data = acc_map, group = ELM_SCN,
                fillColor = pal(acc_map[[col_elm]]),
                fillOpacity = 0.8, color = "white", weight = 0.3,
                label = acc_lab) %>%
    # Drawn over the choropleth rather than filled, so the surface stays
    # readable underneath. Off by default: the point of these maps is
    # that reachability does not follow the boundaries.
    addPolygons(data = catch, group = "Catchment boundaries",
                fill = FALSE, color = "#111111", weight = 2.2, opacity = 0.85,
                label = ~unname(CATCH_LABELS[catchment])) %>%
    addLayersControl(baseGroups = c(NOW_LAB, ELM_SCN),
                     overlayGroups = "Catchment boundaries",
                     options = layersControlOptions(collapsed = FALSE)) %>%
    hideGroup("Catchment boundaries") %>%
    addLegend(pal = pal, values = c(acc_map[[col_now]], acc_map[[col_elm]]),
              title = legend_title, position = "bottomright", opacity = 0.85)
}
```

### First, what makes a school attractive? {#sec-attractiveness}

Both measures need a weight for each school --- some number saying how
much school is there. The choice is not obvious, and it is worth setting
out before the maps rather than burying it in a footnote afterwards.

There are four candidates, each answering a different question.

| Specification | What it measures | Question it answers |
|:--|:--|:--|
| **Admission number** | places the school is allowed to fill | What did the authority decide to provide? |
| **First preferences** | families naming it first | What did families ask for? |
| **Places allocated** | offers actually made | What did the system deliver? |
| **Weighted preferences** | ranks 1--3 with geometric decay | What did families want, allowing for the ordering the rules force on them? |

The **admission number** is the obvious choice and the weakest. It is an
administrative decision, revised occasionally by the authority, and it
says nothing about whether anyone wants the places. Longhill has the
third-largest admission number in the city and the fewest first
preferences of any school; on a PAN weighting it counts as one of
Brighton's more attractive destinations, which is plainly wrong.

**First preferences** fix that but introduce a different distortion. In
the two paired catchments a family has to rank the two local schools
against each other, so the catchment's first preferences are split
between them while its second preferences accumulate. The signature is
unmistakable in the data, and it means counting firsts alone marks down
precisely the schools whose catchment obliges families to put them
second.

**Places allocated** measures what happened rather than what was wanted.
That makes it a poor attractiveness measure for exactly the schools this
document is about: an undersubscribed school allocates every place it is
asked for, so allocation tracks capacity rather than desire at the
bottom of the range.

**Weighted preferences** --- counting all three ranks with a geometric
decay of $\alpha = `r acc$pref$decay`$ --- is what the maps below use. It
keeps the information in the second and third choices without treating
them as equal to a first.

```{r tbl-attractiveness}
#| tbl-cap: "The four specifications, normalised so each averages 1 across the eleven schools. The ratio of second to first preferences separates the paired catchments from the single-school ones almost perfectly."

acc$attract %>%
  left_join(oi$schools %>% select(name, catchment), by = "name") %>%
  mutate(kind = case_when(
           is.na(catchment) ~ "Faith, city-wide",
           catchment %in% c("DS_Varndean", "Hove_Blatch") ~ "Paired",
           TRUE ~ "Single-school")) %>%
  arrange(desc(W_pref)) %>%
  transmute(School = name, Catchment = kind,
            `2nd / 1st` = ifelse(imputed, "--", sprintf("%.2f", p2 / p1)),
            `Admission no.` = sprintf("%.2f", W_pan),
            `1st prefs` = sprintf("%.2f", W_p1),
            `Allocated` = sprintf("%.2f", W_alloc),
            `Weighted prefs` = sprintf("%.2f", W_pref)) %>%
  knitr::kable(align = "llrrrrr")
```

The four schools in paired catchments have second-to-first ratios
between 1.0 and 2.0; the four in single-school catchments, between 0.4
and 0.6.

```{r fig-attractiveness}
#| fig-cap: "The four specifications compared. Schools are ordered by the weighted-preference weight. Where the points for a school spread out, the choice of specification matters for that school."
#| fig-height: 5.6

wlong <- acc$attract %>%
  select(name, W_pan, W_p1, W_alloc, W_pref) %>%
  pivot_longer(-name, names_to = "spec", values_to = "w") %>%
  mutate(spec = factor(spec, c("W_pan", "W_p1", "W_alloc", "W_pref"),
                       c("Admission number", "First preferences",
                         "Places allocated", "Weighted preferences")))

ord <- acc$attract %>% arrange(W_pref) %>% pull(name)

wlong %>%
  mutate(name = factor(name, ord)) %>%
  ggplot(aes(w, name)) +
  geom_line(aes(group = name), colour = "grey72", linewidth = 0.8) +
  geom_point(aes(colour = spec), size = 2.8) +
  geom_vline(xintercept = 1, linetype = "31", colour = "grey45") +
  scale_colour_manual(values = c("Admission number"    = "#999999",
                                 "First preferences"   = "#2166ac",
                                 "Places allocated"    = "#4daf4a",
                                 "Weighted preferences" = "#b2182b"),
                      name = NULL) +
  labs(x = "Attractiveness weight (1 = city average)", y = NULL,
       title = "The four specifications broadly agree, and disagree where it matters",
       subtitle = "Each school's four weights, joined by a grey line. The dashed line is the city average.",
       caption = "Sources: BHCC allocation factsheets; published admission numbers.") +
  theme_bh(11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

```{r w-stats}
#| include: false
wsp <- acc$attract %>%
  mutate(spread = pmax(W_pan, W_p1, W_alloc, W_pref) -
                  pmin(W_pan, W_p1, W_alloc, W_pref)) %>%
  arrange(desc(spread))
w_off <- acc$w_cor[upper.tri(acc$w_cor)]   # the six off-diagonal pairs
```

The lines are mostly short: for most schools the four specifications
agree closely, and the choice would not matter much. They spread
furthest for **`r and_list(head(wsp$name, 3))`**, for three different
reasons --- Longhill because its admission number is generous relative
to any measure of demand, Varndean because the paired catchment splits
its first preferences with Dorothy Stringer, and Cardinal Newman because
it admits city-wide on faith criteria, so what it is asked for and what
it allocates diverge from what it is sized for.

::: {.callout-note appearance="simple"}
## And the surface barely notices

It is worth knowing how little of this propagates. Recomputing the
accessibility surface under each of the four weightings gives
neighbourhood-level rankings that correlate between
`r sprintf("%.3f", min(w_off))` and
`r sprintf("%.3f", max(w_off))` across all six pairs. The weighting
matters a great deal for what you say about an individual school and
very little for the map, because the measure sums over eleven of them.

Two limitations do matter more than the choice between the four.
**Peacehaven Community School is in Lewes district** and does not appear
in the council's factsheets, so it has no preference or allocation data;
its weight is imputed from its admission number at the city-average
rate. And none of these counts is divided by the admission number,
because what should be within reach is *places families want*, which
scales with the size of the school.
:::

### Measure one: gravity accessibility {#sec-acc-gravity}

$$A_i = \sum_j W_j \, c_{ij}^{-\beta}$$

| Term | Meaning |
|:--|:--|
| $A_i$ | the accessibility of neighbourhood $i$ --- the quantity mapped below |
| $j$ | a school; the sum runs over all `r nrow(acc$attract)` in the study area |
| $W_j$ | the weighted-preference attractiveness set out above |
| $c_{ij}$ | routed walk-and-bus minutes from $i$ to $j$, weekday morning |
| $\beta$ | distance decay: how sharply demand falls away with travel time, set at `r sprintf("%.1f", acc$beta$ref)` |

Every school contributes to every neighbourhood's score, discounted by
how far away it is. A neighbourhood with one school ten minutes away and
nothing else scores lower than one with three schools at fifteen
minutes, which is the point of the measure. The result has no units, so
it is indexed with today's child-weighted city average at 100.

::: {.callout-note appearance="simple"}
## Where $\beta$ comes from

$\beta$ is set at **`r sprintf("%.1f", acc$beta$ref)`**. This is a
judgement rather than an estimate: calibrating it properly needs
pupil-level flows, which the council has not released. It is a
reasonable central value for a system of this kind --- journeys are
short and the alternatives sit close together --- and it lies inside the
`r sprintf("%.1f", acc$beta$lo)`--`r sprintf("%.1f", acc$beta$hi)` range
the open model sweeps in section 7. Nothing in this section turns on the
exact figure; the surface is stored at the ends of that range as well,
and the ordering of neighbourhoods is stable across it.
:::

```{r fig-acc-gravity}
#| fig-cap: "Gravity accessibility to secondary school places. Indexed so today's child-weighted city average is 100, and both scenarios use that same base, so the Elm Grove layer can be read against it directly."
#| fig-height: 6

acc_scenario_map("A_index", "A_index_elm", pal_grav,
                 "Gravity index<br>(today's city average = 100)")
```

### Measure two: cumulative opportunity {#sec-acc-cumulative}

$$P_i(t) = \sum_j \mathrm{PAN}_j \cdot \mathbf{1}\!\left(c_{ij} \le t\right)$$

| Term | Meaning |
|:--|:--|
| $P_i(t)$ | school places reachable from neighbourhood $i$ inside $t$ minutes |
| $j$ | a school; the sum again runs over all `r nrow(acc$attract)` |
| $\mathrm{PAN}_j$ | school $j$'s admission number --- the places it actually has |
| $\mathbf{1}(\cdot)$ | an indicator: 1 if the journey is within $t$ minutes, 0 if not |
| $c_{ij}$ | routed walk-and-bus minutes from $i$ to $j$, as above |
| $t$ | the time threshold, set with the slider below |

Where the gravity measure discounts smoothly, this one cuts off. A
school 29 minutes away counts in full; the same school at 31 minutes
counts for nothing. That is crude, but it is also how a threshold
actually works for a family judging whether a journey is feasible ---
and unlike the gravity index, the number means something on its own.

The threshold is doing all the work, so it is worth being able to move
it.

```{r fig-acc-cumulative}
#| fig-cap: "School places reachable within the chosen number of minutes by walking and bus. Drag the slider to change the threshold, switch Longhill between its current site and Elm Grove, and toggle whether the figure is divided by the number of children living there. Darker means fewer places within reach; the darkest areas can reach none at all."
#| fig-height: 7.4

# The polygons are drawn in JavaScript rather than by addPolygons(),
# because the slider has to restyle them on every move. Going through R
# would mean reaching into leaflet's own layer registry, which is an
# undocumented internal; building the layer here means the handles are
# ours and the interaction is straightforward.
#
# Colours are precomputed in R for every combination of threshold,
# scenario and weighting, so the browser only swaps fills.

tl <- acc$thresh_long %>% mutate(scenario = factor(scenario, c("now", "elm")))

pal_t  <- colorNumeric("viridis", domain = c(0, max(tl$places)), reverse = TRUE)
# Places per child are heavily right-skewed, so that scale is built on
# the square root; without it nearly every neighbourhood sits in the
# bottom colour and the map says nothing.
pal_pc <- colorNumeric("viridis", domain = c(0, sqrt(max(tl$per_child))),
                       reverse = TRUE)

tl <- tl %>% mutate(col = pal_t(places), col_pc = pal_pc(sqrt(per_child)))

as_lookup <- function(col) {
  split(tl, tl$scenario) %>%
    lapply(function(s) split(s, s$t) %>%
             lapply(function(d) setNames(as.list(d[[col]]), d$lsoa)))
}

# A coarser simplification than the leaflet maps use, because this
# geometry is embedded in the page as JSON. 60 m is still finer than the
# rendered pixels at any zoom this map offers.
geo_js <- lsoa %>%
  st_transform(27700) %>%
  st_simplify(dTolerance = 60, preserveTopology = TRUE) %>%
  st_transform(4326) %>%
  select(lsoa21cd, lsoa21nm)

lh_pt <- sch_pts[sch_pts$short == "longhill", ]

payload <- list(
  geo        = jsonlite::fromJSON(geojsonsf::sf_geojson(geo_js),
                                  simplifyVector = FALSE),
  thresholds = acc$thresh_grid,
  colours    = as_lookup("col"),
  values     = as_lookup("places"),
  colours_pc = as_lookup("col_pc"),
  values_pc  = as_lookup("per_child"),
  city = split(acc$thresh_city, acc$thresh_city$scenario) %>%
    lapply(function(s) split(s, s$t) %>%
             lapply(function(d) list(mean = round(d$mean_places[1]),
                                     stranded = d$stranded[1],
                                     children = round(d$stranded_children[1])))),
  longhill = list(now = list(lat = lh_pt$lat, lng = lh_pt$lon),
                  elm = list(lat = elm_xy[, 2], lng = elm_xy[, 1])),
  catchments = jsonlite::fromJSON(
    geojsonsf::sf_geojson(catch %>% select(catchment)),
    simplifyVector = FALSE))

m_cum <- leaflet(width = "100%", height = 580) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  fitBounds(lng1 = min(sch_pts$lon) - 0.03, lat1 = min(sch_pts$lat) - 0.02,
            lng2 = max(sch_pts$lon) + 0.03, lat2 = max(sch_pts$lat) + 0.02) %>%
  addCircleMarkers(data = sch_pts %>% filter(short != "longhill"),
                   lng = ~lon, lat = ~lat, radius = 4,
                   fillColor = "#e31a1c", fillOpacity = 1,
                   color = "white", weight = 1.2, label = ~name) %>%
  addLegend(pal = pal_t, values = c(0, max(tl$places)),
            title = "Places within<br>the threshold",
            position = "bottomright", opacity = 0.85)

m_cum <- htmlwidgets::onRender(m_cum, "
function(el, x, data) {
  var map = this;

  var layers = {};
  L.geoJSON(data.geo, {
    style: function() {
      return {weight: 0.3, color: 'white', fillOpacity: 0.82};
    },
    onEachFeature: function(f, layer) {
      layers[f.properties.lsoa21cd] = layer;
      layer.bindTooltip('');
    }
  }).addTo(map);

  var lh = L.circleMarker([data.longhill.now.lat, data.longhill.now.lng],
    {radius: 6, fillColor: '#1a9850', fillOpacity: 1,
     color: 'white', weight: 2}).addTo(map)
    .bindTooltip('Longhill High School');

  // Outlines only, so the surface stays readable underneath. Kept off
  // until asked for: the point of the map is that reachability does not
  // follow the boundaries.
  var catchLayer = L.geoJSON(data.catchments, {
    style: function() {
      return {fill: false, color: '#111111', weight: 2.2, opacity: 0.85};
    },
    onEachFeature: function(f, layer) {
      layer.bindTooltip(f.properties.catchment);
    }
  });

  var ui = L.DomUtil.create('div', 'acc-slider');
  ui.style.cssText = 'background:white;padding:10px 14px;border-radius:4px;' +
    'box-shadow:0 1px 5px rgba(0,0,0,.3);font-family:sans-serif;' +
    'font-size:13px;min-width:290px';
  ui.innerHTML =
    '<div style=\"font-weight:bold;margin-bottom:6px\">Within ' +
      '<span id=\"tval\">30</span> minutes</div>' +
    '<input id=\"tslider\" type=\"range\" min=\"0\" max=\"' +
      (data.thresholds.length - 1) + '\" value=\"' +
      data.thresholds.indexOf(30) + '\" step=\"1\" style=\"width:100%\">' +
    '<div style=\"margin-top:8px\">' +
      '<label style=\"margin-right:10px\"><input type=\"radio\" name=\"scn\" ' +
        'value=\"now\" checked> Today</label>' +
      '<label><input type=\"radio\" name=\"scn\" value=\"elm\"> ' +
        'Longhill at Elm Grove</label></div>' +
    '<div style=\"margin-top:6px;padding-top:6px;border-top:1px solid #eee\">' +
      '<label><input type=\"checkbox\" id=\"perchild\"> ' +
        'Divide by children living there</label><br>' +
      '<label><input type=\"checkbox\" id=\"showcatch\"> ' +
        'Catchment boundaries</label></div>' +
    '<div id=\"citystat\" style=\"margin-top:8px;color:#444;line-height:1.45\">' +
    '</div>';

  var ctl = L.control({position: 'topright'});
  ctl.onAdd = function() { return ui; };
  ctl.addTo(map);
  L.DomEvent.disableClickPropagation(ui);
  L.DomEvent.disableScrollPropagation(ui);

  var names = {};
  data.geo.features.forEach(function(f) {
    names[f.properties.lsoa21cd] = f.properties.lsoa21nm;
  });

  function redraw() {
    var idx = +ui.querySelector('#tslider').value;
    var t = data.thresholds[idx];
    var scn = ui.querySelector('input[name=scn]:checked').value;
    var pc = ui.querySelector('#perchild').checked;
    ui.querySelector('#tval').textContent = t;

    var cols = pc ? data.colours_pc[scn][t] : data.colours[scn][t];
    var vals = pc ? data.values_pc[scn][t]  : data.values[scn][t];
    var unit = pc ? ' places per child' : ' places';

    Object.keys(layers).forEach(function(id) {
      if (cols[id] === undefined) return;
      layers[id].setStyle({fillColor: cols[id]});
      layers[id].setTooltipContent(
        names[id] + '<br>' +
        (pc ? (+vals[id]).toFixed(1) : Math.round(vals[id])) +
        unit + ' within ' + t + ' min');
    });

    lh.setLatLng([data.longhill[scn].lat, data.longhill[scn].lng]);
    lh.setTooltipContent(scn === 'elm' ?
      'Longhill (at Elm Grove)' : 'Longhill High School');

    // Restyling the fills can leave the outlines behind them.
    if (map.hasLayer(catchLayer)) catchLayer.bringToFront();

    var c = data.city[scn][t], o = data.city['now'][t];
    ui.querySelector('#citystat').innerHTML =
      '<b>' + c.mean + '</b> places for the average child' +
      (scn === 'elm' ? ' <span style=\"color:#888\">(' + o.mean +
        ' today)</span>' : '') +
      '<br><b>' + c.stranded + '</b> neighbourhoods reach none (' +
      c.children + ' children)' +
      (scn === 'elm' ? ' <span style=\"color:#888\">(' + o.stranded +
        ' today)</span>' : '');
  }

  ui.querySelector('#tslider').addEventListener('input', redraw);
  ui.querySelector('#perchild').addEventListener('change', redraw);
  ui.querySelector('#showcatch').addEventListener('change', function() {
    if (this.checked) { catchLayer.addTo(map); catchLayer.bringToFront(); }
    else { map.removeLayer(catchLayer); }
  });
  ui.querySelectorAll('input[name=scn]').forEach(function(r) {
    r.addEventListener('change', redraw);
  });
  redraw();
}
", data = payload)

m_cum
```

Two things are worth doing with that slider.

**Drag it down to 20 minutes.** The map empties. At a threshold most
parents would call a reasonable school-run, much of the city can reach
almost nothing --- and the pattern is not centred on the places the
admissions debate concentrates on.

**Switch scenarios at different thresholds.** Below about 30 minutes,
moving Longhill to Elm Grove is unambiguously better for the city. Above
about 35 it stops being so: the far south east loses the only school it
could previously reach at all, and neighbourhoods start appearing in the
"reaching none" count that were not there before. The relocation is not
a free improvement --- it is a trade that is strongly favourable at short
thresholds and turns against a small number of places at long ones.

```{r acc-disagree}
#| include: false
dec <- acc$lsoa %>%
  mutate(d_grav = ntile(A_index, 10), d_cum = ntile(places_30, 10),
         diff = d_cum - d_grav)
mean_dec <- mean(abs(dec$diff))
# The sharpest kind of disagreement: comfortable on the gravity measure,
# nothing at all reachable on the cumulative one.
stranded <- dec %>% filter(places_30 < 1, d_grav >= 6)
```

The two maps agree on the broad shape of the city --- they correlate at
`r sprintf("%.2f", acc$agreement$spearman)` --- and disagree
substantially in the detail, which is exactly what makes showing both
worthwhile. A neighbourhood sits an average of
**`r sprintf("%.1f", mean_dec)` deciles apart** between the two, and some
sit six deciles apart.

The disagreement is not random. The cumulative map has a hard edge the
gravity map does not: **`r nrow(stranded)` neighbourhoods score in the
better half of the city on the gravity measure and can still reach no
school place at all within 30 minutes.** Their nearest school sits a few
minutes the wrong side of the threshold, so the gravity measure --- which
discounts smoothly rather than cutting off --- records them as adequately
served while a parent with a bus timetable would not.

Which map is right depends on the question. If you are asking how much
school is notionally within reach, the gravity surface is the better
description. If you are asking whether a child can get to a school in a
morning, the threshold is the thing that matters, and the cumulative map
is showing you something real that the gravity map hides.

### What moving one school would do to the whole city {#sec-elm-access}

Both maps carry a second layer placing Longhill at Elm Grove. Because
accessibility sums over every school, moving one of them changes the
figure for every neighbourhood --- not only for Longhill's own catchment.

```{r tbl-elm-city}
#| tbl-cap: "City-wide accessibility today and with Longhill relocated to Elm Grove. All figures are weighted by where cohort-aged children live."

ct <- acc$city
tibble::tribble(
  ~Measure, ~Today, ~`At Elm Grove`, ~Change,
  "Gravity accessibility (index)",
    sprintf("%.0f", 100), sprintf("%.1f", 100 * ct$A_elm / ct$A_now),
    sprintf("%+.1f%%", ct$A_pct_change),
  "Places reachable within 30 minutes",
    sprintf("%.0f", ct$p30_now), sprintf("%.0f", ct$p30_elm),
    sprintf("%+.0f", ct$p30_elm - ct$p30_now),
  "Journey to the nearest school (minutes)",
    sprintf("%.1f", ct$near_now), sprintf("%.1f", ct$near_elm),
    sprintf("%+.1f", ct$near_elm - ct$near_now),
  "Neighbourhoods reaching no place in 30 minutes",
    sprintf("%d", ct$zero30_now), sprintf("%d", ct$zero30_elm),
    sprintf("%+d", ct$zero30_elm - ct$zero30_now),
  "Children in those neighbourhoods",
    sprintf("%.0f", ct$zero30_children_now), sprintf("%.0f", ct$zero30_children_elm),
    sprintf("%+.0f", ct$zero30_children_elm - ct$zero30_children_now)
) %>%
  knitr::kable(align = "lrrr")
```

The two measures disagree about how much this matters, and the
disagreement is informative.

**On the gravity measure the effect is small** ---
`r sprintf("%+.1f%%", ct$A_pct_change)` city-wide. That is because
Longhill carries a low attractiveness weight: moving an unpopular school
closer to people does not add much *wanted* school to the map.

**On the cumulative measure it is substantial.** Places reachable within
half an hour rise from `r sprintf("%.0f", ct$p30_now)` to
`r sprintf("%.0f", ct$p30_elm)` for the average child, and the number of
neighbourhoods that can reach **nothing** inside 30 minutes falls from
`r ct$zero30_now` to `r ct$zero30_elm`. In children, that is
**`r sprintf("%.0f", ct$zero30_children_now - ct$zero30_children_elm)`
who would gain a reachable school place they do not currently have.**

`r ct$better` of the `r ct$n` neighbourhoods are better off under the
move and `r ct$worse` are worse off --- the latter being the far south
east, which loses its closest school. This is a genuine trade, but it is
a lopsided one, and it is the strongest argument in this document for
relocation on grounds that have nothing to do with Longhill's own roll.

```{r acc-stats}
#| include: false
zero30 <- acc$lsoa %>% filter(places_30 < 1)
rng <- range(acc$lsoa$A_index)
```

Three findings come out of this, and the first is the one that should
change the conversation.

**`r nrow(zero30)` of the `r nrow(acc$lsoa)` neighbourhoods --- about
`r fmt_n(sum(zero30$Oi))` cohort-aged children --- cannot reach a single
secondary school place within 30 minutes** by walking and bus on a
weekday morning. Not their catchment school; *any* school. For those
neighbourhoods the admissions debate is somewhat beside the point.

**The spread is nearly fivefold.** The best-served neighbourhood scores
`r fmt_n(rng[2])` on the index against `r fmt_n(rng[1])` for the worst.
A child in the centre of the city has several schools genuinely within
reach; a child at either end has one, or none.

**The two measures broadly agree, but not closely.** They correlate at
`r sprintf("%.2f", acc$agreement$spearman)` (Spearman), yet only
`r fmt_pct(100 * acc$agreement$within_one, 0)` of neighbourhoods land
within one decile of each other on both. Where they agree, the finding
is robust to how you define access. Where they disagree, it is because
one is counting a school just over the threshold that the other is
discounting smoothly.

```{r fig-acc-compare}
#| fig-cap: "The two measures against each other. Neighbourhoods along the bottom can reach no places at all inside 30 minutes."
#| fig-height: 4.8

acc$lsoa %>%
  mutate(grp = ifelse(places_30 < 1, "No places within 30 min", "Some places within 30 min")) %>%
  ggplot(aes(A_index, places_30, colour = grp, size = Oi)) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(values = c("No places within 30 min" = "#b2182b",
                                 "Some places within 30 min" = "#2166ac"),
                      name = NULL) +
  scale_size_area(max_size = 6, guide = "none") +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Gravity accessibility index (city average = 100)",
       y = "Places reachable within 30 minutes",
       title = "Two ways of asking the same question",
       subtitle = "Each point is a neighbourhood, sized by the number of cohort-aged children",
       caption = "Routed walk-and-bus times, weekday morning arrival.") +
  theme_bh()
```


::: {.callout-warning appearance="simple"}
## Two caveats that bear on these numbers

**Journey times are modelled, not measured.** They come from `r5r`
routing over the merged street network and published timetables for one
weekday morning. They do not know about school buses, parental lifts,
walking routes children actually use, or a bus that is full when it
arrives.

**One school's times are suspect.** Dorothy Stringer and Varndean are
470 metres apart, yet the routed matrix makes Stringer slower from most
of the city by a median of about five and a half minutes. That may be
real, or it may be an artefact of where the router snapped each school
onto the network. It has not been established which, and it inflates the
apparent inaccessibility of the areas whose nearest school is Stringer.
:::

## Accessibility and child poverty together {#sec-bivariate}

The question that matters for policy is not where access is poor, nor
where deprivation is high, but where the two coincide --- because those
are the neighbourhoods where a difficult journey is least likely to be
solved by a car in the family.

```{r fig-bivariate}
#| fig-cap: "Accessibility and income deprivation affecting children, mapped together. The darkest blue is the corner that matters: least reachable school places, highest child poverty."
#| fig-height: 7

# Stevens-style bivariate palette. Keyed "access-deprivation", each on a
# tercile: access 1 is the least reachable third, deprivation 3 the most
# deprived third. The concerning corner is therefore 1-3, and it is the
# darkest colour.
BIV <- c(
  "1-1" = "#be64ac", "1-2" = "#8c62aa", "1-3" = "#3b4994",
  "2-1" = "#dfb0d6", "2-2" = "#a5add3", "2-3" = "#5698b9",
  "3-1" = "#e8e8e8", "3-2" = "#ace4e4", "3-3" = "#5ac8c8")

biv_map <- lsoa %>%
  inner_join(acc$bivariate %>% select(lsoa, acc_t, dep_t, biv_key,
                                      idaci_score, A_index, places_30, Oi),
             by = c("lsoa21cd" = "lsoa")) %>%
  mutate(fill = unname(BIV[biv_key]))

main <- ggplot(biv_map) +
  geom_sf(aes(fill = fill), colour = "white", linewidth = 0.12) +
  scale_fill_identity() +
  labs(title = "Where poor access and child poverty coincide",
       subtitle = "Each neighbourhood classed on both axes at once",
       caption = "Sources: routed walk-and-bus accessibility; IDACI 2019.") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey30"),
        plot.caption = element_text(colour = "grey45", hjust = 0))

key <- expand_grid(acc_t = 1:3, dep_t = 1:3) %>%
  mutate(fill = unname(BIV[paste0(acc_t, "-", dep_t)])) %>%
  ggplot(aes(acc_t, dep_t, fill = fill)) +
  geom_tile(colour = "white", linewidth = 0.7) +
  scale_fill_identity() +
  scale_x_continuous(breaks = c(1, 3), labels = c("least", "most"),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = c(1, 3), labels = c("least", "most"),
                     expand = c(0, 0)) +
  labs(x = "Accessible →", y = "Deprived →") +
  coord_equal() +
  theme_minimal(base_size = 8) +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 7.5, colour = "grey25"),
        axis.text = element_text(size = 6.5, colour = "grey40"))

main + inset_element(key, left = 0.72, bottom = 0.02, right = 0.99, top = 0.32)
```

```{r biv-stats}
#| include: false
corner <- acc$worst_corner
best_corner <- acc$bivariate %>% filter(acc_t == 3, dep_t == 3)
cells <- acc$bivariate %>%
  count(acc_t, dep_t, wt = Oi, name = "children")
low_acc_high_dep <- cells %>% filter(acc_t == 1, dep_t == 3) %>% pull(children)
hi_acc_high_dep  <- cells %>% filter(acc_t == 3, dep_t == 3) %>% pull(children)
```

There is a real gradient here, and it runs the wrong way. Accessibility
and child poverty correlate at
`r sprintf("%.2f", acc$acc_dep_spearman)` (Spearman): the more deprived
a neighbourhood, the less school it can reach.

**`r nrow(corner)` neighbourhoods --- about `r fmt_n(low_acc_high_dep)`
cohort-aged children --- sit in the worst corner**, in the least
accessible third of the city and the most deprived third at the same
time. At the other end, only `r nrow(best_corner)` neighbourhoods
(`r fmt_n(hi_acc_high_dep)` children) combine high deprivation with good
access.

This is the strongest argument in the document for treating transport as
a school-admissions instrument rather than a separate departmental
concern. A city that wanted to widen real choice for its poorest
families could do more with a bus timetable than with a catchment
boundary --- and unlike a boundary change, it would take nothing away
from anyone else.

::: {.callout-note appearance="simple"}
## What pupil-level data would add here

This section works entirely with modelled journey times between
neighbourhood centroids and school gates. The council holds the home
postcode of every applicant and the school each was allocated. With
that, the same analysis could be run on **journeys children actually
make** rather than journeys they could in principle make --- and the gap
between the two is precisely the quantity that would tell the city
whether its transport network is a real constraint on choice or merely a
theoretical one. Section 9 sets out what that request would look like.
:::

# Admissions rules, and what families actually do {#sec-demand}

## How a place is allocated {#sec-rules}

Families name up to three schools in order. Each school then ranks its
applicants against its own criteria, and the authority runs an
equal-preference matching so that a child is offered the highest-ranked
school that will take them. The catchment criterion sits partway down
most schools' lists, below looked-after children and siblings and, since
2023, below a quota for children eligible for free school meals.

The critical point is one the rules themselves obscure. **A criterion
only does something if a school has to turn someone away.**

```{r fig-rationing}
#| fig-cap: "For each school, the proportion of first-preference applicants who were offered a place. A school at 100% turned nobody away, so its admission criteria did no work at all."
#| fig-height: 5

conv <- bh_data("adjudicator_conversion.rds")

rat <- conv$conv %>%
  group_by(school) %>%
  summarise(p1_named = sum(p1_named), p1_offered = sum(p1_offered),
            .groups = "drop") %>%
  mutate(rate = p1_offered / p1_named,
         rations = rate < 0.999) %>%
  arrange(rate)

ggplot(rat, aes(reorder(school, rate), rate, fill = rations)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, colour = "grey40", linetype = "31") +
  geom_text(aes(label = fmt_pct(100 * rate, 0)), hjust = -0.15, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1.12),
                     breaks = seq(0, 1, 0.25)) +
  scale_fill_manual(values = c(`TRUE` = "#b2182b", `FALSE` = "#4d9221"),
                    labels = c(`TRUE` = "Rations places",
                               `FALSE` = "Turns nobody away"), name = NULL) +
  labs(x = NULL, y = "First preferences that received an offer",
       title = "Only some of these schools ration anything",
       subtitle = "Pooled across three admissions rounds",
       caption = "Source: BHCC evidence to the Schools Adjudicator.") +
  theme_bh() +
  theme(panel.grid.major.y = element_blank())
```

```{r rationing-stats}
#| include: false
n_open  <- sum(!rat$rations)
n_ration <- sum(rat$rations)
open_names <- rat %>% filter(!rations) %>% pull(school)
```

**`r n_open` of the `r nrow(rat)` schools offered a place to every single
first-preference applicant.** For those schools --- `r paste(open_names, collapse = ", ")`
--- the catchment criterion, the sibling criterion and the free school
meals quota are all inert. They admit everyone who asks.

The city therefore runs two admissions systems under one set of rules.
In `r n_ration` schools the criteria bind and the lottery matters. In
the other `r n_open` they are decoration. Debate about admissions
criteria is, in practice, debate about a minority of the city's schools
--- and never about the schools with the most spare capacity.

## Sixteen years of demand {#sec-demand-history}

```{r fig-demand-history}
#| fig-cap: "First preferences expressed for each school, 2010 onwards. The two church schools admit city-wide and are shown separately."
#| fig-height: 6.5

fp <- bh_data("factsheet_panel.rds")

fs <- fp$factsheets %>%
  filter(!is.na(pref1), name != "Total") %>%
  mutate(kind = ifelse(name %in% c("King's School", "Cardinal Newman Catholic School"),
                       "Faith schools (city-wide)", "Catchment schools"))

ggplot(fs, aes(year, pref1, colour = name)) +
  geom_line(linewidth = 0.95) +
  geom_point(size = 1.4) +
  facet_wrap(~ kind, ncol = 1, scales = "free_y") +
  scale_colour_brewer(palette = "Paired", name = NULL) +
  labs(x = NULL, y = "First preferences",
       title = "Demand has moved a long way between schools",
       subtitle = "First preferences expressed, by school",
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh() +
  theme(legend.position = "right", legend.text = element_text(size = 8))
```

## Who gained and who lost {#sec-winners-losers}

```{r fig-demand-change}
#| fig-cap: "Change in first preferences between the first and last five-year windows in the published series."
#| fig-height: 5

yr_rng <- range(fs$year)
early <- fs %>% filter(year <= yr_rng[1] + 4) %>%
  group_by(name) %>% summarise(early = mean(pref1), .groups = "drop")
late  <- fs %>% filter(year >= yr_rng[2] - 4) %>%
  group_by(name) %>% summarise(late = mean(pref1), .groups = "drop")

chg <- inner_join(early, late, by = "name") %>%
  mutate(change = late - early, pct = 100 * change / early) %>%
  arrange(change)

ggplot(chg, aes(reorder(name, change), change, fill = change > 0)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, colour = "grey30") +
  coord_flip() +
  scale_fill_manual(values = c(`TRUE` = "#2166ac", `FALSE` = "#b2182b"),
                    guide = "none") +
  labs(x = NULL, y = "Change in average first preferences",
       title = "The winners and losers of sixteen years of choice",
       subtitle = sprintf("Average of %d-%d against average of %d-%d",
                          yr_rng[1], yr_rng[1] + 4, yr_rng[2] - 4, yr_rng[2]),
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh() +
  theme(panel.grid.major.y = element_blank())
```

```{r chg-stats}
#| include: false
biggest_loss <- chg %>% slice_min(change, n = 2)
biggest_gain <- chg %>% slice_max(change, n = 2)
```

The largest falls are at **`r paste(biggest_loss$name, collapse = " and ")`**;
the largest rises at **`r paste(biggest_gain$name, collapse = " and ")`**.
This is not a system in equilibrium. It is one where a shrinking cohort
is being redistributed towards a small number of schools, and away from
others, faster than the cohort itself is shrinking.

## What families are actually choosing on {#sec-choice}

The council's admissions guide points families at published performance
data. It is worth asking which published number they respond to.

```{r fig-choice}
#| fig-cap: "First preferences per place against headline attainment and against value added. Each point is a school."
#| fig-height: 4.8

os <- bh_data("open_scenarios.rds")
ch <- os$choice

p1 <- ggplot(ch, aes(att8, prefs_per_place)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey35", linewidth = 0.7, linetype = "22") +
  geom_point(colour = "#b2182b", size = 3) +
  ggrepel::geom_text_repel(aes(label = school), size = 2.7, seed = 1,
                           max.overlaps = 20) +
  labs(x = "Attainment 8 (headline)", y = "First preferences per place",
       subtitle = sprintf("Headline attainment (R² = %.2f)",
                          os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"])) +
  theme_bh(11)

p2 <- ggplot(ch, aes(va, prefs_per_place)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey35", linewidth = 0.7, linetype = "22") +
  geom_point(colour = "#2166ac", size = 3) +
  ggrepel::geom_text_repel(aes(label = school), size = 2.7, seed = 1,
                           max.overlaps = 20) +
  labs(x = "Value added (Lever report)", y = NULL,
       subtitle = sprintf("Value added (R² = %.2f)",
                          os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"])) +
  theme_bh(11)

(p1 | p2) +
  plot_annotation(
    title = "Families track the headline score, not the school's contribution",
    caption = "Sources: BHCC factsheets; DfE performance tables; How to Pull the Right Lever.",
    theme = theme_bh())
```

```{r tbl-choice}
#| tbl-cap: "Explaining first preferences per place."
os$choice_tbl %>%
  transmute(Specification = spec,
            `R²` = sprintf("%.2f", r2),
            Coefficients = terms) %>%
  knitr::kable()
```

The result is stark. Headline Attainment 8 explains
**`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"], 0)`**
of the variation in how heavily each school is preferred. The
value-added measure --- the one that actually isolates what the school
contributes --- explains
**`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"], 0)`**.
Adding value added to the attainment model barely moves the fit, and its
coefficient turns *negative*.

Families are choosing on a number that is mostly a description of the
existing intake. That makes school choice partly self-fulfilling: a
school with a difficult intake posts a low headline score, is chosen
less, and receives a still more difficult intake next year. Section 2
showed that Hove Park and BACA are among the schools most penalised by
this conflation. This is a communications problem before it is an
admissions problem, and the authority is not a neutral party in it,
because the authority publishes the guide.

::: {.callout-note appearance="simple"}
## What pupil-level data would add here

This is a school-level association across ten schools --- ten points.
It cannot distinguish families responding to attainment from families
responding to something correlated with it, and it cannot say whether
different kinds of family respond differently. The council holds the
preference ordering of every applicant. With it, the same question
becomes a discrete-choice model over thousands of decisions, and the
answer would be worth acting on rather than merely worth noting.
:::

# Money {#sec-money}

A school's finances follow its roll, because funding is overwhelmingly
per-pupil. A school losing children is therefore losing income against a
cost base that does not shrink at the same rate --- the building, the
leadership team and the curriculum offer are all substantially fixed.

```{r fig-size-finance}
#| fig-cap: "Income per pupil and the proportion of schools in deficit, by size band, nationally."
#| fig-height: 4.8

sf_ <- os$size_fin %>%
  mutate(size_band = factor(size_band, levels = size_band))

pa <- ggplot(sf_, aes(size_band, income_pp)) +
  geom_col(fill = "#2166ac", width = 0.7) +
  scale_y_continuous(labels = label_dollar(prefix = "£", accuracy = 1)) +
  labs(x = NULL, y = NULL, subtitle = "Income per pupil") +
  theme_bh(11) + theme(axis.text.x = element_text(angle = 40, hjust = 1))

pb <- ggplot(sf_, aes(size_band, pct_deficit)) +
  geom_col(fill = "#b2182b", width = 0.7) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(x = NULL, y = NULL, subtitle = "Schools in deficit") +
  theme_bh(11) + theme(axis.text.x = element_text(angle = 40, hjust = 1))

(pa | pb) +
  plot_annotation(
    title = "Small schools receive more per pupil and are still likelier to be in deficit",
    subtitle = "Secondary schools in England, by roll",
    caption = "Source: DfE school income and expenditure data.",
    theme = theme_bh())
```

Small schools are not starved of money per head --- they receive
**`r fmt_n(sf_$income_pp[1] - sf_$income_pp[nrow(sf_)])` more per pupil**
than the largest band. They are nonetheless the likeliest to be in
deficit: `r fmt_pct(sf_$pct_deficit[1], 0)` of schools under 500 pupils,
against `r fmt_pct(min(sf_$pct_deficit), 0)` at the most favourable size.
Extra funding per head does not compensate for the loss of scale.

## Longhill's position {#sec-longhill-money}

```{r tbl-longhill-finance}
#| tbl-cap: "Longhill High School: roll, per-pupil income and expenditure, in-year balance and accumulated reserve."
os$lh_fin %>%
  transmute(Year = year_label,
            Roll = ifelse(is.na(roll), "--", fmt_n(roll)),
            `Income / pupil` = paste0("£", fmt_n(income_pp)),
            `Spend / pupil` = paste0("£", fmt_n(expenditure_pp)),
            `In-year balance` = paste0("£", fmt_n(balance)),
            `Reserve` = paste0("£", fmt_n(reserve))) %>%
  knitr::kable(align = "lrrrrr")
```

Longhill's roll fell from `r fmt_n(os$lh_fin$roll[1])` to
`r fmt_n(os$lh_fin$roll[nrow(os$lh_fin)])` across four years. Its
per-pupil income rose over the same period --- the funding system is
doing what it is meant to do --- and its per-pupil spending rose
faster. The accumulated reserve has fallen from
`r paste0("£", fmt_n(os$lh_fin$reserve[1]))` to
`r paste0("£", fmt_n(os$lh_fin$reserve[nrow(os$lh_fin)]))`.

At the recent rate of depletion --- about
**`r paste0("£", fmt_n(os$reserve_burn))` a year** --- the reserve
supports roughly **`r sprintf("%.1f", os$years_left)` more years**.
That is the constraint that makes this a decision with a deadline rather
than a decision that can be deferred.

# Brightopia: a model of the whole system {#sec-brightopia}

## Why model at all {#sec-why-model}

Every argument in the sections above is about one part of the system:
where the children are, how far the schools are, what families prefer.
A model is what lets you ask what happens when you change one part and
everything else adjusts. Close a school and its children do not vanish;
they appear somewhere else, displacing others. Shrink one school's
admission number and demand redistributes to its neighbours. Nothing in
a spreadsheet of admission numbers captures that.

**Brightopia** is a production-constrained spatial interaction model. It
takes the children in each neighbourhood, the attractiveness of each
school, and the travel cost between them, and predicts the flow:

$$T_{ij} = A_i O_i W_j^{\alpha} c_{ij}^{-\beta}$$

$O_i$ is the number of cohort-aged children in neighbourhood $i$; $W_j$
is school $j$'s attractiveness; $c_{ij}$ is the routed walk-and-bus
cost; $\beta$ governs how sharply demand falls away with travel time;
and $A_i$ is a balancing factor ensuring every child goes somewhere.

The model deliberately contains **no catchment term**. That is the
point. It answers the question "where would children go if only
geography and school size mattered?" --- and the gap between that answer
and reality is a measure of what the admissions system is doing.

```{r fig-brightopia}
#| fig-cap: "Modelled intake against published admission number. A school above the line is one that geography alone would fill; below it, one that geography alone would not."
#| fig-height: 5.5

bt <- brt$at_original_beta %>%
  mutate(fill_rate = modelled / pan2026,
         short = str_remove(name, " (School|High School|Community Academy|Catholic School).*"))

ggplot(bt, aes(pan2026, modelled)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey45", linetype = "31") +
  geom_point(aes(colour = fill_rate >= 1), size = 3.5) +
  ggrepel::geom_text_repel(aes(label = name), size = 2.8, seed = 1,
                           max.overlaps = 20) +
  scale_colour_manual(values = c(`TRUE` = "#2166ac", `FALSE` = "#b2182b"),
                      labels = c(`TRUE` = "Geography would fill it",
                                 `FALSE` = "Geography would not"), name = NULL) +
  scale_x_continuous(limits = c(0, NA)) + scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Published admission number, 2026", y = "Modelled intake",
       title = "What the map alone would do",
       subtitle = sprintf("Distance decay β = %.1f. The dashed line is a school exactly full.",
                          brt$beta_original),
       caption = "Brightopia, open-data specification.") +
  theme_bh()
```

```{r brightopia-stats}
#| include: false
lh_bt <- bt %>%
  filter(str_detect(name, "Longhill")) %>%
  mutate(fill_2024 = modelled / pan2024)
hp_bt <- bt %>% filter(str_detect(name, "Hove Park"))
lh_rng <- range(brt$lh_sweep$modelled_now)
```

City-wide there are `r fmt_n(brt$city_places)` places for
`r fmt_n(brt$city_children)` children --- a fill rate of
`r fmt_pct(100 * brt$city_fill, 0)` before any child expresses a
preference. The system is not short of capacity; it is short of children.

Two schools are worth reading off this chart carefully, and they point
in opposite directions.

**Longhill has already been resized to roughly what geography supports.**
Its modelled intake is **`r fmt_n(lh_bt$modelled)`** children. Against
its former admission number of `r fmt_n(lh_bt$pan2024)` that is
`r fmt_pct(100 * lh_bt$fill_2024, 0)`; against the reduced number of
`r fmt_n(lh_bt$pan2026)` now in force it is
`r fmt_pct(100 * lh_bt$fill_rate, 0)`. The distinction matters, and it
is easy to quote the wrong one. Geography can fill 210 places at
Ovingdean; it could not fill 270. The reduction has already done most of
the work that a reduction can do --- which is why section 8 is about
what remains, and why the modelled intake is remarkably insensitive to
the decay parameter, varying only between
`r fmt_n(lh_rng[1])` and `r fmt_n(lh_rng[2])` across the entire swept
range. Longhill's catchment simply does not contain many more children
to attract.

**Hove Park's problem is not its location.** Geography alone would fill
`r fmt_pct(100 * hp_bt$fill_rate, 0)` of its places --- it is the most
over-subscribed school on the map, sitting in the densest part of the
child population. Yet section 5 shows it among the largest losers of
real demand over sixteen years. A school that the map says should be
full and that families are nonetheless leaving is not a geography
problem, and no catchment or relocation policy reaches it. That gap
between what location predicts and what families do is the clearest
evidence in this document that choice is operating on something other
than distance.

::: {.callout-warning appearance="simple"}
## This is an open-data model, and it is not calibrated

$\beta$ cannot be estimated without pupil-level flows, so it is swept
across a plausible range rather than fitted. Figures here use the
reference value; the range is
`r sprintf("%.1f to %.1f", min(env$betas), max(env$betas))`, and section
8 reports the band rather than a point wherever the conclusion depends
on it. A calibrated model built on the council's own allocation records
would replace every band in this document with an estimate.
:::

# Longhill {#sec-longhill}

Longhill High School sits at Ovingdean, at the eastern edge of both its
catchment and the city. Its roll has fallen by
`r fmt_pct(100 * (os$lh_fin$roll[1] - os$lh_fin$roll[nrow(os$lh_fin)]) / os$lh_fin$roll[1], 0)`
in four years and its reserve supports about
`r sprintf("%.1f", os$years_left)` more years.

Section 7 established the constraint this section has to work inside.
The school draws about `r fmt_n(lh_bt$modelled)` children on geography
alone, and that figure barely moves across the swept parameter range.
Reducing the admission number to `r fmt_n(lh_bt$pan2026)` has already
closed most of the gap between capacity and catchment. What remains is a
cohort that keeps falling (section 3) and a reserve that runs out
(section 6).

So the question is not whether something must change, nor whether the
admission number should come down again --- that lever is largely spent.
It is whether any *combination* of the available changes produces a
school that is viable at the end of the projection period rather than
merely smaller.

```{r fig-longhill-configs}
#| fig-cap: "Modelled Longhill intake under seven configurations, across the projection period. Bands show the range across the swept parameter space."
#| fig-height: 6

lb <- os$lh_band %>%
  mutate(config = factor(config, levels = os$configs))

ggplot(lb, aes(entry_year, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#2166ac", alpha = 0.2) +
  geom_line(colour = "#2166ac", linewidth = 1) +
  geom_point(colour = "#2166ac", size = 1.8) +
  facet_wrap(~ config, ncol = 2, labeller = label_wrap_gen(38)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year of Year 7 entry", y = "Modelled intake",
       title = "Seven futures for Longhill",
       subtitle = "Median and full range across the swept parameter space",
       caption = "Brightopia, open-data specification.") +
  theme_bh(11) +
  theme(strip.text = element_text(size = 8.5))
```

```{r longhill-config-stats}
#| include: false
cfg_2030 <- lb %>% filter(entry_year == 2030) %>% arrange(desc(med))
best <- cfg_2030 %>% slice_max(med, n = 1)
base <- cfg_2030 %>% filter(str_detect(as.character(config), "^A\\."))
```

By 2030 the configurations separate clearly. The status quo
(`r as.character(base$config)`) gives a median modelled intake of
**`r fmt_n(base$med)`**; the strongest option
(`r as.character(best$config)`) gives **`r fmt_n(best$med)`**.

The important point is not which single option wins. It is that **no
single instrument is sufficient, and the instruments compose.** Moving
the school changes its position in the travel network. Reducing its
admission number changes what "full" means. Redrawing catchments changes
which children are steered towards it. Each on its own leaves a school
that is still under-filled on the central specification; taken together
they reach a configuration that works. A decision that picks one lever
and rejects the others is choosing the worst of the available futures.

::: {.callout-important appearance="simple"}
## Read these figures with the open model's limits in mind

These are modelled intakes from an uncalibrated model with no catchment
term, and they are systematically more optimistic about relocation than
a calibrated model would be. The direction of every comparison here is
robust across the swept range; the magnitudes are not. Nothing in this
section should be read as a prediction of a roll.
:::

# What the council needs to do {#sec-council}

## Decisions {#sec-decisions}

**Decide about Longhill on a deadline, not on a consultation cycle.**
The reserve supports about `r sprintf("%.1f", os$years_left)` more years.
A decision deferred past that point is a decision taken by insolvency
rather than by the council.

**Treat the instruments as a package.** Section 8 shows they compose.
Section 3 shows the cohort keeps falling regardless. Adjusting a single
admission number and waiting to see what happens will consume years the
finances do not have.

**Stop pointing families at Attainment 8.** The council publishes the
admissions guide. Section 5 shows families track the headline score at
`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"], 0)`
and the value-added measure at
`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"], 0)`.
Publishing a contextualised measure alongside the headline would cost
nothing and would begin to unwind a self-fulfilling loop the authority
currently sustains.

**Treat transport as an admissions instrument.** Section 4 found
`r nrow(acc$lsoa %>% filter(places_30 < 1))` neighbourhoods that can
reach no school place inside 30 minutes, concentrated in the more
deprived third of the city. A bus timetable can widen real choice
without taking anything from anyone --- which is more than can be said
for any boundary change.

## Questions raised in this document {#sec-questions}

Numbered questions appear in the sections above where the evidence
raises them. They are gathered here so they can be put as a set. None
requires new research; each is answerable from records the council
already holds.

1. **Why is the catchment forecast wrong in Longhill and in Hove Park /
   Blatchington Mill, in opposite directions?** (§3.4) In 2026 Longhill
   was forecast `r lh26$council` and received `r lh26$actual`, while
   Hove Park / Blatchington Mill was `r abs(hb26$council_err)` places
   under. The second has a candidate explanation in the neighbouring
   PACA over-forecast; the first does not.

## Data the council could simply publish {#sec-data-asks}

Every "this cannot be done on published data" note in this document
resolves to a small number of releases. None requires new collection;
all are aggregates of data the council already holds.

| # | What | What it would settle |
|---|------|---------------------|
| 1 | Home LSOA of every applicant, by year and allocated school | Calibrate $\beta$; replace every band in section 8 with an estimate |
| 2 | Full preference ordering, anonymised, by year | Turn section 5's ten-point association into a discrete-choice model |
| 3 | Allocated school by home LSOA, including schools outside the city | Measure the actual outflow the council's forecasts assume |
| 4 | Criterion under which each place was allocated | Establish directly which criteria bind, rather than inferring it |
| 5 | Children allocated a school they did not name, and which | Test the council's stated "nearest school with places" practice |
| 6 | Journey actually made, or at least distance, per allocation | Separate potential accessibility from realised accessibility |

LSOAs average about 1,500 residents and 650 households. A cohort-year
count at LSOA level is not disclosive in the way a postcode-level count
would be, and the council already publishes LSOA-level material
elsewhere.

## Why now {#sec-why-now}

The cohort projections in section 3 are not forecasts. Those children
are in the city's primary schools today. The financial position in
section 6 has a horizon of about
`r sprintf("%.1f", os$years_left)` years. The decisions in front of the
council will shape the school system for the next fifteen to twenty
years, and they are being taken without a picture of the whole system.

This document is an attempt to supply as much of that picture as
published data allows. It stops well short of what the council's own
records would support --- and the offer to do that work, properly and in
the open, stands.
