---
title: "The Brighton Secondary School System"
subtitle: "A strategic view"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
abstract: |
  Brighton & Hove is deciding what to do about a secondary school system
  that is losing children faster than it is losing places. This document
  assembles what can be established about that system from published
  data alone: where the schools and the children are, how reachable the
  schools actually are by bus, how demand has moved over sixteen years,
  and what a model of the whole system says about the choices in front
  of the council. It is written to be argued with.
---

```{r setup}
#| include: false

source(here::here("R", "00_core.R"))

library(scales)
library(patchwork)

knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

# ---- Everything the document draws on --------------------------------
oi     <- bh_data("open_inputs.rds")
acc    <- bh_data("accessibility.rds")
dep    <- bh_data("deprivation_open.rds")
brt    <- bh_data("brightopia.rds")
env    <- bh_data("sensitivity_envelope.rds")
perf   <- bh_data("performance_panel.rds")
lsoa   <- bh_data("lsoa.geojson")
catch  <- bh_data("catchments_current.geojson") %>% st_transform(4326)

fmt_n   <- function(x, d = 0) formatC(round(x, d), big.mark = ",", format = "f", digits = d)
fmt_pct <- function(x, d = 0) paste0(formatC(x, format = "f", digits = d), "%")
```

::: {.callout-note appearance="simple"}
## Everything here is built from published data

No pupil-level record is used anywhere in this document. Every figure
comes from data the council, the DfE or the ONS has already published,
or from a model built on top of it. Where a question can only be
answered with data the council holds but has not released, that is said
plainly rather than glossed over --- and section 9 lists what those
releases would be.
:::

# Executive summary {#sec-exec}

*To be written once the sections below are settled --- it should be the
last thing drafted, not the first.*

# The secondary school system {#sec-system}

Ten secondary schools serve Brighton & Hove: six community schools run
by the local authority, two academies (Brighton Aldridge and Portslade
Aldridge) and two church schools (King's and Cardinal Newman) that set
their own admissions. An eleventh, Peacehaven Community School, sits
just outside the boundary but inside the travel-to-school geography, and
is included throughout because families on the eastern edge of the city
treat it as a real option.

The city allocates places by **catchment area with a lottery tie-break**
for oversubscribed schools --- an arrangement that is very unusual in
England, where distance from the school gate usually decides priority.
It dates from the closure of CoMArt in the east of the city in 2005, and
the catchment system that replaced distance-based allocation in 2008.
CoMArt is marked on the map below because the shape of the system is
still partly the shape of that closure.

## Where the schools, the catchments and the children are {#sec-orientation}

```{r fig-orientation}
#| fig-cap: "Schools, catchment boundaries and the cohort-aged child population. Use the control at the top right to turn each layer on and off."

# ---- Schools ---------------------------------------------------------
schools_sf <- oi$schools %>%
  filter(!is.na(easting), !is.na(northing)) %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>%
  st_transform(4326)
sch_xy <- st_coordinates(schools_sf)
schools_sf$lon <- sch_xy[, 1]; schools_sf$lat <- sch_xy[, 2]

schools_sf <- schools_sf %>%
  mutate(boundary_catchment = as_boundary_catchment(catchment),
         marker_col = ifelse(is.na(boundary_catchment), "#444444",
                             unname(CATCH_COLOURS[boundary_catchment])))

# ---- Children per LSOA ----------------------------------------------
# Oi is the cohort-aged child estimate per zone; zones are LSOA x
# catchment, so an LSOA split by a boundary has to be summed back up.
children <- oi$zones %>%
  group_by(lsoa) %>%
  summarise(children = sum(Oi, na.rm = TRUE), .groups = "drop")

lsoa_pts <- lsoa %>%
  left_join(children, by = c("lsoa21cd" = "lsoa")) %>%
  filter(!is.na(children), children > 0) %>%
  st_point_on_surface() %>%
  suppressWarnings()
pt_xy <- st_coordinates(lsoa_pts)
lsoa_pts$lon <- pt_xy[, 1]; lsoa_pts$lat <- pt_xy[, 2]

# Radius on the square root of the count, so circle *area* is
# proportional to the number of children rather than the radius being -
# the latter exaggerates big areas by a factor of the count.
r_scale <- function(n) 3 + 17 * sqrt(n / max(n, na.rm = TRUE))

pal_catch <- colorFactor(unname(CATCH_COLOURS), names(CATCH_COLOURS))

m <- leaflet(width = "100%", height = 620) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -0.10, lat = 50.84, zoom = 12) %>%
  addPolygons(
    data = catch, group = "Catchments",
    fillColor = ~pal_catch(catchment), fillOpacity = 0.13,
    color = ~pal_catch(catchment), weight = 2.5, opacity = 0.85,
    label = ~unname(CATCH_LABELS[catchment]),
    highlightOptions = highlightOptions(weight = 4, fillOpacity = 0.25,
                                        bringToFront = FALSE)) %>%
  addCircleMarkers(
    data = lsoa_pts, lng = ~lon, lat = ~lat, group = "Children",
    radius = ~r_scale(children),
    fillColor = "#1f4e79", fillOpacity = 0.35,
    color = "#1f4e79", weight = 1, opacity = 0.55,
    label = ~sprintf("%s: about %s children in the cohort",
                     lsoa21nm, round(children)))

# School markers, then logos on top. Both go in the same layer group so
# the control turns the pair on and off together.
for (i in seq_len(nrow(schools_sf))) {
  s <- schools_sf[i, ]
  popup <- sprintf(
    "<b>%s</b><br>Admission number 2024: %s<br>Planned 2030: %s%s",
    s$name, s$pan2024, s$pan2030,
    if (is.na(s$catchment)) "<br><i>Faith school, admits city-wide</i>" else
      paste0("<br>Catchment: ", CATCH_LABELS[s$boundary_catchment]))

  m <- m %>% addCircleMarkers(
    lng = s$lon, lat = s$lat, group = "Schools",
    radius = 6, fillColor = s$marker_col, fillOpacity = 0.95,
    color = "white", weight = 2, popup = popup, label = s$name)

  ic <- logo_icon(LOGO_FILES[as.character(s$urn)])
  if (!is.null(ic))
    m <- m %>% addMarkers(lng = s$lon, lat = s$lat, icon = ic,
                          group = "Schools", popup = popup, label = s$name)
}

# CoMArt, closed 2005
m <- m %>% addCircleMarkers(
  lng = COMART$lon, lat = COMART$lat, group = "CoMArt (closed 2005)",
  radius = 6, fillColor = "#b2182b", fillOpacity = 0.9,
  color = "white", weight = 2,
  popup = "<b>CoMArt</b><br>Closed 2005. Its closure is why the city moved to catchments with a lottery tie-break.",
  label = COMART$name)
ic <- logo_icon(COMART$logo)
if (!is.null(ic))
  m <- m %>% addMarkers(lng = COMART$lon, lat = COMART$lat, icon = ic,
                        group = "CoMArt (closed 2005)",
                        label = COMART$name)

legend_html <- paste0(
  '<div style="background:white;padding:8px 12px;border-radius:4px;',
  'box-shadow:0 1px 5px rgba(0,0,0,.3);font-family:sans-serif;font-size:12px;line-height:1.6">',
  '<div style="font-weight:bold;margin-bottom:4px">Catchment areas, 2025/26</div>',
  paste0('<div style="display:flex;align-items:center;gap:6px">',
         '<span style="display:inline-block;width:14px;height:14px;background:',
         unname(CATCH_COLOURS), ';border-radius:2px;flex-shrink:0"></span><span>',
         unname(CATCH_LABELS), '</span></div>', collapse = ''),
  '<div style="margin-top:6px;padding-top:6px;border-top:1px solid #ddd">',
  '<span style="display:inline-block;width:14px;height:14px;background:#1f4e79;',
  'opacity:.35;border-radius:50%;margin-right:6px"></span>Children per LSOA',
  ' <span style="color:#777">(area &prop; count)</span></div></div>')

m %>%
  addLayersControl(
    overlayGroups = c("Catchments", "Schools", "Children", "CoMArt (closed 2005)"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("CoMArt (closed 2005)") %>%
  addControl(html = legend_html, position = "bottomright")
```

Three things are worth noticing on that map before any analysis begins.

**The catchments are not equivalent objects.** Four of them contain a
single school; two contain a pair. A child in the Varndean / Dorothy
Stringer or Hove Park / Blatchington Mill catchment has two schools to
name and a realistic chance of one of them. A child in the Longhill,
Patcham, BACA or PACA catchment has one. That asymmetry does more work
in this system than almost anything else, and section 5 returns to it.

**The children are not spread evenly.** The cohort is concentrated in
the centre and west; the east of the city, which contains Longhill's
entire catchment, is thin. Longhill's problem is visible on this map
before any model is run.

**Two schools sit outside the geography entirely.** King's and Cardinal
Newman admit city-wide on faith criteria. Together they hold
`r fmt_n(sum(oi$schools$pan2026[is.na(oi$schools$catchment)]))` places,
about
`r fmt_pct(100 * sum(oi$schools$pan2026[is.na(oi$schools$catchment)]) / sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]))`
of the city total. Any statement about what catchments do has to be read
against that.

## What the attainment data actually says {#sec-lever}

The debate about Brighton's schools is usually conducted in the currency
of headline attainment --- Attainment 8, league table position, "good"
and "bad" schools. That currency is close to worthless for the question
the council actually faces, and it is worth being precise about why,
because the usual explanation is also wrong.

```{r lever-models}
#| include: false

nat <- perf$national %>% filter(!is.na(ATT8SCR))

m_fsm   <- lm(ATT8SCR ~ PTFSM6CLA1A, data = nat)
m_prior <- lm(ATT8SCR ~ PTPRIORLO, data = nat)
m_both  <- lm(ATT8SCR ~ PTFSM6CLA1A + PTPRIORLO, data = nat)
m_abs   <- lm(ATT8SCR ~ PTFSM6CLA1A + PTPRIORLO + PPERSABS10, data = nat)
R2 <- function(m) 100 * summary(m)$r.squared

bh_latest <- perf$bh %>%
  filter(year_label == perf$latest, !is.na(ATT8SCR), !is.na(PTPRIORLO))
bh_resid <- bh_latest %>%
  mutate(pred = predict(m_both, newdata = .), resid = ATT8SCR - pred)
above <- sum(bh_resid$resid > 0, na.rm = TRUE)
```

The common account is that attainment tracks deprivation. It does, but
weakly, and stopping there points the city at the wrong lever. Across
`r fmt_n(nobs(m_fsm))` mainstream secondaries in England, the proportion
of disadvantaged pupils explains **`r fmt_pct(R2(m_fsm))`** of the
variation in Attainment 8. The proportion arriving with *low prior
attainment* explains **`r fmt_pct(R2(m_prior))`** --- twice as much. Put
both in together and you get `r fmt_pct(R2(m_both))`: deprivation adds
almost nothing once you know what the intake could already do at age
eleven. Add persistent absence and the figure reaches
**`r fmt_pct(R2(m_abs))`**.

```{r fig-att8-prior}
#| fig-cap: "Attainment 8 against two features of intake, every mainstream secondary in England in the latest year, with Brighton & Hove schools highlighted. Prior attainment and absence both order schools far more sharply than deprivation does."
#| fig-height: 5

panel <- function(xvar, xlab, ttl) {
  d <- nat %>% filter(!is.na(.data[[xvar]]))
  b <- bh_latest %>% filter(!is.na(.data[[xvar]]))
  ggplot(d, aes(.data[[xvar]], ATT8SCR)) +
    geom_point(colour = "grey80", size = 0.9, alpha = 0.5) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
                colour = "grey30", linewidth = 0.7, linetype = "22") +
    geom_point(data = b, colour = "#b2182b", size = 2.4) +
    ggrepel::geom_text_repel(
      data = b %>% filter(SCHNAME %in% c("Longhill High School", "King's School")),
      aes(label = SCHNAME), size = 2.9, colour = "#7f0f22",
      min.segment.length = 0, segment.size = 0.3, seed = 1) +
    scale_x_continuous(labels = label_percent(scale = 1)) +
    labs(x = xlab, y = "Attainment 8", subtitle = ttl) +
    theme_bh(11)
}

(panel("PTFSM6CLA1A", "Disadvantaged pupils",
       sprintf("Deprivation (R² = %.2f)", summary(m_fsm)$r.squared) ) |
 panel("PTPRIORLO", "Pupils with low prior attainment",
       sprintf("Prior attainment (R² = %.2f)", summary(m_prior)$r.squared)) |
 panel("PPERSABS10", "Persistently absent pupils",
       sprintf("Absence (R² = %.2f)",
               summary(lm(ATT8SCR ~ PPERSABS10, data = nat))$r.squared))) +
  plot_annotation(
    title = "Attainment is a description of intake, and of who turns up",
    subtitle = sprintf("Mainstream secondaries in England, %s. Red points are Brighton & Hove.", perf$latest),
    caption = "Source: DfE performance tables.",
    theme = theme_bh())
```

Two consequences follow, and they point in the same direction.

**A headline score is mostly a description of who walks through the
door.** Once prior attainment and deprivation are accounted for,
**`r above` of the `r nrow(bh_resid)`** Brighton & Hove schools score
above what the national relationship predicts for their intake. On the
measure that isolates the school's own contribution, this is not a city
with a school quality problem.

**The lever that moves is attendance, not social mix.** Persistent
absence alone correlates with Attainment 8 at
`r sprintf("%.2f", cor(nat$ATT8SCR, nat$PPERSABS10, use = "complete.obs"))`
--- more strongly than any intake measure here. Absence and deprivation
travel together, and absence is the mechanism through which much of the
deprivation effect actually operates, which is why redistributing
disadvantaged children between schools is a far weaker instrument than
it is usually assumed to be.

::: {.callout-important appearance="simple"}
## One caution about that last figure

Absence is partly an outcome as well as a cause: a child who is
disengaged attends less *and* attains less, so a single-equation fit
overstates how much of that `r fmt_pct(R2(m_abs))` is absence doing the
causal work. The Lever report models this properly, with school and year
effects, rather than reading it off a scatterplot. The ordering of the
three panels above survives that treatment; the exact numbers should not
be quoted as causal effects.
:::

::: {.callout-tip}
## The detail behind this is published separately

The contextualised analysis --- value-added modelling across every local
authority in England, and what it says about where Brighton's schools
really sit --- is set out in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).
Its central findings are that Brighton's schools perform well once
intake is accounted for, and that **absence, not social mixing, is the
largest single lever available to the city**. This document takes those
findings as given and turns to the spatial and demographic questions the
Lever report does not address.
:::

::: {.callout-tip}
## The detail behind this is published separately

The contextualised analysis --- value-added modelling across every
local authority in England, and what it says about where Brighton's
schools really sit --- is set out in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).
Its central findings are that Brighton's schools perform well once
intake is accounted for, that **absence, not social mixing, is the
largest single lever available to the city**, and that absence is the
mechanism through which much of the deprivation effect operates. This
document takes those findings as given and turns to the spatial and
demographic questions the Lever report does not address.
:::

```{r fig-ranking-flip}
#| fig-cap: "The same ten schools ranked two ways: by headline Attainment 8, and by the value-added measure from the Lever report. Lines crossing means the two measures disagree about which schools are doing well."
#| fig-height: 5.5

flip <- oi$attract %>%
  filter(!is.na(att8), !is.na(va_lever), name != "Peacehaven Community School") %>%
  mutate(rank_att8 = rank(-att8), rank_va = rank(-va_lever)) %>%
  select(name, rank_att8, rank_va) %>%
  pivot_longer(starts_with("rank_"), names_to = "measure", values_to = "rank") %>%
  mutate(measure = factor(measure, c("rank_att8", "rank_va"),
                          c("Attainment 8\n(headline)", "Value added\n(Lever report)")))

moved <- flip %>%
  pivot_wider(names_from = measure, values_from = rank) %>%
  rename(att8 = 2, va = 3) %>%
  mutate(shift = att8 - va)

ggplot(flip, aes(measure, rank, group = name)) +
  geom_line(aes(colour = name), linewidth = 1.1, alpha = 0.85, show.legend = FALSE) +
  geom_point(aes(colour = name), size = 3, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = flip %>% filter(measure == levels(flip$measure)[1]),
    aes(label = name), hjust = 1, nudge_x = -0.08, size = 3,
    direction = "y", segment.size = 0.2, seed = 1) +
  ggrepel::geom_text_repel(
    data = flip %>% filter(measure == levels(flip$measure)[2]),
    aes(label = name), hjust = 0, nudge_x = 0.08, size = 3,
    direction = "y", segment.size = 0.2, seed = 1) +
  scale_y_reverse(breaks = 1:10) +
  scale_x_discrete(expand = expansion(add = c(0.85, 0.85))) +
  labs(x = NULL, y = "Rank (1 = highest)",
       title = "Which schools look good depends on which measure you use",
       subtitle = "Ranked by headline attainment, and by contribution to progress once intake is accounted for",
       caption = "Sources: DfE performance tables; How to Pull the Right Lever.") +
  theme_bh() +
  theme(panel.grid.major.x = element_blank())
```

```{r flip-stats}
#| include: false
# Ties are real here - two schools fall by the same four places - so
# collapse to a list rather than letting slice_min pick one silently.
and_list <- function(x) if (length(x) < 2) x else
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))

up   <- moved %>% filter(shift == max(shift))
down <- moved %>% filter(shift == min(shift))
n_moved3 <- sum(abs(moved$shift) >= 3)
```

The lines cross a great deal. **`r and_list(up$name)`** rises
`r abs(up$shift[1])` places when you move from the headline measure to
the value-added one; **`r and_list(down$name)`** falls
`r abs(down$shift[1])`. `r n_moved3` of the ten schools move by three
places or more.

The schools that fall are not bad schools, and the schools that rise are
not suddenly good ones. The two measures are answering different
questions: one asks what pupils at this school achieved, the other asks
how much of that the school is responsible for. Families reading the
published headline are answering the first question while believing they
are answering the second --- and section 5 shows that this has a
measurable effect on where demand actually goes.

# Demographic futures {#sec-demography}

The single most important fact about this system is that it is shrinking,
and that the shrinkage is already locked in. The children who will enter
Year 7 in 2033 have already started primary school. Their number is not
a forecast in any meaningful sense --- it is a count, adjusted for a
transfer rate that has been remarkably stable for a decade.

## What has already happened {#sec-cohort-history}

```{r cohort-data}
#| include: false
rc <- bh_data("reception_cohort.rds")

y7    <- rc$secondary_city
rec   <- rc$reception_city
peak  <- y7 %>% slice_max(y7_offers, n = 1)
now   <- y7 %>% slice_max(year, n = 1)
proj  <- rc$projection
last_proj <- proj %>% slice_max(entry_year, n = 1)
drop_pct <- 100 * (last_proj$central - now$y7_offers) / now$y7_offers
```

Year 7 offers across the city peaked at **`r fmt_n(peak$y7_offers)`** in
`r peak$year`. In `r now$year` the figure was
**`r fmt_n(now$y7_offers)`** --- a fall of
`r fmt_pct(100 * (peak$y7_offers - now$y7_offers) / peak$y7_offers, 1)`
from the peak. That decline is not a projection. It has happened.

```{r fig-cohort-history}
#| fig-cap: "Reception and Year 7 offers made across Brighton & Hove. The reception line leads the Year 7 line by seven years, so it is a preview of secondary demand rather than a separate story."
#| fig-height: 5

bind_rows(
  y7  %>% transmute(year, offers = y7_offers, phase = "Year 7 offers"),
  rec %>% transmute(year, offers = rec_offers, phase = "Reception offers")) %>%
  ggplot(aes(year, offers, colour = phase)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Reception offers" = "#2c7fb8",
                                 "Year 7 offers" = "#b2182b"), name = NULL) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Offers made",
       title = "Both phases are past their peak",
       subtitle = "Reception offers turn down first, and Year 7 follows seven years later",
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh()
```

## The cohort already in school {#sec-projection}

The transfer rate from a reception cohort to the Year 7 cohort seven
years later has been strikingly stable. Across the six cohorts where
both ends are observable, **`r fmt_pct(100 * rc$survival_recent, 1)`** of
reception offers reappear as Year 7 offers, with a standard deviation of
just `r fmt_pct(100 * rc$survival_sd, 1)`. The city loses about one child
in nine between ages four and eleven --- to independent schools, to
moves out of the area, and to schools across the boundary.

Applying that rate to reception cohorts already in school gives a
projection that depends on almost no assumptions:

```{r fig-projection}
#| fig-cap: "Year 7 demand projected from reception cohorts already in school. The band is the range implied by the observed year-to-year variation in the transfer rate."
#| fig-height: 5

ggplot() +
  geom_ribbon(data = proj, aes(entry_year, ymin = lo, ymax = hi),
              fill = "#b2182b", alpha = 0.18) +
  geom_line(data = proj, aes(entry_year, central),
            colour = "#b2182b", linewidth = 1.1, linetype = "22") +
  geom_point(data = proj, aes(entry_year, central), colour = "#b2182b", size = 2) +
  geom_line(data = y7, aes(year, y7_offers), colour = "grey25", linewidth = 1.1) +
  geom_point(data = y7, aes(year, y7_offers), colour = "grey25", size = 2) +
  annotate("text", x = 2016, y = 2480, label = "Observed", colour = "grey25",
           hjust = 0, fontface = "bold", size = 3.6) +
  annotate("text", x = 2029.5, y = 2180, label = "Projected from\nchildren already in school",
           colour = "#b2182b", hjust = 0, size = 3.4, lineheight = 0.95) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Year 7 offers",
       title = sprintf("A further %s fall is already in the primary schools",
                       fmt_pct(abs(drop_pct), 0)),
       subtitle = sprintf("From %s in %s to about %s by %s",
                          fmt_n(now$y7_offers), now$year,
                          fmt_n(last_proj$central), last_proj$entry_year),
       caption = "Sources: BHCC allocation factsheets; transfer rate estimated from six observed cohorts.") +
  theme_bh()
```

By **`r last_proj$entry_year`** the city should expect about
**`r fmt_n(last_proj$central)`** Year 7 children, against
`r fmt_n(now$y7_offers)` today --- a further fall of
`r fmt_pct(abs(drop_pct), 0)`. Against a current published admission
capacity of about
`r fmt_n(sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]))`
places, that is roughly
`r fmt_n(sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]) - last_proj$central)`
surplus places --- the equivalent of two and a half average-sized
secondary schools standing empty.

## Where the council's forecast differs {#sec-council-forecast}

The council publishes its own catchment-level forecasts. They are
consistently **higher** than what the primary cohorts support, and the
gap widens with distance.

```{r fig-forecast-gap}
#| fig-cap: "The council's own demand forecast against a projection built from reception cohorts already in school. Positive values mean the council expects more children than the primary registers contain."
#| fig-height: 4.6

rc$comparison %>%
  mutate(gap = ons_demand - reception_based) %>%
  ggplot(aes(entry_year, gap)) +
  geom_col(fill = "#7b3294", alpha = 0.85, width = 0.65) +
  geom_hline(yintercept = 0, colour = "grey30") +
  geom_text(aes(label = sprintf("%+.0f", gap),
                vjust = ifelse(gap >= 0, -0.4, 1.3)), size = 3.2) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.18))) +
  labs(x = "Year of Year 7 entry", y = "Council forecast minus cohort projection",
       title = "The council expects more children than the primary registers hold",
       subtitle = "Difference in Year 7 places, city-wide",
       caption = "Sources: BHCC published forecasts; reception-based projection as above.") +
  theme_bh()
```

```{r lh-gap}
#| include: false
lh_council <- rc$council %>%
  mutate(catchment = as_model_catchment(CatchmentGroup)) %>%
  filter(catchment == "Longhill")
lh_cohort <- rc$by_catchment %>% filter(catchment == "Longhill")
lh_join <- lh_council %>%
  inner_join(lh_cohort %>% select(entry_year, projected), by = "entry_year") %>%
  mutate(gap = council - projected)
stopifnot(nrow(lh_join) > 0)
```

The city-wide gap is modest in most years. Broken down by catchment it
is not, and Longhill is where it opens widest.

```{r fig-longhill-gap}
#| fig-cap: "Longhill catchment: the council's forecast, the projection from its own primary cohorts, and the school's admission number."
#| fig-height: 4.6

lh_join %>%
  select(entry_year, Council = council, `From primary cohorts` = projected) %>%
  pivot_longer(-entry_year, names_to = "series", values_to = "children") %>%
  ggplot(aes(entry_year, children, colour = series)) +
  geom_hline(yintercept = rc$council_lh_pan, linetype = "31", colour = "grey45") +
  annotate("text", x = min(lh_join$entry_year), y = rc$council_lh_pan + 7,
           label = sprintf("Admission number (%s)", rc$council_lh_pan),
           hjust = 0, size = 3.1, colour = "grey35") +
  geom_line(linewidth = 1.1) + geom_point(size = 2.4) +
  scale_colour_manual(values = c("Council" = "#7b3294",
                                 "From primary cohorts" = "#008837"), name = NULL) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year of Year 7 entry", y = "Children",
       title = "Longhill: the two forecasts disagree, and both sit below the admission number",
       subtitle = sprintf("Mean gap of %.0f children a year",
                          mean(lh_join$gap, na.rm = TRUE)),
       caption = "Sources: BHCC catchment forecasts; reception-based projection.") +
  theme_bh()
```

The council's forecast for the Longhill catchment runs an average of
**`r fmt_n(mean(lh_join$gap, na.rm = TRUE))` children a year above** what
that catchment's own primary cohorts support. Both series sit below the
school's admission number of `r rc$council_lh_pan` in every year. This
matters because decisions about Longhill's future are being taken
against the higher of two numbers, and the higher one is the one the
primary registers do not support.

## The Peacehaven accession {#sec-peacehaven}

```{r peacehaven}
#| include: false
area_children <- oi$zones %>%
  group_by(area) %>%
  summarise(lsoas = n_distinct(lsoa), children = sum(Oi, na.rm = TRUE),
            .groups = "drop")
exp_area <- area_children %>% filter(area == "Expansion area")
tot_children <- sum(area_children$children)
ph_pan <- oi$schools$pan2026[oi$schools$name == "Peacehaven Community School"]
```

The catchment system reaches beyond the city boundary. The Peacehaven
and Telscombe area contributes **`r exp_area$lsoas` LSOAs** and about
**`r fmt_n(exp_area$children)` cohort-aged children**, roughly
`r fmt_pct(100 * exp_area$children / tot_children, 0)`
of the total the model covers.

In headcount terms this is close to self-supporting: Peacehaven
Community School has an admission number of `r ph_pan`, against about
`r fmt_n(exp_area$children)` children in its area. The difficulty is not
the arithmetic but the direction of travel. These families sit at the
eastern end of a transport network that runs east--west along the coast,
and section 4 shows that they are among the least well served in the
whole study area --- not because they are far from *a* school, but
because they are far from *most* schools, and therefore have the fewest
alternatives if their first choice is full.

# How reachable are the schools? {#sec-access}

Almost all of the public argument about Brighton's schools is conducted
in terms of admissions rules --- catchments, priorities, tie-breaks.
Underneath the rules sits a physical constraint that no rule can undo:
how long it actually takes a eleven-year-old to get to each school on a
bus. This section establishes that constraint before section 5 returns
to the rules.

All journey times here are routed with `r5r` over the merged street
network and bus timetable for the area, for a weekday morning arrival.
They include walking to the stop, waiting, the ride itself, and any
transfer.

## Journey time to a single school {#sec-one-school}

```{r fig-one-school}
#| fig-cap: "Walk-and-bus journey time to Longhill High School from every neighbourhood in the study area, for a weekday morning arrival."
#| fig-height: 6

nm <- bh_data("network_maps.rds")

# The surface object carries one point per zone, and zones are
# LSOA x catchment. Collapse to LSOA on the child population and join
# the times onto the LSOA polygons, so this draws as a surface rather
# than as a scatter of centroids.
surf_lsoa <- nm$surface %>%
  st_drop_geometry() %>%
  group_by(lsoa) %>%
  summarise(mins = weighted.mean(ovingdean, pmax(Oi, 1e-6), na.rm = TRUE),
            .groups = "drop")

surf <- lsoa %>%
  inner_join(surf_lsoa, by = c("lsoa21cd" = "lsoa")) %>%
  filter(is.finite(mins))

pal_time <- colorNumeric("magma", domain = surf$mins, reverse = TRUE)

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = surf,
              fillColor = ~pal_time(mins), fillOpacity = 0.78,
              color = "white", weight = 0.4,
              label = ~sprintf("%s: %.0f minutes", lsoa21nm, mins)) %>%
  addCircleMarkers(
    lng = st_coordinates(st_transform(
      st_as_sf(oi$schools %>% filter(short == "longhill"),
               coords = c("easting", "northing"), crs = 27700), 4326))[, 1],
    lat = st_coordinates(st_transform(
      st_as_sf(oi$schools %>% filter(short == "longhill"),
               coords = c("easting", "northing"), crs = 27700), 4326))[, 2],
    radius = 7, fillColor = "#1a9850", fillOpacity = 1,
    color = "white", weight = 2, label = "Longhill High School") %>%
  addLegend(pal = pal_time, values = surf$mins, title = "Minutes",
            position = "bottomright", opacity = 0.85)
```

The median neighbourhood is
**`r fmt_n(median(nm$surface$ovingdean, na.rm = TRUE))` minutes** from
Longhill's current site. The school sits at the edge of its own
catchment and at the edge of the city, which is the geometric fact
underneath most of section 8.

## What one of those journeys actually looks like {#sec-legs}

A single number in minutes hides what the journey involves. These are
seven real routed journeys, drawn leg by leg: walking legs follow the
street network, bus legs follow the service's own shape.

```{r fig-legs}
#| fig-cap: "Seven worked journeys, routed leg by leg. Walking legs in orange, bus legs in blue. Each can be checked against local knowledge."
#| fig-height: 6

rg <- bh_data("route_geometries.rds")
legs <- rg$legs %>% st_transform(4326)

pal_mode <- colorFactor(c("BUS" = "#2166ac", "WALK" = "#e08214"),
                        domain = c("BUS", "WALK"))

mleg <- leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron)

for (j in sort(unique(legs$journey))) {
  d <- legs %>% filter(journey == j)
  tot <- rg$summary$total_minutes[rg$summary$journey == j]
  mleg <- mleg %>%
    addPolylines(data = d, group = j,
                 color = ~pal_mode(leg_mode), weight = 5, opacity = 0.85,
                 label = ~sprintf("%s - %s leg, %.0f min%s", j, leg_mode, minutes,
                                  ifelse(nzchar(route), paste0(" (service ", route, ")"), "")),
                 popup = ~sprintf("<b>%s</b><br>%s minutes in total", j, tot))
}

mleg %>%
  addLayersControl(overlayGroups = sort(unique(legs$journey)),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  addLegend(pal = pal_mode, values = c("BUS", "WALK"), title = "Leg",
            position = "bottomright", opacity = 0.9)
```

```{r tbl-journeys}
#| tbl-cap: "The same seven journeys, summarised."
rg$summary %>%
  transmute(Journey = journey,
            `Total minutes` = total_minutes,
            Legs = legs,
            `Bus services used` = ifelse(nzchar(services), services, "walk only")) %>%
  arrange(`Total minutes`) %>%
  knitr::kable()
```

Woodingdean to Longhill takes 11 minutes. Whitehawk to Dorothy Stringer
takes 52, for a journey of about the same straight-line distance. The
network is not symmetric, and neither is opportunity.

## Which school is actually nearest {#sec-nearest}

```{r fig-nearest}
#| fig-cap: "The nearest secondary school to each neighbourhood by routed walk-and-bus time, regardless of catchment."
#| fig-height: 6

near <- lsoa %>%
  left_join(acc$lsoa %>% select(lsoa, nearest_school, nearest_min),
            by = c("lsoa21cd" = "lsoa")) %>%
  filter(!is.na(nearest_school))

pal_near <- colorFactor("Set3", domain = sort(unique(near$nearest_school)))

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = near,
              fillColor = ~pal_near(nearest_school), fillOpacity = 0.72,
              color = "white", weight = 0.4,
              label = ~sprintf("%s: %s, %.0f min", lsoa21nm, nearest_school, nearest_min)) %>%
  addLegend(pal = pal_near, values = near$nearest_school, title = "Nearest school",
            position = "bottomright", opacity = 0.85)
```

```{r nearest-stats}
#| include: false
near_tab <- acc$lsoa %>%
  count(nearest_school, wt = Oi, name = "children") %>%
  arrange(desc(children)) %>%
  mutate(pct = 100 * children / sum(children))
```

This map is worth comparing with the catchment map in section 2. They
are not the same shape. The nearest school by bus is not always the
catchment school, and for
`r fmt_pct(100 * mean(acc$lsoa$nearest_min > 25), 0)`
of neighbourhoods even the nearest school is more than 25 minutes away.

## Potential accessibility: how much school is within reach {#sec-potential}

Nearest-school distance answers a narrow question. It says nothing about
whether a neighbourhood has *one* school within reach or *five* --- and
that difference is the whole of what a family's realistic options
amount to when their first preference is full.

Two measures are used here, deliberately, because they fail in opposite
directions.

::: {.panel-tabset}
### Gravity

$$A_i = \sum_j W_j \, c_{ij}^{-\beta}$$

Every school counts, discounted by how far away it is. $W_j$ is the
school's attractiveness (its admission number), $c_{ij}$ the routed
walk-and-bus time, and $\beta$ the distance decay --- set here at
`r sprintf("%.1f", acc$beta$ref)`, the reference value from the model in
section 7. The number has no units, so it is indexed with the
child-weighted city average at 100.

### Cumulative opportunity

$$P_i(t) = \sum_j \mathrm{PAN}_j \cdot \mathbf{1}(c_{ij} \le t)$$

The number of school places reachable inside $t$ minutes. Trivial to
explain --- "how many places can you get to in half an hour" --- but it
treats a 29-minute journey and a 31-minute journey as entirely different
things.
:::

```{r fig-accessibility}
#| fig-cap: "Potential accessibility to secondary school places, by neighbourhood. Indexed so that the child-weighted city average is 100. Darker means less school within reach."
#| fig-height: 6

acc_map <- lsoa %>%
  inner_join(acc$lsoa, by = c("lsoa21cd" = "lsoa"))

pal_acc <- colorNumeric("viridis", domain = acc_map$A_index, reverse = TRUE)

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = acc_map,
              fillColor = ~pal_acc(A_index), fillOpacity = 0.8,
              color = "white", weight = 0.4,
              label = ~sprintf(
                "%s | index %.0f | %.0f places within 30 min | nearest %s, %.0f min",
                lsoa21nm, A_index, places_30, nearest_school, nearest_min)) %>%
  addLegend(pal = pal_acc, values = acc_map$A_index,
            title = "Accessibility<br>index", position = "bottomright",
            opacity = 0.85)
```

```{r acc-stats}
#| include: false
zero30 <- acc$lsoa %>% filter(places_30 < 1)
rng <- range(acc$lsoa$A_index)
```

Three findings come out of this, and the first is the one that should
change the conversation.

**`r nrow(zero30)` of the `r nrow(acc$lsoa)` neighbourhoods --- about
`r fmt_n(sum(zero30$Oi))` cohort-aged children --- cannot reach a single
secondary school place within 30 minutes** by walking and bus on a
weekday morning. Not their catchment school; *any* school. For those
neighbourhoods the admissions debate is somewhat beside the point.

**The spread is nearly fivefold.** The best-served neighbourhood scores
`r fmt_n(rng[2])` on the index against `r fmt_n(rng[1])` for the worst.
A child in the centre of the city has several schools genuinely within
reach; a child at either end has one, or none.

**The two measures broadly agree, but not closely.** They correlate at
`r sprintf("%.2f", acc$agreement$spearman)` (Spearman), yet only
`r fmt_pct(100 * acc$agreement$within_one, 0)` of neighbourhoods land
within one decile of each other on both. Where they agree, the finding
is robust to how you define access. Where they disagree, it is because
one is counting a school just over the threshold that the other is
discounting smoothly.

```{r fig-acc-compare}
#| fig-cap: "The two measures against each other. Neighbourhoods along the bottom can reach no places at all inside 30 minutes."
#| fig-height: 4.8

acc$lsoa %>%
  mutate(grp = ifelse(places_30 < 1, "No places within 30 min", "Some places within 30 min")) %>%
  ggplot(aes(A_index, places_30, colour = grp, size = Oi)) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(values = c("No places within 30 min" = "#b2182b",
                                 "Some places within 30 min" = "#2166ac"),
                      name = NULL) +
  scale_size_area(max_size = 6, guide = "none") +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Gravity accessibility index (city average = 100)",
       y = "Places reachable within 30 minutes",
       title = "Two ways of asking the same question",
       subtitle = "Each point is a neighbourhood, sized by the number of cohort-aged children",
       caption = "Routed walk-and-bus times, weekday morning arrival.") +
  theme_bh()
```

::: {.callout-warning appearance="simple"}
## Two caveats that bear on these numbers

**Journey times are modelled, not measured.** They come from `r5r`
routing over the merged street network and published timetables for one
weekday morning. They do not know about school buses, parental lifts,
walking routes children actually use, or a bus that is full when it
arrives.

**One school's times are suspect.** Dorothy Stringer and Varndean are
470 metres apart, yet the routed matrix makes Stringer slower from most
of the city by a median of about five and a half minutes. That may be
real, or it may be an artefact of where the router snapped each school
onto the network. It has not been established which, and it inflates the
apparent inaccessibility of the areas whose nearest school is Stringer.
:::

## Accessibility and child poverty together {#sec-bivariate}

The question that matters for policy is not where access is poor, nor
where deprivation is high, but where the two coincide --- because those
are the neighbourhoods where a difficult journey is least likely to be
solved by a car in the family.

```{r fig-bivariate}
#| fig-cap: "Accessibility and income deprivation affecting children, mapped together. The darkest blue is the corner that matters: least reachable school places, highest child poverty."
#| fig-height: 7

# Stevens-style bivariate palette. Keyed "access-deprivation", each on a
# tercile: access 1 is the least reachable third, deprivation 3 the most
# deprived third. The concerning corner is therefore 1-3, and it is the
# darkest colour.
BIV <- c(
  "1-1" = "#be64ac", "1-2" = "#8c62aa", "1-3" = "#3b4994",
  "2-1" = "#dfb0d6", "2-2" = "#a5add3", "2-3" = "#5698b9",
  "3-1" = "#e8e8e8", "3-2" = "#ace4e4", "3-3" = "#5ac8c8")

biv_map <- lsoa %>%
  inner_join(acc$bivariate %>% select(lsoa, acc_t, dep_t, biv_key,
                                      idaci_score, A_index, places_30, Oi),
             by = c("lsoa21cd" = "lsoa")) %>%
  mutate(fill = unname(BIV[biv_key]))

main <- ggplot(biv_map) +
  geom_sf(aes(fill = fill), colour = "white", linewidth = 0.12) +
  scale_fill_identity() +
  labs(title = "Where poor access and child poverty coincide",
       subtitle = "Each neighbourhood classed on both axes at once",
       caption = "Sources: routed walk-and-bus accessibility; IDACI 2019.") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey30"),
        plot.caption = element_text(colour = "grey45", hjust = 0))

key <- expand_grid(acc_t = 1:3, dep_t = 1:3) %>%
  mutate(fill = unname(BIV[paste0(acc_t, "-", dep_t)])) %>%
  ggplot(aes(acc_t, dep_t, fill = fill)) +
  geom_tile(colour = "white", linewidth = 0.7) +
  scale_fill_identity() +
  scale_x_continuous(breaks = c(1, 3), labels = c("least", "most"),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = c(1, 3), labels = c("least", "most"),
                     expand = c(0, 0)) +
  labs(x = "Accessible →", y = "Deprived →") +
  coord_equal() +
  theme_minimal(base_size = 8) +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 7.5, colour = "grey25"),
        axis.text = element_text(size = 6.5, colour = "grey40"))

main + inset_element(key, left = 0.72, bottom = 0.02, right = 0.99, top = 0.32)
```

```{r biv-stats}
#| include: false
corner <- acc$worst_corner
best_corner <- acc$bivariate %>% filter(acc_t == 3, dep_t == 3)
cells <- acc$bivariate %>%
  count(acc_t, dep_t, wt = Oi, name = "children")
low_acc_high_dep <- cells %>% filter(acc_t == 1, dep_t == 3) %>% pull(children)
hi_acc_high_dep  <- cells %>% filter(acc_t == 3, dep_t == 3) %>% pull(children)
```

There is a real gradient here, and it runs the wrong way. Accessibility
and child poverty correlate at
`r sprintf("%.2f", acc$acc_dep_spearman)` (Spearman): the more deprived
a neighbourhood, the less school it can reach.

**`r nrow(corner)` neighbourhoods --- about `r fmt_n(low_acc_high_dep)`
cohort-aged children --- sit in the worst corner**, in the least
accessible third of the city and the most deprived third at the same
time. At the other end, only `r nrow(best_corner)` neighbourhoods
(`r fmt_n(hi_acc_high_dep)` children) combine high deprivation with good
access.

This is the strongest argument in the document for treating transport as
a school-admissions instrument rather than a separate departmental
concern. A city that wanted to widen real choice for its poorest
families could do more with a bus timetable than with a catchment
boundary --- and unlike a boundary change, it would take nothing away
from anyone else.

::: {.callout-note appearance="simple"}
## What pupil-level data would add here

This section works entirely with modelled journey times between
neighbourhood centroids and school gates. The council holds the home
postcode of every applicant and the school each was allocated. With
that, the same analysis could be run on **journeys children actually
make** rather than journeys they could in principle make --- and the gap
between the two is precisely the quantity that would tell the city
whether its transport network is a real constraint on choice or merely a
theoretical one. Section 9 sets out what that request would look like.
:::
