---
title: "The Brighton Secondary School System"
subtitle: "A strategic view"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
abstract: |
  Brighton & Hove is deciding what to do about a secondary school system
  that is losing children faster than it is losing places. This document
  assembles what can be established about that system from published
  data alone: where the schools and the children are, how reachable the
  schools actually are by bus, how demand has moved over sixteen years,
  and what a model of the whole system says about the choices in front
  of the council. It is written to be argued with.
---

```{r setup}
#| include: false

source(here::here("R", "00_core.R"))

library(scales)
library(patchwork)

knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

# ---- Everything the document draws on --------------------------------
oi     <- bh_data("open_inputs.rds")
acc    <- bh_data("accessibility.rds")
dep    <- bh_data("deprivation_open.rds")
brt    <- bh_data("brightopia.rds")
env    <- bh_data("sensitivity_envelope.rds")
perf   <- bh_data("performance_panel.rds")
lsoa   <- bh_data("lsoa.geojson")
catch  <- bh_data("catchments_current.geojson") %>% st_transform(4326)

# Loaded here as well as in the sections that use them, so the executive
# summary can quote figures that are computed further down the document.
rc     <- bh_data("reception_cohort.rds")
os     <- bh_data("open_scenarios.rds")
conv   <- bh_data("adjudicator_conversion.rds")
fp     <- bh_data("factsheet_panel.rds")

fmt_n   <- function(x, d = 0) formatC(round(x, d), big.mark = ",", format = "f", digits = d)
fmt_pct <- function(x, d = 0) paste0(formatC(x, format = "f", digits = d), "%")
```

```{r exec-figures}
#| include: false

# Headline figures for the executive summary. Every one of these is
# recomputed in the section it belongs to; they are gathered here only
# so the summary can quote them before those sections run.

ex_now      <- rc$secondary_city %>% slice_max(year, n = 1)
ex_peak     <- rc$secondary_city %>% slice_max(y7_offers, n = 1)
ex_proj     <- rc$projection %>% slice_max(entry_year, n = 1)
ex_fall     <- 100 * (ex_proj$central - ex_now$y7_offers) / ex_now$y7_offers
ex_places   <- sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"])
ex_surplus  <- ex_places - ex_proj$central

ex_zero30   <- acc$lsoa %>% filter(places_30 < 1)
ex_corner   <- acc$worst_corner

ex_rat <- conv$conv %>%
  group_by(school) %>%
  summarise(rate = sum(p1_offered) / sum(p1_named), .groups = "drop")
ex_open <- sum(ex_rat$rate > 0.999)

ex_r2_att8 <- os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"]
ex_r2_va   <- os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"]

# Same specification as section 2.3, which follows the Lever report:
# log outcome, logged rates, KS2 score entered linearly.
ex_nat <- perf$national %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  mutate(ks2_c = KS2ASS - 100)
ex_fsm   <- summary(lm(log(ATT8SCR) ~ log(PTFSM6CLA1A), data = ex_nat))$r.squared
ex_prior <- summary(lm(log(ATT8SCR) ~ ks2_c, data = ex_nat))$r.squared
ex_absn  <- summary(lm(log(ATT8SCR) ~ log(PERCTOT), data = ex_nat))$r.squared

ex_lh <- brt$at_original_beta %>%
  filter(str_detect(name, "Longhill")) %>%
  mutate(fill_2026 = modelled / pan2026, fill_2024 = modelled / pan2024)

# The between-school share of *pupil-level* variance. This cannot be
# computed from school-level aggregates - it needs pupil records - so it
# is quoted from the school-effectiveness literature, not estimated
# here. Defined at the top because both the executive summary and
# section 2.4 use it.
SCHOOL_LO <- 8; SCHOOL_HI <- 15
```

::: {.callout-note appearance="simple"}
## Everything here is built from published data

No pupil-level record is used anywhere in this document. Every figure
comes from data the council, the DfE or the ONS has already published,
or from a model built on top of it. Where a question can only be
answered with data the council holds but has not released, that is said
plainly rather than glossed over --- and section 9 lists what those
releases would be.
:::

# Executive summary {#sec-exec}

## The situation

Brighton & Hove has a secondary school system built for more children
than it now has, and the shortfall is going to get worse before anything
the council decides can affect it. Year 7 offers peaked at
`r fmt_n(ex_peak$y7_offers)` in `r ex_peak$year` and stood at
`r fmt_n(ex_now$y7_offers)` in `r ex_now$year`. The children who will
enter Year 7 in `r ex_proj$entry_year` are already in the city's primary
schools, and there are about **`r fmt_n(ex_proj$central)`** of them ---
a further fall of `r fmt_pct(abs(ex_fall), 0)`. Against roughly
`r fmt_n(ex_places)` published places, that implies something like
**`r fmt_n(ex_surplus)` surplus places**, or two and a half average
secondary schools' worth.

The council's own catchment forecasts sit above that projection in
almost every year, and the gap is widest in the Longhill catchment ---
the place where the consequences of being wrong are largest.

## What the evidence says

**This is not a school quality problem.** Across English secondaries,
absence explains `r fmt_pct(100 * ex_absn, 0)` of the variation in
Attainment 8 and prior attainment `r fmt_pct(100 * ex_prior, 0)`, while
deprivation explains only `r fmt_pct(100 * ex_fsm, 0)`. Once intake is
accounted for, Brighton's schools mostly perform above what their intake
predicts. The largest lever available to the city is **absence**, not
social mixing --- established in detail in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).

**Moving children between schools is a weak instrument, and this is the
single most important point for admissions policy.** Sixty years of
school-effectiveness research puts the share of variation in pupil
outcomes attributable to differences *between schools* at
`r SCHOOL_LO`--`r SCHOOL_HI` per cent. Most of what separates one
school's results from another's arrives with the children. A
redistribution policy therefore works on the smallest term available
while leaving the intake characteristics and circumstances that account
for the other `r 100 - SCHOOL_HI`--`r 100 - SCHOOL_LO` per cent
untouched. Nor can the measured drivers be separated: of the school-level
variation deprivation, absence and prior attainment jointly explain,
about three quarters is variance they *share*. This is not an argument
that schools do not matter --- it is an argument that the council should
not expect admissions changes to move attainment much.

**Admissions criteria only bind at some schools.**
`r ex_open` of the `r nrow(ex_rat)` schools offered a place to every
first-preference applicant. For those schools the catchment rule, the
sibling rule and the free school meals quota do nothing at all. The city
runs two admissions systems under one set of rules, and public argument
about criteria concerns only one of them.

**Families are choosing on the wrong number.** Headline Attainment 8
explains `r fmt_pct(100 * ex_r2_att8, 0)` of the variation in how
heavily each school is preferred. The value-added measure --- what the
school actually contributes --- explains
`r fmt_pct(100 * ex_r2_va, 0)`. Because the headline score is largely a
description of the existing intake, choice becomes self-fulfilling.

**Physical access is unequal, and unequal in the worst direction.**
`r nrow(ex_zero30)` of `r nrow(acc$lsoa)` neighbourhoods --- about
`r fmt_n(sum(ex_zero30$Oi))` cohort-aged children --- can reach **no**
secondary school place within 30 minutes by walking and bus.
Accessibility and child poverty correlate at
`r sprintf("%.2f", acc$acc_dep_spearman)`: `r nrow(ex_corner)`
neighbourhoods sit in both the least-reachable third and the most
deprived third.

**Longhill has already been shrunk to about what geography supports.**
A model of where children would go if only distance and school size
mattered gives it `r fmt_n(ex_lh$modelled)` children ---
`r fmt_pct(100 * ex_lh$fill_2024, 0)` of its former admission number of
`r fmt_n(ex_lh$pan2024)`, but `r fmt_pct(100 * ex_lh$fill_2026, 0)` of
the reduced `r fmt_n(ex_lh$pan2026)` now in force. The reduction has
already done most of what a reduction can do, and the remaining problem
is a falling cohort against a reserve that supports about
`r sprintf("%.1f", os$years_left)` more years.

## What follows

1. **Decide about Longhill on the financial timetable**, not the
   consultation one. The horizon is about
   `r sprintf("%.1f", os$years_left)` years.
2. **Treat the instruments as a package.** Moving the school, changing
   its admission number and redrawing catchments each fall short alone
   and compose when combined. Picking one and waiting is the worst
   available option.
3. **Publish a contextualised measure alongside Attainment 8.** It costs
   nothing and begins to unwind a loop the authority currently sustains
   through its own admissions guide.
4. **Treat the bus network as an admissions instrument.** It is the only
   lever here that widens choice without taking anything from anyone.
5. **Release six tables** (§9) that would convert most of the bands in
   this document into estimates.

## How confident to be

The demographic findings are the firmest: they are counts of children
already in school, adjusted by a transfer rate that has varied by under
two percentage points in a decade. The accessibility findings are firm
in direction and approximate in magnitude --- modelled journey times,
not measured ones. The model results in sections 7 and 8 are the
softest: the model is swept rather than calibrated, because calibration
needs data the council has not released. Every comparison there is
robust across the swept range; no individual number should be read as a
prediction.

# The secondary school system {#sec-system}

Ten secondary schools serve Brighton & Hove: six community schools run
by the local authority, two academies (Brighton Aldridge and Portslade
Aldridge) and two church schools (King's and Cardinal Newman) that set
their own admissions. An eleventh, Peacehaven Community School, sits
just outside the boundary but inside the travel-to-school geography, and
is included throughout because families on the eastern edge of the city
treat it as a real option.

The city allocates places by **catchment area with a lottery tie-break**
for oversubscribed schools --- an arrangement that is very unusual in
England, where distance from the school gate usually decides priority.
It dates from the closure of CoMArt in the east of the city in 2005, and
the catchment system that replaced distance-based allocation in 2008.
CoMArt is marked on the map below because the shape of the system is
still partly the shape of that closure.

## Where the schools, the catchments and the children are {#sec-orientation}

```{r fig-orientation}
#| fig-cap: "Schools, catchment boundaries and the cohort-aged child population. Use the control at the top right to turn each layer on and off."

# ---- Schools ---------------------------------------------------------
sch <- schools_sf(oi$schools)

sch_popup <- sprintf(
  "<b>%s</b><br>Admission number 2024: %s<br>Planned 2030: %s%s",
  sch$name, sch$pan2024, sch$pan2030,
  ifelse(is.na(sch$catchment),
         "<br><i>Faith school, admits city-wide</i>",
         paste0("<br>Catchment: ", CATCH_LABELS[sch$boundary_catchment])))

# ---- Children per LSOA ----------------------------------------------
# Oi is the cohort-aged child estimate per zone; zones are LSOA x
# catchment, so an LSOA split by a boundary has to be summed back up.
children <- oi$zones %>%
  group_by(lsoa) %>%
  summarise(children = sum(Oi, na.rm = TRUE), .groups = "drop")

lsoa_pts <- lsoa %>%
  left_join(children, by = c("lsoa21cd" = "lsoa")) %>%
  filter(!is.na(children), children > 0) %>%
  st_point_on_surface() %>%
  suppressWarnings()
pt_xy <- st_coordinates(lsoa_pts)
lsoa_pts$lon <- pt_xy[, 1]; lsoa_pts$lat <- pt_xy[, 2]

# Radius on the square root of the count, so circle *area* is
# proportional to the number of children rather than the radius being -
# the latter exaggerates big areas by a factor of the count.
r_scale <- function(n) 3 + 17 * sqrt(n / max(n, na.rm = TRUE))

pal_catch <- colorFactor(unname(CATCH_COLOURS), names(CATCH_COLOURS))

m <- leaflet(width = "100%", height = 620) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -0.10, lat = 50.84, zoom = 12) %>%
  addPolygons(
    data = catch, group = "Catchments",
    fillColor = ~pal_catch(catchment), fillOpacity = 0.13,
    color = ~pal_catch(catchment), weight = 2.5, opacity = 0.85,
    label = ~unname(CATCH_LABELS[catchment]),
    highlightOptions = highlightOptions(weight = 4, fillOpacity = 0.25,
                                        bringToFront = FALSE)) %>%
  addCircleMarkers(
    data = lsoa_pts, lng = ~lon, lat = ~lat, group = "Children",
    radius = ~r_scale(children),
    fillColor = "#1f4e79", fillOpacity = 0.35,
    color = "#1f4e79", weight = 1, opacity = 0.55,
    label = ~sprintf("%s: about %s children in the cohort",
                     lsoa21nm, round(children)))

m <- add_school_layer(m, sch, group = "Schools", popup = sch_popup)

# CoMArt, closed 2005
m <- m %>% addCircleMarkers(
  lng = COMART$lon, lat = COMART$lat, group = "CoMArt (closed 2005)",
  radius = 6, fillColor = "#b2182b", fillOpacity = 0.9,
  color = "white", weight = 2,
  popup = "<b>CoMArt</b><br>Closed 2005. Its closure is why the city moved to catchments with a lottery tie-break.",
  label = COMART$name)
ic <- logo_icon(COMART$logo)
if (!is.null(ic))
  m <- m %>% addMarkers(lng = COMART$lon, lat = COMART$lat, icon = ic,
                        group = "CoMArt (closed 2005)",
                        label = COMART$name)

legend_html <- paste0(
  '<div style="background:white;padding:8px 12px;border-radius:4px;',
  'box-shadow:0 1px 5px rgba(0,0,0,.3);font-family:sans-serif;font-size:12px;line-height:1.6">',
  '<div style="font-weight:bold;margin-bottom:4px">Catchment areas, 2025/26</div>',
  paste0('<div style="display:flex;align-items:center;gap:6px">',
         '<span style="display:inline-block;width:14px;height:14px;background:',
         unname(CATCH_COLOURS), ';border-radius:2px;flex-shrink:0"></span><span>',
         unname(CATCH_LABELS), '</span></div>', collapse = ''),
  '<div style="margin-top:6px;padding-top:6px;border-top:1px solid #ddd">',
  '<span style="display:inline-block;width:14px;height:14px;background:#1f4e79;',
  'opacity:.35;border-radius:50%;margin-right:6px"></span>Children per LSOA',
  ' <span style="color:#777">(area &prop; count)</span></div></div>')

m %>%
  addLayersControl(
    overlayGroups = c("Catchments", "Schools", "Children", "CoMArt (closed 2005)"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("CoMArt (closed 2005)") %>%
  addControl(html = legend_html, position = "bottomright")
```

Three things are worth noticing on that map before any analysis begins.

**The catchments are not equivalent objects.** Four of them contain a
single school; two contain a pair. A child in the Varndean / Dorothy
Stringer or Hove Park / Blatchington Mill catchment has two schools to
name and a realistic chance of one of them. A child in the Longhill,
Patcham, BACA or PACA catchment has one. That asymmetry does more work
in this system than almost anything else, and section 5 returns to it.

**The children are not spread evenly.** The cohort is concentrated in
the centre and west; the east of the city, which contains Longhill's
entire catchment, is thin. Longhill's problem is visible on this map
before any model is run.

**Two schools sit outside the geography entirely.** King's and Cardinal
Newman admit city-wide on faith criteria. Together they hold
`r fmt_n(sum(oi$schools$pan2026[is.na(oi$schools$catchment)]))` places,
about
`r fmt_pct(100 * sum(oi$schools$pan2026[is.na(oi$schools$catchment)]) / sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]))`
of the city total. Any statement about what catchments do has to be read
against that.

## Where the children actually are {#sec-child-density}

The circles on the map above are LSOA totals, and an LSOA is a blunt
unit --- about 1,500 people, drawn for administrative convenience rather
than to describe a neighbourhood. Two finer views of the same population
follow. They are shown side by side because they answer slightly
different questions, and it is worth deciding which one this document
should carry.

```{r pcd-data}
#| include: false

pcd <- bh_data("postcode_children.csv") %>%
  filter(!is.na(easting), !is.na(northing), children > 0)

# Land mask: the union of the city's LSOAs. Without it a smoothed
# surface runs cheerfully out into the Channel.
city <- lsoa %>%
  filter(lsoa21cd %in% unique(pcd$lsoa)) %>%
  st_transform(27700) %>%
  st_union() %>%
  st_make_valid()
```

::: {.panel-tabset}

### Density surface

```{r fig-child-density}
#| fig-cap: "Children aged 0-18 per hectare, smoothed from postcode-level estimates with a 400 m kernel and clipped to the built-up area. Census 2021 household composition apportioned to postcode centroids."

BW <- 400          # kernel bandwidth in metres - a walkable neighbourhood
GRID <- 320        # grid resolution

# A weighted kernel density: each child contributes one point, so the
# surface is the density of children rather than of postcodes. Counts are
# apportioned estimates and so fractional; rounding for replication
# changes the total by well under a percent.
reps <- round(pcd$children)
kx <- rep(pcd$easting,  reps)
ky <- rep(pcd$northing, reps)

bb <- st_bbox(city)
k <- MASS::kde2d(kx, ky, h = c(BW, BW), n = GRID,
                 lims = c(bb["xmin"] - 500, bb["xmax"] + 500,
                          bb["ymin"] - 500, bb["ymax"] + 500))

# kde2d integrates to 1 over square metres. Scale to children per
# hectare so the legend means something to a reader.
dens <- k$z * sum(reps) * 1e4

brks <- c(0, 2, 5, 10, 15, 22, 30, max(45, ceiling(max(dens))))
pal_d <- viridisLite::viridis(length(brks) - 1, option = "magma",
                              direction = -1, begin = 0.08)

# isoband wants z indexed [y, x]; kde2d returns [x, y].
bands <- isoband::isobands(x = k$x, y = k$y, z = t(dens),
                           levels_low = brks[-length(brks)],
                           levels_high = brks[-1])

band_sf <- st_sf(
  lo = brks[-length(brks)], hi = brks[-1],
  geometry = st_sfc(isoband::iso_to_sfg(bands), crs = 27700)) %>%
  st_make_valid() %>%
  st_intersection(city) %>%
  filter(!st_is_empty(geometry)) %>%
  st_transform(4326)

band_sf$col <- pal_d[seq_len(nrow(band_sf))]
band_sf$lab <- sprintf("%g - %g children per hectare", band_sf$lo, band_sf$hi)

leaflet(width = "100%", height = 600) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = band_sf, fillColor = ~col, fillOpacity = 0.72,
              color = NA, weight = 0, label = ~lab, group = "Children") %>%
  addPolygons(data = catch, fill = FALSE, color = "#333333",
              weight = 1.6, opacity = 0.7, group = "Catchments",
              label = ~unname(CATCH_LABELS[catchment])) %>%
  add_school_layer(sch, group = "Schools", popup = sch_popup) %>%
  addLayersControl(overlayGroups = c("Children", "Catchments", "Schools"),
                   options = layersControlOptions(collapsed = TRUE)) %>%
  addLegend(colors = band_sf$col, labels = band_sf$lab,
            title = "Children 0-18<br>per hectare",
            position = "bottomright", opacity = 0.8)
```

### Postcode detail

```{r fig-pcd-dots}
#| fig-cap: "Every residential postcode in the city, sized by the number of children aged 0-18 and coloured by income deprivation affecting children. Decile 1 is the most deprived tenth of neighbourhoods nationally."

pcd_sf <- pcd %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>%
  st_transform(4326) %>%
  filter(!is.na(idaci_decile))

pal_idaci <- colorFactor("RdYlBu", domain = sort(unique(pcd_sf$idaci_decile)))

# Radius on the square root of the count so circle area is proportional
# to children. Scaled well down from the LSOA circles in 2.1: there are
# 4,229 postcodes against 179 LSOAs, so the same per-circle size would
# cover the city in a single wash of ink. The median postcode draws at
# about 3.5 px and the largest at about 9.
leaflet(width = "100%", height = 600) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = catch, fill = FALSE, color = "#333333",
              weight = 1.6, opacity = 0.7, group = "Catchments",
              label = ~unname(CATCH_LABELS[catchment])) %>%
  addCircleMarkers(
    data = pcd_sf, group = "Postcodes",
    radius = ~0.5 + sqrt(children),
    fillColor = ~pal_idaci(idaci_decile), fillOpacity = 0.6,
    stroke = FALSE,
    label = ~sprintf("%s | %.0f children 0-18 | IDACI decile %s",
                     postcode, children, idaci_decile)) %>%
  add_school_layer(sch, group = "Schools", popup = sch_popup) %>%
  addLayersControl(overlayGroups = c("Postcodes", "Catchments", "Schools"),
                   options = layersControlOptions(collapsed = TRUE)) %>%
  addLegend(pal = pal_idaci, values = sort(unique(pcd_sf$idaci_decile)),
            title = "IDACI decile<br><span style='font-weight:normal'>1 = most deprived</span>",
            position = "bottomright", opacity = 0.9)
```

:::

```{r density-check}
#| include: false
# Sanity check on the surface. The comparison has to be like for like:
# the surface is children per hectare, so testing it against the raw
# count in each postcode mostly measures how much postcodes differ in
# size (that correlation is only ~0.36 and means little). Compare it
# instead with the children within 500 m of each postcode, which is the
# same quantity the surface is estimating. A transposed grid would show
# up here as a collapse towards the swapped-index value.
ix <- findInterval(pcd$easting,  k$x)
iy <- findInterval(pcd$northing, k$y)
ok <- ix > 0 & ix <= length(k$x) & iy > 0 & iy <= length(k$y)
sampled <- dens[cbind(ix[ok], iy[ok])]

xy <- as.matrix(pcd[ok, c("easting", "northing")])
ch <- pcd$children[ok]
local500 <- vapply(seq_len(nrow(xy)), function(i)
  sum(ch[(xy[, 1] - xy[i, 1])^2 + (xy[, 2] - xy[i, 2])^2 <= 500^2]),
  numeric(1))

dens_cor  <- cor(sampled, local500, method = "spearman")
dens_swap <- suppressWarnings(
  cor(dens[cbind(iy[ok], ix[ok])], local500, method = "spearman"))
stopifnot(dens_cor > 0.6, dens_cor > dens_swap + 0.2)

n_pcd  <- nrow(pcd)
```

The surface and the dots are built from the same
`r fmt_n(n_pcd)` postcodes and the same
`r fmt_n(sum(pcd$children))` children, and they show the same city ---
the smoothed surface tracks the number of children within 500 metres of
each postcode at a rank correlation of
`r sprintf("%.2f", dens_cor)`. What differs is what each one makes easy
to see.

The **surface** shows where children are concentrated: a dense central
and western band, thinning sharply east of the marina. It answers "where
is the demand?" and it does not pretend to precision it lacks.

The **postcode map** shows the same concentration but adds the social
gradient, and it exposes something the surface cannot --- that
deprivation in this city is fine-grained. Deprived and comfortable
postcodes sit within a few hundred metres of each other, particularly in
the centre. That texture matters later: section 4 shows that poor
transport access and child poverty coincide, and section 5 that
catchment boundaries drawn at this scale will always cut through mixed
ground rather than separating rich areas from poor ones.

::: {.callout-note appearance="simple"}
## Two limits on both maps

These are **Census 2021 output-area household counts apportioned to
postcode centroids**, not a register of children. They are estimates,
and at postcode level they are noisy estimates --- reliable in aggregate,
not for any individual postcode.

They also cover **Brighton & Hove only**. The Peacehaven and Telscombe
area, which contributes about `r fmt_n(sum(oi$zones$Oi[oi$zones$area == "Expansion area"]))`
cohort-aged children to the study area, is absent from both. The
LSOA-level figures used everywhere else in this document do include it.
:::

## What the attainment data actually says {#sec-lever}

The debate about Brighton's schools is usually conducted in the currency
of headline attainment --- Attainment 8, league table position, "good"
and "bad" schools. That currency is close to worthless for the question
the council actually faces, and it is worth being precise about why,
because the usual explanation is also wrong.

```{r lever-models}
#| include: false

# These follow the specification used in "How to Pull the Right Lever"
# and the RPE paper:
#
#   log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL)
#                  + ks2_c + ...
#
# Rates are logged - deprivation, absence, EAL - so their coefficients
# are elasticities. The prior-attainment term is a mean KS2 score
# centred at 100 and entered linearly, because it is a score rather
# than a rate: logging it is neither interpretable nor better fitting.
# That distinction matters. An earlier draft of this section logged a
# *proportion* with low prior attainment instead, which fitted worse and
# silently dropped the 132 highest-attaining intakes in the country,
# since a log cannot take their zero.
#
# Absence here is PERCTOT, the overall absence rate, again matching the
# house model rather than the persistent-absence measure.

# Every model in this section is fitted on the same rows. lm() drops
# incomplete cases silently, so without this the figures would not be
# comparable with each other or with the decomposition below - and the
# difference is not cosmetic. Deprivation's R2 alone is 0.21 across all
# schools reporting it, and 0.37 across the schools that also report a
# KS2 score, because the ones missing a KS2 score are not a random
# sample of schools.
nat <- perf$national %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  mutate(ks2_c = KS2ASS - 100) %>%
  filter(!is.na(ks2_c), !is.na(PTFSM6CLA1A), !is.na(PERCTOT),
         PTFSM6CLA1A > 0, PERCTOT > 0)

m_fsm   <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A), data = nat)
m_prior <- lm(log(ATT8SCR) ~ ks2_c, data = nat)
m_abs1  <- lm(log(ATT8SCR) ~ log(PERCTOT), data = nat)
m_both  <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT), data = nat)
m_full  <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + ks2_c,
              data = nat)

R2 <- function(m) 100 * summary(m)$r.squared
cf <- coef(m_full)

bh_latest <- perf$bh %>%
  filter(year_label == perf$latest, !is.na(ATT8SCR), !is.na(PTFSM6CLA1A),
         !is.na(PERCTOT), !is.na(KS2ASS)) %>%
  mutate(ks2_c = KS2ASS - 100)
bh_resid <- bh_latest %>%
  mutate(pred = exp(predict(m_full, newdata = .)), resid = ATT8SCR - pred)
above <- sum(bh_resid$resid > 0, na.rm = TRUE)
```

The common account is that attainment tracks deprivation. It does, but
weakly, and stopping there points the city at the wrong lever.

The specification below follows the model in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html):
Attainment 8 and the rate variables are logged, so their coefficients
are elasticities, while prior attainment enters as a mean KS2 score
because it is a score and not a rate.

Across `r fmt_n(nobs(m_fsm))` mainstream secondaries in England, the
proportion of disadvantaged pupils explains **`r fmt_pct(R2(m_fsm))`**
of the variation in Attainment 8. Prior attainment explains
**`r fmt_pct(R2(m_prior))`**. And the single strongest of the three is
neither: **absence alone explains `r fmt_pct(R2(m_abs1))`**. All three
together reach `r fmt_pct(R2(m_full))`.

The elasticities make the point more sharply than the R² values do. In
the combined model a one per cent rise in absence is associated with a
**`r sprintf("%.2f", abs(cf[["log(PERCTOT)"]]))` per cent** fall in
Attainment 8; a one per cent rise in the disadvantaged share with a fall
of just **`r sprintf("%.2f", abs(cf[["log(PTFSM6CLA1A)"]]))` per cent**
--- roughly
`r sprintf("%.0f", abs(cf[["log(PERCTOT)"]]) / abs(cf[["log(PTFSM6CLA1A)"]]))`
times smaller.

That ordering is the finding. Who is poor matters least of the three.
What children could already do at eleven matters more. Whether they are
in the room matters most.

```{r fig-att8-prior}
#| fig-cap: "Attainment 8 against two features of intake, every mainstream secondary in England in the latest year, with Brighton & Hove schools highlighted. Prior attainment and absence both order schools far more sharply than deprivation does."
#| fig-height: 5

# Attainment 8 is on a log axis in every panel, matching the model. The
# x axis is logged for the two rates and left linear for the KS2 score,
# so each panel is drawn on the scale its term is actually fitted on.
panel <- function(xvar, xlab, ttl, logx) {
  d <- nat %>% filter(!is.na(.data[[xvar]]))
  b <- bh_latest %>% filter(!is.na(.data[[xvar]]))
  if (logx) { d <- d %>% filter(.data[[xvar]] > 0); b <- b %>% filter(.data[[xvar]] > 0) }
  p <- ggplot(d, aes(.data[[xvar]], ATT8SCR)) +
    geom_point(colour = "grey80", size = 0.9, alpha = 0.5) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
                colour = "grey30", linewidth = 0.7, linetype = "22") +
    geom_point(data = b, colour = "#b2182b", size = 2.4) +
    ggrepel::geom_text_repel(
      data = b %>% filter(SCHNAME %in% c("Longhill High School", "King's School")),
      aes(label = SCHNAME), size = 2.9, colour = "#7f0f22",
      min.segment.length = 0, segment.size = 0.3, seed = 1) +
    scale_y_log10(breaks = c(10, 20, 30, 40, 50, 60, 70)) +
    labs(x = xlab, y = "Attainment 8", subtitle = ttl) +
    theme_bh(11)
  if (logx)
    p <- p + scale_x_log10(labels = label_percent(scale = 1),
                           breaks = c(1, 2, 5, 10, 20, 50))
  p
}

(panel("PTFSM6CLA1A", "Disadvantaged pupils",
       sprintf("Deprivation, logged (R² = %.2f)", summary(m_fsm)$r.squared), TRUE) |
 panel("ks2_c", "Mean KS2 prior attainment, centred",
       sprintf("Prior attainment, linear (R² = %.2f)", summary(m_prior)$r.squared), FALSE) |
 panel("PERCTOT", "Overall absence rate",
       sprintf("Absence, logged (R² = %.2f)", summary(m_abs1)$r.squared), TRUE)) +
  plot_annotation(
    title = "Attainment is a description of intake, and of who turns up",
    subtitle = sprintf("Mainstream secondaries in England, %s. Attainment 8 log scaled throughout; rates logged, the KS2 score linear. Red points are Brighton & Hove.",
                       perf$latest),
    caption = "Source: DfE performance tables. Specification follows How to Pull the Right Lever.",
    theme = theme_bh())
```

Two consequences follow, and they point in the same direction.

**A headline score is mostly a description of who walks through the
door.** Once prior attainment and deprivation are accounted for,
**`r above` of the `r nrow(bh_resid)`** Brighton & Hove schools score
above what the national relationship predicts for their intake. On the
measure that isolates the school's own contribution, this is not a city
with a school quality problem.

**The lever that moves is attendance, not social mix.** Absence and
deprivation travel together, and absence is the mechanism through which
much of the deprivation effect actually operates --- which is why
redistributing disadvantaged children between schools is a far weaker
instrument than it is usually assumed to be.

::: {.callout-important appearance="simple"}
## Two cautions

**Absence is partly an outcome as well as a cause.** A child who is
disengaged attends less *and* attains less, so a single-equation fit
overstates how much of that `r fmt_pct(R2(m_full))` is absence doing the
causal work. The Lever report models this properly --- with school, year,
region and inspection-rating effects, and a first stage that models
absence itself --- rather than reading it off a scatterplot. These three
panels are an illustration of that model's premise, not a substitute for
it.

**These are single-year, single-equation fits.** The full model is
fitted on a multi-year panel with random effects and a wider set of
school characteristics, including English as an additional language,
teacher retention and admissions policy. The ordering of the three terms
here matches it; the individual R² values will not, and none of them
should be quoted as a causal effect.
:::

::: {.callout-tip}
## The detail behind this is published separately

The contextualised analysis --- value-added modelling across every local
authority in England, and what it says about where Brighton's schools
really sit --- is set out in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).
Its central findings are that Brighton's schools perform well once
intake is accounted for, and that **absence, not social mixing, is the
largest single lever available to the city**. This document takes those
findings as given and turns to the spatial and demographic questions the
Lever report does not address.
:::

::: {.callout-tip}
## The detail behind this is published separately

The contextualised analysis --- value-added modelling across every
local authority in England, and what it says about where Brighton's
schools really sit --- is set out in
[How to Pull the Right Lever](https://adamdennett.github.io/school_attainment_tool/index.html).
Its central findings are that Brighton's schools perform well once
intake is accounted for, that **absence, not social mixing, is the
largest single lever available to the city**, and that absence is the
mechanism through which much of the deprivation effect operates. This
document takes those findings as given and turns to the spatial and
demographic questions the Lever report does not address.
:::

## How much can a school change at all? {#sec-school-effects}

There is a prior question underneath all of this, and it is the one that
bears most directly on what the council can achieve by moving children
between schools.

```{r variance-decomp}
#| include: false

# Shapley (LMG) decomposition, following RPE_Paper.qmd. A t-value in a
# multiple regression describes each predictor's *unique* contribution,
# so where predictors are as correlated as these are it cannot rank
# them. The Shapley share instead averages each variable's incremental
# R2 across every order in which the three could enter the model, which
# is the standard way of apportioning variance that predictors share.

r2_of <- function(vs) {
  rhs <- if (!length(vs)) "1" else paste(vs, collapse = " + ")
  summary(lm(as.formula(paste("log(ATT8SCR) ~", rhs)), data = nat))$r.squared
}
three <- c("log(PTFSM6CLA1A)", "log(PERCTOT)", "ks2_c")
perms <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))

shap <- setNames(numeric(3), three)
for (p in perms) {
  sofar <- character(0)
  for (i in p) {
    v <- three[i]
    shap[v] <- shap[v] + (r2_of(c(sofar, v)) - r2_of(sofar))
    sofar <- c(sofar, v)
  }
}
shap <- shap / length(perms)
shap_share <- 100 * shap / sum(shap)

# How much of the explained variance is shared rather than unique: the
# gap between the full model and the sum of each term's last-in
# contribution.
uniq   <- vapply(three, function(v) r2_of(three) - r2_of(setdiff(three, v)),
                 numeric(1))
r2_all <- r2_of(three)
shared_pct <- 100 * (r2_all - sum(uniq)) / r2_all

LAB <- c("log(PTFSM6CLA1A)" = "Disadvantage",
         "log(PERCTOT)"     = "Absence",
         "ks2_c"            = "Prior attainment")

```

Since @Coleman1966, study after study has found that the share of
variation in pupils' outcomes attributable to differences *between
schools* --- as opposed to differences in the pupils they admit and the
circumstances those pupils bring with them --- is somewhere in the order
of **`r SCHOOL_LO` to `r SCHOOL_HI` per cent**. The landmark English
study settled on about a tenth [@SmithTomlinson1989], and the finding
has held up across countries and phases [@TeddlieReynolds2000]. The
decomposition in *How to Pull the Right Lever* recovers the same order
of magnitude.

```{r fig-school-share}
#| fig-cap: "Left: the share of variation in pupil outcomes attributable to differences between schools, from the school-effectiveness literature. Right: how the explained variance in school-level Attainment 8 divides between the three main intake measures, by Shapley decomposition."
#| fig-height: 4.2

share_df <- tibble(
  part = factor(c("Between schools", "Pupils, homes, circumstances"),
                levels = c("Between schools", "Pupils, homes, circumstances")),
  lo = c(SCHOOL_LO, 100 - SCHOOL_HI), hi = c(SCHOOL_HI, 100 - SCHOOL_LO),
  mid = (lo + hi) / 2)

pA <- ggplot(share_df, aes(x = 1, y = mid, fill = part)) +
  geom_col(width = 0.55, colour = "white") +
  geom_text(aes(label = ifelse(part == "Between schools",
                               sprintf("%g-%g%%", lo, hi),
                               sprintf("%g-%g%%", lo, hi))),
            position = position_stack(vjust = 0.5),
            colour = "white", fontface = "bold", size = 3.6) +
  scale_fill_manual(values = c("Between schools" = "#b2182b",
                               "Pupils, homes, circumstances" = "#9aa5b1"),
                    name = NULL) +
  coord_flip() +
  labs(x = NULL, y = "Share of variation in pupil outcomes (%)",
       subtitle = "What a school itself accounts for\n(pupil level, from the literature)") +
  theme_bh(11) +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank(),
        legend.position = "bottom", legend.text = element_text(size = 8))

pB <- tibble(term = LAB[three], share = shap_share) %>%
  ggplot(aes(reorder(term, share), share)) +
  geom_col(fill = "#2166ac", width = 0.62) +
  geom_text(aes(label = fmt_pct(share, 0)), hjust = -0.18, size = 3.4) +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(shap_share) * 1.22)) +
  labs(x = NULL, y = "Share of explained variance (%)",
       subtitle = sprintf("How that intake divides\n(school level, R² = %.2f)", r2_all)) +
  theme_bh(11) +
  theme(panel.grid.major.y = element_blank())

(pA | pB) +
  plot_annotation(
    title = "Most of what separates schools' results arrives with the children",
    caption = "Sources: Coleman (1966); Smith & Tomlinson (1989); Teddlie & Reynolds (2000); DfE performance tables.",
    theme = theme_bh())
```

Two things follow, and together they are the most important finding in
this section for the decisions in front of the council.

**Moving children between schools operates on the small term.** If
`r SCHOOL_LO`--`r SCHOOL_HI` per cent of the variation in outcomes is
attributable to schools, then redistributing pupils works on that share
while leaving untouched the `r 100 - SCHOOL_HI`--`r 100 - SCHOOL_LO` per
cent that arrives with them. It changes *which* school a child attends
without changing the intake characteristics and social circumstances
that account for most of the rest. A policy of social mixing through
admissions is therefore working on the smaller term by construction ---
not because mixing is undesirable, but because the arithmetic limits
what it can deliver.

**And the measured drivers cannot be cleanly separated anyway.** Of the
`r fmt_pct(100 * r2_all, 0)` of school-level variation the three
measures jointly explain, **`r fmt_pct(shared_pct, 0)` is variance they
*share*** rather than variance any one of them owns. Disadvantage,
absence and prior attainment correlate at
`r sprintf("%.2f", cor(log(nat$PTFSM6CLA1A), nat$ks2_c))`,
`r sprintf("%.2f", cor(log(nat$PERCTOT), nat$ks2_c))` and
`r sprintf("%.2f", cor(log(nat$PTFSM6CLA1A), log(nat$PERCTOT)))`
because they are substantially the same phenomenon seen three ways. Any
policy that targets one of them in isolation --- and an admissions
policy targets disadvantage in isolation --- is pulling on a rope
attached to the other two.

::: {.callout-important}
## What this does and does not say

It does **not** say schools do not matter. A tenth of the variance
across three and a half thousand secondary schools is a great deal, and
section 2.3 showed Brighton's schools mostly performing above what their
intakes predict. Nor does it say that segregation between schools is
acceptable; there are fairness arguments for mixed intakes that have
nothing to do with attainment.

What it says is narrower and harder to argue with: **the council should
not expect a redistribution policy to move attainment much**, because
the mechanism it operates through is the smallest of the terms
available. The instruments that reach the larger terms --- attendance
above all --- are not admissions instruments at all.
:::

```{r fig-ranking-flip}
#| fig-cap: "The same ten schools ranked two ways: by headline Attainment 8, and by the value-added measure from the Lever report. Lines crossing means the two measures disagree about which schools are doing well."
#| fig-height: 5.5

flip <- oi$attract %>%
  filter(!is.na(att8), !is.na(va_lever), name != "Peacehaven Community School") %>%
  mutate(rank_att8 = rank(-att8), rank_va = rank(-va_lever)) %>%
  select(name, rank_att8, rank_va) %>%
  pivot_longer(starts_with("rank_"), names_to = "measure", values_to = "rank") %>%
  mutate(measure = factor(measure, c("rank_att8", "rank_va"),
                          c("Attainment 8\n(headline)", "Value added\n(Lever report)")))

moved <- flip %>%
  pivot_wider(names_from = measure, values_from = rank) %>%
  rename(att8 = 2, va = 3) %>%
  mutate(shift = att8 - va)

ggplot(flip, aes(measure, rank, group = name)) +
  geom_line(aes(colour = name), linewidth = 1.1, alpha = 0.85, show.legend = FALSE) +
  geom_point(aes(colour = name), size = 3, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data = flip %>% filter(measure == levels(flip$measure)[1]),
    aes(label = name), hjust = 1, nudge_x = -0.08, size = 3,
    direction = "y", segment.size = 0.2, seed = 1) +
  ggrepel::geom_text_repel(
    data = flip %>% filter(measure == levels(flip$measure)[2]),
    aes(label = name), hjust = 0, nudge_x = 0.08, size = 3,
    direction = "y", segment.size = 0.2, seed = 1) +
  scale_y_reverse(breaks = 1:10) +
  scale_x_discrete(expand = expansion(add = c(0.85, 0.85))) +
  labs(x = NULL, y = "Rank (1 = highest)",
       title = "Which schools look good depends on which measure you use",
       subtitle = "Ranked by headline attainment, and by contribution to progress once intake is accounted for",
       caption = "Sources: DfE performance tables; How to Pull the Right Lever.") +
  theme_bh() +
  theme(panel.grid.major.x = element_blank())
```

```{r flip-stats}
#| include: false
# Ties are real here - two schools fall by the same four places - so
# collapse to a list rather than letting slice_min pick one silently.
and_list <- function(x) if (length(x) < 2) x else
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))

up   <- moved %>% filter(shift == max(shift))
down <- moved %>% filter(shift == min(shift))
n_moved3 <- sum(abs(moved$shift) >= 3)
```

The lines cross a great deal. **`r and_list(up$name)`** rises
`r abs(up$shift[1])` places when you move from the headline measure to
the value-added one; **`r and_list(down$name)`** falls
`r abs(down$shift[1])`. `r n_moved3` of the ten schools move by three
places or more.

The schools that fall are not bad schools, and the schools that rise are
not suddenly good ones. The two measures are answering different
questions: one asks what pupils at this school achieved, the other asks
how much of that the school is responsible for. Families reading the
published headline are answering the first question while believing they
are answering the second --- and section 5 shows that this has a
measurable effect on where demand actually goes.

# Demographic futures {#sec-demography}

The single most important fact about this system is that it is shrinking,
and that the shrinkage is already locked in. The children who will enter
Year 7 in 2033 have already started primary school. Their number is not
a forecast in any meaningful sense --- it is a count, adjusted for a
transfer rate that has been remarkably stable for a decade.

## What has already happened {#sec-cohort-history}

```{r cohort-data}
#| include: false
rc <- bh_data("reception_cohort.rds")

y7    <- rc$secondary_city
rec   <- rc$reception_city
peak  <- y7 %>% slice_max(y7_offers, n = 1)
now   <- y7 %>% slice_max(year, n = 1)
proj  <- rc$projection
last_proj <- proj %>% slice_max(entry_year, n = 1)
drop_pct <- 100 * (last_proj$central - now$y7_offers) / now$y7_offers
```

Year 7 offers across the city peaked at **`r fmt_n(peak$y7_offers)`** in
`r peak$year`. In `r now$year` the figure was
**`r fmt_n(now$y7_offers)`** --- a fall of
`r fmt_pct(100 * (peak$y7_offers - now$y7_offers) / peak$y7_offers, 1)`
from the peak. That decline is not a projection. It has happened.

```{r fig-cohort-history}
#| fig-cap: "Reception and Year 7 offers made across Brighton & Hove. The reception line leads the Year 7 line by seven years, so it is a preview of secondary demand rather than a separate story."
#| fig-height: 5

bind_rows(
  y7  %>% transmute(year, offers = y7_offers, phase = "Year 7 offers"),
  rec %>% transmute(year, offers = rec_offers, phase = "Reception offers")) %>%
  ggplot(aes(year, offers, colour = phase)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Reception offers" = "#2c7fb8",
                                 "Year 7 offers" = "#b2182b"), name = NULL) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Offers made",
       title = "Both phases are past their peak",
       subtitle = "Reception offers turn down first, and Year 7 follows seven years later",
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh()
```

## The cohort already in school {#sec-projection}

The transfer rate from a reception cohort to the Year 7 cohort seven
years later has been strikingly stable. Across the six cohorts where
both ends are observable, **`r fmt_pct(100 * rc$survival_recent, 1)`** of
reception offers reappear as Year 7 offers, with a standard deviation of
just `r fmt_pct(100 * rc$survival_sd, 1)`. The city loses about one child
in nine between ages four and eleven --- to independent schools, to
moves out of the area, and to schools across the boundary.

Applying that rate to reception cohorts already in school gives a
projection that depends on almost no assumptions:

```{r fig-projection}
#| fig-cap: "Year 7 demand projected from reception cohorts already in school. The band is the range implied by the observed year-to-year variation in the transfer rate."
#| fig-height: 5

ggplot() +
  geom_ribbon(data = proj, aes(entry_year, ymin = lo, ymax = hi),
              fill = "#b2182b", alpha = 0.18) +
  geom_line(data = proj, aes(entry_year, central),
            colour = "#b2182b", linewidth = 1.1, linetype = "22") +
  geom_point(data = proj, aes(entry_year, central), colour = "#b2182b", size = 2) +
  geom_line(data = y7, aes(year, y7_offers), colour = "grey25", linewidth = 1.1) +
  geom_point(data = y7, aes(year, y7_offers), colour = "grey25", size = 2) +
  annotate("text", x = 2016, y = 2480, label = "Observed", colour = "grey25",
           hjust = 0, fontface = "bold", size = 3.6) +
  annotate("text", x = 2029.5, y = 2180, label = "Projected from\nchildren already in school",
           colour = "#b2182b", hjust = 0, size = 3.4, lineheight = 0.95) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Year 7 offers",
       title = sprintf("A further %s fall is already in the primary schools",
                       fmt_pct(abs(drop_pct), 0)),
       subtitle = sprintf("From %s in %s to about %s by %s",
                          fmt_n(now$y7_offers), now$year,
                          fmt_n(last_proj$central), last_proj$entry_year),
       caption = "Sources: BHCC allocation factsheets; transfer rate estimated from six observed cohorts.") +
  theme_bh()
```

By **`r last_proj$entry_year`** the city should expect about
**`r fmt_n(last_proj$central)`** Year 7 children, against
`r fmt_n(now$y7_offers)` today --- a further fall of
`r fmt_pct(abs(drop_pct), 0)`. Against a current published admission
capacity of about
`r fmt_n(sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]))`
places, that is roughly
`r fmt_n(sum(oi$schools$pan2026[oi$schools$name != "Peacehaven Community School"]) - last_proj$central)`
surplus places --- the equivalent of two and a half average-sized
secondary schools standing empty.

## Where the council's forecast differs {#sec-council-forecast}

The council publishes its own catchment-level forecasts. They are
consistently **higher** than what the primary cohorts support, and the
gap widens with distance.

```{r fig-forecast-gap}
#| fig-cap: "The council's own demand forecast against a projection built from reception cohorts already in school. Positive values mean the council expects more children than the primary registers contain."
#| fig-height: 4.6

rc$comparison %>%
  mutate(gap = ons_demand - reception_based) %>%
  ggplot(aes(entry_year, gap)) +
  geom_col(fill = "#7b3294", alpha = 0.85, width = 0.65) +
  geom_hline(yintercept = 0, colour = "grey30") +
  geom_text(aes(label = sprintf("%+.0f", gap),
                vjust = ifelse(gap >= 0, -0.4, 1.3)), size = 3.2) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.18))) +
  labs(x = "Year of Year 7 entry", y = "Council forecast minus cohort projection",
       title = "The council expects more children than the primary registers hold",
       subtitle = "Difference in Year 7 places, city-wide",
       caption = "Sources: BHCC published forecasts; reception-based projection as above.") +
  theme_bh()
```

```{r lh-gap}
#| include: false
lh_council <- rc$council %>%
  mutate(catchment = as_model_catchment(CatchmentGroup)) %>%
  filter(catchment == "Longhill")
lh_cohort <- rc$by_catchment %>% filter(catchment == "Longhill")
lh_join <- lh_council %>%
  inner_join(lh_cohort %>% select(entry_year, projected), by = "entry_year") %>%
  mutate(gap = council - projected)
stopifnot(nrow(lh_join) > 0)
```

The city-wide gap is modest in most years. Broken down by catchment it
is not, and Longhill is where it opens widest.

```{r fig-longhill-gap}
#| fig-cap: "Longhill catchment: the council's forecast, the projection from its own primary cohorts, and the school's admission number."
#| fig-height: 4.6

lh_join %>%
  select(entry_year, Council = council, `From primary cohorts` = projected) %>%
  pivot_longer(-entry_year, names_to = "series", values_to = "children") %>%
  ggplot(aes(entry_year, children, colour = series)) +
  geom_hline(yintercept = rc$council_lh_pan, linetype = "31", colour = "grey45") +
  annotate("text", x = min(lh_join$entry_year), y = rc$council_lh_pan + 7,
           label = sprintf("Admission number (%s)", rc$council_lh_pan),
           hjust = 0, size = 3.1, colour = "grey35") +
  geom_line(linewidth = 1.1) + geom_point(size = 2.4) +
  scale_colour_manual(values = c("Council" = "#7b3294",
                                 "From primary cohorts" = "#008837"), name = NULL) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year of Year 7 entry", y = "Children",
       title = "Longhill: the two forecasts disagree, and both sit below the admission number",
       subtitle = sprintf("Mean gap of %.0f children a year",
                          mean(lh_join$gap, na.rm = TRUE)),
       caption = "Sources: BHCC catchment forecasts; reception-based projection.") +
  theme_bh()
```

The council's forecast for the Longhill catchment runs an average of
**`r fmt_n(mean(lh_join$gap, na.rm = TRUE))` children a year above** what
that catchment's own primary cohorts support. Both series sit below the
school's admission number of `r rc$council_lh_pan` in every year. This
matters because decisions about Longhill's future are being taken
against the higher of two numbers, and the higher one is the one the
primary registers do not support.

## The Peacehaven accession {#sec-peacehaven}

```{r peacehaven}
#| include: false
area_children <- oi$zones %>%
  group_by(area) %>%
  summarise(lsoas = n_distinct(lsoa), children = sum(Oi, na.rm = TRUE),
            .groups = "drop")
exp_area <- area_children %>% filter(area == "Expansion area")
tot_children <- sum(area_children$children)
ph_pan <- oi$schools$pan2026[oi$schools$name == "Peacehaven Community School"]
```

The catchment system reaches beyond the city boundary. The Peacehaven
and Telscombe area contributes **`r exp_area$lsoas` LSOAs** and about
**`r fmt_n(exp_area$children)` cohort-aged children**, roughly
`r fmt_pct(100 * exp_area$children / tot_children, 0)`
of the total the model covers.

In headcount terms this is close to self-supporting: Peacehaven
Community School has an admission number of `r ph_pan`, against about
`r fmt_n(exp_area$children)` children in its area. The difficulty is not
the arithmetic but the direction of travel. These families sit at the
eastern end of a transport network that runs east--west along the coast,
and section 4 shows that they are among the least well served in the
whole study area --- not because they are far from *a* school, but
because they are far from *most* schools, and therefore have the fewest
alternatives if their first choice is full.

# How reachable are the schools? {#sec-access}

Almost all of the public argument about Brighton's schools is conducted
in terms of admissions rules --- catchments, priorities, tie-breaks.
Underneath the rules sits a physical constraint that no rule can undo:
how long it actually takes a eleven-year-old to get to each school on a
bus. This section establishes that constraint before section 5 returns
to the rules.

All journey times here are routed with `r5r` over the merged street
network and bus timetable for the area, for a weekday morning arrival.
They include walking to the stop, waiting, the ride itself, and any
transfer.

## Journey time to a single school {#sec-one-school}

```{r fig-one-school}
#| fig-cap: "Walk-and-bus journey time to Longhill High School from every neighbourhood in the study area, for a weekday morning arrival."
#| fig-height: 6

nm <- bh_data("network_maps.rds")

# The surface object carries one point per zone, and zones are
# LSOA x catchment. Collapse to LSOA on the child population and join
# the times onto the LSOA polygons, so this draws as a surface rather
# than as a scatter of centroids.
surf_lsoa <- nm$surface %>%
  st_drop_geometry() %>%
  group_by(lsoa) %>%
  summarise(mins = weighted.mean(ovingdean, pmax(Oi, 1e-6), na.rm = TRUE),
            .groups = "drop")

surf <- lsoa %>%
  inner_join(surf_lsoa, by = c("lsoa21cd" = "lsoa")) %>%
  filter(is.finite(mins))

pal_time <- colorNumeric("magma", domain = surf$mins, reverse = TRUE)

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = surf,
              fillColor = ~pal_time(mins), fillOpacity = 0.78,
              color = "white", weight = 0.4,
              label = ~sprintf("%s: %.0f minutes", lsoa21nm, mins)) %>%
  addCircleMarkers(
    lng = st_coordinates(st_transform(
      st_as_sf(oi$schools %>% filter(short == "longhill"),
               coords = c("easting", "northing"), crs = 27700), 4326))[, 1],
    lat = st_coordinates(st_transform(
      st_as_sf(oi$schools %>% filter(short == "longhill"),
               coords = c("easting", "northing"), crs = 27700), 4326))[, 2],
    radius = 7, fillColor = "#1a9850", fillOpacity = 1,
    color = "white", weight = 2, label = "Longhill High School") %>%
  addLegend(pal = pal_time, values = surf$mins, title = "Minutes",
            position = "bottomright", opacity = 0.85)
```

The median neighbourhood is
**`r fmt_n(median(nm$surface$ovingdean, na.rm = TRUE))` minutes** from
Longhill's current site. The school sits at the edge of its own
catchment and at the edge of the city, which is the geometric fact
underneath most of section 8.

## What one of those journeys actually looks like {#sec-legs}

A single number in minutes hides what the journey involves. These are
seven real routed journeys, drawn leg by leg: walking legs follow the
street network, bus legs follow the service's own shape.

```{r fig-legs}
#| fig-cap: "Seven worked journeys, routed leg by leg. Walking legs in orange, bus legs in blue. Each can be checked against local knowledge."
#| fig-height: 6

rg <- bh_data("route_geometries.rds")
legs <- rg$legs %>% st_transform(4326)

pal_mode <- colorFactor(c("BUS" = "#2166ac", "WALK" = "#e08214"),
                        domain = c("BUS", "WALK"))

mleg <- leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron)

for (j in sort(unique(legs$journey))) {
  d <- legs %>% filter(journey == j)
  tot <- rg$summary$total_minutes[rg$summary$journey == j]
  mleg <- mleg %>%
    addPolylines(data = d, group = j,
                 color = ~pal_mode(leg_mode), weight = 5, opacity = 0.85,
                 label = ~sprintf("%s - %s leg, %.0f min%s", j, leg_mode, minutes,
                                  ifelse(nzchar(route), paste0(" (service ", route, ")"), "")),
                 popup = ~sprintf("<b>%s</b><br>%s minutes in total", j, tot))
}

mleg %>%
  addLayersControl(overlayGroups = sort(unique(legs$journey)),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  addLegend(pal = pal_mode, values = c("BUS", "WALK"), title = "Leg",
            position = "bottomright", opacity = 0.9)
```

```{r tbl-journeys}
#| tbl-cap: "The same seven journeys, summarised."
rg$summary %>%
  transmute(Journey = journey,
            `Total minutes` = total_minutes,
            Legs = legs,
            `Bus services used` = ifelse(nzchar(services), services, "walk only")) %>%
  arrange(`Total minutes`) %>%
  knitr::kable()
```

Woodingdean to Longhill takes 11 minutes. Whitehawk to Dorothy Stringer
takes 52, for a journey of about the same straight-line distance. The
network is not symmetric, and neither is opportunity.

## Which school is actually nearest {#sec-nearest}

```{r fig-nearest}
#| fig-cap: "The nearest secondary school to each neighbourhood by routed walk-and-bus time, regardless of catchment."
#| fig-height: 6

near <- lsoa %>%
  left_join(acc$lsoa %>% select(lsoa, nearest_school, nearest_min),
            by = c("lsoa21cd" = "lsoa")) %>%
  filter(!is.na(nearest_school))

pal_near <- colorFactor("Set3", domain = sort(unique(near$nearest_school)))

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = near,
              fillColor = ~pal_near(nearest_school), fillOpacity = 0.72,
              color = "white", weight = 0.4,
              label = ~sprintf("%s: %s, %.0f min", lsoa21nm, nearest_school, nearest_min)) %>%
  addLegend(pal = pal_near, values = near$nearest_school, title = "Nearest school",
            position = "bottomright", opacity = 0.85)
```

```{r nearest-stats}
#| include: false
near_tab <- acc$lsoa %>%
  count(nearest_school, wt = Oi, name = "children") %>%
  arrange(desc(children)) %>%
  mutate(pct = 100 * children / sum(children))
```

This map is worth comparing with the catchment map in section 2. They
are not the same shape. The nearest school by bus is not always the
catchment school, and for
`r fmt_pct(100 * mean(acc$lsoa$nearest_min > 25), 0)`
of neighbourhoods even the nearest school is more than 25 minutes away.

## Potential accessibility: how much school is within reach {#sec-potential}

Nearest-school distance answers a narrow question. It says nothing about
whether a neighbourhood has *one* school within reach or *five* --- and
that difference is the whole of what a family's realistic options
amount to when their first preference is full.

Two measures are used here, deliberately, because they fail in opposite
directions.

::: {.panel-tabset}
### Gravity

$$A_i = \sum_j W_j \, c_{ij}^{-\beta}$$

Every school counts, discounted by how far away it is. $W_j$ is the
school's attractiveness (its admission number), $c_{ij}$ the routed
walk-and-bus time, and $\beta$ the distance decay --- set here at
`r sprintf("%.1f", acc$beta$ref)`, the reference value from the model in
section 7. The number has no units, so it is indexed with the
child-weighted city average at 100.

### Cumulative opportunity

$$P_i(t) = \sum_j \mathrm{PAN}_j \cdot \mathbf{1}(c_{ij} \le t)$$

The number of school places reachable inside $t$ minutes. Trivial to
explain --- "how many places can you get to in half an hour" --- but it
treats a 29-minute journey and a 31-minute journey as entirely different
things.
:::

```{r fig-accessibility}
#| fig-cap: "Potential accessibility to secondary school places, by neighbourhood. Indexed so that the child-weighted city average is 100. Darker means less school within reach."
#| fig-height: 6

acc_map <- lsoa %>%
  inner_join(acc$lsoa, by = c("lsoa21cd" = "lsoa"))

pal_acc <- colorNumeric("viridis", domain = acc_map$A_index, reverse = TRUE)

leaflet(width = "100%", height = 560) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = acc_map,
              fillColor = ~pal_acc(A_index), fillOpacity = 0.8,
              color = "white", weight = 0.4,
              label = ~sprintf(
                "%s | index %.0f | %.0f places within 30 min | nearest %s, %.0f min",
                lsoa21nm, A_index, places_30, nearest_school, nearest_min)) %>%
  addLegend(pal = pal_acc, values = acc_map$A_index,
            title = "Accessibility<br>index", position = "bottomright",
            opacity = 0.85)
```

```{r acc-stats}
#| include: false
zero30 <- acc$lsoa %>% filter(places_30 < 1)
rng <- range(acc$lsoa$A_index)
```

Three findings come out of this, and the first is the one that should
change the conversation.

**`r nrow(zero30)` of the `r nrow(acc$lsoa)` neighbourhoods --- about
`r fmt_n(sum(zero30$Oi))` cohort-aged children --- cannot reach a single
secondary school place within 30 minutes** by walking and bus on a
weekday morning. Not their catchment school; *any* school. For those
neighbourhoods the admissions debate is somewhat beside the point.

**The spread is nearly fivefold.** The best-served neighbourhood scores
`r fmt_n(rng[2])` on the index against `r fmt_n(rng[1])` for the worst.
A child in the centre of the city has several schools genuinely within
reach; a child at either end has one, or none.

**The two measures broadly agree, but not closely.** They correlate at
`r sprintf("%.2f", acc$agreement$spearman)` (Spearman), yet only
`r fmt_pct(100 * acc$agreement$within_one, 0)` of neighbourhoods land
within one decile of each other on both. Where they agree, the finding
is robust to how you define access. Where they disagree, it is because
one is counting a school just over the threshold that the other is
discounting smoothly.

```{r fig-acc-compare}
#| fig-cap: "The two measures against each other. Neighbourhoods along the bottom can reach no places at all inside 30 minutes."
#| fig-height: 4.8

acc$lsoa %>%
  mutate(grp = ifelse(places_30 < 1, "No places within 30 min", "Some places within 30 min")) %>%
  ggplot(aes(A_index, places_30, colour = grp, size = Oi)) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(values = c("No places within 30 min" = "#b2182b",
                                 "Some places within 30 min" = "#2166ac"),
                      name = NULL) +
  scale_size_area(max_size = 6, guide = "none") +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Gravity accessibility index (city average = 100)",
       y = "Places reachable within 30 minutes",
       title = "Two ways of asking the same question",
       subtitle = "Each point is a neighbourhood, sized by the number of cohort-aged children",
       caption = "Routed walk-and-bus times, weekday morning arrival.") +
  theme_bh()
```

::: {.callout-warning appearance="simple"}
## Two caveats that bear on these numbers

**Journey times are modelled, not measured.** They come from `r5r`
routing over the merged street network and published timetables for one
weekday morning. They do not know about school buses, parental lifts,
walking routes children actually use, or a bus that is full when it
arrives.

**One school's times are suspect.** Dorothy Stringer and Varndean are
470 metres apart, yet the routed matrix makes Stringer slower from most
of the city by a median of about five and a half minutes. That may be
real, or it may be an artefact of where the router snapped each school
onto the network. It has not been established which, and it inflates the
apparent inaccessibility of the areas whose nearest school is Stringer.
:::

## Accessibility and child poverty together {#sec-bivariate}

The question that matters for policy is not where access is poor, nor
where deprivation is high, but where the two coincide --- because those
are the neighbourhoods where a difficult journey is least likely to be
solved by a car in the family.

```{r fig-bivariate}
#| fig-cap: "Accessibility and income deprivation affecting children, mapped together. The darkest blue is the corner that matters: least reachable school places, highest child poverty."
#| fig-height: 7

# Stevens-style bivariate palette. Keyed "access-deprivation", each on a
# tercile: access 1 is the least reachable third, deprivation 3 the most
# deprived third. The concerning corner is therefore 1-3, and it is the
# darkest colour.
BIV <- c(
  "1-1" = "#be64ac", "1-2" = "#8c62aa", "1-3" = "#3b4994",
  "2-1" = "#dfb0d6", "2-2" = "#a5add3", "2-3" = "#5698b9",
  "3-1" = "#e8e8e8", "3-2" = "#ace4e4", "3-3" = "#5ac8c8")

biv_map <- lsoa %>%
  inner_join(acc$bivariate %>% select(lsoa, acc_t, dep_t, biv_key,
                                      idaci_score, A_index, places_30, Oi),
             by = c("lsoa21cd" = "lsoa")) %>%
  mutate(fill = unname(BIV[biv_key]))

main <- ggplot(biv_map) +
  geom_sf(aes(fill = fill), colour = "white", linewidth = 0.12) +
  scale_fill_identity() +
  labs(title = "Where poor access and child poverty coincide",
       subtitle = "Each neighbourhood classed on both axes at once",
       caption = "Sources: routed walk-and-bus accessibility; IDACI 2019.") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey30"),
        plot.caption = element_text(colour = "grey45", hjust = 0))

key <- expand_grid(acc_t = 1:3, dep_t = 1:3) %>%
  mutate(fill = unname(BIV[paste0(acc_t, "-", dep_t)])) %>%
  ggplot(aes(acc_t, dep_t, fill = fill)) +
  geom_tile(colour = "white", linewidth = 0.7) +
  scale_fill_identity() +
  scale_x_continuous(breaks = c(1, 3), labels = c("least", "most"),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = c(1, 3), labels = c("least", "most"),
                     expand = c(0, 0)) +
  labs(x = "Accessible →", y = "Deprived →") +
  coord_equal() +
  theme_minimal(base_size = 8) +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 7.5, colour = "grey25"),
        axis.text = element_text(size = 6.5, colour = "grey40"))

main + inset_element(key, left = 0.72, bottom = 0.02, right = 0.99, top = 0.32)
```

```{r biv-stats}
#| include: false
corner <- acc$worst_corner
best_corner <- acc$bivariate %>% filter(acc_t == 3, dep_t == 3)
cells <- acc$bivariate %>%
  count(acc_t, dep_t, wt = Oi, name = "children")
low_acc_high_dep <- cells %>% filter(acc_t == 1, dep_t == 3) %>% pull(children)
hi_acc_high_dep  <- cells %>% filter(acc_t == 3, dep_t == 3) %>% pull(children)
```

There is a real gradient here, and it runs the wrong way. Accessibility
and child poverty correlate at
`r sprintf("%.2f", acc$acc_dep_spearman)` (Spearman): the more deprived
a neighbourhood, the less school it can reach.

**`r nrow(corner)` neighbourhoods --- about `r fmt_n(low_acc_high_dep)`
cohort-aged children --- sit in the worst corner**, in the least
accessible third of the city and the most deprived third at the same
time. At the other end, only `r nrow(best_corner)` neighbourhoods
(`r fmt_n(hi_acc_high_dep)` children) combine high deprivation with good
access.

This is the strongest argument in the document for treating transport as
a school-admissions instrument rather than a separate departmental
concern. A city that wanted to widen real choice for its poorest
families could do more with a bus timetable than with a catchment
boundary --- and unlike a boundary change, it would take nothing away
from anyone else.

::: {.callout-note appearance="simple"}
## What pupil-level data would add here

This section works entirely with modelled journey times between
neighbourhood centroids and school gates. The council holds the home
postcode of every applicant and the school each was allocated. With
that, the same analysis could be run on **journeys children actually
make** rather than journeys they could in principle make --- and the gap
between the two is precisely the quantity that would tell the city
whether its transport network is a real constraint on choice or merely a
theoretical one. Section 9 sets out what that request would look like.
:::

# Admissions rules, and what families actually do {#sec-demand}

## How a place is allocated {#sec-rules}

Families name up to three schools in order. Each school then ranks its
applicants against its own criteria, and the authority runs an
equal-preference matching so that a child is offered the highest-ranked
school that will take them. The catchment criterion sits partway down
most schools' lists, below looked-after children and siblings and, since
2023, below a quota for children eligible for free school meals.

The critical point is one the rules themselves obscure. **A criterion
only does something if a school has to turn someone away.**

```{r fig-rationing}
#| fig-cap: "For each school, the proportion of first-preference applicants who were offered a place. A school at 100% turned nobody away, so its admission criteria did no work at all."
#| fig-height: 5

conv <- bh_data("adjudicator_conversion.rds")

rat <- conv$conv %>%
  group_by(school) %>%
  summarise(p1_named = sum(p1_named), p1_offered = sum(p1_offered),
            .groups = "drop") %>%
  mutate(rate = p1_offered / p1_named,
         rations = rate < 0.999) %>%
  arrange(rate)

ggplot(rat, aes(reorder(school, rate), rate, fill = rations)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, colour = "grey40", linetype = "31") +
  geom_text(aes(label = fmt_pct(100 * rate, 0)), hjust = -0.15, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1.12),
                     breaks = seq(0, 1, 0.25)) +
  scale_fill_manual(values = c(`TRUE` = "#b2182b", `FALSE` = "#4d9221"),
                    labels = c(`TRUE` = "Rations places",
                               `FALSE` = "Turns nobody away"), name = NULL) +
  labs(x = NULL, y = "First preferences that received an offer",
       title = "Only some of these schools ration anything",
       subtitle = "Pooled across three admissions rounds",
       caption = "Source: BHCC evidence to the Schools Adjudicator.") +
  theme_bh() +
  theme(panel.grid.major.y = element_blank())
```

```{r rationing-stats}
#| include: false
n_open  <- sum(!rat$rations)
n_ration <- sum(rat$rations)
open_names <- rat %>% filter(!rations) %>% pull(school)
```

**`r n_open` of the `r nrow(rat)` schools offered a place to every single
first-preference applicant.** For those schools --- `r paste(open_names, collapse = ", ")`
--- the catchment criterion, the sibling criterion and the free school
meals quota are all inert. They admit everyone who asks.

The city therefore runs two admissions systems under one set of rules.
In `r n_ration` schools the criteria bind and the lottery matters. In
the other `r n_open` they are decoration. Debate about admissions
criteria is, in practice, debate about a minority of the city's schools
--- and never about the schools with the most spare capacity.

## Sixteen years of demand {#sec-demand-history}

```{r fig-demand-history}
#| fig-cap: "First preferences expressed for each school, 2010 onwards. The two church schools admit city-wide and are shown separately."
#| fig-height: 6.5

fp <- bh_data("factsheet_panel.rds")

fs <- fp$factsheets %>%
  filter(!is.na(pref1), name != "Total") %>%
  mutate(kind = ifelse(name %in% c("King's School", "Cardinal Newman Catholic School"),
                       "Faith schools (city-wide)", "Catchment schools"))

ggplot(fs, aes(year, pref1, colour = name)) +
  geom_line(linewidth = 0.95) +
  geom_point(size = 1.4) +
  facet_wrap(~ kind, ncol = 1, scales = "free_y") +
  scale_colour_brewer(palette = "Paired", name = NULL) +
  labs(x = NULL, y = "First preferences",
       title = "Demand has moved a long way between schools",
       subtitle = "First preferences expressed, by school",
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh() +
  theme(legend.position = "right", legend.text = element_text(size = 8))
```

## Who gained and who lost {#sec-winners-losers}

```{r fig-demand-change}
#| fig-cap: "Change in first preferences between the first and last five-year windows in the published series."
#| fig-height: 5

yr_rng <- range(fs$year)
early <- fs %>% filter(year <= yr_rng[1] + 4) %>%
  group_by(name) %>% summarise(early = mean(pref1), .groups = "drop")
late  <- fs %>% filter(year >= yr_rng[2] - 4) %>%
  group_by(name) %>% summarise(late = mean(pref1), .groups = "drop")

chg <- inner_join(early, late, by = "name") %>%
  mutate(change = late - early, pct = 100 * change / early) %>%
  arrange(change)

ggplot(chg, aes(reorder(name, change), change, fill = change > 0)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, colour = "grey30") +
  coord_flip() +
  scale_fill_manual(values = c(`TRUE` = "#2166ac", `FALSE` = "#b2182b"),
                    guide = "none") +
  labs(x = NULL, y = "Change in average first preferences",
       title = "The winners and losers of sixteen years of choice",
       subtitle = sprintf("Average of %d-%d against average of %d-%d",
                          yr_rng[1], yr_rng[1] + 4, yr_rng[2] - 4, yr_rng[2]),
       caption = "Source: BHCC published allocation factsheets.") +
  theme_bh() +
  theme(panel.grid.major.y = element_blank())
```

```{r chg-stats}
#| include: false
biggest_loss <- chg %>% slice_min(change, n = 2)
biggest_gain <- chg %>% slice_max(change, n = 2)
```

The largest falls are at **`r paste(biggest_loss$name, collapse = " and ")`**;
the largest rises at **`r paste(biggest_gain$name, collapse = " and ")`**.
This is not a system in equilibrium. It is one where a shrinking cohort
is being redistributed towards a small number of schools, and away from
others, faster than the cohort itself is shrinking.

## What families are actually choosing on {#sec-choice}

The council's admissions guide points families at published performance
data. It is worth asking which published number they respond to.

```{r fig-choice}
#| fig-cap: "First preferences per place against headline attainment and against value added. Each point is a school."
#| fig-height: 4.8

os <- bh_data("open_scenarios.rds")
ch <- os$choice

p1 <- ggplot(ch, aes(att8, prefs_per_place)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey35", linewidth = 0.7, linetype = "22") +
  geom_point(colour = "#b2182b", size = 3) +
  ggrepel::geom_text_repel(aes(label = school), size = 2.7, seed = 1,
                           max.overlaps = 20) +
  labs(x = "Attainment 8 (headline)", y = "First preferences per place",
       subtitle = sprintf("Headline attainment (R² = %.2f)",
                          os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"])) +
  theme_bh(11)

p2 <- ggplot(ch, aes(va, prefs_per_place)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey35", linewidth = 0.7, linetype = "22") +
  geom_point(colour = "#2166ac", size = 3) +
  ggrepel::geom_text_repel(aes(label = school), size = 2.7, seed = 1,
                           max.overlaps = 20) +
  labs(x = "Value added (Lever report)", y = NULL,
       subtitle = sprintf("Value added (R² = %.2f)",
                          os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"])) +
  theme_bh(11)

(p1 | p2) +
  plot_annotation(
    title = "Families track the headline score, not the school's contribution",
    caption = "Sources: BHCC factsheets; DfE performance tables; How to Pull the Right Lever.",
    theme = theme_bh())
```

```{r tbl-choice}
#| tbl-cap: "Explaining first preferences per place."
os$choice_tbl %>%
  transmute(Specification = spec,
            `R²` = sprintf("%.2f", r2),
            Coefficients = terms) %>%
  knitr::kable()
```

The result is stark. Headline Attainment 8 explains
**`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"], 0)`**
of the variation in how heavily each school is preferred. The
value-added measure --- the one that actually isolates what the school
contributes --- explains
**`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"], 0)`**.
Adding value added to the attainment model barely moves the fit, and its
coefficient turns *negative*.

Families are choosing on a number that is mostly a description of the
existing intake. That makes school choice partly self-fulfilling: a
school with a difficult intake posts a low headline score, is chosen
less, and receives a still more difficult intake next year. Section 2
showed that Hove Park and BACA are among the schools most penalised by
this conflation. This is a communications problem before it is an
admissions problem, and the authority is not a neutral party in it,
because the authority publishes the guide.

::: {.callout-note appearance="simple"}
## What pupil-level data would add here

This is a school-level association across ten schools --- ten points.
It cannot distinguish families responding to attainment from families
responding to something correlated with it, and it cannot say whether
different kinds of family respond differently. The council holds the
preference ordering of every applicant. With it, the same question
becomes a discrete-choice model over thousands of decisions, and the
answer would be worth acting on rather than merely worth noting.
:::

# Money {#sec-money}

A school's finances follow its roll, because funding is overwhelmingly
per-pupil. A school losing children is therefore losing income against a
cost base that does not shrink at the same rate --- the building, the
leadership team and the curriculum offer are all substantially fixed.

```{r fig-size-finance}
#| fig-cap: "Income per pupil and the proportion of schools in deficit, by size band, nationally."
#| fig-height: 4.8

sf_ <- os$size_fin %>%
  mutate(size_band = factor(size_band, levels = size_band))

pa <- ggplot(sf_, aes(size_band, income_pp)) +
  geom_col(fill = "#2166ac", width = 0.7) +
  scale_y_continuous(labels = label_dollar(prefix = "£", accuracy = 1)) +
  labs(x = NULL, y = NULL, subtitle = "Income per pupil") +
  theme_bh(11) + theme(axis.text.x = element_text(angle = 40, hjust = 1))

pb <- ggplot(sf_, aes(size_band, pct_deficit)) +
  geom_col(fill = "#b2182b", width = 0.7) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(x = NULL, y = NULL, subtitle = "Schools in deficit") +
  theme_bh(11) + theme(axis.text.x = element_text(angle = 40, hjust = 1))

(pa | pb) +
  plot_annotation(
    title = "Small schools receive more per pupil and are still likelier to be in deficit",
    subtitle = "Secondary schools in England, by roll",
    caption = "Source: DfE school income and expenditure data.",
    theme = theme_bh())
```

Small schools are not starved of money per head --- they receive
**`r fmt_n(sf_$income_pp[1] - sf_$income_pp[nrow(sf_)])` more per pupil**
than the largest band. They are nonetheless the likeliest to be in
deficit: `r fmt_pct(sf_$pct_deficit[1], 0)` of schools under 500 pupils,
against `r fmt_pct(min(sf_$pct_deficit), 0)` at the most favourable size.
Extra funding per head does not compensate for the loss of scale.

## Longhill's position {#sec-longhill-money}

```{r tbl-longhill-finance}
#| tbl-cap: "Longhill High School: roll, per-pupil income and expenditure, in-year balance and accumulated reserve."
os$lh_fin %>%
  transmute(Year = year_label,
            Roll = ifelse(is.na(roll), "--", fmt_n(roll)),
            `Income / pupil` = paste0("£", fmt_n(income_pp)),
            `Spend / pupil` = paste0("£", fmt_n(expenditure_pp)),
            `In-year balance` = paste0("£", fmt_n(balance)),
            `Reserve` = paste0("£", fmt_n(reserve))) %>%
  knitr::kable(align = "lrrrrr")
```

Longhill's roll fell from `r fmt_n(os$lh_fin$roll[1])` to
`r fmt_n(os$lh_fin$roll[nrow(os$lh_fin)])` across four years. Its
per-pupil income rose over the same period --- the funding system is
doing what it is meant to do --- and its per-pupil spending rose
faster. The accumulated reserve has fallen from
`r paste0("£", fmt_n(os$lh_fin$reserve[1]))` to
`r paste0("£", fmt_n(os$lh_fin$reserve[nrow(os$lh_fin)]))`.

At the recent rate of depletion --- about
**`r paste0("£", fmt_n(os$reserve_burn))` a year** --- the reserve
supports roughly **`r sprintf("%.1f", os$years_left)` more years**.
That is the constraint that makes this a decision with a deadline rather
than a decision that can be deferred.

# Brightopia: a model of the whole system {#sec-brightopia}

## Why model at all {#sec-why-model}

Every argument in the sections above is about one part of the system:
where the children are, how far the schools are, what families prefer.
A model is what lets you ask what happens when you change one part and
everything else adjusts. Close a school and its children do not vanish;
they appear somewhere else, displacing others. Shrink one school's
admission number and demand redistributes to its neighbours. Nothing in
a spreadsheet of admission numbers captures that.

**Brightopia** is a production-constrained spatial interaction model. It
takes the children in each neighbourhood, the attractiveness of each
school, and the travel cost between them, and predicts the flow:

$$T_{ij} = A_i O_i W_j^{\alpha} c_{ij}^{-\beta}$$

$O_i$ is the number of cohort-aged children in neighbourhood $i$; $W_j$
is school $j$'s attractiveness; $c_{ij}$ is the routed walk-and-bus
cost; $\beta$ governs how sharply demand falls away with travel time;
and $A_i$ is a balancing factor ensuring every child goes somewhere.

The model deliberately contains **no catchment term**. That is the
point. It answers the question "where would children go if only
geography and school size mattered?" --- and the gap between that answer
and reality is a measure of what the admissions system is doing.

```{r fig-brightopia}
#| fig-cap: "Modelled intake against published admission number. A school above the line is one that geography alone would fill; below it, one that geography alone would not."
#| fig-height: 5.5

bt <- brt$at_original_beta %>%
  mutate(fill_rate = modelled / pan2026,
         short = str_remove(name, " (School|High School|Community Academy|Catholic School).*"))

ggplot(bt, aes(pan2026, modelled)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey45", linetype = "31") +
  geom_point(aes(colour = fill_rate >= 1), size = 3.5) +
  ggrepel::geom_text_repel(aes(label = name), size = 2.8, seed = 1,
                           max.overlaps = 20) +
  scale_colour_manual(values = c(`TRUE` = "#2166ac", `FALSE` = "#b2182b"),
                      labels = c(`TRUE` = "Geography would fill it",
                                 `FALSE` = "Geography would not"), name = NULL) +
  scale_x_continuous(limits = c(0, NA)) + scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Published admission number, 2026", y = "Modelled intake",
       title = "What the map alone would do",
       subtitle = sprintf("Distance decay β = %.1f. The dashed line is a school exactly full.",
                          brt$beta_original),
       caption = "Brightopia, open-data specification.") +
  theme_bh()
```

```{r brightopia-stats}
#| include: false
lh_bt <- bt %>%
  filter(str_detect(name, "Longhill")) %>%
  mutate(fill_2024 = modelled / pan2024)
hp_bt <- bt %>% filter(str_detect(name, "Hove Park"))
lh_rng <- range(brt$lh_sweep$modelled_now)
```

City-wide there are `r fmt_n(brt$city_places)` places for
`r fmt_n(brt$city_children)` children --- a fill rate of
`r fmt_pct(100 * brt$city_fill, 0)` before any child expresses a
preference. The system is not short of capacity; it is short of children.

Two schools are worth reading off this chart carefully, and they point
in opposite directions.

**Longhill has already been resized to roughly what geography supports.**
Its modelled intake is **`r fmt_n(lh_bt$modelled)`** children. Against
its former admission number of `r fmt_n(lh_bt$pan2024)` that is
`r fmt_pct(100 * lh_bt$fill_2024, 0)`; against the reduced number of
`r fmt_n(lh_bt$pan2026)` now in force it is
`r fmt_pct(100 * lh_bt$fill_rate, 0)`. The distinction matters, and it
is easy to quote the wrong one. Geography can fill 210 places at
Ovingdean; it could not fill 270. The reduction has already done most of
the work that a reduction can do --- which is why section 8 is about
what remains, and why the modelled intake is remarkably insensitive to
the decay parameter, varying only between
`r fmt_n(lh_rng[1])` and `r fmt_n(lh_rng[2])` across the entire swept
range. Longhill's catchment simply does not contain many more children
to attract.

**Hove Park's problem is not its location.** Geography alone would fill
`r fmt_pct(100 * hp_bt$fill_rate, 0)` of its places --- it is the most
over-subscribed school on the map, sitting in the densest part of the
child population. Yet section 5 shows it among the largest losers of
real demand over sixteen years. A school that the map says should be
full and that families are nonetheless leaving is not a geography
problem, and no catchment or relocation policy reaches it. That gap
between what location predicts and what families do is the clearest
evidence in this document that choice is operating on something other
than distance.

::: {.callout-warning appearance="simple"}
## This is an open-data model, and it is not calibrated

$\beta$ cannot be estimated without pupil-level flows, so it is swept
across a plausible range rather than fitted. Figures here use the
reference value; the range is
`r sprintf("%.1f to %.1f", min(env$betas), max(env$betas))`, and section
8 reports the band rather than a point wherever the conclusion depends
on it. A calibrated model built on the council's own allocation records
would replace every band in this document with an estimate.
:::

# Longhill {#sec-longhill}

Longhill High School sits at Ovingdean, at the eastern edge of both its
catchment and the city. Its roll has fallen by
`r fmt_pct(100 * (os$lh_fin$roll[1] - os$lh_fin$roll[nrow(os$lh_fin)]) / os$lh_fin$roll[1], 0)`
in four years and its reserve supports about
`r sprintf("%.1f", os$years_left)` more years.

Section 7 established the constraint this section has to work inside.
The school draws about `r fmt_n(lh_bt$modelled)` children on geography
alone, and that figure barely moves across the swept parameter range.
Reducing the admission number to `r fmt_n(lh_bt$pan2026)` has already
closed most of the gap between capacity and catchment. What remains is a
cohort that keeps falling (section 3) and a reserve that runs out
(section 6).

So the question is not whether something must change, nor whether the
admission number should come down again --- that lever is largely spent.
It is whether any *combination* of the available changes produces a
school that is viable at the end of the projection period rather than
merely smaller.

```{r fig-longhill-configs}
#| fig-cap: "Modelled Longhill intake under seven configurations, across the projection period. Bands show the range across the swept parameter space."
#| fig-height: 6

lb <- os$lh_band %>%
  mutate(config = factor(config, levels = os$configs))

ggplot(lb, aes(entry_year, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#2166ac", alpha = 0.2) +
  geom_line(colour = "#2166ac", linewidth = 1) +
  geom_point(colour = "#2166ac", size = 1.8) +
  facet_wrap(~ config, ncol = 2, labeller = label_wrap_gen(38)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year of Year 7 entry", y = "Modelled intake",
       title = "Seven futures for Longhill",
       subtitle = "Median and full range across the swept parameter space",
       caption = "Brightopia, open-data specification.") +
  theme_bh(11) +
  theme(strip.text = element_text(size = 8.5))
```

```{r longhill-config-stats}
#| include: false
cfg_2030 <- lb %>% filter(entry_year == 2030) %>% arrange(desc(med))
best <- cfg_2030 %>% slice_max(med, n = 1)
base <- cfg_2030 %>% filter(str_detect(as.character(config), "^A\\."))
```

By 2030 the configurations separate clearly. The status quo
(`r as.character(base$config)`) gives a median modelled intake of
**`r fmt_n(base$med)`**; the strongest option
(`r as.character(best$config)`) gives **`r fmt_n(best$med)`**.

The important point is not which single option wins. It is that **no
single instrument is sufficient, and the instruments compose.** Moving
the school changes its position in the travel network. Reducing its
admission number changes what "full" means. Redrawing catchments changes
which children are steered towards it. Each on its own leaves a school
that is still under-filled on the central specification; taken together
they reach a configuration that works. A decision that picks one lever
and rejects the others is choosing the worst of the available futures.

::: {.callout-important appearance="simple"}
## Read these figures with the open model's limits in mind

These are modelled intakes from an uncalibrated model with no catchment
term, and they are systematically more optimistic about relocation than
a calibrated model would be. The direction of every comparison here is
robust across the swept range; the magnitudes are not. Nothing in this
section should be read as a prediction of a roll.
:::

# What the council needs to do {#sec-council}

## Decisions {#sec-decisions}

**Decide about Longhill on a deadline, not on a consultation cycle.**
The reserve supports about `r sprintf("%.1f", os$years_left)` more years.
A decision deferred past that point is a decision taken by insolvency
rather than by the council.

**Treat the instruments as a package.** Section 8 shows they compose.
Section 3 shows the cohort keeps falling regardless. Adjusting a single
admission number and waiting to see what happens will consume years the
finances do not have.

**Stop pointing families at Attainment 8.** The council publishes the
admissions guide. Section 5 shows families track the headline score at
`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"], 0)`
and the value-added measure at
`r fmt_pct(100 * os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"], 0)`.
Publishing a contextualised measure alongside the headline would cost
nothing and would begin to unwind a self-fulfilling loop the authority
currently sustains.

**Treat transport as an admissions instrument.** Section 4 found
`r nrow(acc$lsoa %>% filter(places_30 < 1))` neighbourhoods that can
reach no school place inside 30 minutes, concentrated in the more
deprived third of the city. A bus timetable can widen real choice
without taking anything from anyone --- which is more than can be said
for any boundary change.

## Data the council could simply publish {#sec-data-asks}

Every "this cannot be done on published data" note in this document
resolves to a small number of releases. None requires new collection;
all are aggregates of data the council already holds.

| # | What | What it would settle |
|---|------|---------------------|
| 1 | Home LSOA of every applicant, by year and allocated school | Calibrate $\beta$; replace every band in section 8 with an estimate |
| 2 | Full preference ordering, anonymised, by year | Turn section 5's ten-point association into a discrete-choice model |
| 3 | Allocated school by home LSOA, including schools outside the city | Measure the actual outflow the council's forecasts assume |
| 4 | Criterion under which each place was allocated | Establish directly which criteria bind, rather than inferring it |
| 5 | Children allocated a school they did not name, and which | Test the council's stated "nearest school with places" practice |
| 6 | Journey actually made, or at least distance, per allocation | Separate potential accessibility from realised accessibility |

LSOAs average about 1,500 residents and 650 households. A cohort-year
count at LSOA level is not disclosive in the way a postcode-level count
would be, and the council already publishes LSOA-level material
elsewhere.

## Why now {#sec-why-now}

The cohort projections in section 3 are not forecasts. Those children
are in the city's primary schools today. The financial position in
section 6 has a horizon of about
`r sprintf("%.1f", os$years_left)` years. The decisions in front of the
council will shape the school system for the next fifteen to twenty
years, and they are being taken without a picture of the whole system.

This document is an attempt to supply as much of that picture as
published data allows. It stops well short of what the council's own
records would support --- and the offer to do that work, properly and in
the open, stands.
