---
title: "The model, stage by stage"
subtitle: "A technical guide for checking each element"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
abstract: |
  This document takes the model the simulator runs and opens it up one
  element at a time. Each section states what a term is, where its data
  comes from, exactly how it is computed, what it looks like for a worked
  example, the checks that can be run against it, and what it assumes. It
  is written to be audited: every number is produced by the code in this
  repository, and every step should be reproducible by hand.
---

```{r setup}
#| include: false
source(here::here("R", "00_core.R"))
library(scales)
knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

oi  <- bh_data("open_inputs.rds")
mt  <- bh_data("model_terms.rds")
inp <- readRDS(here::here("app", "data", "sim_inputs.rds"))
z   <- oi$zones
zc  <- z %>% filter(area != "Expansion area")
dem <- inp$demand
cal <- mt$calibrated
rc  <- bh_data("reception_cohort.rds")
acc <- bh_data("accessibility.rds")

# The state-sector share is a single constant in the open bundle's build.
# It is recovered here from the projection rather than retyped, so this
# document cannot drift from the number the model used.
dem_raw <- readr::read_csv(here::here("data", "open_demand_projection.csv"),
                           show_col_types = FALSE)
STATE_SHARE <- with(dem_raw[dem_raw$area == "Brighton & Hove" &
                              dem_raw$entry_year == 2026, ],
                    state_demand / children)
W5 <- cal$W / mean(cal$W)
```

# What this document is

The simulator runs one model. Everything in the strategic view, and every
scenario in it, is that model with different settings. This guide takes
it apart in the order the model is written, so that each piece can be
checked on its own.

$$
T_{ij}^{\,y} \;=\; A_i^{\,y}\, O_i^{\,y}\, W_j\, c_{ij}^{-\beta}\,
e^{\gamma_{h(i)} \cdot \mathbb{1}[\,i \in \text{catchment}(j)\,]}\, C_j^{\delta},
\qquad
A_i^{\,y} = \frac{1}{\sum_k W_k\, c_{ik}^{-\beta}\, e^{\gamma_{h(i)} \mathbb{1}[\cdot]}\, C_k^{\delta}}
$$

| Symbol | Meaning | Section |
|:--|:--|:--|
| $T_{ij}$ | children from neighbourhood $i$ modelled as attending school $j$ | the interior of the matrix |
| $O_i$ | children in neighbourhood $i$ entering Year 7 | §2 |
| $W_j$ | how attractive school $j$ is, net of where it stands | §3 |
| $c_{ij}$ | routed walk-and-bus journey time from $i$ to $j$ | §4 |
| $\beta$ | distance decay | to follow |
| $\gamma_h$ | catchment term, fitted for each catchment $h$ | to follow |
| $C_j$, $\delta$ | competing destinations | to follow |
| $A_i$ | balancing factor: every neighbourhood places all of its children | to follow |
| $y$ | entry year, carried as a superscript | §2.7 |

Two further stages are applied to the $T_{ij}$ this produces: the
admission-number ceiling with the council's oversubscription priorities,
and the re-offering of refused children. Each gets its own section after
the terms above.

::: {.callout-note}
## The year is the only dynamic element, and it enters in one place

Nothing in the equation carries a time subscript except $O_i$. Journey
times, attractiveness, catchment terms and admission numbers are all held
at a single point in time, and the year chosen on the simulator's slider
changes only the **size** of each neighbourhood's cohort (§2.7). Whether
that is the right treatment — and what an explicit time dimension would
have to carry — is a question this guide returns to once the static terms
are set out.
:::

```{r matrix-blank}
#| tbl-cap: "The shape of the model, for four neighbourhoods and four schools. Sections 2 and 3 fill in the two margins shown; the rest follow, and the interior cells come last."
ex_zones <- zc %>% arrange(desc(Oi)) %>% slice(1:4)
sch4 <- c("Dorothy Stringer School", "Varndean School",
          "Longhill High School", "Cardinal Newman Catholic School")
data.frame(
  Neighbourhood = c(sub("[|].*$", "", ex_zones$zone), "$W_j$"),
  `Dorothy Stringer` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[1]]])),
  Varndean = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[2]]])),
  Longhill = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[3]]])),
  `Cardinal Newman` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[4]]])),
  `$A_i$` = c(rep("?", 4), ""),
  `$O_i$ (2026)` = c(sprintf("%.1f", ex_zones$Oi), ""),
  check.names = FALSE) %>%
  knitr::kable(align = "lccccrr")
```

The interior of that matrix is what the model produces. The margins are
what it is given, or what it fits. $O_i$ is the one margin that is a
count of real people, and it is where this guide starts.

# $O_i$: the children in each neighbourhood {#sec-oi}

## What it is

$O_i$ is the number of children living in zone $i$ who will enter Year 7
at a **state** secondary school in the year being modelled. The model is
production-constrained on it, which means that whatever else happens,

$$\sum_j T_{ij} = O_i \quad \text{for every } i.$$

Every child is placed somewhere. Nothing is created and nothing is lost,
which is why $O_i$ has to be right before anything else can be.

## What a zone is

A zone is **an LSOA cut by catchment**, keyed as
`LSOA | pre-2024 catchment | current catchment`. The city has
`r fmt_n(nrow(zc))` zones drawn from `r fmt_n(n_distinct(zc$lsoa))`
LSOAs.

The cut exists because the 2024 boundary change runs *through* LSOAs.
Assigning a whole LSOA to whichever catchment contains its centroid would
register only a fraction of the children the redraw actually moves. So
every postcode in the ONS Postcode Directory is assigned a catchment on
each of the two maps by point-in-polygon, and the LSOA is split
accordingly. Carrying **both** maps in the key means the two regimes sit
on identical geography and can be compared without rebuilding anything.

```{r zone-example}
#| tbl-cap: "Two LSOAs on the Longhill / Stringer-Varndean boundary, each split into two zones: the same neighbourhood, a different catchment on the current map."
z %>%
  filter(lsoa %in% c("E01016862", "E01016863")) %>%
  transmute(Zone = zone, LSOA = lsoa, `Pre-2024` = catch_pre2024,
            `Current map` = catch_optionZ, Postcodes = n_pcd,
            `$O_i$ (2026)` = sprintf("%.1f", Oi)) %>%
  knitr::kable(align = "llllrr")
```

## Where the numbers come from {#sec-oi-sources}

| Input | Source | What it supplies |
|:--|:--|:--|
| Ward population by single year of age, mid-2022 | [ONS ward-level mid-year population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/wardlevelmidyearpopulationestimatesexperimental) (`sapewardstablefinal.xlsx`) | how many children of each age live in each ward |
| LSOA population by single year of age | [ONS lower super output area mid-year population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates) (`lsoa_age_11_props_totals.xlsx`) | how a ward's children divide between its LSOAs |
| Postcode directory, February 2024 | [ONS Postcode Directory, ONS Open Geography Portal](https://geoportal.statistics.gov.uk/) | which ward and LSOA each postcode is in |
| Catchment boundaries | [Brighton & Hove City Council, school admissions](https://www.brighton-hove.gov.uk/schools-and-learning/school-admissions) — the option Z map in force from 2026 entry, and the map it replaced | the catchment half of the zone key |
| Year 7 offers, 2024 | [Allocation factsheet, Year 7 places September 2024](https://www.brighton-hove.gov.uk/schools-and-learning/school-policies-reports-strategies-and-other-documents/allocation-factsheet-year-7-places-september-2024) | the state-sector share |
| LSOA boundaries | [ONS Open Geography Portal](https://geoportal.statistics.gov.uk/) | drawing, and joining the age data to zones |

None of it is pupil-level. The smallest thing any of these sources
describes is a postcode count or an LSOA age band.

## How $O_i$ is built, step by step

The code is `public/R/01_open_inputs.R` in the private bundle, sections
1b and 1c. Its result travels to the simulator as `data/open_inputs.rds`.

**Step 1 — ages become entry years.** A child aged $a$ in the mid-2022
estimates enters Year 7 in $2022 + (11 - a)$. Ages 0 to 11 therefore give
entry years 2022 to 2033, every one of them children ONS has already
counted rather than births anyone has projected.

**Step 2 — two extrapolated years.** Children entering in 2034 and 2035
were not born when the mid-2022 estimate was taken. They are a linear
continuation of the five most recent observed cohorts, 2029 to 2033, and
are flagged as extrapolated wherever they appear.

**Step 3 — resident children become state-sector demand.** Not every
resident child takes a state Year 7 place: some go to independent
schools, some leave the city. A single constant carries that:

$$\text{state share} \;=\; \frac{\text{Year 7 offers made, 2024}}{\text{resident children, 2024 entry}} \;=\; `r sprintf("%.3f", STATE_SHARE)`$$

taken from the council's published offer total against the ONS cohort of
the same year.

**Step 4 — a ward is divided between its zones.** ONS publishes the age
detail at ward level, so each ward's children are shared out between the
zones in it. The weight for a zone is its LSOA's age-11 population share,
divided between the zones that LSOA is cut into by their postcode counts:

$$u_i \;=\; \underbrace{p_{\ell(i)}}_{\text{LSOA age-11 share}} \times
\underbrace{\frac{n_i}{\sum_{k \in \ell(i)} n_k}}_{\text{postcodes in this zone}},
\qquad
O_i^{\,2026} \;=\; \text{ward children}_{w(i)} \times \frac{u_i}{\sum_{k \in w(i)} u_k} \times \text{state share}$$

where $\ell(i)$ is the zone's LSOA, $w(i)$ its ward and $n_i$ its
postcode count. `r fmt_n(sum(!is.na(z$prop)))` of the `r fmt_n(nrow(z))`
zones use the ONS age share; the `r fmt_n(sum(is.na(z$prop)))` in the
expansion area outside the city fall back to postcode counts alone.

::: {.callout-note appearance="simple"}
## Why not divide a ward by postcode count

Splitting a ward's children by how many postcodes each LSOA holds assumes
every postcode contains the same number of eleven-year-olds. Family
housing, student housing and retirement housing differ enormously on
that, and the difference runs along exactly the lines catchment
boundaries follow. An earlier version of this work used postcode counts
between LSOAs, and the deprivation contrast between the two halves of a
paired catchment was almost entirely smoothed away. Postcode counts do
the narrower job above — dividing one LSOA between its own fragments,
where nothing finer is published.
:::

::: {.callout-important}
## Corrected: a split LSOA used to be counted twice

Until this version, each zone of a split LSOA inherited the **whole**
LSOA's age-11 share as its weight. Ward and city totals were unaffected,
because the denominator grew too, but inside a ward a split LSOA took
about 1.7 to 1.9 times the share its population implied and the unsplit
LSOAs around it took correspondingly less. The splits sit exactly on the
catchment boundaries, so the error was concentrated in the neighbourhoods
these questions turn on — Whitehawk and Kemptown among them.

The weight is now divided between a split LSOA's zones by their postcode
counts, as in the formula above. The table below is the check that used
to fail.
:::

```{r split-check}
#| tbl-cap: "Each LSOA's modelled share of its ward against the share its age-11 population implies, grouped by how many zones the LSOA is split into. A ratio of 1 is agreement."
chk <- z %>% filter(!is.na(prop)) %>%
  group_by(ward) %>% mutate(zone_share = Oi / sum(Oi)) %>% ungroup() %>%
  group_by(ward, lsoa) %>%
  summarise(prop = first(prop), zone_share = sum(zone_share),
            n_zones = n(), .groups = "drop") %>%
  group_by(ward) %>% mutate(prop_share = prop / sum(prop)) %>% ungroup() %>%
  mutate(ratio = zone_share / prop_share)
chk %>%
  group_by(`Zones the LSOA is split into` = n_zones) %>%
  summarise(LSOAs = n(), `Median ratio` = sprintf("%.2f", median(ratio)),
            Lowest = sprintf("%.2f", min(ratio)),
            Highest = sprintf("%.2f", max(ratio)), .groups = "drop") %>%
  knitr::kable(align = "rrrrr")
```

`r fmt_n(sum(z$Oi[z$lsoa %in% chk$lsoa[chk$n_zones > 1]]))` of the
`r fmt_n(sum(z$Oi))` children in the model live in a split LSOA, so the
correction moves children between zones inside those wards without
changing any ward or city total.

```{r fig-split-lsoa}
#| fig-cap: "One LSOA cut by the catchment boundary in force from 2026. Each point is a live postcode, coloured by the catchment it falls in; the black line is the LSOA boundary and the heavier line the catchment boundary. The model splits the LSOA's children between the two sides in proportion to the postcodes on each."
#| fig-height: 5
pcd <- readr::read_csv(here::here("data", "zone_postcode_points.csv"),
                       show_col_types = FALSE)
# The clearest example to draw: the split LSOA with the most postcodes on
# its smaller side, so neither half is a handful of points.
pick <- pcd %>% count(lsoa, catch_optionZ) %>% group_by(lsoa) %>%
  filter(n() == 2) %>% summarise(smaller = min(n), .groups = "drop") %>%
  slice_max(smaller, n = 1) %>% pull(lsoa)
pp <- pcd %>% filter(lsoa == pick) %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>% st_transform(4326) %>%
  mutate(Catchment = unname(CATCH_LABELS[as_boundary_catchment(catch_optionZ)]))
ls_one <- bh_data("lsoa.geojson") %>% filter(lsoa21cd == pick) %>% st_transform(4326)
catch_sf <- bh_data("catchments_current.geojson") %>% st_transform(4326)
bb <- st_bbox(st_buffer(st_transform(ls_one, 27700), 250)) %>%
  st_as_sfc() %>% st_transform(4326) %>% st_bbox()
ggplot() +
  geom_sf(data = catch_sf, aes(fill = catchment), colour = "grey25", alpha = 0.10,
          linewidth = 0.9, show.legend = FALSE) +
  geom_sf(data = ls_one, fill = NA, colour = "black", linewidth = 0.9) +
  geom_sf(data = pp, aes(colour = Catchment), size = 2.6, alpha = 0.9) +
  scale_fill_manual(values = CATCH_COLOURS) +
  scale_colour_manual(
    values = setNames(unname(CATCH_COLOURS[as_boundary_catchment(unique(pp$catch_optionZ))]),
                      unname(CATCH_LABELS[as_boundary_catchment(unique(pp$catch_optionZ))])),
    name = "Catchment of the postcode") +
  coord_sf(xlim = bb[c("xmin", "xmax")], ylim = bb[c("ymin", "ymax")]) +
  labs(title = sprintf("%s, split between two catchments", pick),
       subtitle = sprintf("%d postcodes on one side, %d on the other",
                          max(table(pp$catch_optionZ)), min(table(pp$catch_optionZ)))) +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"), legend.position = "top")
```

```{r split-worked}
#| tbl-cap: "The same LSOA, as the model sees it. The age-11 share belongs to the whole LSOA; each zone takes the part of it its postcodes imply."
z %>% filter(lsoa == pick) %>%
  transmute(Zone = zone, Postcodes = n_pcd,
            `Share of the LSOA` = sprintf("%.0f%%", 100 * n_pcd / sum(n_pcd)),
            `Age-11 share of the LSOA` = sprintf("%.4f", prop),
            `Weight used` = sprintf("%.4f", prop * n_pcd / sum(n_pcd)),
            `$O_i^{\\,2026}$` = sprintf("%.1f", Oi)) %>%
  knitr::kable(align = "lrrrrr")
```

Before the correction both rows of that table carried the whole LSOA's
age-11 share, so the neighbourhood counted twice in its ward's
denominator and took roughly double the children it should.

## A worked example

```{r oi-worked}
#| tbl-cap: "One ward, divided between its zones. Ward children is the ONS estimate for this cohort; the share is the zone's weight within the ward; O is the two multiplied together with the state-sector share."
wd <- z %>% filter(ward == (z %>% arrange(desc(Oi)) %>% slice(1) %>% pull(ward)))
wd %>%
  arrange(desc(Oi)) %>%
  transmute(Zone = sub("[|]", " | ", zone), LSOA = lsoa, Postcodes = n_pcd,
            `Age-11 share of LSOA` = sprintf("%.4f", prop),
            `Share of the ward` = sprintf("%.3f", share_in_ward),
            `Ward children` = fmt_n(ward_children),
            `$O_i$ (2026)` = sprintf("%.1f", Oi)) %>%
  knitr::kable(align = "llrrrrr")
```

Reading the first row: the ward holds `r fmt_n(wd$ward_children[1])`
children of this cohort; that zone carries
`r sprintf("%.1f%%", 100 * max(wd$share_in_ward))` of the ward's weight;
multiplied by the state-sector share of
`r sprintf("%.3f", STATE_SHARE)`, the zone holds
`r sprintf("%.1f", max(wd$Oi))` Year 7 children in 2026.

## The city total

```{r oi-total}
#| tbl-cap: "The modelled 2026 cohort by catchment. These are children living in the catchment, not children attending its school."
zc %>% group_by(Catchment = catchment) %>%
  summarise(children = sum(Oi), Zones = n(), .groups = "drop") %>%
  arrange(desc(children)) %>%
  transmute(Catchment, `Children, 2026` = fmt_n(children), Zones) %>%
  knitr::kable(align = "lrr")
```

The city's zones hold `r fmt_n(sum(zc$Oi))` children for 2026 entry. That
is the model's entire demand before any child chooses anything, and every
intake the simulator reports is drawn from it.

### Where the city totals come from {#sec-oi-provenance}

Two different city-wide totals appear in this work, and they have
different provenance. It is worth being explicit about which is which.

**Offers actually made** are the council's own published figures. Its
[allocation factsheets](https://www.brighton-hove.gov.uk/schools-and-learning/school-admissions)
are issued after each March allocation round, and the series from 2010 to
2026 was transcribed school by school in the earlier
[Brighton and Hove allocations](https://adamdennett.github.io/BH_Schools_2/brighton_and_hove_allocations.html)
analysis. This bundle does not retype them: `public/R/01a_factsheet_panel.R`
re-extracts the year-by-year tables from that document, so every offer,
preference and admission number here traces back to a published council
factsheet.

**Resident children** are ONS, as set out in @sec-oi-sources. They are a
different quantity from offers: a resident child may go to an independent
school, or out of the city.

**The model's city total is the first anchored on the second.** The
state-sector share of `r sprintf("%.3f", STATE_SHARE)` is the council's
published 2024 offer total divided by the ONS resident cohort for the same
entry year, so the model's demand is ONS children scaled to the level the
council's own offers imply. It is not a free-floating estimate.

```{r oi-provenance}
#| tbl-cap: "The two totals, and the model's demand between them."
fp <- bh_data("factsheet_panel.rds")$factsheets %>% filter(name != "Total")
off <- fp %>% group_by(year) %>% summarise(offers = sum(off_total), .groups = "drop")
tibble::tribble(
  ~Quantity, ~Source, ~Value,
  "Year 7 offers made, 2024", "Council allocation factsheet",
  fmt_n(off$offers[off$year == 2024]),
  "Year 7 offers made, latest round", "Council allocation factsheet",
  sprintf("%s (%d)", fmt_n(off$offers[which.max(off$year)]), max(off$year)),
  "Resident children, 2024 entry", "ONS ward estimates, mid-2022",
  fmt_n(dem_raw$children[dem_raw$area == "Brighton & Hove" & dem_raw$entry_year == 2024]),
  "State-sector share", "The first divided by the second",
  sprintf("%.3f", STATE_SHARE),
  "Model demand, 2026 entry", "ONS children x state share",
  fmt_n(sum(zc$Oi))) %>%
  knitr::kable(align = "llr")
```

The model's 2026 demand and the offers made in the 2026 round are not the
same number and should not be expected to match exactly: the offers
include children placed at a school they did not name, and exclude
children who took a place outside the city. The comparison is still worth
making, and it is the first check listed in @sec-oi-checks.

## The slider: how $O_i$ changes with the entry year {#sec-oi-year}

The year slider does exactly one thing to $O_i$. The city's projected
cohort is turned into an index on 2026, and every zone is scaled by it:

$$O_i^{\,y} \;=\; O_i^{\,2026} \times d^{\,y}, \qquad
d^{\,y} \;=\; \frac{\text{state demand}^{\,y}}{\text{state demand}^{\,2026}}$$

```{r oi-index}
#| tbl-cap: "The demand index behind the slider. The last two years are the extrapolated cohorts."
dem %>%
  transmute(`Entry year` = year, `City cohort` = fmt_n(cohort),
            `Index on 2026` = sprintf("%.3f", index),
            Basis = ifelse(extrapolated, "extrapolated",
                           "children already counted")) %>%
  knitr::kable(align = "rrrl")
```

```{r fig-oi-index}
#| fig-cap: "The cohort the model scales to. Every zone is multiplied by the same factor, so the map of where children live is held still and only the size of the cohort moves."
#| fig-height: 4.2
dem %>%
  ggplot(aes(year, cohort)) +
  geom_col(aes(fill = extrapolated), width = 0.66) +
  geom_text(aes(label = fmt_n(cohort)), vjust = -0.5, size = 3.2, colour = "grey25") +
  scale_fill_manual(values = c(`FALSE` = "#1f3b57", `TRUE` = "#9aa7b4"),
                    labels = c(`FALSE` = "children already counted",
                               `TRUE` = "extrapolated"), name = NULL) +
  scale_x_continuous(breaks = dem$year) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "State-sector Year 7 demand",
       title = "The cohort behind the slider, 2026 to 2035") +
  theme_bh() + theme(legend.position = "top")
```

### Which projection this is, and the one beside it {#sec-oi-projection}

Two projections of the city's Year 7 cohort appear in this work. They are
built from different data and they are used for different things, so it
is worth being plain about which drives the model.

**The one the slider uses** is the ONS estimate carried forward: the
children counted in the mid-2022 small-area estimates, aged forward to
their Year 7 entry year (§2.4, steps 1 and 2), and multiplied by the
state-sector share. It is used because it is the only one that exists at
**ward** level, and the model needs the geography as well as the total —
$O_i$ is a ward figure divided between zones.

**The one beside it** is built from children already in the city's
primary schools: Reception offers made seven years earlier, multiplied by
the share of a Reception cohort that reappears as a Year 7 offer. That
ratio is measured rather than assumed, from the council's own published
factsheets, and it has been strikingly steady —
`r fmt_pct(100 * rc$survival_recent, 1)` across the five most recent
cohorts, with a standard deviation of
`r fmt_pct(100 * rc$survival_sd, 1)`. The method is set out in full in
[how our projection method works](https://adamdennett.github.io/BH_Schools_2/brighton_and_hove_allocations.html#how-our-projection-method-works),
and section 3 of the strategic view uses it as its headline projection
because it rests on children already registered in a school rather than
on a population estimate.

```{r oi-two-projections}
#| tbl-cap: "The two projections side by side. The first drives the model's zone-level demand; the second is the check on it, and the strategic view's headline series."
rc$comparison %>%
  transmute(`Entry year` = entry_year,
            `From ONS estimates (drives the slider)` = fmt_n(ons_demand),
            `From Reception offers` = fmt_n(reception_based),
            Difference = sprintf("%+.0f", reception_based - ons_demand),
            `%` = sprintf("%+.1f%%", 100 * (reception_based / ons_demand - 1))) %>%
  knitr::kable(align = "rrrrr")
```

They agree within a few per cent for most years, which is reassuring for
both, and they diverge most in the years where the ONS series is itself
extrapolated. Neither is a forecast in the sense of predicting behaviour:
one counts children who exist, the other counts children already in
school and applies a measured survival rate.

**What this does not do.** The projection is city-wide, so the slider
cannot say that one neighbourhood shrinks faster than another: every zone
keeps its 2026 share of the city's children in every year. The council's
own forecasts differ from this mostly in exactly that respect — its
Longhill catchment line falls faster than the city's does — so comparing
catchment totals year by year compares two different claims rather than
two estimates of one thing.

## Checks you can run {#sec-oi-checks}

```{r oi-checks}
#| tbl-cap: "Identities that should hold, and what they give as this document is built."
ward_tot <- z %>% group_by(ward) %>%
  summarise(oi = sum(Oi), implied = first(ward_children) * STATE_SHARE,
            .groups = "drop") %>%
  mutate(d = abs(oi - implied))
tibble::tribble(
  ~Check, ~Result,
  "Every zone has a positive cohort",
  sprintf("smallest is %.2f children", min(z$Oi)),
  "Zone cohorts sum to each ward's estimate times the state share",
  sprintf("largest difference %.3f children", max(ward_tot$d)),
  "The city's zones sum to the 2026 projection",
  sprintf("%s against %s", fmt_n(sum(zc$Oi)), fmt_n(dem$cohort[dem$year == 2026])),
  "The index is exactly 1 in the base year",
  sprintf("%.6f", dem$index[dem$year == 2026]),
  "A split LSOA takes the share its population implies",
  sprintf("largest ratio %.2f", max(chk$ratio))) %>%
  knitr::kable(align = "lr")
```

Two further comparisons are worth doing by hand, because they test $O_i$
against something the model never saw: the city total against the
council's published Year 7 offers for the same round, and each
catchment's total against the catchment forecasts in the council's place
planning appendix.

## What to be careful about

- **The base is mid-2022.** Migration into or out of the city since then
  is not in the counts, and a child who has moved between wards is still
  counted where they were.
- **One state-sector share, for every year and every neighbourhood.** It
  is `r sprintf("%.3f", STATE_SHARE)`, from a single round. If
  independent-school take-up varies across the city — and it plainly does
  — then $O_i$ is too high in some places and too low in others.
- **Ward-level age detail.** Within a ward, the LSOA age-11 distribution
  does the work; how children are distributed inside an LSOA is not
  observed, so postcode counts stand in for it where a boundary cuts one.
- **Peacehaven and the expansion area** are carried in the open bundle
  but excluded from the simulator's zones. Peacehaven Community School is
  a destination in the model, not a source of demand.
- **No composition.** $O_i$ is a headcount. Which of those children are
  disadvantaged is a separate layer applied later, not part of this term.

## Questions this raises

1. Should the state-sector share vary by neighbourhood, given that
   independent-school take-up does?
2. Should the year slider move the *geography* of demand as well as its
   size — that is, should each catchment carry its own projection, as the
   council's forecasts do?

# $W_j$: how attractive each school is {#sec-wj}

## What it is

$W_j$ is the pull of school $j$ on a family, **net of where the school
stands**. It is dimensionless, scaled so the city average is 1, and it
enters the model multiplicatively: doubling $W_j$ doubles the numerator
of every flow into $j$ before distance, catchment and competition have
their say. It also enters the competing-destinations term, because a
school's rivals count for more when they are schools families want.

It is the hardest term in the model to pin down, because everything
observable about a school is entangled with where it is. A school with
high attainment may be sought after because it teaches well, or because
it sits where better-off families already live.

## Two layers: the open specifications, and the fitted values

**Layer one — the open specifications.** The open bundle carries five
candidate measures of attractiveness, so that a conclusion can be tested
across all of them rather than resting on one.

```{r wj-specs}
#| tbl-cap: "The five open specifications, each scaled so the city average is 1. Weighted preferences count ranks 1 to 3, a second preference worth half a first and a third a quarter."
oi$attract %>%
  transmute(School = short_sch(name), PAN = pan, `Attainment 8` = round(att8, 1),
            Equal = sprintf("%.2f", W_equal), `By PAN` = sprintf("%.2f", W_pan),
            `First preferences` = sprintf("%.2f", W_prefs),
            `Weighted preferences` = sprintf("%.2f", W_wprefs),
            `By Attainment 8` = sprintf("%.2f", W_att8)) %>%
  arrange(desc(as.numeric(`Weighted preferences`))) %>%
  knitr::kable(align = "lrrrrrrr")
```

The central one is **weighted preferences per place**: the council's
published factsheets, pooled over the most recent rounds, counting first,
second and third preferences with a geometric discount of 0.5 per rank,
divided by the admission number. The paired catchments are why: a family
in Stringer / Varndean or Hove Park / Blatchington Mill must rank two
local schools against each other, so a first-preference count reads a
school families put second as one nobody wanted.

**Layer two — the fitted values the simulator uses.** The full model does
not take any of those five as given. It **balances** $W_j$ so that the
model reproduces the share of the city's first preferences each school
actually receives, using the council's catchment-level preference table
from its evidence to the Schools Adjudicator. The catchment terms
$\gamma_h$ are fitted in the same loop (§4, to follow), because the two
trade off against each other.

$$W_j \;\leftarrow\; W_j \left(\frac{\text{target share}_j}{\text{modelled share}_j}\right)^{0.8},
\qquad \text{repeat until } \max_j |\text{modelled} - \text{target}| < 10^{-6}$$

starting from the weighted-preference values, with the result rescaled to
a city mean of 1. The exponent 0.8 is damping: it stops the balancing
oscillating around the target. This is the same device as the balancing
factor $A_i$, applied to the other margin of the matrix — $A_i$ makes
every neighbourhood place all its children, and this makes every school
receive the share of first preferences it is observed to receive.

```{r wj-fitted}
#| tbl-cap: "The fitted attractiveness the simulator's sliders multiply, beside the weighted-preference measure it starts from, and each school's target share of the city's first preferences."
tibble(name = names(W5), W5 = as.numeric(W5)) %>%
  left_join(oi$attract %>% select(name, W_wprefs, att8), by = "name") %>%
  left_join(tibble(name = names(cal$target_first_prefs),
                   target = as.numeric(cal$target_first_prefs)), by = "name") %>%
  arrange(desc(W5)) %>%
  transmute(School = short_sch(name),
            `Weighted preferences` = sprintf("%.2f", W_wprefs),
            `Fitted $W_j$` = sprintf("%.2f", W5),
            `Target share of first preferences` = sprintf("%.1f%%", 100 * target),
            `Attainment 8` = sprintf("%.1f", att8)) %>%
  knitr::kable(align = "lrrrr")
```

```{r fig-wj}
#| fig-cap: "The fitted attractiveness against the measure it starts from. A school above the line is one the model has to make more attractive than its weighted preferences suggest to reproduce the preferences each catchment gives it."
#| fig-height: 4.6
tibble(name = names(W5), W5 = as.numeric(W5)) %>%
  left_join(oi$attract %>% select(name, W_wprefs), by = "name") %>%
  ggplot(aes(W_wprefs, W5)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "31") +
  geom_point(size = 3.4, colour = "#1f3b57") +
  ggrepel::geom_text_repel(aes(label = short_sch(name)), size = 3.4, seed = 1,
                           colour = "#1f3b57") +
  expand_limits(x = 0, y = 0) +
  labs(x = "Weighted preferences per place (city mean = 1)",
       y = "Fitted attractiveness (city mean = 1)",
       title = "What the balancing does to the starting values") +
  theme_bh()
```

## Where it sits in the matrix

```{r matrix-wj}
#| tbl-cap: "The same four neighbourhoods and schools. Two margins are now filled: the children in each neighbourhood, and the pull of each school."
data.frame(
  Neighbourhood = c(sub("[|].*$", "", ex_zones$zone), "$W_j$"),
  `Dorothy Stringer` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[1]]])),
  Varndean = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[2]]])),
  Longhill = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[3]]])),
  `Cardinal Newman` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[4]]])),
  `$A_i$` = c(rep("?", 4), ""),
  `$O_i$ (2026)` = c(sprintf("%.1f", ex_zones$Oi), ""),
  check.names = FALSE) %>%
  knitr::kable(align = "lccccrr")
```

## What the attractiveness slider does

Each school's slider multiplies its fitted $W_j$, from
`r sprintf("%.2f", 0.25)` to `r sprintf("%.0f", 4)` times. Nothing else
changes: the same journey times, the same catchment terms, the same
admission numbers. Three details are worth knowing.

- **The multiplier is applied to the fitted value**, not to weighted
  preferences, so a slider at 1 reproduces the model exactly.
- **The out-of-city schools are not touched by it.** Priory, Peacehaven,
  Seahaven and Seaford Head carry their own fitted attractiveness and a
  decay on straight-line distance, and the sliders do not reach them.
- **CoMArt has no preferences**, because it does not exist. When it is
  opened, it borrows the attractiveness of
  `r inp$comart$w_from_short` as its starting value, and its slider moves
  from there.

## Turning a multiplier into something meaningful

A slider position is not a quantity anyone outside a model can judge, so
the app converts it to the one published number families appear to
respond to. Across the ten city schools, the fitted attractiveness is
log-linear in Attainment 8:

$$\log W_j = a + b \cdot \text{Attainment 8}_j,
\qquad b = `r sprintf("%.4f", inp$attain$slope)`,
\quad R^2 = `r sprintf("%.2f", inp$attain$r2)`,
\quad n = `r inp$attain$n`$$

so a point of Attainment 8 is worth about
`r fmt_pct(100 * inp$attain$per_point, 0)` more attractiveness, and a
multiplier $m$ inverts to $\log(m) / b$ points. That is how the simulator
answers "how good would this school have to be to fill?" — and it is an
association, not a lever: Attainment 8 is itself largely a description of
who a school admits.

## Checks you can run {#sec-oi-checks}

```{r wj-checks}
#| tbl-cap: "Identities and fits that should hold for the attractiveness term."
own <- cal$own
tibble::tribble(
  ~Check, ~Result,
  "Fitted attractiveness averages 1 across the city",
  sprintf("%.6f", mean(W5)),
  "Balanced first-preference shares match the council's table",
  sprintf("largest gap %.4f percentage points",
          100 * max(abs(cal$cells$m5_share - cal$cells$obs_share))),
  "Share of each catchment's first preferences to its own school",
  sprintf("observed %.0f%% to %.0f%%, modelled %.0f%% to %.0f%%",
          100 * min(own$observed), 100 * max(own$observed),
          100 * min(own$M5), 100 * max(own$M5)),
  "Attainment 8 explains the fitted values",
  sprintf("R2 %.2f on %d schools", inp$attain$r2, inp$attain$n),
  "A slider at 1 reproduces the published model",
  "checked in app/tests/check.R") %>%
  knitr::kable(align = "lr")
```

## What to be careful about

- **Ten observations.** Every statement about *why* a school is
  attractive rests on ten schools. The balancing is an accounting
  identity across those ten; the Attainment 8 fit is a description of
  them, not evidence about causes.
- **Attractiveness and location are only partly separated.** The
  balancing asks what pull each school needs *given* where it stands and
  how far families will travel, which is a real separation — but it
  inherits whatever $\beta$ and $\gamma_h$ are set to. If distance decay
  were weaker, the fitted $W_j$ would be different.
- **The faith schools have no catchment**, so nothing in the model
  restricts who may choose them. Their fitted attractiveness absorbs the
  religious criteria that really ration those places, which is why they
  carry high values.
- **Peacehaven's preferences are not published** by the council, so both
  preference measures are filled at the lowest observed rate in the city.
  That is an assumption, and it flatters Longhill: a weak Peacehaven
  keeps eastern children in the city.
- **$W_j$ does not move with time.** A school that improves, or a new
  building, is a slider position rather than anything the model
  generates. Whether attractiveness should respond to a school's roll,
  its intake or its results is exactly the dynamic question §2.7 flags.

## Questions this raises

1. Should $W_j$ be allowed to vary by family type — by deprivation, say —
   rather than being one number per school for the whole city?
2. Should the balancing target first preferences, as now, or all three
   ranks discounted as the open specification does?
3. Should a school's attractiveness respond to its own intake over time,
   which is the feedback loop the strategic view describes but the model
   does not contain?
# $c_{ij}$: the journey from a neighbourhood to a school {#sec-cij}

```{r cij-setup}
#| include: false
costs <- oi$costs_now
cost_elm <- oi$costs_elm
tm <- readRDS(here::here("data", "travel", "build_metadata.rds"))
lsoa_sf <- bh_data("lsoa.geojson") %>% st_transform(4326)
catch_sf <- bh_data("catchments_current.geojson") %>% st_transform(4326)
sch_sf <- schools_sf(oi$schools) %>% filter(name != "Peacehaven Community School")
zone_lsoa <- z %>% select(zone, lsoa, Oi)
# Zone costs carried up to LSOA level for drawing, weighted by children.
to_lsoa <- function(d, school) {
  d %>% filter(name == school) %>%
    inner_join(zone_lsoa, by = "zone") %>%
    group_by(lsoa) %>%
    summarise(cij = weighted.mean(cij, Oi), .groups = "drop")
}
LH <- "Longhill High School"
```

## What it is

$c_{ij}$ is the **generalised cost** of getting from neighbourhood $i$ to
school $j$: minutes of walking and bus, door to door, on a weekday
morning. It is not a straight-line distance, and it is not a car journey.
It enters the model as $c_{ij}^{-\beta}$, so it is the term that makes a
school ten minutes away far more likely to be chosen than one an hour
away — and the exponent $\beta$ that governs how fast that falls away is
its own section, to follow.

The city has `r fmt_n(nrow(inp$cost$now))` zone-school pairs in the
simulator's cost table, spanning
`r sprintf("%.0f", min(inp$cost$now$cij))` to
`r sprintf("%.0f", max(inp$cost$now$cij))` minutes.

## Where the numbers come from

Every journey is routed with [`r5r`](https://ipeagit.github.io/r5r/) over
two networks: the street network from
[OpenStreetMap](https://www.openstreetmap.org/copyright), and the bus
timetable as a [GTFS feed](https://data.bus-data.dft.gov.uk/) from the
Department for Transport's Bus Open Data Service. The build is
`public/R/00b_build_travel_matrix.R` in the open bundle.

```{r cij-params}
#| tbl-cap: "The routing parameters. Every journey in this work uses these; changing any of them changes every journey time."
tibble::tribble(
  ~Setting, ~Value,
  "Modes", paste(tm$params$mode, collapse = " and "),
  "Departure", format(tm$params$departure_datetime, "%A %d %B %Y, %H:%M"),
  "Longest walk allowed, either end", sprintf("%d minutes", tm$params$max_walk_time),
  "Longest journey allowed", sprintf("%d minutes", tm$params$max_trip_duration),
  "Departure window", sprintf("%d minute", tm$params$time_window),
  "Percentile of journeys in the window", sprintf("%dth", tm$params$percentiles),
  "Origins", sprintf("%s postcodes", fmt_n(tm$n_origins)),
  "Destinations", sprintf("%d, including the Elm Grove site", nrow(tm$destinations))) %>%
  knitr::kable(align = "lr")
```

The departure date is chosen rather than assumed: the build counts the
trips each candidate weekday actually carries in the feed and takes a
Tuesday, Wednesday or Thursday in service, because a date the feed does
not serve returns no journeys at all and a Monday or Friday timetable is
lighter than a normal school day.

## From postcodes to zones

The matrix is routed **postcode to school**. A zone's cost is the mean
over the live postcodes in it:

$$c_{ij} \;=\; \frac{1}{|P_i|}\sum_{p \in P_i} c_{pj}$$

where $P_i$ is the set of postcodes in zone $i$. Because a zone is an
LSOA cut by catchment (§2.2), the two halves of a split LSOA get their
own journey times — which is the point of cutting them.

## Two corrections the raw routing needs

**Walking is always available.** Where the bus network is awkward the
router returns a poor itinerary: about a fifth of pairs came back slower
than simply walking the distance, in one case 99 minutes for 4 km. That
is a property of the router, not of the city, so every journey is capped
at the time it would take to walk:

$$c_{ij} \;=\; \min\!\left(c^{\text{routed}}_{ij},\;
\frac{d_{ij} \times `r sprintf("%.1f", acc$walk_fallback$circuity)`}{`r sprintf("%.2f", acc$walk_fallback$kmh)` \text{ km/h}} \times 60\right)$$

The walking pace comes from a systematic review of active school travel:
children's mean walking trip is
`r fmt_n(acc$walk_fallback$review$dist_m)` m in
`r sprintf("%.1f", acc$walk_fallback$review$time_min)` minutes, or
`r sprintf("%.2f", acc$walk_fallback$kmh)` km/h, and the circuity factor
of `r sprintf("%.1f", acc$walk_fallback$circuity)` turns straight-line
distance into street distance. The cap binds on
`r fmt_pct(100 * mean(costs$walk_capped), 0)` of pairs. **It is a ceiling
on modelled time, not a claim that children walk that far** — the same
review finds they rarely walk beyond 1.8 km.

**A distance model fills the gaps.** Where the matrix has no routed value
at all, a fitted model of time on straight-line distance stands in. It
reaches `r fmt_pct(100 - 100 * mean(costs$routed), 1)` of pairs in the
city, so almost everything here is routed rather than predicted.

```{r fig-cij-walk}
#| fig-cap: "Every zone-school pair: routed journey time against straight-line distance. The line is the walking time the cap uses. Points above it are journeys the router made slower than walking, and they are pulled down onto the line."
#| fig-height: 4.6
wk <- costs %>% mutate(walk_min = km * acc$walk_fallback$circuity / acc$walk_fallback$kmh * 60)
ggplot(wk, aes(km, cij)) +
  geom_point(aes(colour = walk_capped), size = 1.1, alpha = 0.5) +
  geom_line(aes(km, walk_min), colour = "#1f3b57", linewidth = 0.9) +
  scale_colour_manual(values = c(`FALSE` = "grey65", `TRUE` = "#d95f02"),
                      labels = c(`FALSE` = "routed time used",
                                 `TRUE` = "capped at walking time"), name = NULL) +
  labs(x = "Straight-line distance (km)", y = "Journey time used (minutes)",
       title = "Where the cap binds",
       subtitle = sprintf("%s of %s pairs are capped",
                          fmt_n(sum(costs$walk_capped)), fmt_n(nrow(costs)))) +
  theme_bh() + theme(legend.position = "top")
```

## What a journey-time surface looks like

```{r fig-cij-longhill}
#| fig-cap: "Journey time to Longhill from every neighbourhood, at its current site and at the Elm Grove site. Both panels use the same scale, so they can be read against each other. The white dot is the school."
#| fig-height: 5
lh_now <- to_lsoa(costs, LH) %>% mutate(site = "Ovingdean (today)")
lh_elm <- to_lsoa(cost_elm, LH) %>% mutate(site = "Elm Grove")
lh_map <- lsoa_sf %>%
  inner_join(bind_rows(lh_now, lh_elm), by = c("lsoa21cd" = "lsoa")) %>%
  mutate(site = factor(site, c("Ovingdean (today)", "Elm Grove")))
pts <- bind_rows(
  sch_sf %>% filter(name == LH) %>% mutate(site = "Ovingdean (today)"),
  st_as_sf(data.frame(e = oi$elm_grove$easting, n = oi$elm_grove$northing),
           coords = c("e", "n"), crs = 27700) %>% st_transform(4326) %>%
    mutate(site = "Elm Grove")) %>%
  mutate(site = factor(site, levels(lh_map$site)))
ggplot() +
  geom_sf(data = lh_map, aes(fill = cij), colour = NA) +
  geom_sf(data = catch_sf, fill = NA, colour = "white", linewidth = 0.25) +
  geom_sf(data = pts, shape = 21, fill = "white", colour = "black", size = 3, stroke = 1.1) +
  facet_wrap(~ site) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "Minutes") +
  labs(title = "The same school, two sites",
       subtitle = "Walk-and-bus journey time from each neighbourhood, weekday morning") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"),
        strip.text = element_text(face = "bold"))
```

The median neighbourhood is
`r sprintf("%.0f", median(lh_now$cij))` minutes from Ovingdean and
`r sprintf("%.0f", median(lh_elm$cij))` from Elm Grove. That difference
is the whole of what relocation does to $c_{ij}$ — and, because nothing
else in the model moves with it except the competition term, most of what
relocation does at all.

```{r fig-cij-nearest}
#| fig-cap: "Which school each neighbourhood can reach quickest, by routed walk-and-bus time. Boundaries are the catchments in force; where a colour crosses a boundary, the catchment school is not the nearest one."
#| fig-height: 5.2
nearest <- costs %>%
  inner_join(zone_lsoa, by = "zone") %>%
  group_by(lsoa, name) %>% summarise(cij = weighted.mean(cij, Oi), .groups = "drop_last") %>%
  slice_min(cij, n = 1, with_ties = FALSE) %>% ungroup() %>%
  mutate(School = short_sch(name))
near_map <- lsoa_sf %>% inner_join(nearest, by = c("lsoa21cd" = "lsoa"))
ggplot() +
  geom_sf(data = near_map, aes(fill = School), colour = "white", linewidth = 0.1) +
  geom_sf(data = catch_sf, fill = NA, colour = "grey15", linewidth = 0.6) +
  geom_sf(data = sch_sf, shape = 21, fill = "white", colour = "black", size = 2.6) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  labs(title = "The quickest school from each neighbourhood",
       subtitle = "Routed walk-and-bus time, not distance, and not who is admitted") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"))
```

## A worked example

```{r cij-worked}
#| tbl-cap: "One neighbourhood's journey to every city school, with the straight-line distance beside it. The two are not the same ordering, which is the reason for routing at all."
ex_zone <- costs %>% filter(zone == ex_zones$zone[1])
ex_zone %>%
  arrange(cij) %>%
  transmute(School = short_sch(name), `Journey time (minutes)` = sprintf("%.0f", cij),
            `Straight line (km)` = sprintf("%.1f", km),
            Basis = ifelse(walk_capped, "capped at walking time",
                           ifelse(routed, "routed", "distance model"))) %>%
  knitr::kable(align = "lrrl")
```

## The journeys the model does not route

- **The four schools outside the city** — Priory, Peacehaven, Seahaven
  and Seaford Head — are chosen on **straight-line kilometres**, not
  routed minutes. The bus network has no direct eastern service to Lewes,
  so the routed journey from Woodingdean goes through Brighton and takes
  over an hour, which is not what the families who make that journey do.
  Their decay is fitted on km against the published 2024 destinations.
- **CoMArt**, when opened, is routed from its old site on the same
  network and with the same parameters as every other school.
- **Longhill at Elm Grove** has a routed column of its own. An earlier
  version had no routed times for the site and gave it the average
  accessibility of a Brighton secondary, which was wrong by about sixty
  children; the build now stops rather than let that happen silently.

## Checks you can run

```{r cij-checks}
#| tbl-cap: "What the cost table should satisfy, and what it gives."
st <- acc$statutory
tibble::tribble(
  ~Check, ~Result,
  "Every zone can reach every city school",
  sprintf("%s pairs, none missing", fmt_n(nrow(inp$cost$now))),
  "Journeys are routed rather than predicted",
  sprintf("%.1f%% routed", 100 * mean(costs$routed)),
  "No journey is slower than walking",
  sprintf("%.0f%% capped, none above the walk line", 100 * mean(costs$walk_capped)),
  "Mean journey to the catchment school",
  sprintf("%.1f minutes", st$mean_min),
  "Against the statutory limit for secondary (75 minutes)",
  sprintf("%.0f children over it", st$over_75_children),
  "Against the national average school journey (19 minutes)",
  sprintf("city mean %.1f minutes", st$mean_min)) %>%
  knitr::kable(align = "lr")
```

The strongest check is not in the table: a handful of journeys can be
compared against local knowledge. Woodingdean to Longhill comes out at
about 17 minutes and Woodingdean to Dorothy Stringer at about 68;
Whitehawk to Longhill about 28 and to Stringer about 62. Those are the
journeys that decide the eastern half of this work, and they are worth
checking against what a family there would say.

## What to be careful about

- **One departure, one day.** Every journey is a single weekday morning
  in the timetable as it stood when the matrix was built. Term-time
  variation, a missed connection and a cancelled bus are all outside it.
- **A mean over postcodes.** A zone's cost is the average of the
  postcodes in it, so a neighbourhood with one badly connected corner
  shows as slightly worse rather than showing both.
- **No cars, no lifts, no walking buses.** The model prices the journey a
  child could make on foot and by bus. A family with a car makes a
  different journey, and that is one reason the east's figures should be
  read as a ceiling on difficulty rather than a description of it.
- **Time is symmetric here.** One morning journey stands for the return
  as well, and no allowance is made for after-school travel.
- **The cap is generous at distance.** A 6 km walk is not a real school
  journey, but capping at walking time is deliberately crude: it removes
  router artefacts without inventing a bus that does not run.

## Questions this raises

1. Should the zone cost be the mean over its postcodes, or the
   child-weighted mean, or the best available journey rather than the
   average one?
2. Should journeys be routed at more than one departure time, so that a
   thin early timetable is visible?
3. Should the out-of-city schools be routed rather than measured on
   straight-line distance, if a feed covering East Sussex services were
   used?
