---
title: "The model, stage by stage"
subtitle: "A technical guide for checking each element"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
abstract: |
  This document takes the model the simulator runs and opens it up one
  element at a time. Each section states what a term is, where its data
  comes from, exactly how it is computed, what it looks like for a worked
  example, the checks that can be run against it, and what it assumes. It
  is written to be audited: every number is produced by the code in this
  repository, and every step should be reproducible by hand.
---

```{r setup}
#| include: false
source(here::here("R", "00_core.R"))
library(scales)
knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

oi  <- bh_data("open_inputs.rds")
mt  <- bh_data("model_terms.rds")
inp <- readRDS(here::here("app", "data", "sim_inputs.rds"))
z   <- oi$zones
zc  <- z %>% filter(area != "Expansion area")
dem <- inp$demand
cal <- mt$calibrated
rc  <- bh_data("reception_cohort.rds")
acc <- bh_data("accessibility.rds")

# The simulator's own engine, so the guide can run the model it describes.
source(here::here("app", "R", "model.R"))
source(here::here("app", "R", "outcomes.R"))

# The state-sector share is a single constant in the open bundle's build.
# It is recovered here from the projection rather than retyped, so this
# document cannot drift from the number the model used.
dem_raw <- readr::read_csv(here::here("data", "open_demand_projection.csv"),
                           show_col_types = FALSE)
STATE_SHARE <- with(dem_raw[dem_raw$area == "Brighton & Hove" &
                              dem_raw$entry_year == 2026, ],
                    state_demand / children)
W5 <- cal$W / mean(cal$W)
```

# What this document is

The simulator runs one model. Everything in the strategic view, and every
scenario in it, is that model with different settings. This guide takes
it apart in the order the model is written, so that each piece can be
checked on its own.

$$
T_{ij}^{\,y} \;=\; A_i^{\,y}\, O_i^{\,y}\, W_j\, c_{ij}^{-\beta}\,
e^{\gamma_{h(i)} \cdot \mathbb{1}[\,i \in \text{catchment}(j)\,]}\, C_j^{\delta},
\qquad
A_i^{\,y} = \frac{1}{\sum_k W_k\, c_{ik}^{-\beta}\, e^{\gamma_{h(i)} \mathbb{1}[\cdot]}\, C_k^{\delta}}
$$

| Symbol | Meaning | Section |
|:--|:--|:--|
| $T_{ij}$ | children from neighbourhood $i$ modelled as attending school $j$ | the interior of the matrix |
| $O_i$ | children in neighbourhood $i$ entering Year 7 | §2 |
| $W_j$ | how attractive school $j$ is, net of where it stands | §3 |
| $c_{ij}$ | routed walk-and-bus journey time from $i$ to $j$ | §4 |
| $\beta$ | distance decay | §5 |
| $\gamma_h$ | catchment term, fitted for each catchment $h$ | §6 |
| $C_j$, $\delta$ | competing destinations | §7 |
| $A_i$ | balancing factor: every neighbourhood places all of its children | §8 |
| $y$ | entry year, carried as a superscript | §2.7 |

Two further stages are applied to the $T_{ij}$ this produces: the
admission-number ceiling with the council's oversubscription priorities,
and the re-offering of refused children. Each gets its own section after
the terms above.

::: {.callout-note}

## The year is the only dynamic element, and it enters in one place

Nothing in the equation carries a time subscript except $O_i$. Journey
times, attractiveness, catchment terms and admission numbers are all held
at a single point in time, and the year chosen on the simulator's slider
changes only the **size** of each neighbourhood's cohort (§2.7). Whether
that is the right treatment — and what an explicit time dimension would
have to carry — is a question this guide returns to once the static terms
are set out.
:::

```{r matrix-blank}
#| tbl-cap: "The shape of the model, for four neighbourhoods and four schools. Sections 2 and 3 fill in the two margins shown; the rest follow, and the interior cells come last."
ex_zones <- zc %>% arrange(desc(Oi)) %>% slice(1:4)
sch4 <- c("Dorothy Stringer School", "Varndean School",
          "Longhill High School", "Cardinal Newman Catholic School")
data.frame(
  Neighbourhood = c(sub("[|].*$", "", ex_zones$zone), "$W_j$"),
  `Dorothy Stringer` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[1]]])),
  Varndean = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[2]]])),
  Longhill = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[3]]])),
  `Cardinal Newman` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[4]]])),
  `$A_i$` = c(rep("?", 4), ""),
  `$O_i$ (2026)` = c(sprintf("%.1f", ex_zones$Oi), ""),
  check.names = FALSE) %>%
  knitr::kable(align = "lccccrr")
```

The interior of that matrix is what the model produces. The margins are
what it is given, or what it fits. $O_i$ is the one margin that is a
count of real people, and it is where this guide starts.

# $O_i$: the children in each neighbourhood {#sec-oi}

## What it is

$O_i$ is the number of children living in zone $i$ who will enter Year 7
at a **state** secondary school in the year being modelled. The model is
production-constrained on it, which means that whatever else happens,

$$\sum_j T_{ij} = O_i \quad \text{for every } i.$$

Every child is placed somewhere. Nothing is created and nothing is lost,
which is why $O_i$ has to be right before anything else can be.

## What a zone is

A zone is **an LSOA cut by catchment**, keyed as
`LSOA | pre-2024 catchment | current catchment`. The city has
`r fmt_n(nrow(zc))` zones drawn from `r fmt_n(n_distinct(zc$lsoa))`
LSOAs.

The cut exists because the 2024 boundary change runs *through* LSOAs.
Assigning a whole LSOA to whichever catchment contains its centroid would
register only a fraction of the children the redraw actually moves. So
every postcode in the ONS Postcode Directory is assigned a catchment on
each of the two maps by point-in-polygon, and the LSOA is split
accordingly. Carrying **both** maps in the key means the two regimes sit
on identical geography and can be compared without rebuilding anything.

```{r zone-example}
#| tbl-cap: "Two LSOAs on the Longhill / Stringer-Varndean boundary, each split into two zones: the same neighbourhood, a different catchment on the current map."
z %>%
  filter(lsoa %in% c("E01016862", "E01016863")) %>%
  transmute(Zone = zone, LSOA = lsoa, `Pre-2024` = catch_pre2024,
            `Current map` = catch_optionZ, Postcodes = n_pcd,
            `$O_i$ (2026)` = sprintf("%.1f", Oi)) %>%
  knitr::kable(align = "llllrr")
```

## Where the numbers come from {#sec-oi-sources}

| Input | Source | What it supplies |
|:--|:--|:--|
| Ward population by single year of age, mid-2022 | [ONS ward-level mid-year population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/wardlevelmidyearpopulationestimatesexperimental) (`sapewardstablefinal.xlsx`) | how many children of each age live in each ward |
| LSOA population by single year of age | [ONS lower super output area mid-year population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates) (`lsoa_age_11_props_totals.xlsx`) | how a ward's children divide between its LSOAs |
| Postcode directory, February 2024 | [ONS Postcode Directory, ONS Open Geography Portal](https://geoportal.statistics.gov.uk/) | which ward and LSOA each postcode is in |
| Catchment boundaries | [Brighton & Hove City Council, school admissions](https://www.brighton-hove.gov.uk/schools-and-learning/school-admissions) — the option Z map in force from 2026 entry, and the map it replaced | the catchment half of the zone key |
| Year 7 offers, 2024 | [Allocation factsheet, Year 7 places September 2024](https://www.brighton-hove.gov.uk/schools-and-learning/school-policies-reports-strategies-and-other-documents/allocation-factsheet-year-7-places-september-2024) | the state-sector share |
| LSOA boundaries | [ONS Open Geography Portal](https://geoportal.statistics.gov.uk/) | drawing, and joining the age data to zones |

None of it is pupil-level. The smallest thing any of these sources
describes is a postcode count or an LSOA age band.

## How $O_i$ is built, step by step

The code is `public/R/01_open_inputs.R` in the private bundle, sections
1b and 1c. Its result travels to the simulator as `data/open_inputs.rds`.

**Step 1 — ages become entry years.** A child aged $a$ in the mid-2022
estimates enters Year 7 in $2022 + (11 - a)$. Ages 0 to 11 therefore give
entry years 2022 to 2033, every one of them children ONS has already
counted rather than births anyone has projected.

**Step 2 — two extrapolated years.** Children entering in 2034 and 2035
were not born when the mid-2022 estimate was taken. They are a linear
continuation of the five most recent observed cohorts, 2029 to 2033, and
are flagged as extrapolated wherever they appear.

**Step 3 — resident children become state-sector demand.** Not every
resident child takes a state Year 7 place: some go to independent
schools, some leave the city. A single constant carries that:

$$\text{state share} \;=\; \frac{\text{Year 7 offers made, 2024}}{\text{resident children, 2024 entry}} \;=\; `r sprintf("%.3f", STATE_SHARE)`$$

taken from the council's published offer total against the ONS cohort of
the same year.

**Step 4 — a ward is divided between its zones.** ONS publishes the age
detail at ward level, so each ward's children are shared out between the
zones in it. The weight for a zone is its LSOA's age-11 population share,
divided between the zones that LSOA is cut into by their postcode counts:

$$u_i \;=\; \underbrace{p_{\ell(i)}}_{\text{LSOA age-11 share}} \times
\underbrace{\frac{n_i}{\sum_{k \in \ell(i)} n_k}}_{\text{postcodes in this zone}},
\qquad
O_i^{\,2026} \;=\; \text{ward children}_{w(i)} \times \frac{u_i}{\sum_{k \in w(i)} u_k} \times \text{state share}$$

where $\ell(i)$ is the zone's LSOA, $w(i)$ its ward and $n_i$ its
postcode count. `r fmt_n(sum(!is.na(z$prop)))` of the `r fmt_n(nrow(z))`
zones use the ONS age share; the `r fmt_n(sum(is.na(z$prop)))` in the
expansion area outside the city fall back to postcode counts alone.

::: {.callout-note appearance="simple"}

## Why not divide a ward by postcode count

Splitting a ward's children by how many postcodes each LSOA holds assumes
every postcode contains the same number of eleven-year-olds. Family
housing, student housing and retirement housing differ enormously on
that, and the difference runs along exactly the lines catchment
boundaries follow. An earlier version of this work used postcode counts
between LSOAs, and the deprivation contrast between the two halves of a
paired catchment was almost entirely smoothed away. Postcode counts do
the narrower job above — dividing one LSOA between its own fragments,
where nothing finer is published.
:::

::: {.callout-important}

## Corrected: a split LSOA used to be counted twice

Until this version, each zone of a split LSOA inherited the **whole**
LSOA's age-11 share as its weight. Ward and city totals were unaffected,
because the denominator grew too, but inside a ward a split LSOA took
about 1.7 to 1.9 times the share its population implied and the unsplit
LSOAs around it took correspondingly less. The splits sit exactly on the
catchment boundaries, so the error was concentrated in the neighbourhoods
these questions turn on — Whitehawk and Kemptown among them.

The weight is now divided between a split LSOA's zones by their postcode
counts, as in the formula above. The table below is the check that used
to fail.
:::

```{r split-check}
#| tbl-cap: "Each LSOA's modelled share of its ward against the share its age-11 population implies, grouped by how many zones the LSOA is split into. A ratio of 1 is agreement."
chk <- z %>% filter(!is.na(prop)) %>%
  group_by(ward) %>% mutate(zone_share = Oi / sum(Oi)) %>% ungroup() %>%
  group_by(ward, lsoa) %>%
  summarise(prop = first(prop), zone_share = sum(zone_share),
            n_zones = n(), .groups = "drop") %>%
  group_by(ward) %>% mutate(prop_share = prop / sum(prop)) %>% ungroup() %>%
  mutate(ratio = zone_share / prop_share)
chk %>%
  group_by(`Zones the LSOA is split into` = n_zones) %>%
  summarise(LSOAs = n(), `Median ratio` = sprintf("%.2f", median(ratio)),
            Lowest = sprintf("%.2f", min(ratio)),
            Highest = sprintf("%.2f", max(ratio)), .groups = "drop") %>%
  knitr::kable(align = "rrrrr")
```

`r fmt_n(sum(z$Oi[z$lsoa %in% chk$lsoa[chk$n_zones > 1]]))` of the
`r fmt_n(sum(z$Oi))` children in the model live in a split LSOA, so the
correction moves children between zones inside those wards without
changing any ward or city total.

```{r fig-split-lsoa}
#| fig-cap: "One LSOA cut by the catchment boundary in force from 2026. Each point is a live postcode, coloured by the catchment it falls in; the black line is the LSOA boundary and the heavier line the catchment boundary. The model splits the LSOA's children between the two sides in proportion to the postcodes on each."
#| fig-height: 5
pcd <- readr::read_csv(here::here("data", "zone_postcode_points.csv"),
                       show_col_types = FALSE)
# The clearest example to draw: the split LSOA with the most postcodes on
# its smaller side, so neither half is a handful of points.
pick <- pcd %>% count(lsoa, catch_optionZ) %>% group_by(lsoa) %>%
  filter(n() == 2) %>% summarise(smaller = min(n), .groups = "drop") %>%
  slice_max(smaller, n = 1) %>% pull(lsoa)
pp <- pcd %>% filter(lsoa == pick) %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>% st_transform(4326) %>%
  mutate(Catchment = unname(CATCH_LABELS[as_boundary_catchment(catch_optionZ)]))
ls_one <- bh_data("lsoa.geojson") %>% filter(lsoa21cd == pick) %>% st_transform(4326)
catch_sf <- bh_data("catchments_current.geojson") %>% st_transform(4326)
bb <- st_bbox(st_buffer(st_transform(ls_one, 27700), 250)) %>%
  st_as_sfc() %>% st_transform(4326) %>% st_bbox()
ggplot() +
  geom_sf(data = catch_sf, aes(fill = catchment), colour = "grey25", alpha = 0.10,
          linewidth = 0.9, show.legend = FALSE) +
  geom_sf(data = ls_one, fill = NA, colour = "black", linewidth = 0.9) +
  geom_sf(data = pp, aes(colour = Catchment), size = 2.6, alpha = 0.9) +
  scale_fill_manual(values = CATCH_COLOURS) +
  scale_colour_manual(
    values = setNames(unname(CATCH_COLOURS[as_boundary_catchment(unique(pp$catch_optionZ))]),
                      unname(CATCH_LABELS[as_boundary_catchment(unique(pp$catch_optionZ))])),
    name = "Catchment of the postcode") +
  coord_sf(xlim = bb[c("xmin", "xmax")], ylim = bb[c("ymin", "ymax")]) +
  labs(title = sprintf("%s, split between two catchments", pick),
       subtitle = sprintf("%d postcodes on one side, %d on the other",
                          max(table(pp$catch_optionZ)), min(table(pp$catch_optionZ)))) +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"), legend.position = "top")
```

```{r split-worked}
#| tbl-cap: "The same LSOA, as the model sees it. The age-11 share belongs to the whole LSOA; each zone takes the part of it its postcodes imply."
z %>% filter(lsoa == pick) %>%
  transmute(Zone = zone, Postcodes = n_pcd,
            `Share of the LSOA` = sprintf("%.0f%%", 100 * n_pcd / sum(n_pcd)),
            `Age-11 share of the LSOA` = sprintf("%.4f", prop),
            `Weight used` = sprintf("%.4f", prop * n_pcd / sum(n_pcd)),
            `$O_i^{\\,2026}$` = sprintf("%.1f", Oi)) %>%
  knitr::kable(align = "lrrrrr")
```

Before the correction both rows of that table carried the whole LSOA's
age-11 share, so the neighbourhood counted twice in its ward's
denominator and took roughly double the children it should.

## A worked example

```{r oi-worked}
#| tbl-cap: "One ward, divided between its zones. Ward children is the ONS estimate for this cohort; the share is the zone's weight within the ward; O is the two multiplied together with the state-sector share."
wd <- z %>% filter(ward == (z %>% arrange(desc(Oi)) %>% slice(1) %>% pull(ward)))
wd %>%
  arrange(desc(Oi)) %>%
  transmute(Zone = sub("[|]", " | ", zone), LSOA = lsoa, Postcodes = n_pcd,
            `Age-11 share of LSOA` = sprintf("%.4f", prop),
            `Share of the ward` = sprintf("%.3f", share_in_ward),
            `Ward children` = fmt_n(ward_children),
            `$O_i$ (2026)` = sprintf("%.1f", Oi)) %>%
  knitr::kable(align = "llrrrrr")
```

Reading the first row: the ward holds `r fmt_n(wd$ward_children[1])`
children of this cohort; that zone carries
`r sprintf("%.1f%%", 100 * max(wd$share_in_ward))` of the ward's weight;
multiplied by the state-sector share of
`r sprintf("%.3f", STATE_SHARE)`, the zone holds
`r sprintf("%.1f", max(wd$Oi))` Year 7 children in 2026.

## The city total

```{r oi-total}
#| tbl-cap: "The modelled 2026 cohort by catchment. These are children living in the catchment, not children attending its school."
zc %>% group_by(Catchment = catchment) %>%
  summarise(children = sum(Oi), Zones = n(), .groups = "drop") %>%
  arrange(desc(children)) %>%
  transmute(Catchment, `Children, 2026` = fmt_n(children), Zones) %>%
  knitr::kable(align = "lrr")
```

The city's zones hold `r fmt_n(sum(zc$Oi))` children for 2026 entry. That
is the model's entire demand before any child chooses anything, and every
intake the simulator reports is drawn from it.

### Where the city totals come from {#sec-oi-provenance}

Two different city-wide totals appear in this work, and they have
different provenance. It is worth being explicit about which is which.

**Offers actually made** are the council's own published figures. Its
[allocation factsheets](https://www.brighton-hove.gov.uk/schools-and-learning/school-admissions)
are issued after each March allocation round, and the series from 2010 to
2026 was transcribed school by school in the earlier
[Brighton and Hove allocations](https://adamdennett.github.io/BH_Schools_2/brighton_and_hove_allocations.html)
analysis. This bundle does not retype them: `public/R/01a_factsheet_panel.R`
re-extracts the year-by-year tables from that document, so every offer,
preference and admission number here traces back to a published council
factsheet.

**Resident children** are ONS, as set out in @sec-oi-sources. They are a
different quantity from offers: a resident child may go to an independent
school, or out of the city.

**The model's city total is the first anchored on the second.** The
state-sector share of `r sprintf("%.3f", STATE_SHARE)` is the council's
published 2024 offer total divided by the ONS resident cohort for the same
entry year, so the model's demand is ONS children scaled to the level the
council's own offers imply. It is not a free-floating estimate.

```{r oi-provenance}
#| tbl-cap: "The two totals, and the model's demand between them."
fp <- bh_data("factsheet_panel.rds")$factsheets %>% filter(name != "Total")
off <- fp %>% group_by(year) %>% summarise(offers = sum(off_total), .groups = "drop")
tibble::tribble(
  ~Quantity, ~Source, ~Value,
  "Year 7 offers made, 2024", "Council allocation factsheet",
  fmt_n(off$offers[off$year == 2024]),
  "Year 7 offers made, latest round", "Council allocation factsheet",
  sprintf("%s (%d)", fmt_n(off$offers[which.max(off$year)]), max(off$year)),
  "Resident children, 2024 entry", "ONS ward estimates, mid-2022",
  fmt_n(dem_raw$children[dem_raw$area == "Brighton & Hove" & dem_raw$entry_year == 2024]),
  "State-sector share", "The first divided by the second",
  sprintf("%.3f", STATE_SHARE),
  "Model demand, 2026 entry", "ONS children x state share",
  fmt_n(sum(zc$Oi))) %>%
  knitr::kable(align = "llr")
```

The model's 2026 demand and the offers made in the 2026 round are not the
same number and should not be expected to match exactly: the offers
include children placed at a school they did not name, and exclude
children who took a place outside the city. The comparison is still worth
making, and it is the first check listed in @sec-oi-checks.

## The slider: how $O_i$ changes with the entry year {#sec-oi-year}

The year slider does exactly one thing to $O_i$. The city's projected
cohort is turned into an index on 2026, and every zone is scaled by it:

$$O_i^{\,y} \;=\; O_i^{\,2026} \times d^{\,y}, \qquad
d^{\,y} \;=\; \frac{\text{state demand}^{\,y}}{\text{state demand}^{\,2026}}$$

```{r oi-index}
#| tbl-cap: "The demand index behind the slider. The last two years are the extrapolated cohorts."
dem %>%
  transmute(`Entry year` = year, `City cohort` = fmt_n(cohort),
            `Index on 2026` = sprintf("%.3f", index),
            Basis = ifelse(extrapolated, "extrapolated",
                           "children already counted")) %>%
  knitr::kable(align = "rrrl")
```

```{r fig-oi-index}
#| fig-cap: "The cohort the model scales to. Every zone is multiplied by the same factor, so the map of where children live is held still and only the size of the cohort moves."
#| fig-height: 4.2
dem %>%
  ggplot(aes(year, cohort)) +
  geom_col(aes(fill = extrapolated), width = 0.66) +
  geom_text(aes(label = fmt_n(cohort)), vjust = -0.5, size = 3.2, colour = "grey25") +
  scale_fill_manual(values = c(`FALSE` = "#1f3b57", `TRUE` = "#9aa7b4"),
                    labels = c(`FALSE` = "children already counted",
                               `TRUE` = "extrapolated"), name = NULL) +
  scale_x_continuous(breaks = dem$year) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "State-sector Year 7 demand",
       title = "The cohort behind the slider, 2026 to 2035") +
  theme_bh() + theme(legend.position = "top")
```

### Which projection this is, and the one beside it {#sec-oi-projection}

Two projections of the city's Year 7 cohort appear in this work. They are
built from different data and they are used for different things, so it
is worth being plain about which drives the model.

**The one the slider uses** is the ONS estimate carried forward: the
children counted in the mid-2022 small-area estimates, aged forward to
their Year 7 entry year (§2.4, steps 1 and 2), and multiplied by the
state-sector share. It is used because it is the only one that exists at
**ward** level, and the model needs the geography as well as the total —
$O_i$ is a ward figure divided between zones.

**The one beside it** is built from children already in the city's
primary schools: Reception offers made seven years earlier, multiplied by
the share of a Reception cohort that reappears as a Year 7 offer. That
ratio is measured rather than assumed, from the council's own published
factsheets, and it has been strikingly steady —
`r fmt_pct(100 * rc$survival_recent, 1)` across the five most recent
cohorts, with a standard deviation of
`r fmt_pct(100 * rc$survival_sd, 1)`. The method is set out in full in
[how our projection method works](https://adamdennett.github.io/BH_Schools_2/brighton_and_hove_allocations.html#how-our-projection-method-works),
and section 3 of the strategic view uses it as its headline projection
because it rests on children already registered in a school rather than
on a population estimate.

```{r oi-two-projections}
#| tbl-cap: "The two projections side by side. The first drives the model's zone-level demand; the second is the check on it, and the strategic view's headline series."
rc$comparison %>%
  transmute(`Entry year` = entry_year,
            `From ONS estimates (drives the slider)` = fmt_n(ons_demand),
            `From Reception offers` = fmt_n(reception_based),
            Difference = sprintf("%+.0f", reception_based - ons_demand),
            `%` = sprintf("%+.1f%%", 100 * (reception_based / ons_demand - 1))) %>%
  knitr::kable(align = "rrrrr")
```

They agree within a few per cent for most years, which is reassuring for
both, and they diverge most in the years where the ONS series is itself
extrapolated. Neither is a forecast in the sense of predicting behaviour:
one counts children who exist, the other counts children already in
school and applies a measured survival rate.

**What this does not do.** The projection is city-wide, so the slider
cannot say that one neighbourhood shrinks faster than another: every zone
keeps its 2026 share of the city's children in every year. The council's
own forecasts differ from this mostly in exactly that respect — its
Longhill catchment line falls faster than the city's does — so comparing
catchment totals year by year compares two different claims rather than
two estimates of one thing.

## Checks you can run {#sec-oi-checks}

```{r oi-checks}
#| tbl-cap: "Identities that should hold, and what they give as this document is built."
ward_tot <- z %>% group_by(ward) %>%
  summarise(oi = sum(Oi), implied = first(ward_children) * STATE_SHARE,
            .groups = "drop") %>%
  mutate(d = abs(oi - implied))
tibble::tribble(
  ~Check, ~Result,
  "Every zone has a positive cohort",
  sprintf("smallest is %.2f children", min(z$Oi)),
  "Zone cohorts sum to each ward's estimate times the state share",
  sprintf("largest difference %.3f children", max(ward_tot$d)),
  "The city's zones sum to the 2026 projection",
  sprintf("%s against %s", fmt_n(sum(zc$Oi)), fmt_n(dem$cohort[dem$year == 2026])),
  "The index is exactly 1 in the base year",
  sprintf("%.6f", dem$index[dem$year == 2026]),
  "A split LSOA takes the share its population implies",
  sprintf("largest ratio %.2f", max(chk$ratio))) %>%
  knitr::kable(align = "lr")
```

Two further comparisons are worth doing by hand, because they test $O_i$
against something the model never saw: the city total against the
council's published Year 7 offers for the same round, and each
catchment's total against the catchment forecasts in the council's place
planning appendix.

## What to be careful about

- **The base is mid-2022.** Migration into or out of the city since then
  is not in the counts, and a child who has moved between wards is still
  counted where they were.
- **One state-sector share, for every year and every neighbourhood.** It
  is `r sprintf("%.3f", STATE_SHARE)`, from a single round. If
  independent-school take-up varies across the city — and it plainly does
  — then $O_i$ is too high in some places and too low in others.
- **Ward-level age detail.** Within a ward, the LSOA age-11 distribution
  does the work; how children are distributed inside an LSOA is not
  observed, so postcode counts stand in for it where a boundary cuts one.
- **Peacehaven and the expansion area** are carried in the open bundle
  but excluded from the simulator's zones. Peacehaven Community School is
  a destination in the model, not a source of demand.
- **No composition.** $O_i$ is a headcount. Which of those children are
  disadvantaged is a separate layer applied later, not part of this term.

## Questions this raises

1. Should the state-sector share vary by neighbourhood, given that
   independent-school take-up does?
2. Should the year slider move the *geography* of demand as well as its
   size — that is, should each catchment carry its own projection, as the
   council's forecasts do?

# $W_j$: how attractive each school is {#sec-wj}

## What it is

$W_j$ is the pull of school $j$ on a family, **net of where the school
stands**. It is dimensionless, scaled so the city average is 1, and it
enters the model multiplicatively: doubling $W_j$ doubles the numerator
of every flow into $j$ before distance, catchment and competition have
their say. It also enters the competing-destinations term, because a
school's rivals count for more when they are schools families want.

It is the hardest term in the model to pin down, because everything
observable about a school is entangled with where it is. A school with
high attainment may be sought after because it teaches well, or because
it sits where better-off families already live.

## Two layers: the open specifications, and the fitted values

**Layer one — the open specifications.** The open bundle carries five
candidate measures of attractiveness, so that a conclusion can be tested
across all of them rather than resting on one.

```{r wj-specs}
#| tbl-cap: "The five open specifications, each scaled so the city average is 1. Weighted preferences count ranks 1 to 3, a second preference worth half a first and a third a quarter."
oi$attract %>%
  transmute(School = short_sch(name), PAN = pan, `Attainment 8` = round(att8, 1),
            Equal = sprintf("%.2f", W_equal), `By PAN` = sprintf("%.2f", W_pan),
            `First preferences` = sprintf("%.2f", W_prefs),
            `Weighted preferences` = sprintf("%.2f", W_wprefs),
            `By Attainment 8` = sprintf("%.2f", W_att8)) %>%
  arrange(desc(as.numeric(`Weighted preferences`))) %>%
  knitr::kable(align = "lrrrrrrr")
```

The central one is **weighted preferences per place**: the council's
published factsheets, pooled over the most recent rounds, counting first,
second and third preferences with a geometric discount of 0.5 per rank,
divided by the admission number. The paired catchments are why: a family
in Stringer / Varndean or Hove Park / Blatchington Mill must rank two
local schools against each other, so a first-preference count reads a
school families put second as one nobody wanted.

**Layer two — the fitted values the simulator uses.** The full model does
not take any of those five as given. It **balances** $W_j$ so that the
model reproduces the share of the city's first preferences each school
actually receives, using the council's catchment-level preference table
from its evidence to the Schools Adjudicator. The catchment terms
$\gamma_h$ are fitted in the same loop (§4, to follow), because the two
trade off against each other.

$$W_j \;\leftarrow\; W_j \left(\frac{\text{target share}_j}{\text{modelled share}_j}\right)^{0.8},
\qquad \text{repeat until } \max_j |\text{modelled} - \text{target}| < 10^{-6}$$

starting from the weighted-preference values, with the result rescaled to
a city mean of 1. The exponent 0.8 is damping: it stops the balancing
oscillating around the target. This is the same device as the balancing
factor $A_i$, applied to the other margin of the matrix — $A_i$ makes
every neighbourhood place all its children, and this makes every school
receive the share of first preferences it is observed to receive.

```{r wj-fitted}
#| tbl-cap: "The fitted attractiveness the simulator's sliders multiply, beside the weighted-preference measure it starts from, and each school's target share of the city's first preferences."
tibble(name = names(W5), W5 = as.numeric(W5)) %>%
  left_join(oi$attract %>% select(name, W_wprefs, att8), by = "name") %>%
  left_join(tibble(name = names(cal$target_first_prefs),
                   target = as.numeric(cal$target_first_prefs)), by = "name") %>%
  arrange(desc(W5)) %>%
  transmute(School = short_sch(name),
            `Weighted preferences` = sprintf("%.2f", W_wprefs),
            `Fitted $W_j$` = sprintf("%.2f", W5),
            `Target share of first preferences` = sprintf("%.1f%%", 100 * target),
            `Attainment 8` = sprintf("%.1f", att8)) %>%
  knitr::kable(align = "lrrrr")
```

```{r fig-wj}
#| fig-cap: "The fitted attractiveness against the measure it starts from. A school above the line is one the model has to make more attractive than its weighted preferences suggest to reproduce the preferences each catchment gives it."
#| fig-height: 4.6
tibble(name = names(W5), W5 = as.numeric(W5)) %>%
  left_join(oi$attract %>% select(name, W_wprefs), by = "name") %>%
  ggplot(aes(W_wprefs, W5)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "31") +
  geom_point(size = 3.4, colour = "#1f3b57") +
  ggrepel::geom_text_repel(aes(label = short_sch(name)), size = 3.4, seed = 1,
                           colour = "#1f3b57") +
  expand_limits(x = 0, y = 0) +
  labs(x = "Weighted preferences per place (city mean = 1)",
       y = "Fitted attractiveness (city mean = 1)",
       title = "What the balancing does to the starting values") +
  theme_bh()
```

## What point in time does $W_j$ belong to? {#sec-wj-time}

$W_j$ is **fixed**. One number per school, no year subscript, used for
2026 and for 2035 alike. That deserves stating plainly, because it is the
assumption a reader is least likely to notice and the one that a school
can falsify fastest.

Two windows go into it:

```{r wj-window}
#| tbl-cap: "The observations behind the attractiveness term."
fp <- bh_data("factsheet_panel.rds")
tibble::tribble(
  ~Layer, ~Source, ~Window,
  "Starting values: weighted preferences per place",
  "Council allocation factsheets",
  sprintf("%d rounds, %d onwards", fp$pref_years, fp$attract_from),
  "Fitted values: balanced to first preferences",
  "Council's catchment-level preference table",
  sprintf("%d rounds, %s", length(cal$rounds), paste(cal$rounds, collapse = ", ")),
  "Rank discount inside the window",
  "A second preference counts this much of a first",
  sprintf("%.2f, and a third %.2f", fp$wpref_decay, fp$wpref_decay^2)) %>%
  knitr::kable(align = "lll")
```

So the model's attractiveness describes the city as families have treated
it over the last three to five admissions rounds. It is a recent average,
not a long-run constant, and not a snapshot of one year either — a single
round moves too much to fit to.

**Why that matters: the ordering is not stable over a decade.**

```{r fig-wj-time}
#| fig-cap: "First preferences for each school, from the published allocation factsheets. Nothing here is modelled. The shaded band is the window the attractiveness term is built from."
#| fig-height: 5
fs <- fp$factsheets %>% filter(name != "Total") %>%
  mutate(School = short_sch(name),
         hl = School %in% c("King's", "Hove Park", "Longhill"))
ggplot(fs, aes(year, pref1, group = School)) +
  annotate("rect", xmin = fp$attract_from - 0.5, xmax = max(fs$year) + 0.4,
           ymin = -Inf, ymax = Inf, fill = "#1f3b57", alpha = 0.07) +
  geom_line(data = fs %>% filter(!hl), colour = "grey75", linewidth = 0.7) +
  geom_line(data = fs %>% filter(hl), aes(colour = School), linewidth = 1.3) +
  ggrepel::geom_text_repel(
    data = fs %>% filter(hl, year == max(year)), aes(label = School, colour = School),
    hjust = 0, nudge_x = 0.4, direction = "y", size = 4, seed = 2, show.legend = FALSE) +
  scale_colour_manual(values = c(`King's` = "#7570b3", `Hove Park` = "#d95f02",
                                Longhill = "#b2182b"), guide = "none") +
  scale_x_continuous(breaks = seq(2010, 2026, 4), limits = c(2010, 2029)) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "First preferences",
       title = "How a school is regarded is not a constant",
       subtitle = sprintf("Published factsheets, %d to %d. Shaded: the rounds the attractiveness term uses.",
                          min(fs$year), max(fs$year))) +
  theme_bh()
```

Over sixteen years King's first preferences roughly doubled while Hove
Park's roughly halved, in the same part of the same city. A $W_j$ built
from 2014 would put those two schools the other way round from a $W_j$
built from 2026. Reputations move, and they can move quickly: a new head,
a turn in attendance or results, a building programme, an Ofsted
judgement.

::: {.callout-important}

## How the model handles this, and what it cannot do

**What it does.** It fixes attractiveness at its recent value and gives
you a slider to move it. Every scenario in the strategic view that says
"a school half as attractive again" is that slider, and §3.6 converts a
slider position into points of Attainment 8 so the size of the
assumption is legible rather than abstract.

**What it does not do.** Attractiveness never responds to anything the
model itself produces. A school whose roll falls does not become less
wanted in the next year of the run; a school whose intake changes does
not see its results, and therefore its reputation, move with it. The
feedback loop the strategic view describes — a more disadvantaged intake,
lower headline scores, fewer preferences, a more disadvantaged intake
still — is **outside** this model. Where that loop is real, a run over ten
years understates how far apart schools drift.

That is the clearest case in this document for making time explicit
rather than implicit: $O_i$ already carries a year, and the natural next
step would be to let $W_j$ carry one too.
:::

## Why there is no exponent on $W_j$ {#sec-wj-alpha}

The general form of a production-constrained model carries a power on
attractiveness as well as on cost:

$$T_{ij} \;=\; A_i\, O_i\, W_j^{\alpha}\, c_{ij}^{-\beta}$$

Here $\alpha$ is absent, which is the same as fixing it at 1. That is a
deliberate choice and it follows from what $W_j$ is.

**A fitted quantity does not need a parameter to scale it.** $W_j$ is not
a proxy standing in for attractiveness; it *is* attractiveness, balanced
until the model reproduces each school's share of the city's first
preferences. An exponent on it would be a second free parameter doing the
same job as the balancing, and the two would trade off against each other:
any $\alpha$ can be absorbed by re-balancing $W$, because
$(W^{1/\alpha})^{\alpha} = W$. Fitting both would be fitting the same
thing twice, and the pair would not be identified.

**It would be needed if $W_j$ were a proxy.** Suppose attractiveness were
taken crudely as the number of pupils on roll, or as the admission
number, or as Attainment 8 — all of which this work carries as
alternative specifications (§3.2). None of those *is* pull; each is a
measurable thing believed to move with it, and there is no reason the
relationship should be one-for-one. A school twice the size is not
necessarily twice as wanted. An exponent would then be exactly the right
device: $\alpha$ fitted to observed flows would say how strongly demand
responds to the proxy, and $\alpha = 1$ would be a finding rather than an
assumption. In the classic retail application that is precisely what the
exponent on floorspace does.

So the rule is: **balance it, or scale it, but not both.** This model
balances. A version of it built on roll or on published attainment, which
is what anyone without the council's preference table would have to do,
should carry an $\alpha$ and fit it.

## Where it sits in the matrix

```{r matrix-wj}
#| tbl-cap: "The same four neighbourhoods and schools. Two margins are now filled: the children in each neighbourhood, and the pull of each school."
data.frame(
  Neighbourhood = c(sub("[|].*$", "", ex_zones$zone), "$W_j$"),
  `Dorothy Stringer` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[1]]])),
  Varndean = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[2]]])),
  Longhill = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[3]]])),
  `Cardinal Newman` = c(rep("$T_{ij}$", 4), sprintf("%.2f", W5[[sch4[4]]])),
  `$A_i$` = c(rep("?", 4), ""),
  `$O_i$ (2026)` = c(sprintf("%.1f", ex_zones$Oi), ""),
  check.names = FALSE) %>%
  knitr::kable(align = "lccccrr")
```

## What the attractiveness slider does

Each school's slider multiplies its fitted $W_j$, from
`r sprintf("%.2f", 0.25)` to `r sprintf("%.0f", 4)` times. Nothing else
changes: the same journey times, the same catchment terms, the same
admission numbers. Three details are worth knowing.

- **The multiplier is applied to the fitted value**, not to weighted
  preferences, so a slider at 1 reproduces the model exactly.
- **The out-of-city schools are not touched by it.** Priory, Peacehaven,
  Seahaven and Seaford Head carry their own fitted attractiveness and a
  decay on straight-line distance, and the sliders do not reach them.
- **CoMArt has no preferences**, because it does not exist. When it is
  opened, it borrows the attractiveness of
  `r inp$comart$w_from_short` as its starting value, and its slider moves
  from there.

## Turning a multiplier into something meaningful

A slider position is not a quantity anyone outside a model can judge, so
the app converts it to the one published number families appear to
respond to. Across the ten city schools, the fitted attractiveness is
log-linear in Attainment 8:

$$\log W_j = a + b \cdot \text{Attainment 8}_j,
\qquad b = `r sprintf("%.4f", inp$attain$slope)`,
\quad R^2 = `r sprintf("%.2f", inp$attain$r2)`,
\quad n = `r inp$attain$n`$$

so a point of Attainment 8 is worth about
`r fmt_pct(100 * inp$attain$per_point, 0)` more attractiveness, and a
multiplier $m$ inverts to $\log(m) / b$ points. That is how the simulator
answers "how good would this school have to be to fill?" — and it is an
association, not a lever: Attainment 8 is itself largely a description of
who a school admits.

## Checks you can run {#sec-wj-checks}

```{r wj-checks}
#| tbl-cap: "Identities and fits that should hold for the attractiveness term."
own <- cal$own
tibble::tribble(
  ~Check, ~Result,
  "Fitted attractiveness averages 1 across the city",
  sprintf("%.6f", mean(W5)),
  "Balanced first-preference shares match the council's table",
  sprintf("largest gap %.4f percentage points",
          100 * max(abs(cal$cells$m5_share - cal$cells$obs_share))),
  "Share of each catchment's first preferences to its own school",
  sprintf("observed %.0f%% to %.0f%%, modelled %.0f%% to %.0f%%",
          100 * min(own$observed), 100 * max(own$observed),
          100 * min(own$M5), 100 * max(own$M5)),
  "Attainment 8 explains the fitted values",
  sprintf("R2 %.2f on %d schools", inp$attain$r2, inp$attain$n),
  "A slider at 1 reproduces the published model",
  "checked in app/tests/check.R") %>%
  knitr::kable(align = "lr")
```

## What to be careful about

- **Ten observations.** Every statement about *why* a school is
  attractive rests on ten schools. The balancing is an accounting
  identity across those ten; the Attainment 8 fit is a description of
  them, not evidence about causes.
- **Attractiveness and location are only partly separated.** The
  balancing asks what pull each school needs *given* where it stands and
  how far families will travel, which is a real separation — but it
  inherits whatever $\beta$ and $\gamma_h$ are set to. If distance decay
  were weaker, the fitted $W_j$ would be different.
- **The faith schools have no catchment**, so nothing in the model
  restricts who may choose them. Their fitted attractiveness absorbs the
  religious criteria that really ration those places, which is why they
  carry high values.
- **Peacehaven's preferences are not published** by the council, so both
  preference measures are filled at the lowest observed rate in the city.
  That is an assumption, and it flatters Longhill: a weak Peacehaven
  keeps eastern children in the city.
- **$W_j$ does not move with time.** A school that improves, or a new
  building, is a slider position rather than anything the model
  generates. Whether attractiveness should respond to a school's roll,
  its intake or its results is exactly the dynamic question §2.7 flags.

## Questions this raises

1. Should $W_j$ be allowed to vary by family type — by deprivation, say —
   rather than being one number per school for the whole city?
2. Should the balancing target first preferences, as now, or all three
   ranks discounted as the open specification does?
3. Should a school's attractiveness respond to its own intake over time,
   which is the feedback loop the strategic view describes but the model
   does not contain?

# $c_{ij}$: the journey from a neighbourhood to a school {#sec-cij}

```{r cij-setup}
#| include: false
costs <- oi$costs_now
cost_elm <- oi$costs_elm
tm <- readRDS(here::here("data", "travel", "build_metadata.rds"))
lsoa_sf <- bh_data("lsoa.geojson") %>% st_transform(4326)
catch_sf <- bh_data("catchments_current.geojson") %>% st_transform(4326)
sch_sf <- schools_sf(oi$schools) %>% filter(name != "Peacehaven Community School")
zone_lsoa <- z %>% select(zone, lsoa, Oi)
# Zone costs carried up to LSOA level for drawing, weighted by children.
to_lsoa <- function(d, school) {
  d %>% filter(name == school) %>%
    inner_join(zone_lsoa, by = "zone") %>%
    group_by(lsoa) %>%
    summarise(cij = weighted.mean(cij, Oi), .groups = "drop")
}
LH <- "Longhill High School"
```

## What it is

$c_{ij}$ is the **generalised cost** of getting from neighbourhood $i$ to
school $j$: minutes of walking and bus, door to door, on a weekday
morning. It is not a straight-line distance, and it is not a car journey.
It enters the model as $c_{ij}^{-\beta}$, so it is the term that makes a
school ten minutes away far more likely to be chosen than one an hour
away — and the exponent $\beta$ that governs how fast that falls away is
its own section, to follow.

The city has `r fmt_n(nrow(inp$cost$now))` zone-school pairs in the
simulator's cost table, spanning
`r sprintf("%.0f", min(inp$cost$now$cij))` to
`r sprintf("%.0f", max(inp$cost$now$cij))` minutes.

## Where the numbers come from

Every journey is routed with [`r5r`](https://ipeagit.github.io/r5r/) over
two networks: the street network from
[OpenStreetMap](https://www.openstreetmap.org/copyright), and the bus
timetable as a [GTFS feed](https://data.bus-data.dft.gov.uk/) from the
Department for Transport's Bus Open Data Service. The build is
`public/R/00b_build_travel_matrix.R` in the open bundle.

```{r cij-params}
#| tbl-cap: "The routing parameters. Every journey in this work uses these; changing any of them changes every journey time."
tibble::tribble(
  ~Setting, ~Value,
  "Modes", paste(tm$params$mode, collapse = " and "),
  "Departure", format(tm$params$departure_datetime, "%A %d %B %Y, %H:%M"),
  "Longest walk allowed, either end", sprintf("%d minutes", tm$params$max_walk_time),
  "Longest journey allowed", sprintf("%d minutes", tm$params$max_trip_duration),
  "Departure window", sprintf("%d minute", tm$params$time_window),
  "Percentile of journeys in the window", sprintf("%dth", tm$params$percentiles),
  "Origins", sprintf("%s postcodes", fmt_n(tm$n_origins)),
  "Destinations", sprintf("%d, including the Elm Grove site", nrow(tm$destinations))) %>%
  knitr::kable(align = "lr")
```

The departure date is chosen rather than assumed: the build counts the
trips each candidate weekday actually carries in the feed and takes a
Tuesday, Wednesday or Thursday in service, because a date the feed does
not serve returns no journeys at all and a Monday or Friday timetable is
lighter than a normal school day.

## From postcodes to zones

The matrix is routed **postcode to school**. A zone's cost is the mean
over the live postcodes in it:

$$c_{ij} \;=\; \frac{1}{|P_i|}\sum_{p \in P_i} c_{pj}$$

where $P_i$ is the set of postcodes in zone $i$. Because a zone is an
LSOA cut by catchment (§2.2), the two halves of a split LSOA get their
own journey times — which is the point of cutting them.

## Two corrections the raw routing needs

**Walking is always available.** Where the bus network is awkward the
router returns a poor itinerary: about a fifth of pairs came back slower
than simply walking the distance, in one case 99 minutes for 4 km. That
is a property of the router, not of the city, so every journey is capped
at the time it would take to walk:

$$c_{ij} \;=\; \min\!\left(c^{\text{routed}}_{ij},\;
\frac{d_{ij} \times `r sprintf("%.1f", acc$walk_fallback$circuity)`}{`r sprintf("%.2f", acc$walk_fallback$kmh)` \text{ km/h}} \times 60\right)$$

The walking pace comes from a systematic review of active school travel:
children's mean walking trip is
`r fmt_n(acc$walk_fallback$review$dist_m)` m in
`r sprintf("%.1f", acc$walk_fallback$review$time_min)` minutes, or
`r sprintf("%.2f", acc$walk_fallback$kmh)` km/h, and the circuity factor
of `r sprintf("%.1f", acc$walk_fallback$circuity)` turns straight-line
distance into street distance. The cap binds on
`r fmt_pct(100 * mean(costs$walk_capped), 0)` of pairs. **It is a ceiling
on modelled time, not a claim that children walk that far** — the same
review finds they rarely walk beyond 1.8 km.

**A distance model fills the gaps.** Where the matrix has no routed value
at all, a fitted model of time on straight-line distance stands in. It
reaches `r fmt_pct(100 - 100 * mean(costs$routed), 1)` of pairs in the
city, so almost everything here is routed rather than predicted.

```{r fig-cij-walk}
#| fig-cap: "Every zone-school pair: routed journey time against straight-line distance. The line is the walking time the cap uses. Points above it are journeys the router made slower than walking, and they are pulled down onto the line."
#| fig-height: 4.6
wk <- costs %>% mutate(walk_min = km * acc$walk_fallback$circuity / acc$walk_fallback$kmh * 60)
ggplot(wk, aes(km, cij)) +
  geom_point(aes(colour = walk_capped), size = 1.1, alpha = 0.5) +
  geom_line(aes(km, walk_min), colour = "#1f3b57", linewidth = 0.9) +
  scale_colour_manual(values = c(`FALSE` = "grey65", `TRUE` = "#d95f02"),
                      labels = c(`FALSE` = "routed time used",
                                 `TRUE` = "capped at walking time"), name = NULL) +
  labs(x = "Straight-line distance (km)", y = "Journey time used (minutes)",
       title = "Where the cap binds",
       subtitle = sprintf("%s of %s pairs are capped",
                          fmt_n(sum(costs$walk_capped)), fmt_n(nrow(costs)))) +
  theme_bh() + theme(legend.position = "top")
```

## What a journey-time surface looks like

```{r fig-cij-longhill}
#| fig-cap: "Journey time to Longhill from every neighbourhood, at its current site and at the Elm Grove site. Both panels use the same scale, so they can be read against each other. The white dot is the school."
#| fig-height: 5
lh_now <- to_lsoa(costs, LH) %>% mutate(site = "Ovingdean (today)")
lh_elm <- to_lsoa(cost_elm, LH) %>% mutate(site = "Elm Grove")
lh_map <- lsoa_sf %>%
  inner_join(bind_rows(lh_now, lh_elm), by = c("lsoa21cd" = "lsoa")) %>%
  mutate(site = factor(site, c("Ovingdean (today)", "Elm Grove")))
pts <- bind_rows(
  sch_sf %>% filter(name == LH) %>% mutate(site = "Ovingdean (today)"),
  st_as_sf(data.frame(e = oi$elm_grove$easting, n = oi$elm_grove$northing),
           coords = c("e", "n"), crs = 27700) %>% st_transform(4326) %>%
    mutate(site = "Elm Grove")) %>%
  mutate(site = factor(site, levels(lh_map$site)))
ggplot() +
  geom_sf(data = lh_map, aes(fill = cij), colour = NA) +
  geom_sf(data = catch_sf, fill = NA, colour = "white", linewidth = 0.25) +
  geom_sf(data = pts, shape = 21, fill = "white", colour = "black", size = 3, stroke = 1.1) +
  facet_wrap(~ site) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "Minutes") +
  labs(title = "The same school, two sites",
       subtitle = "Walk-and-bus journey time from each neighbourhood, weekday morning") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"),
        strip.text = element_text(face = "bold"))
```

The median neighbourhood is
`r sprintf("%.0f", median(lh_now$cij))` minutes from Ovingdean and
`r sprintf("%.0f", median(lh_elm$cij))` from Elm Grove. That difference
is the whole of what relocation does to $c_{ij}$ — and, because nothing
else in the model moves with it except the competition term, most of what
relocation does at all.

```{r fig-cij-nearest}
#| fig-cap: "Which school each neighbourhood can reach quickest, by routed walk-and-bus time. Boundaries are the catchments in force; where a colour crosses a boundary, the catchment school is not the nearest one."
#| fig-height: 5.2
nearest <- costs %>%
  inner_join(zone_lsoa, by = "zone") %>%
  group_by(lsoa, name) %>% summarise(cij = weighted.mean(cij, Oi), .groups = "drop_last") %>%
  slice_min(cij, n = 1, with_ties = FALSE) %>% ungroup() %>%
  mutate(School = short_sch(name))
near_map <- lsoa_sf %>% inner_join(nearest, by = c("lsoa21cd" = "lsoa"))
ggplot() +
  geom_sf(data = near_map, aes(fill = School), colour = "white", linewidth = 0.1) +
  geom_sf(data = catch_sf, fill = NA, colour = "grey15", linewidth = 0.6) +
  geom_sf(data = sch_sf, shape = 21, fill = "white", colour = "black", size = 2.6) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  labs(title = "The quickest school from each neighbourhood",
       subtitle = "Routed walk-and-bus time, not distance, and not who is admitted") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"))
```

## A worked example

```{r cij-worked}
#| tbl-cap: "One neighbourhood's journey to every city school, with the straight-line distance beside it. The two are not the same ordering, which is the reason for routing at all."
ex_zone <- costs %>% filter(zone == ex_zones$zone[1])
ex_zone %>%
  arrange(cij) %>%
  transmute(School = short_sch(name), `Journey time (minutes)` = sprintf("%.0f", cij),
            `Straight line (km)` = sprintf("%.1f", km),
            Basis = ifelse(walk_capped, "capped at walking time",
                           ifelse(routed, "routed", "distance model"))) %>%
  knitr::kable(align = "lrrl")
```

## The journeys the model does not route

- **The four schools outside the city** — Priory, Peacehaven, Seahaven
  and Seaford Head — are chosen on **straight-line kilometres**, not
  routed minutes. The bus network has no direct eastern service to Lewes,
  so the routed journey from Woodingdean goes through Brighton and takes
  over an hour, which is not what the families who make that journey do.
  Their decay is fitted on km against the published 2024 destinations.
- **CoMArt**, when opened, is routed from its old site on the same
  network and with the same parameters as every other school.
- **Longhill at Elm Grove** has a routed column of its own. An earlier
  version had no routed times for the site and gave it the average
  accessibility of a Brighton secondary, which was wrong by about sixty
  children; the build now stops rather than let that happen silently.

## Checks you can run

```{r cij-checks}
#| tbl-cap: "What the cost table should satisfy, and what it gives."
st <- acc$statutory
tibble::tribble(
  ~Check, ~Result,
  "Every zone can reach every city school",
  sprintf("%s pairs, none missing", fmt_n(nrow(inp$cost$now))),
  "Journeys are routed rather than predicted",
  sprintf("%.1f%% routed", 100 * mean(costs$routed)),
  "No journey is slower than walking",
  sprintf("%.0f%% capped, none above the walk line", 100 * mean(costs$walk_capped)),
  "Mean journey to the catchment school",
  sprintf("%.1f minutes", st$mean_min),
  "Against the statutory limit for secondary (75 minutes)",
  sprintf("%.0f children over it", st$over_75_children),
  "Against the national average school journey (19 minutes)",
  sprintf("city mean %.1f minutes", st$mean_min)) %>%
  knitr::kable(align = "lr")
```

The strongest check is not in the table: a handful of journeys can be
compared against local knowledge. Woodingdean to Longhill comes out at
about 17 minutes and Woodingdean to Dorothy Stringer at about 68;
Whitehawk to Longhill about 28 and to Stringer about 62. Those are the
journeys that decide the eastern half of this work, and they are worth
checking against what a family there would say.

## What to be careful about

- **One departure, one day.** Every journey is a single weekday morning
  in the timetable as it stood when the matrix was built. Term-time
  variation, a missed connection and a cancelled bus are all outside it.
- **A mean over postcodes.** A zone's cost is the average of the
  postcodes in it, so a neighbourhood with one badly connected corner
  shows as slightly worse rather than showing both.
- **No cars, no lifts, no walking buses.** The model prices the journey a
  child could make on foot and by bus. A family with a car makes a
  different journey, and that is one reason the east's figures should be
  read as a ceiling on difficulty rather than a description of it.
- **Time is symmetric here.** One morning journey stands for the return
  as well, and no allowance is made for after-school travel.
- **The cap is generous at distance.** A 6 km walk is not a real school
  journey, but capping at walking time is deliberately crude: it removes
  router artefacts without inventing a bus that does not run.

## Questions this raises

1. Should the zone cost be the mean over its postcodes, or the
   child-weighted mean, or the best available journey rather than the
   average one?
2. Should journeys be routed at more than one departure time, so that a
   thin early timetable is visible?
3. Should the out-of-city schools be routed rather than measured on
   straight-line distance, if a feed covering East Sussex services were
   used?

# $\beta$: how sharply distance deters {#sec-beta}

```{r beta-setup}
#| include: false
env <- bh_data("sensitivity_envelope.rds")
BETA <- inp$params$beta
ext <- cal$outside
```

## What it is

$\beta$ is the exponent on journey time. It converts the minutes of §4
into a pull:

$$\text{pull of school } j \text{ on neighbourhood } i \;\propto\; c_{ij}^{-\beta}$$

A larger $\beta$ means families discount distance more sharply. It is a
single number for the whole city, fixed at **`r sprintf("%.1f", BETA)`**,
and it is the same in every run of the simulator: no slider moves it.

The form matters as much as the value. This is a **power** decay, not an
exponential one: doubling the journey divides the pull by
$2^{`r sprintf("%.1f", BETA)`} \approx `r sprintf("%.1f", 2^BETA)`$,
whatever the starting point. That is the standard form in spatial
interaction modelling, and it behaves sensibly over the range of journeys
a city this size contains.

```{r fig-beta-curve}
#| fig-cap: "What the exponent does. Each curve is the pull of a school against journey time, scaled to 1 at ten minutes. The model uses 1.7; the sweep in the open model runs from 1.5 to 3.2."
#| fig-height: 4.4
bs <- c(1.5, BETA, 2.4, 3.2)
tidyr::expand_grid(beta = bs, mins = seq(10, 70, by = 0.5)) %>%
  mutate(pull = (mins^-beta) / (10^-beta),
         lab = sprintf("β = %.1f", beta),
         model = beta == BETA) %>%
  ggplot(aes(mins, pull, colour = lab, linewidth = model)) +
  geom_line() +
  scale_linewidth_manual(values = c(`FALSE` = 0.7, `TRUE` = 1.6), guide = "none") +
  scale_colour_manual(values = c("grey70", "#1f3b57", "grey55", "grey40"), name = NULL) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Journey time (minutes)", y = "Pull, relative to a ten-minute journey",
       title = "A school twice as far away is a fraction as likely to be chosen",
       subtitle = sprintf("At β = %.1f, a 20-minute journey has %.0f%% of the pull of a 10-minute one, and 40 minutes has %.0f%%",
                          BETA, 100 * 2^-BETA, 100 * 4^-BETA)) +
  theme_bh() + theme(legend.position = "top")
```

## Where the value comes from, and what it is not

**It is a judgement, not an estimate.** Fitting $\beta$ properly needs
observed flows — how many children actually travel from each
neighbourhood to each school — and the council has not released them. A
catchment-level preference table identifies the catchment terms (§6, to
follow) but not distance decay, because within a catchment every family
is at a different distance and the table does not say who is where.

So `r sprintf("%.1f", BETA)` is a central value chosen for a system of
this kind: journeys are short, the alternatives sit close together, and
it is squarely inside the range this literature reports for school
choice. The open model does not rely on it being right. It runs
everything across `r length(env$betas)` values from
`r sprintf("%.1f", min(env$betas))` to `r sprintf("%.1f", max(env$betas))`
and reports which conclusions survive the whole range.

::: {.callout-note appearance="simple"}

## What would pin it down

A Poisson regression of flows on log journey time with an origin fixed
effect returns $\beta$ with a standard error, from records the admissions
process already produces. It is half a day's work on data the council
holds. Every band in the open model that comes from sweeping $\beta$
would become a number with an interval around it.
:::

## How much rests on it

```{r beta-importance}
#| tbl-cap: "What the open model's answer is actually sensitive to. Share of the variation in Longhill's modelled recruitment across the full sweep, by input."
env$importance %>%
  filter(term != "Residuals") %>%
  mutate(Input = recode(term,
    "w_spec" = "Which attractiveness specification",
    "factor(entry_year)" = "Which entry year",
    "site_label" = "Longhill's site",
    "factor(beta)" = "Distance decay (β)",
    "factor(gamma)" = "Strength of the catchment term (γ)")) %>%
  arrange(desc(pct)) %>%
  transmute(Input, `Share of the variation` = sprintf("%.1f%%", pct)) %>%
  knitr::kable(align = "lr")
```

**Distance decay is not the big unknown.** It accounts for
`r sprintf("%.0f%%", env$importance$pct[env$importance$term == "factor(beta)"])`
of the variation in the answer, against
`r sprintf("%.0f%%", env$importance$pct[env$importance$term == "w_spec"])`
for how attractive each school is taken to be. That is worth knowing
because it says what data would be worth having: school-level revealed
preference, more than travel behaviour.

```{r fig-beta-sweep}
#| fig-cap: "Longhill's modelled recruitment across the swept range of distance decay, at each attractiveness specification, 2026. The vertical line is the value the model uses. The spread between the coloured lines is the attractiveness question; the slope along each line is the distance-decay question."
#| fig-height: 4.8
env$envelope %>%
  filter(entry_year == 2026, gamma == 1.6) %>%
  mutate(site = ifelse(site == "now", "Ovingdean (today)", "Elm Grove")) %>%
  ggplot(aes(beta, natural_intake, colour = w_label, group = w_label)) +
  geom_vline(xintercept = BETA, colour = "grey45", linetype = "31") +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ site) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  expand_limits(y = 0) +
  labs(x = "Distance decay (β)", y = "Children recruited",
       title = "The slope along each line is what β decides",
       subtitle = "Longhill's recruitment with its own admission number unbinding, catchment term at 1.6") +
  theme_bh() + theme(legend.position = "top")
```

## What it looks like from one neighbourhood

```{r fig-beta-shares}
#| fig-cap: "One neighbourhood's children, shared between schools by journey time alone, at three values of the exponent. Attractiveness, catchment and capacity are all switched off here, so the only thing acting is distance."
#| fig-height: 4.6
zex <- costs %>% filter(zone == ex_zones$zone[1])
tidyr::expand_grid(beta = c(1.5, BETA, 3.2), zex) %>%
  group_by(beta) %>%
  mutate(share = cij^-beta / sum(cij^-beta),
         School = short_sch(name)) %>%
  ungroup() %>%
  mutate(lab = sprintf("β = %.1f", beta),
         School = forcats::fct_reorder(School, cij)) %>%
  ggplot(aes(share, School, fill = lab)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  scale_x_continuous(labels = label_percent()) +
  scale_fill_manual(values = c("grey72", "#1f3b57", "grey45"), name = NULL) +
  labs(x = "Share of the neighbourhood's children", y = NULL,
       title = sprintf("Distance alone, from %s", sub("[|].*$", "", ex_zones$zone[1])),
       subtitle = "Schools ordered by journey time. A steeper decay concentrates children on the nearest schools.") +
  theme_bh() + theme(legend.position = "top")
```

That is the whole of what the exponent does: at
`r sprintf("%.1f", 1.5)` the nearest school takes
`r sprintf("%.0f%%", 100 * with(zex, (cij^-1.5 / sum(cij^-1.5))[which.min(cij)]))`
of this neighbourhood's children on distance alone; at
`r sprintf("%.1f", 3.2)`, `r sprintf("%.0f%%", 100 * with(zex, (cij^-3.2 / sum(cij^-3.2))[which.min(cij)]))`.

## A second decay, on a different scale

The four schools outside the city do not use $\beta$ at all. They are
chosen on straight-line kilometres with a decay of their own, fitted to
the published 2024 destinations:

$$\text{pull of an out-of-city school} \;=\; W^{\text{ext}}_j \; d_{ij}^{-`r sprintf("%.1f", ext$decay)`}$$

`r sprintf("%.1f", ext$decay)` is far steeper than
`r sprintf("%.1f", BETA)`, and the two are not comparable: this one is on
kilometres rather than minutes, and it is doing a different job. It is
absorbing everything that stops a Brighton family choosing a school in
another authority — the admissions arrangements of a different council,
the absence of a school bus, the sheer unfamiliarity — and compressing it
into the one dimension the fit has. It reproduces the published pattern,
which is a low bar for four schools and one year of data, and it should
not be read as a behavioural estimate at all.

## Checks you can run

```{r beta-checks}
#| tbl-cap: "What the decay should produce, and what it gives at the value the model uses."
r_now <- run_sim(inp, year = 2026, rules = list(rule = "priorities"))
m_now <- outcomes(inp, r_now)
tibble::tribble(
  ~Check, ~Result,
  "The exponent the model runs at",
  sprintf("%.1f, unchanged in every scenario", BETA),
  "Mean modelled journey to the school a child attends",
  sprintf("%.1f minutes", m_now$mean_min),
  "Against the national average school journey",
  sprintf("%.0f minutes (National Travel Survey)", acc$benchmarks$nts_min),
  "Share of children modelled at a school in their own catchment",
  sprintf("%.0f%%", 100 * r_now$in_catch_share),
  "Sensitivity of the answer to the exponent",
  sprintf("%.0f%% of the variation in the sweep", env$importance$pct[env$importance$term == "factor(beta)"])) %>%
  knitr::kable(align = "lr")
```

The mean modelled journey sits above the national average, which is what
a city with a coastal edge and one school in the far east should produce.
If it came out at or below the national figure, either the decay or the
journey times would be wrong.

## What to be careful about

- **One exponent for the whole city.** Families differ in how far they
  will travel, and the literature is consistent that the willingness
  varies with income and with whether a car is available. A single
  $\beta$ averages over that, and it is the assumption most likely to
  matter for the equity questions in the strategic view.
- **Everything else is fitted *given* this value.** The attractiveness
  balancing (§3) and the catchment terms (§6) are estimated with $\beta$
  held at `r sprintf("%.1f", BETA)`. A different decay would give
  different fitted values for both, which is why the sweep re-runs the
  whole model rather than swapping the exponent at the end.
- **Time, not distance, and not cost.** A journey that is quick but
  requires two buses and a fare is treated as a short journey.
- **The floor at short journeys.** A power decay rises steeply as time
  approaches zero, so the very shortest journeys carry a lot of weight.
  The routed minimum in the table is
  `r sprintf("%.0f", min(costs$cij))` minutes, which keeps it in bounds.

## Questions this raises

1. Should $\beta$ vary between groups of families — by deprivation, or by
   whether a catchment has one school or two — rather than being one
   number for the city?
2. Is the power form right, or would an exponential decay fit the pattern
   of Brighton's journeys better? Both are standard; they differ most at
   the long-journey end, which is where the eastern questions sit.
3. Should the out-of-city decay be re-fitted once more than one year of
   destination data is available, or replaced by a proper choice model
   for schools in another authority?

# $\gamma_h$: how much living in the catchment counts {#sec-gamma}

```{r gamma-setup}
#| include: false
gam <- cal$gamma
gam_lab <- setNames(unname(CATCH_LABELS[CATCH_FROM_MODEL[names(gam)]]), names(gam))
r_now <- run_sim(inp, year = 2026, rules = list(rule = "priorities"))
```

## What it is

$\gamma_h$ is the bonus a school gets for being **in the family's own
catchment**. It sits in the exponent, so it multiplies a school's pull:

$$\text{pull} \;\propto\; W_j\, c_{ij}^{-\beta} \times
\begin{cases} e^{\gamma_{h(i)}} & \text{if } j \text{ is a school of } i\text{'s catchment}\\ 1 & \text{otherwise}\end{cases}$$

The subscript matters: $\gamma$ is indexed by **$h(i)$, the catchment the
family lives in**, not by the school. Stringer / Varndean families are
modelled as following their map more closely than Brighton Aldridge's
families follow theirs, and that is a statement about families rather
than about schools.

::: {.callout-important}

## This is about what families name, not about who gets admitted

Admission priority is a separate stage of the model (§8, to follow):
when a school is oversubscribed the council's criteria decide who is
offered a place, and living in the catchment is one of them. $\gamma_h$
is doing something different and earlier. It says how much more likely a
family is to **want** a school because it is the one their map points
them to — through habit, siblings, the school bus, or simply reading the
catchment as a statement about where their child belongs.

The distinction is why the strategic view can find that catchment
priority allocates nothing at an undersubscribed school while the
catchment still matters a great deal there: nobody is turned away, but
the map still shapes what families ask for.
:::

## The fitted values

```{r gamma-table}
#| tbl-cap: "The catchment term fitted for each catchment, and what it does to a school's pull. A term of 0 would mean the map makes no difference to what families name."
tibble(catchment = names(gam), gamma = as.numeric(gam)) %>%
  mutate(Catchment = unname(gam_lab[catchment]),
         `Fitted $\\gamma_h$` = sprintf("%.1f", gamma),
         `Multiplies a catchment school's pull by` = sprintf("%.1f×", exp(gamma))) %>%
  arrange(desc(gamma)) %>%
  select(Catchment, `Fitted $\\gamma_h$`, `Multiplies a catchment school's pull by`) %>%
  knitr::kable(align = "lrr")
```

```{r fig-gamma-map}
#| fig-cap: "The fitted catchment term by catchment. Darker is a population that follows its map more closely, once distance and attractiveness are accounted for."
#| fig-height: 5
gsf <- catch_sf %>%
  left_join(tibble(catchment = unname(CATCH_FROM_MODEL[names(gam)]),
                   gamma = as.numeric(gam)), by = "catchment")
ggplot() +
  geom_sf(data = gsf, aes(fill = gamma), colour = "white", linewidth = 0.5) +
  geom_sf(data = sch_sf, shape = 21, fill = "white", colour = "black", size = 2.6) +
  geom_sf_text(data = gsf %>% st_point_on_surface() %>% suppressWarnings(),
               aes(label = sprintf("%.1f", gamma)), size = 5, fontface = "bold",
               colour = "white", nudge_y = 0.006) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "γ") +
  coord_sf(expand = TRUE) +
  labs(title = "How closely each catchment's families follow their map",
       subtitle = "Fitted to the council's catchment-level preference table") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"))
```

Stringer / Varndean, at `r sprintf("%.1f", gam[["DS_Varndean"]])`, and
Longhill, at `r sprintf("%.1f", gam[["Longhill"]])`, are the two most
attached populations in the city; Brighton Aldridge's, at
`r sprintf("%.1f", gam[["BACA"]])`, the least. Those two facts do much of
the work in every relocation and redraw result in the strategic view,
because a move trades one against the other.

## How they are fitted

The council's evidence to the Schools Adjudicator includes a
catchment-level preference table: how many first, second and third
preferences each catchment's children gave each school, pooled over
`r length(mt$calibrated$rounds)` rounds. That is a flow matrix — coarse,
but a matrix — and it is enough to fit one term per catchment.

The loop is:

1. Take the current catchment terms and **balance $W_j$** so the model
   reproduces each school's share of the city's first preferences (§3).
2. For one catchment, try every $\gamma$ on a grid from
   `r sprintf("%.0f", 0)` to `r sprintf("%.0f", 4)` in steps of 0.1,
   re-balancing $W_j$ at each step, and keep the value that minimises the
   **Poisson deviance** between the modelled and observed shares in that
   table.
3. Repeat for each catchment, then sweep again until nothing moves.

The deviance is the right loss here because the cells are counts of
preferences: it weights a cell by how many families it contains, so a
catchment that gives 400 first preferences counts for more than one that
gives 40.

```{r fig-gamma-profile}
#| fig-cap: "The fit against each candidate value, catchment by catchment. The point is the value chosen: the bottom of each curve. A flat curve would mean the data cannot distinguish one value from another."
#| fig-height: 5
cal$profile %>%
  mutate(Catchment = unname(gam_lab[catchment])) %>%
  group_by(Catchment) %>%
  mutate(rel = deviance - min(deviance)) %>%
  ungroup() %>%
  ggplot(aes(gamma, rel)) +
  geom_line(colour = "#1f3b57", linewidth = 0.9) +
  geom_point(data = ~ .x %>% group_by(Catchment) %>% slice_min(rel, n = 1),
             colour = "#d95f02", size = 2.6) +
  facet_wrap(~ Catchment, scales = "free_y") +
  labs(x = "Catchment term (γ)", y = "Deviance above the best fit",
       title = "Each catchment's own curve, and where it bottoms out",
       subtitle = "Lower is a better fit to the council's preference table") +
  theme_bh(11)
```

## What the fitting buys

```{r gamma-fit}
#| tbl-cap: "How well each specification reproduces the council's preference table. M4 is one catchment term for the whole city, fitted against offers; M5 is one per catchment, fitted against preferences."
cal$fit %>%
  transmute(Model = ifelse(model == "M4", "One term for the whole city (M4)",
                           "One term per catchment (M5)"),
            `Poisson deviance` = fmt_n(deviance),
            `Cell error (percentage points)` = sprintf("%.1f", cell_rmse_pp)) %>%
  knitr::kable(align = "lrr")
```

```{r fig-gamma-own}
#| fig-cap: "The share of each catchment's first preferences that go to its own school or schools: what the council's table records, and what each specification produces."
#| fig-height: 4.6
cal$own %>%
  mutate(Catchment = unname(gam_lab[home])) %>%
  tidyr::pivot_longer(c(observed, M4, M5), names_to = "src", values_to = "share") %>%
  mutate(src = recode(src, observed = "Observed",
                      M4 = "One term for the city", M5 = "One term per catchment"),
         src = factor(src, c("Observed", "One term for the city", "One term per catchment")),
         Catchment = forcats::fct_reorder(Catchment, share, .fun = max)) %>%
  ggplot(aes(share, Catchment, colour = src, shape = src)) +
  geom_point(size = 3.6) +
  scale_colour_manual(values = c("#1f3b57", "grey65", "#d95f02"), name = NULL) +
  scale_shape_manual(values = c(16, 15, 17), name = NULL) +
  scale_x_continuous(labels = label_percent()) +
  labs(x = "Share of the catchment's first preferences going to its own school", y = NULL,
       title = "A single city-wide term cannot reproduce the pattern",
       subtitle = "Fitting one term per catchment does, and it is the same model otherwise") +
  theme_bh() + theme(legend.position = "top")
```

The gain is large and it is concentrated where it matters. A single
city-wide term puts Stringer / Varndean's own-school share at
`r fmt_pct(100 * cal$own$M4[cal$own$home == "DS_Varndean"], 0)` against an
observed `r fmt_pct(100 * cal$own$observed[cal$own$home == "DS_Varndean"], 0)`,
and Longhill's at
`r fmt_pct(100 * cal$own$M4[cal$own$home == "Longhill"], 0)` against
`r fmt_pct(100 * cal$own$observed[cal$own$home == "Longhill"], 0)`. Fitting
each catchment separately reproduces both.

## What the slider does

The simulator's catchment slider multiplies **every** fitted term by the
same factor, from 0 to 2. At 0 the map stops mattering to what families
name and the model becomes distance and attractiveness alone; at 2 every
catchment's attachment doubles in the exponent, which is a very strong
map indeed. At 1 it uses the fitted values.

That single multiplier is deliberately blunt. It answers "what if
families follow their catchments more, or less, than they do now?"
without pretending to know which catchment would change.

```{r gamma-slider}
#| tbl-cap: "What the slider does to the share of children modelled at a school in their own catchment, 2026, council priorities."
purrr::map_dfr(c(0, 0.5, 1, 1.5, 2), function(g) {
  rr <- run_sim(inp, year = 2026, gamma = g, rules = list(rule = "priorities"))
  tibble(`Slider` = sprintf("%.1f×", g),
         `At a school in their own catchment` = fmt_pct(100 * rr$in_catch_share, 0))
}) %>%
  knitr::kable(align = "lr")
```

## Where it sits in the matrix

Unlike $O_i$ and $W_j$, the catchment term does not live in a margin. It
is a **mask over the interior**: for each neighbourhood, the cells of
schools in its own catchment are multiplied by $e^{\gamma_h}$ and the
rest are left alone.

```{r matrix-gamma}
#| tbl-cap: "The four neighbourhoods again. A cell marked with the multiplier is one where the school is in that neighbourhood's catchment; the others are untouched."
dsg <- inp$designs[["Current catchments"]]
cellmark <- function(zone, school) {
  in_c <- !is.na(dsg$school[school]) && dsg$school[school] == dsg$zone[zone]
  h <- unname(dsg$zone[zone])
  if (in_c) sprintf("$T_{ij} \\times %.1f$", exp(gam[[h]])) else "$T_{ij}$"
}
data.frame(
  Neighbourhood = c(sub("[|].*$", "", ex_zones$zone), "$W_j$"),
  `Dorothy Stringer` = c(vapply(ex_zones$zone, cellmark, "", school = sch4[1]),
                         sprintf("%.2f", W5[[sch4[1]]])),
  Varndean = c(vapply(ex_zones$zone, cellmark, "", school = sch4[2]),
               sprintf("%.2f", W5[[sch4[2]]])),
  Longhill = c(vapply(ex_zones$zone, cellmark, "", school = sch4[3]),
               sprintf("%.2f", W5[[sch4[3]]])),
  `Cardinal Newman` = c(vapply(ex_zones$zone, cellmark, "", school = sch4[4]),
                        sprintf("%.2f", W5[[sch4[4]]])),
  `$O_i$ (2026)` = c(sprintf("%.1f", ex_zones$Oi), ""),
  check.names = FALSE) %>%
  knitr::kable(align = "lccccr")
```

Cardinal Newman is in nobody's catchment: it and King's admit city-wide
on faith criteria, so no cell in their columns is ever multiplied. In the
two paired catchments **both** schools are in catchment, which is what
makes those catchments different objects from the four single-school
ones.

## Checks you can run

```{r gamma-checks}
#| tbl-cap: "What the catchment terms should satisfy."
prof_flat <- cal$profile %>% group_by(catchment) %>%
  summarise(span = diff(range(deviance)), .groups = "drop")
tibble::tribble(
  ~Check, ~Result,
  "Every catchment's fit has an interior minimum",
  sprintf("fitted %.1f to %.1f, grid 0 to 4", min(gam), max(gam)),
  "The data can tell one value from another",
  sprintf("deviance spans %s to %s across the grid",
          fmt_n(min(prof_flat$span)), fmt_n(max(prof_flat$span))),
  "Fitting per catchment beats one term for the city",
  sprintf("deviance %s against %s", fmt_n(cal$fit$deviance[cal$fit$model == "M5"]),
          fmt_n(cal$fit$deviance[cal$fit$model == "M4"])),
  "Modelled children at a catchment school, today",
  fmt_pct(100 * r_now$in_catch_share, 0),
  "Slider at zero removes the map entirely",
  fmt_pct(100 * run_sim(inp, year = 2026, gamma = 0,
                        rules = list(rule = "priorities"))$in_catch_share, 0)) %>%
  knitr::kable(align = "lr")
```

## What to be careful about

::: {.callout-warning}

## The assumption every redraw result depends on

The terms are fitted on the map families have lived with, and then
**carried unchanged to any other map**. A redraw in the simulator moves
which schools count as in catchment; it does not change how strongly a
family follows the line.

That is the single most consequential assumption in this work. If
families follow a new boundary less closely than the one they know —
which the strategic view finds some evidence for in what happened after
2024 — then every redraw result here is too optimistic about how many
children the redraw moves. The strategic view reports the halved-term
version of the relocation runs for exactly this reason.
:::

- **The term is fitted on the pre-2024 map**, because that is the map the
  preference table covers.
- **It bundles several mechanisms.** Siblings already at the school, the
  free bus, habit, and the sense that the catchment school is *the*
  school all end up in one number, because the table cannot separate
  them.
- **It is a home-catchment term, not a school term.** A school in two
  catchments — none exist today, but CoMArt opening would create one —
  takes each catchment's own strength.
- **Paired catchments are a single object here.** A Stringer / Varndean
  family gets the bonus on both schools, and which of the two they prefer
  is left to $W_j$, distance, and the exclusivity populations.

## Questions this raises

1. Should the term decay when a boundary changes — a family newly placed
   in a catchment following it less closely than one that has been in it
   for years?
2. Should it vary with how many schools a catchment contains, rather than
   being one number per catchment?
3. Would first preferences by small area, before and after the 2024
   change, let the normative half of this be separated from the
   procedural half?

# $C_j$ and $\delta$: the company a school keeps {#sec-cj}

```{r cj-setup}
#| include: false
DELTA <- inp$params$delta
SIGMA <- inp$params$sigma
sch_city <- inp$schools[inp$schools$city & !(inp$schools$hypothetical %in% TRUE), ]
W_run <- setNames(sch_city$W, sch_city$name)
xy <- sch_city %>% select(name, easting, northing)
pairs_km <- tidyr::expand_grid(name = xy$name, k = xy$name) %>%
  filter(name != k) %>%
  left_join(xy, by = "name") %>%
  left_join(xy %>% rename(k = name, ke = easting, kn = northing), by = "k") %>%
  mutate(km = pmax(sqrt((easting - ke)^2 + (northing - kn)^2) / 1000, 0.1),
         Wk = unname(W_run[k]), contrib = Wk * km^-SIGMA,
         short = short_sch(name), k_short = short_sch(k))

sch_elm <- sch_city
lh_row <- sch_elm$name == "Longhill High School"
sch_elm$easting[lh_row] <- sch_city$elm_easting[lh_row]
sch_elm$northing[lh_row] <- sch_city$elm_northing[lh_row]
C_now <- compete_now(sch_city, W_run, SIGMA)
C_elm <- compete_now(sch_elm, W_run, SIGMA)

Cj <- tibble(name = names(C_now), C = unname(C_now),
             C_elm = unname(C_elm[names(C_now)])) %>%
  mutate(short = short_sch(name), index = 100 * C / mean(C),
         index_elm = 100 * C_elm / mean(C), mult = (C / mean(C))^DELTA)
cj_of <- function(s, col) Cj[[col]][Cj$short == s]
min_km <- function(s) min(pairs_km$km[pairs_km$short == s])
rmse0 <- function(x) x$rmse[abs(x$value) < 1e-8]
r2_0  <- function(x) x$r2[abs(x$value) < 1e-8]
```

## What it is

$C_j$ measures **how much company a school has**: how many other schools
stand close to it, weighted by how wanted each of them is.

$$C_j \;=\; \sum_{k \neq j} W_k\, d_{jk}^{-\sigma}, \qquad \sigma = `r sprintf("%.0f", SIGMA)`$$

$d_{jk}$ is the straight-line distance in kilometres between two
**schools**, and it deliberately carries a different symbol from
$c_{ij}$: nobody travels from one school to another, so routing it would
mean nothing. All that is wanted is a measure of how clustered the
schools are. With $\sigma$ at `r sprintf("%.0f", SIGMA)` the term reduces
to *the sum of the rivals' attractiveness per kilometre they are away*.

It enters the model as a power on the pull:

$$\text{pull} \;\propto\; W_j\, c_{ij}^{-\beta}\, e^{\gamma_{h}\mathbb{1}[\cdot]}\, C_j^{\,\delta}, \qquad \delta = `r sprintf("%+.1f", DELTA)`$$

The reasoning is Fotheringham's: a plain gravity model is misspecified
when destinations cluster, because people choose an area first and a
destination within it second. Left out, the clustering does not go away —
it is absorbed by the distance parameter instead.

## Brighton in one term

```{r fig-cj-map}
#| fig-cap: "Each school sized by the company it keeps, with its score on an index where the city average is 100. The lines open up the two extremes: every rival's contribution to Dorothy Stringer's total in orange, and to Longhill's in navy. A thicker line is a bigger contribution."
#| fig-height: 5.0
sch_pts <- sch_city %>%
  select(name, easting, northing) %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>%
  st_transform(4326) %>%
  mutate(short = short_sch(name),
         lon = st_coordinates(.)[, 1], lat = st_coordinates(.)[, 2]) %>%
  left_join(Cj %>% select(name, C, index), by = "name")
lonlat <- sch_pts %>% st_drop_geometry() %>% select(name, lon, lat)
segs <- pairs_km %>%
  filter(short %in% c("Dorothy Stringer", "Longhill")) %>%
  left_join(lonlat, by = "name") %>%
  left_join(lonlat %>% rename(k = name, klon = lon, klat = lat), by = "k")

ggplot() +
  geom_sf(data = catch_sf, fill = "grey96", colour = "white", linewidth = 0.5) +
  geom_segment(data = segs, aes(x = lon, y = lat, xend = klon, yend = klat,
                                linewidth = contrib, colour = short), alpha = 0.5) +
  geom_sf(data = sch_pts, aes(size = C), shape = 21, fill = "white",
          colour = "black", stroke = 0.9) +
  ggrepel::geom_text_repel(
    data = sch_pts, aes(lon, lat, label = sprintf("%s\n%.0f", short, index)),
    size = 2.9, lineheight = 0.9, colour = "#1f3b57", seed = 11,
    box.padding = 0.45, point.padding = 0.3, min.segment.length = 0.2,
    segment.colour = "grey55", segment.size = 0.3, max.overlaps = Inf) +
  scale_linewidth_continuous(range = c(0.2, 3.4), guide = "none") +
  scale_size_continuous(range = c(2, 9), guide = "none") +
  scale_colour_manual(values = c(`Dorothy Stringer` = "#d95f02", Longhill = "#1f3b57"),
                      name = NULL,
                      labels = function(x) paste("contributions to", x)) +
  coord_sf(expand = TRUE, clip = "off") +
  labs(title = "Five hundred metres from a rival, or five kilometres",
       subtitle = "The competing-destinations term by school, city average = 100") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"),
        legend.position = "top")
```

The two ends of the city are different objects on this measure.

```{r cj-contrib}
#| tbl-cap: "What the term is made of at its two extremes. Each row is one rival: how far away it is, how wanted it is, and what it therefore contributes."
pairs_km %>%
  filter(short %in% c("Dorothy Stringer", "Longhill")) %>%
  group_by(short) %>% arrange(desc(contrib), .by_group = TRUE) %>%
  mutate(rank = row_number(), total = sum(contrib)) %>%
  filter(rank <= 3) %>% ungroup() %>%
  transmute(School = sprintf("%s (total %.2f)", short, total),
            Rival = k_short, `Distance (km)` = sprintf("%.2f", km),
            `Its $W_k$` = sprintf("%.2f", Wk),
            Contribution = sprintf("%.2f", contrib)) %>%
  knitr::kable(align = "llrrr")
```

Varndean stands `r sprintf("%.0f", 1000 * min_km("Dorothy Stringer"))`
metres from Dorothy Stringer, and Cardinal Newman — the most wanted
school in the city on $W_j$ — is under two kilometres from both. Those
two rivals alone contribute more to Stringer's term than **all nine** of
Longhill's rivals contribute to its own. On an index with the city
average at 100, Stringer scores
`r sprintf("%.0f", cj_of("Dorothy Stringer", "index"))` and Longhill
`r sprintf("%.0f", cj_of("Longhill", "index"))`.

Two features of the term follow from how it is built, and both matter in
the simulator:

- **It uses the attractiveness in force, not a fixed table.** Pull an
  attractiveness slider and every *other* school's $C_j$ moves too,
  because $W_k$ sits inside the sum.
- **It is a sum over the schools that exist in a run.** Close a school and
  its neighbours lose the company it was providing; open CoMArt in East
  Brighton and Longhill gains some.

## What the fitted exponent does

$\delta$ was fitted by grid search over
`r sprintf("%+.1f to %+.1f", min(mt$delta_fit$value), max(mt$delta_fit$value))`,
against the offers the council made in the most recent round, with the
capacity ceiling and the catchment term already in place (§6 explains why
the ceiling goes in first).

```{r fig-cj-delta}
#| fig-cap: "How much the term improves the fit, for three versions of the model. Each curve is that version's error against the offers made, relative to itself at δ = 0, so below 1 means the term is earning its place. The shaded half is the sign Fotheringham's argument predicts."
#| fig-height: 4.8
dprof <- bind_rows(
  mt$delta_fit   %>% mutate(v = "With the ceiling and the catchment term"),
  mt$delta_alone %>% mutate(v = "With the ceiling only"),
  mt$delta_nocap %>% mutate(v = "With neither")) %>%
  group_by(v) %>% mutate(rel = rmse / rmse[abs(value) < 1e-8]) %>% ungroup() %>%
  mutate(v = factor(v, c("With the ceiling and the catchment term",
                         "With the ceiling only", "With neither")))
ggplot(dprof, aes(value, rel, colour = v)) +
  annotate("rect", xmin = min(dprof$value), xmax = 0, ymin = -Inf, ymax = Inf,
           fill = "#b2182b", alpha = 0.05) +
  geom_hline(yintercept = 1, colour = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(data = dprof %>% group_by(v) %>% slice_min(rel, n = 1) %>% ungroup(),
             size = 3, show.legend = FALSE) +
  annotate("text", x = min(dprof$value) + 0.05, y = 1.55, hjust = 0, vjust = 1,
           size = 3.5, colour = "#b2182b", lineheight = 0.95,
           label = "rivals take share\n(competing destinations)") +
  scale_colour_manual(values = c("#1f3b57", "#d95f02", "grey60"), name = NULL) +
  coord_cartesian(ylim = c(0.8, 1.6)) +
  labs(x = "δ", y = "Error, relative to the same model at δ = 0",
       title = sprintf("Fitted at δ = %+.1f, on the wrong side of zero", DELTA),
       subtitle = "Dots mark each version's best value") +
  theme_bh() +
  theme(legend.position = "top", legend.direction = "vertical",
        legend.justification = "left")
```

**The sign is not the one the term was added to test.** A negative
$\delta$ is the competing-destinations prediction: rivals nearby take
share. The fit lands at `r sprintf("%+.1f", DELTA)` — a school does
slightly *better* for having company. That is agglomeration, the pattern
of retail parks and restaurant streets, rather than the correction
Fotheringham proposed. Negative values are not marginally worse but much
worse: at $\delta = -1$ the error is about
`r sprintf("%.1f", mt$delta_fit$rmse[which.min(abs(mt$delta_fit$value + 1))] / min(mt$delta_fit$rmse))`
times its best.

Three things keep the finding modest:

- **The gain is small.** RMSE falls from
  `r sprintf("%.1f", rmse0(mt$delta_fit))` at $\delta = 0$ to
  `r sprintf("%.1f", min(mt$delta_fit$rmse))` at the fitted value, and
  $R^2$ from `r sprintf("%.3f", r2_0(mt$delta_fit))` to
  `r sprintf("%.3f", mt$delta_fit$r2[which.min(mt$delta_fit$rmse)])`.
- **Without the ceiling it does nothing at all.** Fitted with no capacity
  constraint the best value is `r sprintf("%+.1f", mt$delta_nocap_hat)`
  and the error moves only from `r sprintf("%.1f", rmse0(mt$delta_nocap))`
  to `r sprintf("%.1f", min(mt$delta_nocap$rmse))` — the grey curve above,
  which barely dips below its own starting point. What the term appears to explain is mostly rationing.
- **It and $\gamma$ are partly substitutes.** With the ceiling but no
  catchment term, $\delta$ on its own fits at
  `r sprintf("%+.1f", mt$delta_alone_hat)` and reaches much the same
  error. Both terms hand extra pull to the same central schools by
  different routes, so neither is a clean estimate of its own mechanism.

```{r cj-multiplier}
#| tbl-cap: "What the term does to each school's pull at the fitted exponent. A multiplier of 1 would be no effect. The last column is the same index with Longhill rebuilt at Elm Grove, which changes the distances every school is measured against."
Cj %>% arrange(desc(index)) %>%
  transmute(School = short,
            `$C_j$ (city average = 100)` = sprintf("%.0f", index),
            `Multiplies its pull by` = sprintf("%.2f", mult),
            `$C_j$ with Longhill at Elm Grove` = sprintf("%.0f", index_elm)) %>%
  knitr::kable(align = "lrrr")
```

So the effect on pull runs from about
`r sprintf("%.2f", min(Cj$mult))` for the most isolated school to
`r sprintf("%.2f", max(Cj$mult))` for the most crowded: Longhill's pull
is cut by roughly `r sprintf("%.0f", 100 * (1 - min(Cj$mult)))`% against
the city average and Stringer's raised by about
`r sprintf("%.0f", 100 * (max(Cj$mult) - 1))`%. Smaller than $\gamma$ or
$\beta$, but not nothing.

## Why it is in the model at all

Three reasons, none of them "it improved the fit a lot":

1. **It is the standard correction** for a gravity model over clustered
   destinations. Leaving it out does not remove the clustering; it loads
   it onto $\beta$, and the distance parameter then means two things at
   once.
2. **It moves when a school moves.** Rebuilt at Elm Grove, Longhill's
   term rises from `r sprintf("%.0f", cj_of("Longhill", "index"))` to
   `r sprintf("%.0f", cj_of("Longhill", "index_elm"))` — still below
   average, but no longer alone on the eastern edge. That is a real
   difference between the two sites and the model should see it. It is
   also a correction made in this version: before it, competition was
   measured from Ovingdean even in runs where the school had moved.
3. **It records a negative finding.** That a compact city of ten schools
   shows no competing-destinations effect is worth reporting, and it is
   at least consistent with the structure: a production-constrained model
   already normalises each neighbourhood's choices across every school
   through $A_i$ (§8), which is where much of the work the term would
   otherwise do has already been done.

## Checks you can run

```{r cj-checks}
#| tbl-cap: "What the competing-destinations term should satisfy."
tibble::tribble(
  ~Check, ~Result,
  "Schools in the sum",
  sprintf("%d city schools; out-of-city destinations carry no C_j at all",
          nrow(sch_city)),
  "Distances are school to school, straight line",
  sprintf("%.2f km to %.2f km", min(pairs_km$km), max(pairs_km$km)),
  "Closest pair in the city",
  sprintf("%s and %s, %.0f m", pairs_km$short[which.min(pairs_km$km)],
          pairs_km$k_short[which.min(pairs_km$km)], 1000 * min(pairs_km$km)),
  "Most isolated school",
  sprintf("%s, nearest rival %.1f km", Cj$short[which.min(Cj$C)],
          min_km(Cj$short[which.min(Cj$C)])),
  "Rivals are attractiveness-weighted",
  "yes: an unwanted neighbour counts for less than a wanted one",
  "Effect on pull at the fitted exponent",
  sprintf("%.2f to %.2f times", min(Cj$mult), max(Cj$mult)),
  "Measured from where each school stands in the run",
  sprintf("yes: Longhill scores %.0f at Ovingdean and %.0f at Elm Grove",
          cj_of("Longhill", "index"), cj_of("Longhill", "index_elm"))) %>%
  knitr::kable(align = "lr")
```

## What to be careful about

- **$\sigma$ is set, not fitted.** The decay on school-to-school distance
  is fixed at `r sprintf("%.0f", SIGMA)`. Fitting it as well as $\delta$
  on ten destination margins would be asking too much of the data.
- **Ten schools.** Every statement here rests on ten points, and two
  tight pairs — Stringer with Varndean, Blatchington Mill with Hove Park
  — dominate the variation.
- **It is entangled with $W_j$ and with $\gamma$.** The sum uses the
  attractiveness that was itself balanced with this term in place, and
  §6's catchment term can do much of the same work.
- **Straight lines, not journeys.** Two schools half a kilometre apart
  across a railway are as close, on this measure, as two half a kilometre
  apart on the same street.
- **Fitted on offers, not preferences.** Unlike $\gamma_h$, which was
  refitted against the catchment preference table, $\delta$ was fitted
  against the offers made — and those are shaped by the ceiling as much
  as by choice.

## Questions this raises

1. Should $\delta$ be refitted against preferences, as $\gamma_h$ now is?
   That is the cleaner question, and this is the last term still fitted
   against the older target.
2. Is the positive sign real, or is it standing in for something the
   model is missing — the faith schools' city-wide draw, say, which
   happens to sit in the crowded middle of the city?
3. Would a nested form — choose an area, then a school within it —
   represent the paired catchments better than an exponent on clustering
   does?

# $A_i$: the balancing factor {#sec-ai}

```{r ai-setup}
#| include: false
r_ai <- run_sim(inp, year = 2026, rules = list(rule = "priorities"))
d_ai <- r_ai$flows
ext_ai <- inp$params$outside
g_ai <- unname(inp$params$gamma[d_ai$catchment]); g_ai[is.na(g_ai)] <- 0
d_ai$util <- d_ai$Wj * d_ai$cij^(-inp$params$beta) *
  exp(g_ai * d_ai$in_catch + inp$params$delta * log(d_ai$Cj))
is_ext <- d_ai$name %in% names(ext_ai$W)
d_ai$util[is_ext] <- unname(ext_ai$W[d_ai$name[is_ext]]) *
  d_ai$km[is_ext]^(-ext_ai$decay)

# The engine's own answer, to check the hand computation against.
own <- tapply(d_ai$util, d_ai$orig, sum)
d_ai$A <- 1 / own[d_ai$orig]
d_ai$by_hand <- d_ai$util * d_ai$Oi_pop * d_ai$A
ai_gap <- max(abs(d_ai$by_hand - d_ai$wanted))

pull_z <- d_ai %>% filter(pop == "either") %>%
  group_by(zone) %>%
  summarise(pull = sum(util), Oi = first(Oi_pop), catchment = first(catchment),
            ext_share = sum(util[name %in% names(ext_ai$W)]) / sum(util),
            .groups = "drop") %>%
  mutate(A = 1 / pull, index = 100 * pull / stats::median(pull))

row_chk <- d_ai %>% group_by(orig) %>%
  summarise(Oi = first(Oi_pop), wanted = sum(wanted), final = sum(flow), .groups = "drop")

ZC <- "E01016901|DS_Varndean|DS_Varndean"
ZE <- "E01017012|Longhill|Longhill"
worked <- function(zz) d_ai %>% filter(zone == zz, pop == "either")
ai_tbl <- function(zz) {
  worked(zz) %>%
    transmute(School = short_sch(name), `Journey (min)` = sprintf("%.0f", cij),
              `$W_j$` = ifelse(is.na(Wj), "—", sprintf("%.2f", Wj)),
              `In catchment` = ifelse(in_catch == 1, "yes", ""),
              Pull = signif(util, 3),
              `Share of the row` = sprintf("%.1f%%", 100 * util / sum(util)),
              Children = sprintf("%.1f", wanted)) %>%
    arrange(desc(Pull)) %>%
    mutate(Pull = format(Pull, scientific = TRUE, digits = 3)) %>%
    head(6)
}
ai_of <- function(zz) 1 / sum(worked(zz)$util)
oi_of <- function(zz) worked(zz)$Oi_pop[1]
```

## What it is

$A_i$ is not a thing anyone chooses or observes. It is whatever number
makes each neighbourhood's row of the matrix add up to the children who
live there:

$$\sum_j T_{ij}^{\,y} = O_i^{\,y} \quad\Longrightarrow\quad
A_i^{\,y} \;=\; \frac{1}{\sum_k W_k\, c_{ik}^{-\beta}\,
e^{\gamma_{h(i)}\mathbb{1}[\cdot]}\, C_k^{\delta}}$$

Call the denominator the neighbourhood's **total pull**: everything the
city offers it, discounted by how hard each school is to reach. $A_i$ is
its reciprocal, and the model then divides the children of $i$ between
schools in proportion to each school's share of that total:

$$T_{ij} \;=\; O_i \times \frac{\text{pull}_{ij}}{\sum_k \text{pull}_{ik}}$$

This is what "production-constrained" means, and it is the single
strongest assumption in the model. **Every child is placed somewhere.**
Demand cannot leak away, and it cannot grow because the city got better:
if every school became less attractive tomorrow, $A_i$ would simply rise
to compensate and exactly the same children would be distributed in
exactly the same proportions.

## Computed per population, not per zone

One subtlety: the model does not balance over zones but over
**origin-populations**. In the two paired catchments some families will
take one of the pair but not the other, so a zone there carries up to
three populations with different choice sets, each balanced on its own.

```{r ai-pops}
#| tbl-cap: "The origin-populations the model balances over. Each is a group of children with its own choice set, and therefore its own balancing factor."
d_ai %>% distinct(orig, pop, zone) %>%
  count(pop, name = "Origins") %>%
  mutate(Population = ifelse(pop == "either", "Everyone else", pop)) %>%
  select(Population, Origins) %>%
  knitr::kable(align = "lr")
```

That is `r fmt_n(nrow(row_chk))` origins over
`r fmt_n(n_distinct(d_ai$zone))` zones, and it matters because a child
who would not take Varndean must have Varndean *out of the denominator*,
not merely low in it. Otherwise refusing them at Dorothy Stringer would
send them next door.

## What the total pull looks like

```{r fig-ai-map}
#| fig-cap: "The denominator of the balancing factor: how much school, weighted by how reachable it is, each neighbourhood has available. Drawn on a log scale, because the range is more than eightyfold. Zone values are carried up to LSOAs, weighted by children."
#| fig-height: 5
pull_map <- lsoa_sf %>%
  inner_join(pull_z %>% inner_join(zone_lsoa %>% select(zone, lsoa), by = "zone") %>%
               group_by(lsoa) %>%
               summarise(index = weighted.mean(index, Oi), .groups = "drop"),
             by = c("lsoa21cd" = "lsoa"))
ggplot() +
  geom_sf(data = pull_map, aes(fill = index), colour = NA) +
  geom_sf(data = catch_sf, fill = NA, colour = "white", linewidth = 0.3) +
  geom_sf(data = sch_sf, shape = 21, fill = "white", colour = "black",
          size = 2.4, stroke = 0.9) +
  scale_fill_viridis_c(option = "viridis", trans = "log10",
                       name = "Total pull\n(median = 100)",
                       breaks = c(30, 100, 300, 1000)) +
  labs(title = "What each neighbourhood has within reach",
       subtitle = "The sum the balancing factor divides by, current catchment map, 2026") +
  theme_void(12) +
  theme(plot.title = element_text(face = "bold", colour = "#1f3b57"),
        plot.subtitle = element_text(colour = "grey35"))
```

The range is extreme — from
`r sprintf("%.0f", min(pull_z$index))` to
`r sprintf("%.0f", max(pull_z$index))` on an index with the median at 100
— and most of it is the catchment term compounding with distance. A
neighbourhood between Dorothy Stringer and Varndean has two schools it
can walk to *and* the catchment bonus on both. A neighbourhood in
Ovingdean has one school within easy reach, with the lowest $W_j$ in the
city.

For the far east of the city the denominator also contains schools
outside it. In a typical Longhill-catchment neighbourhood they are
`r sprintf("%.1f%%", 100 * weighted.mean(pull_z$ext_share[pull_z$catchment == "Longhill"], pull_z$Oi[pull_z$catchment == "Longhill"]))`
of the total pull, and they are under
`r sprintf("%.1f%%", 100 * max(pull_z$ext_share[!pull_z$catchment %in% c("Longhill", "BACA")]))`
everywhere west of Brighton Aldridge's.

## Two rows, worked

```{r ai-worked-central}
#| tbl-cap: "A neighbourhood between the two Fiveways schools. Top six destinations of fourteen; the rest are rounding."
ai_tbl(ZC) %>% knitr::kable(align = "lrrlrrr")
```

`r sprintf("%.0f", oi_of(ZC))` children, total pull
`r sprintf("%.3f", 1 / ai_of(ZC))`, so
$A_i = `r sprintf("%.2f", ai_of(ZC))`$.

```{r ai-worked-east}
#| tbl-cap: "A neighbourhood in Longhill's catchment. The same table, the same city, a different world."
ai_tbl(ZE) %>% knitr::kable(align = "lrrlrrr")
```

`r sprintf("%.0f", oi_of(ZE))` children, total pull
`r sprintf("%.4f", 1 / ai_of(ZE))`, so
$A_i = `r sprintf("%.1f", ai_of(ZE))`$ — about
`r sprintf("%.0f", ai_of(ZE) / ai_of(ZC))` times the first.

Read the two tables side by side and the model's behaviour is visible in
one page. In the first, `r sprintf("%.0f%%", 100 * sum(worked(ZC)$util[worked(ZC)$in_catch == 1]) / sum(worked(ZC)$util))`
of the row goes to the two catchment schools before anything else is
considered. In the second, the catchment school takes
`r sprintf("%.0f%%", 100 * sum(worked(ZE)$util[worked(ZE)$in_catch == 1]) / sum(worked(ZE)$util))`,
and the second destination is Priory School in Lewes, outside the city
altogether.

::: {.callout-note}
## The balancing factor is not the accessibility index, but it is close

The strategic view reports an accessibility index for each
neighbourhood. That is the same sum with the behavioural terms stripped
out — attractiveness and journey time only, no catchment bonus, no
competition term, city schools alone. The two rank neighbourhoods
similarly but not identically (Spearman
$\rho = `r sprintf("%.2f", cor(pull_z$pull, acc$zone$A_hansen[match(pull_z$zone, acc$zone$zone)], method = "spearman"))`$),
and the difference is mostly the catchment term: it rewards the
neighbourhoods that are already close to a wanted school.
:::

## Where it sits in the sequence

$A_i$ is applied to the **uncapped** pull, before any admission number is
consulted. The model first works out what families want, and only then
rations it (§9), re-offering the children who are refused (§10).

The constraint survives that. After the ceiling and the re-offering, the
row still sums to the children who live there — nobody is lost in
rationing, they are only placed elsewhere:

```{r ai-checks}
#| tbl-cap: "What the balancing factor should satisfy. Every number here is computed from the run, not asserted."
tibble::tribble(
  ~Check, ~Result,
  "Origins balanced",
  sprintf("%s populations over %s zones", fmt_n(nrow(row_chk)),
          fmt_n(n_distinct(d_ai$zone))),
  "Rows sum to their children, before the ceiling",
  sprintf("largest error %.0e children", max(abs(row_chk$wanted - row_chk$Oi))),
  "Rows still sum to their children, after it",
  sprintf("largest error %.2f children", max(abs(row_chk$final - row_chk$Oi))),
  "City total placed",
  sprintf("%s children in 2026, all of them", fmt_n(round(sum(d_ai$flow)))),
  "Recomputing the flows by hand reproduces the engine",
  sprintf("largest difference %.0e children", ai_gap),
  "Range of the balancing factor",
  sprintf("%.2f to %.1f", min(pull_z$A), max(pull_z$A)),
  "Out-of-city schools are inside the denominator",
  "yes: they compete for children like any other destination") %>%
  knitr::kable(align = "lr")
```

## What to be careful about

- **No demand response.** Because the row is fixed at $O_i$, nothing the
  model does can make families leave the state system, home-school, move
  house, or appear from elsewhere. A scenario that makes every school
  worse produces the same `r fmt_n(round(sum(d_ai$flow)))` placements as
  one that makes every school better. Where a scenario would plausibly change how many families opt
  out, the model will not show it, and §2's state-sector share is the
  only place that quantity lives.
- **Substitution is proportional.** Close a school, and the children who
  wanted it are redistributed among the others in proportion to those
  schools' existing pull for that neighbourhood — the classic
  independence-of-irrelevant-alternatives behaviour. The exclusivity
  populations are a patch on exactly this, for the one case where it is
  known to be wrong.
- **It absorbs everything the model does not contain.** Any systematic
  pull the equation misses is silently rescaled away by $A_i$ rather than
  showing up as a residual. A model that fits well at the school level
  can still be wrong about why.
- **It is recomputed in every run.** Change a year, a slider, a boundary
  or a school's existence and the denominator changes for every affected
  neighbourhood. $A_i$ is never carried over between runs.

## Questions this raises

1. Should the state-sector share respond to the scenario rather than
   being fixed in $O_i$ — a doubly-constrained or elastic form?
2. Should the choice set be truncated — some neighbourhoods carry
   fourteen destinations when families realistically consider three or
   four?
3. Are the exclusivity populations the right patch for proportional
   substitution, or should the whole thing be nested?