---
title: "Brighton & Hove secondary schools"
subtitle: "What the evidence says, how it was built, and what to do about Longhill"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    width: 1280
    height: 720
    margin: 0.06
    smaller: true
    scrollable: true
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
source(here::here("R", "00_core.R"))
knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

fmt_n   <- function(x, d = 0) formatC(round(x, d), big.mark = ",", format = "f", digits = d)
fmt_pct <- function(x, d = 0) paste0(formatC(x, format = "f", digits = d), "%")
gbp_m   <- function(x) sprintf("%s£%.1fm", ifelse(x < 0, "−", ""), abs(x) / 1e6)
gbp_k   <- function(x) sprintf("%s£%sk", ifelse(x < 0, "−", ""), fmt_n(abs(x) / 1e3))

inp <- readRDS(here::here("app", "data", "sim_inputs.rds"))
cop <- bh_data("council_options.rds")
whx <- bh_data("whitehawk_explained.rds")
acc <- bh_data("accessibility.rds")

co  <- function(id, col, y = 2030) cop$runs[[col]][cop$runs$id == id & cop$runs$year == y]
whe <- function(id, col, y = 2026)
  cop$whitehawk_elm[[col]][cop$whitehawk_elm$id == id & cop$whitehawk_elm$year == y]

city   <- inp$schools[inp$schools$city, ]
places <- sum(city$pan)
kids   <- function(y) inp$demand$cohort[inp$demand$year == y]
no30   <- acc$lsoa[acc$lsoa$places_30 < 1, ]

INK <- "#1f3b57"; MUTED <- "#6b7785"; ACCENT <- "#2a78d6"; WARM <- "#d95f02"; BAD <- "#b2182b"
theme_slide <- function(base = 16)
  ggplot2::theme_minimal(base) +
  ggplot2::theme(panel.grid.major.y = ggplot2::element_blank(),
                 panel.grid.minor = ggplot2::element_blank(),
                 plot.title = ggplot2::element_text(face = "bold", colour = INK),
                 plot.subtitle = ggplot2::element_text(colour = MUTED),
                 plot.title.position = "plot",
                 axis.text = ggplot2::element_text(colour = INK))
theme_map <- function(base = 14)
  ggplot2::theme_void(base) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", colour = INK),
                 plot.subtitle = ggplot2::element_text(colour = MUTED),
                 plot.title.position = "plot",
                 legend.position = "right", strip.text = ggplot2::element_text(face = "bold", size = base))

library(ggplot2)
library(patchwork)

oi    <- bh_data("open_inputs.rds")
lsoa  <- bh_data("lsoa.geojson") |> sf::st_transform(4326)
catch <- bh_data("catchments_current.geojson") |> sf::st_transform(4326)
sch   <- schools_sf(oi$schools) |> dplyr::filter(name != "Peacehaven Community School")
rc    <- bh_data("reception_cohort.rds")
sfin  <- bh_data("school_finance.rds")
brt   <- bh_data("brightopia.rds")
mt    <- bh_data("model_terms.rds")
fm    <- bh_data("flow_map.rds")
p6x   <- bh_data("priority6_sweep.rds")
elm_pt <- sf::st_as_sf(data.frame(e = oi$elm_grove$easting, n = oi$elm_grove$northing),
                       coords = c("e", "n"), crs = 27700) |> sf::st_transform(4326)
CITY_X <- c(-0.25, -0.01); CITY_Y <- c(50.80, 50.885)

# The simulator's own engine, run once for 2026, so the model slides show
# the model as it is.
source(here::here("app", "R", "model.R"))
source(here::here("app", "R", "outcomes.R"))
r26 <- run_sim(inp, year = 2026, rules = list(rule = "priorities"))
d26 <- r26$flows
ext <- inp$params$outside
g26 <- unname(inp$params$gamma[d26$catchment]); g26[is.na(g26)] <- 0
d26$util <- d26$Wj * d26$cij^(-inp$params$beta) *
  exp(g26 * d26$in_catch + inp$params$delta * log(d26$Cj))
is_ext <- d26$name %in% names(ext$W)
d26$util[is_ext] <- unname(ext$W[d26$name[is_ext]]) * d26$km[is_ext]^(-ext$decay)
catch_lab <- function(x) dplyr::coalesce(unname(CATCH_LABELS[CATCH_FROM_MODEL[x]]), x)
sch_city <- inp$schools[inp$schools$city & !(inp$schools$hypothetical %in% TRUE), ]
```

## Before anything else

::: {.callout-warning}
**Beta.** Every number in these slides is **modelled** from published data, not a record of what happened or a forecast to plan on. The model has not been fully validated.
:::

- Built from **published data only** --- no pupil-level records.
- The full working, every source and every caveat: [The Brighton Secondary School System: a strategic view](https://adamdennett.github.io/bh-school-system/)
- Every stage of the model, opened up for checking: [The model, stage by stage](https://adamdennett.github.io/bh-school-system/model-guide.html)
- Try any option yourself: [the policy simulator](https://adamdennett-bh-school-system.share.connect.posit.cloud/)

# The situation

## A system built for more children than it has

::: {.columns}
::: {.column width="55%"}
- The city publishes **`r fmt_n(places)`** Year 7 places across its ten secondary schools.
- The modelled cohort is **`r fmt_n(kids(2026))`** in 2026 and **`r fmt_n(kids(2035))`** by 2035 --- those children are already in primary school.
- By 2035 that is about **`r fmt_n(co("today", "empty", 2035))` empty Year 7 places a year**.
- Nine of the ten schools have been spending their reserves. The money is **a city problem, not only a Longhill problem**.
:::
::: {.column width="45%"}
**Longhill**

- Takes about **`r fmt_n(co("today", "lh_intake", 2026))`** children in 2026 and **`r fmt_n(co("today", "lh_intake"))`** in 2030, on 210 places.
- About **`r gbp_m(-co("today", "lh_gap"))` a year** short of break-even in 2030.
- Sits at the far edge of the city, at Ovingdean.
:::
:::

## Ten schools, six catchments, and where the children are

```{r fig-system-map}
#| fig-width: 11
#| fig-height: 6
kid_n <- oi$zones |> dplyr::group_by(lsoa) |> dplyr::summarise(children = sum(Oi), .groups = "drop")
kid_pts <- lsoa |> dplyr::inner_join(kid_n, by = c("lsoa21cd" = "lsoa")) |>
  sf::st_point_on_surface() |> suppressWarnings()
ggplot() +
  geom_sf(data = catch, aes(fill = catchment), alpha = 0.22, colour = "grey35", linewidth = 0.35) +
  geom_sf(data = kid_pts, aes(size = children), colour = "#1f4e79", alpha = 0.35) +
  geom_sf(data = sch, shape = 21, fill = "white", colour = "black", size = 3.4, stroke = 1.1) +
  ggrepel::geom_text_repel(data = sch, aes(lon, lat, label = short_sch(name)), size = 4,
                           colour = INK, seed = 2, box.padding = 0.5, min.segment.length = 0) +
  annotate("point", x = COMART$lon, y = COMART$lat, colour = BAD, size = 3, shape = 4, stroke = 1.6) +
  annotate("text", x = COMART$lon, y = COMART$lat - 0.004, label = "CoMArt (closed 2005)",
           colour = BAD, size = 3.4) +
  scale_fill_manual(values = CATCH_COLOURS, labels = CATCH_LABELS, name = "Catchment") +
  scale_size_area(max_size = 7, name = "Children per\nneighbourhood") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = "Most children live in the centre and west; Longhill sits at the far eastern edge",
       subtitle = "Two catchments hold a pair of schools; four hold one. King's and Cardinal Newman admit city-wide.") +
  theme_map()
```

## The cohort is already falling

```{r fig-cohort}
#| fig-width: 11
#| fig-height: 5.4
y7 <- rc$secondary_city; recp <- rc$reception_city; proj <- rc$projection
ggplot() +
  geom_ribbon(data = proj, aes(entry_year, ymin = lo, ymax = hi), fill = BAD, alpha = 0.18) +
  geom_line(data = recp, aes(year, rec_offers), colour = "#2c7fb8", linewidth = 1.1) +
  geom_line(data = y7, aes(year, y7_offers), colour = INK, linewidth = 1.3) +
  geom_line(data = proj, aes(entry_year, central), colour = BAD, linewidth = 1.2, linetype = "22") +
  annotate("text", x = min(recp$year) + 0.3, y = max(recp$rec_offers) * 1.03, label = "Reception offers",
           colour = "#2c7fb8", hjust = 0, size = 4.4) +
  annotate("text", x = min(y7$year) + 0.3, y = min(y7$y7_offers) * 0.93, label = "Year 7 offers",
           colour = INK, hjust = 0, size = 4.4, fontface = "bold") +
  annotate("text", x = max(proj$entry_year) - 0.2, y = max(proj$hi) * 1.06,
           label = "Projected from children\nalready in primary school", colour = BAD, hjust = 1, size = 4) +
  scale_y_continuous(labels = scales::label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Offers",
       title = "Reception offers turned down first; Year 7 follows seven years later",
       subtitle = sprintf("Year 7 demand projected to about %s by %s, from children already counted",
                          fmt_n(proj$central[which.max(proj$entry_year)]), max(proj$entry_year))) +
  theme_slide(15)
```

## Who can reach a school within 30 minutes

```{r fig-access}
#| fig-width: 11.5
#| fig-height: 5.6
acc_sf <- lsoa |>
  dplyr::inner_join(acc$lsoa |> dplyr::select(lsoa, places_30, places_30_elm), by = c("lsoa21cd" = "lsoa")) |>
  tidyr::pivot_longer(c(places_30, places_30_elm), names_to = "site", values_to = "places") |>
  dplyr::mutate(site = factor(ifelse(site == "places_30", "Longhill at Ovingdean (today)", "Longhill at the top of Elm Grove"),
                              levels = c("Longhill at Ovingdean (today)", "Longhill at the top of Elm Grove")))
elm_lab <- sf::st_sf(site = factor("Longhill at the top of Elm Grove", levels = levels(acc_sf$site)),
                     geometry = sf::st_geometry(elm_pt))
lh_lab <- sch |> dplyr::filter(grepl("Longhill", name)) |>
  dplyr::mutate(site = factor("Longhill at Ovingdean (today)", levels = levels(acc_sf$site)))
ggplot() +
  geom_sf(data = acc_sf, aes(fill = places), colour = NA) +
  geom_sf(data = catch, fill = NA, colour = "white", linewidth = 0.25) +
  geom_sf(data = lh_lab, shape = 21, fill = "white", colour = "black", size = 3.4, stroke = 1.2) +
  geom_sf(data = elm_lab, shape = 21, fill = "white", colour = "black", size = 3.4, stroke = 1.2) +
  facet_wrap(~ site) +
  scale_fill_viridis_c(option = "viridis", direction = -1, labels = scales::label_comma(),
                       name = "School places\nwithin 30 min") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = sprintf("%s neighbourhoods can reach no school place within 30 minutes by walk and bus", nrow(no30)),
       subtitle = "Darker is more places in reach; yellow is few or none. The white dot is Longhill.") +
  theme_map(14)
```

## Poor access and child poverty in the same places

```{r fig-bivariate}
#| fig-width: 11
#| fig-height: 6
BIV <- c("1-1" = "#3b4994", "1-2" = "#8c62aa", "1-3" = "#be64ac",
         "2-1" = "#5698b9", "2-2" = "#a5add3", "2-3" = "#dfb0d6",
         "3-1" = "#5ac8c8", "3-2" = "#ace4e4", "3-3" = "#e8e8e8")
biv <- lsoa |> dplyr::inner_join(acc$bivariate |> dplyr::select(lsoa, biv_key), by = c("lsoa21cd" = "lsoa"))
main <- ggplot() +
  geom_sf(data = biv, aes(fill = biv_key), colour = "white", linewidth = 0.05) +
  geom_sf(data = catch, fill = NA, colour = "grey15", linewidth = 0.4) +
  geom_sf(data = sch, shape = 21, fill = "white", colour = "black", size = 2.4) +
  scale_fill_manual(values = BIV, guide = "none") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = "Dark blue: the least reachable school places and the most child poverty, together",
       subtitle = sprintf("Accessibility and child poverty correlate at %.2f across neighbourhoods", acc$acc_dep_spearman)) +
  theme_map(14)
key <- expand.grid(a = 1:3, d = 1:3) |>
  dplyr::mutate(k = paste0(a, "-", d)) |>
  ggplot(aes(d, a, fill = k)) + geom_tile(colour = "white") +
  scale_fill_manual(values = BIV, guide = "none") +
  labs(x = "Less deprived →", y = "More reachable →") + coord_fixed() +
  theme_minimal(10) + theme(axis.text = element_blank(), panel.grid = element_blank())
main + inset_element(key, left = 0.0, bottom = 0.02, right = 0.2, top = 0.36)
```

## Nine of the ten schools are spending their reserves

```{r fig-reserves}
#| fig-width: 12
#| fig-height: 5.6
fex <- sfin$exposure |> dplyr::mutate(short = short_sch(name))
ord <- fex |> dplyr::arrange(reserve_pct) |> dplyr::pull(name)
rp <- sfin$panel |> dplyr::mutate(school = factor(short_sch(name), short_sch(ord)), neg = pmin(reserve_pct, 0))
ggplot(rp, aes(year, reserve_pct)) +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +
  geom_ribbon(aes(ymin = neg, ymax = 0), fill = "#d03b3b", alpha = 0.3) +
  geom_line(linewidth = 0.9, colour = INK) + geom_point(size = 1.2, colour = INK) +
  facet_wrap(~ school, ncol = 5) +
  scale_x_continuous(breaks = sort(unique(rp$year)),
                     labels = function(x) paste0(substr(x, 3, 4), "/", substr(x + 1, 3, 4))) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = NULL, y = "Reserve as a share of income",
       title = "Revenue reserves, worst first; red shading is a school in deficit",
       subtitle = "DfE school income and expenditure returns") +
  theme_slide(12) + theme(axis.text.x = element_text(size = 7))
```

## What the evidence says

- **This is not mainly a school quality problem.** Most of what separates schools' headline results arrives with the children; the part a school can reach runs mostly through **attendance**.
- **Families choose on the headline score**, which largely describes the intake --- so choice is self-fulfilling.
- **Admissions criteria only bind at some schools.** Several schools offer a place to every first-preference applicant.
- **Physical access is unequal, in the worst direction.** `r nrow(no30)` neighbourhoods can reach **no** school place within 30 minutes by walking and bus, and access and child poverty go together.
- Children from the most deprived neighbourhoods travel **`r sprintf("%.1f", co("today", "dep_gap", 2026))` minutes further** each way than everyone else, under the map in force.

# How the model works

## From published data to options

```{mermaid}
%%| fig-width: 12
flowchart LR
  d1["<b>Children</b><br/>census, postcodes,<br/>council projection"]
  d2["<b>Journeys</b><br/>walk and bus,<br/>routed"]
  d3["<b>What families ask for</b><br/>factsheets, catchment<br/>preference table"]
  d4["<b>Places and rules</b><br/>admission numbers,<br/>council priorities"]
  m["<b>Demand</b><br/>children shared by distance,<br/>attractiveness, catchment,<br/>competition<br/>⚙ year · map · site · appeal"]
  c{"<b>Full?</b><br/>⚙ places"}
  p["<b>Priorities</b> decide<br/>who is admitted<br/>⚙ rules"]
  r["<b>Refused</b> re-offered:<br/>partner school,<br/>then 2nd preferences"]
  f["<b>Intakes</b> →<br/>journeys<br/>disadvantaged mix<br/>budgets"]
  s["<b>29 options<br/>× 3 years</b><br/>and the<br/>simulator"]
  d1 & d2 & d3 --> m --> c
  d4 --> c
  c -- yes --> p --> r --> c
  c -- no --> f --> s
  classDef data fill:#f3f5f8,stroke:#9aa7b4,color:#1f3b57;
  classDef ctl fill:#fdecd9,stroke:#d95f02,stroke-width:2px,color:#3d1c00;
  classDef run fill:#eef4fb,stroke:#1f3b57,color:#102a43;
  classDef out fill:#edf6ef,stroke:#2b7a4b,color:#102a43;
  class d1,d2,d3,d4 data;
  class m,c,p ctl;
  class r run;
  class f,s out;
```

Orange steps are the ones the simulator lets you change (⚙). Moving any control re-runs everything from the start --- about a twentieth of a second. [Every step, opened up.](https://adamdennett.github.io/bh-school-system/model-guide.html#sec-algorithm)

## A spatial interaction model

::: {.columns}
::: {.column width="48%"}
$$T_{ij} = A_i\, O_i\, W_j\, c_{ij}^{-\beta}\, e^{\gamma_h\, \text{in catchment}}\, C_j^{\delta}$$

- $T_{ij}$ --- children from neighbourhood $i$ choosing school $j$
- $O_i$ --- children living in $i$
- $W_j$ --- how attractive school $j$ is
- $c_{ij}$ --- walk-and-bus minutes; $\beta$ how fast demand falls with them
- $\gamma_h$ --- how strongly each catchment follows its map
- $C_j$ --- rival schools nearby
- $A_i$ --- makes every neighbourhood place all its children
:::
::: {.column width="52%"}
```{mermaid}
flowchart TB
  N["Each neighbourhood's children"] --> Sh["Shared across schools by<br/>distance, attractiveness,<br/>catchment, competition"]
  Sh --> Cap{"School full?"}
  Cap -- "no" --> In["Admitted"]
  Cap -- "yes" --> Pr["Council priorities decide<br/>who is admitted"]
  Pr --> Re["The rest re-offered at their<br/>next choice with room"]
  Re --> Cap
```
:::
:::

## The map alone: every school equally attractive

```{r fig-brightopia}
#| fig-width: 11
#| fig-height: 5.6
bt <- brt$geog_only_at_ref |> dplyr::mutate(fills = modelled >= pan2026)
ggplot(bt, aes(pan2026, modelled)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = "31") +
  geom_point(aes(colour = fills), size = 4.5) +
  ggrepel::geom_text_repel(aes(label = short_sch(name)), size = 4.2, seed = 1, colour = INK) +
  scale_colour_manual(values = c(`TRUE` = "#2166ac", `FALSE` = BAD),
                      labels = c(`TRUE` = "Geography alone would fill it", `FALSE` = "It would not"), name = NULL) +
  scale_x_continuous(limits = c(0, NA)) + scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Admission number, 2026", y = "Children drawn on geography alone",
       title = "Distance alone explains much of who goes where",
       subtitle = "Every school identical, only journey time differs. Dashed line: exactly full.") +
  theme_slide(15) + theme(legend.position = "top")
```

## Adding one thing at a time

```{r fig-ladder}
#| fig-width: 12
#| fig-height: 5.4
lad <- mt$ladder |> dplyr::mutate(model = factor(model, model))
LAD_LAB <- c(M0 = "Distance\nonly", M1 = "+ what families\nask for", M2 = "+ school\ncapacity",
             M3 = "+ one\ncatchment term", M4 = "+ competition", M5 = "Fitted to each\ncatchment")
p1 <- ggplot(lad, aes(model, mae_free)) +
  geom_col(fill = INK, width = 0.62) +
  geom_text(aes(label = round(mae_free)), vjust = -0.4, size = 4.6, colour = INK) +
  scale_x_discrete(labels = LAD_LAB) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Average error, children per school",
       subtitle = "Error on the schools not rationing places") + theme_slide(13)
gam <- tibble::tibble(catchment = names(mt$calibrated$gamma), gamma = unname(mt$calibrated$gamma)) |>
  dplyr::mutate(label = catch_lab(catchment), label = forcats::fct_reorder(label, gamma))
p2 <- ggplot(gam, aes(gamma, label)) +
  geom_col(fill = ACCENT, width = 0.62) +
  geom_text(aes(label = sprintf("%.1f", gamma)), hjust = -0.25, size = 4.6, colour = INK) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Fitted catchment term", y = NULL,
       subtitle = "How strongly each catchment follows its map") + theme_slide(13)
(p1 | p2) + plot_annotation(title = "Capacity does most of the work; families follow their catchment strongly",
                            theme = theme_slide(15))
```

## Some schools have company; Longhill has none

```{r fig-cj}
#| fig-width: 11.5
#| fig-height: 5.8
W_run <- setNames(sch_city$W, sch_city$name)
xy <- sch_city |> dplyr::select(name, easting, northing)
pairs_km <- tidyr::expand_grid(name = xy$name, k = xy$name) |>
  dplyr::filter(name != k) |>
  dplyr::left_join(xy, by = "name") |>
  dplyr::left_join(xy |> dplyr::rename(k = name, ke = easting, kn = northing), by = "k") |>
  dplyr::mutate(km = pmax(sqrt((easting - ke)^2 + (northing - kn)^2) / 1000, 0.1),
                contrib = unname(W_run[k]) * km^-inp$params$sigma,
                short = short_sch(name))
cn <- compete_now(sch_city, W_run, inp$params$sigma)
cj <- tibble::tibble(name = names(cn), C = unname(cn)) |> dplyr::mutate(index = 100 * C / mean(C))
pts <- sch_city |> dplyr::select(name, easting, northing) |>
  sf::st_as_sf(coords = c("easting", "northing"), crs = 27700) |> sf::st_transform(4326)
xy_ll <- sf::st_coordinates(pts)
pts <- pts |> dplyr::mutate(short = short_sch(name), lon = xy_ll[, 1], lat = xy_ll[, 2]) |>
  dplyr::left_join(cj, by = "name")
ll <- sf::st_drop_geometry(pts) |> dplyr::select(name, lon, lat)
segs <- pairs_km |> dplyr::filter(short %in% c("Dorothy Stringer", "Longhill")) |>
  dplyr::left_join(ll, by = "name") |>
  dplyr::left_join(ll |> dplyr::rename(k = name, klon = lon, klat = lat), by = "k")
ggplot() +
  geom_sf(data = catch, fill = "grey95", colour = "white", linewidth = 0.5) +
  geom_segment(data = segs, aes(x = lon, y = lat, xend = klon, yend = klat,
                                linewidth = contrib, colour = short), alpha = 0.5) +
  geom_sf(data = pts, aes(size = C), shape = 21, fill = "white", colour = "black", stroke = 1) +
  ggrepel::geom_text_repel(data = pts, aes(lon, lat, label = sprintf("%s  %.0f", short, index)),
                           size = 4, colour = INK, seed = 11, box.padding = 0.5,
                           segment.colour = "grey60", max.overlaps = Inf) +
  scale_linewidth_continuous(range = c(0.3, 4), guide = "none") +
  scale_size_continuous(range = c(2.5, 11), guide = "none") +
  scale_colour_manual(values = c(`Dorothy Stringer` = WARM, Longhill = INK), name = NULL,
                      labels = function(x) paste("pull of each rival on", x)) +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = sprintf("Stringer and Varndean are %.0f metres apart; Longhill's nearest rival is %.1f km away",
                       1000 * min(pairs_km$km), min(pairs_km$km[pairs_km$short == "Longhill"])),
       subtitle = "Circle size and number: how much company each school has (city average = 100). Company helps a school slightly.") +
  theme_map(14) + theme(legend.position = "top")
```

## What each neighbourhood has within reach

```{r fig-pull}
#| fig-width: 11.5
#| fig-height: 5.8
pull_z <- d26 |> dplyr::filter(pop == "either") |>
  dplyr::group_by(zone) |>
  dplyr::summarise(pull = sum(util), Oi = dplyr::first(Oi_pop), .groups = "drop") |>
  dplyr::mutate(index = 100 * pull / stats::median(pull))
pull_map <- lsoa |>
  dplyr::inner_join(pull_z |>
                      dplyr::inner_join(oi$zones |> dplyr::select(zone, lsoa), by = "zone") |>
                      dplyr::group_by(lsoa) |>
                      dplyr::summarise(index = stats::weighted.mean(index, Oi), .groups = "drop"),
                    by = c("lsoa21cd" = "lsoa"))
ggplot() +
  geom_sf(data = pull_map, aes(fill = index), colour = NA) +
  geom_sf(data = catch, fill = NA, colour = "white", linewidth = 0.35) +
  geom_sf(data = sch, shape = 21, fill = "white", colour = "black", size = 2.8, stroke = 1) +
  scale_fill_viridis_c(trans = "log10", breaks = c(30, 100, 300, 1000),
                       name = "Pull within reach\n(median = 100)") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = "Between Stringer and Varndean families have two schools and a catchment bonus on both",
       subtitle = "The model shares each neighbourhood's children in proportion to this total. Log scale: the range is more than a hundredfold.") +
  theme_map(14)
```

## Longhill's modelled catchment, as the model gets more realistic

```{r fig-flows}
#| fig-width: 12.5
#| fig-height: 5
FM_LABELS <- c(M0 = "Distance only", M1 = "With what families ask for", M5 = "Full model")
FM_COL <- c(Bus = "#2166ac", Walk = WARM)
fl <- fm$net |> dplyr::filter(name == "Longhill High School") |>
  dplyr::mutate(model = factor(model, names(FM_LABELS), FM_LABELS))
fl_bb <- sf::st_bbox(fl)
fm_sch <- schools_sf() |> dplyr::filter(name %in% unique(fm$net$name))
ggplot() +
  geom_sf(data = catch, fill = NA, colour = "grey80", linewidth = 0.3) +
  geom_sf(data = fl |> dplyr::arrange(flow), aes(linewidth = flow, colour = leg_mode),
          alpha = 0.75, lineend = "round") +
  geom_sf(data = fm_sch, colour = "grey45", size = 1.3) +
  geom_sf(data = fm_sch |> dplyr::filter(name == "Longhill High School"),
          colour = "black", fill = "white", shape = 21, size = 3.2, stroke = 1.1) +
  facet_wrap(~ model, nrow = 1) +
  scale_linewidth_continuous(range = c(0.15, 2.8), guide = "none") +
  scale_colour_manual(values = FM_COL, name = NULL) +
  coord_sf(xlim = c(fl_bb["xmin"], fl_bb["xmax"]), ylim = c(fl_bb["ymin"], fl_bb["ymax"])) +
  labs(title = "Routed over the real network, Longhill's draw shrinks back towards its own catchment",
       subtitle = "Line width is modelled children on each street or bus segment. White dot: Longhill.") +
  theme_map(13) + theme(legend.position = "top")
```

## When a school is full

```{r fig-ceiling}
#| fig-width: 12
#| fig-height: 5.8
prefs26 <- bh_data("factsheet_panel.rds")$factsheets |>
  dplyr::filter(year == 2026, name != "Total") |>
  dplyr::transmute(name, pref1, pref12 = pref1 + pref2)
by_sch <- d26 |>
  dplyr::group_by(name) |>
  dplyr::summarise(wanted = sum(wanted), offered = sum(flow), .groups = "drop") |>
  dplyr::inner_join(sch_city |> dplyr::select(name, pan), by = "name") |>
  dplyr::left_join(prefs26, by = "name") |>
  dplyr::mutate(short = forcats::fct_reorder(short_sch(name), pan))
cl <- by_sch |>
  dplyr::select(short, `Model: first choice` = wanted, `Named first` = pref1,
                `Named first or second` = pref12, `Offered, model` = offered) |>
  tidyr::pivot_longer(-short, names_to = "what", values_to = "n") |>
  dplyr::mutate(what = factor(what, c("Model: first choice", "Named first",
                                      "Named first or second", "Offered, model")))
ggplot(cl, aes(n, short)) +
  geom_segment(data = by_sch, aes(x = 0, xend = pan, y = short, yend = short),
               colour = "grey86", linewidth = 10) +
  geom_segment(data = by_sch, aes(x = pref1, xend = pref12, y = short, yend = short),
               colour = WARM, alpha = 0.35, linewidth = 1.6) +
  geom_point(aes(colour = what, shape = what), size = 4.2) +
  geom_text(data = by_sch, aes(x = 0, label = sprintf("%d places ", pan)),
            hjust = 1, size = 3.8, colour = MUTED) +
  scale_colour_manual(values = c(WARM, "#9e6a3b", WARM, INK), name = NULL) +
  scale_shape_manual(values = c(17, 1, 16, 15), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.12, 0.03))) +
  labs(x = "Children or families, 2026 (named: council factsheet)", y = NULL,
       title = sprintf("Counting second choices, %d of the ten schools are asked for by more families than they hold",
                       sum(by_sch$pref12 > by_sch$pan)),
       subtitle = with(by_sch[by_sch$short == "Dorothy Stringer", ],
                       sprintf("Stringer: %d named it first, %d more second. Families in its catchment are happy with either school, and the model treats them so.",
                               pref1, pref12 - pref1))) +
  theme_slide(14) + theme(legend.position = "top")
```

## Where the refused children go

```{r fig-overflow}
#| fig-width: 12
#| fig-height: 5.6
# The engine's own cascade, re-run with bookkeeping (model guide, section 10).
rl <- inp$rules
kern <- inp$params$overflow
cap_ov <- setNames(as.numeric(inp$schools$pan), inp$schools$name)
cap_all <- c(cap_ov[!is.na(cap_ov)], setNames(rep(1e6, length(ext$W)), names(ext$W)))
dsg <- inp$designs[["Current catchments"]]
o_pop <- tapply(d26$Oi_pop, d26$orig, function(x) x[1])
fsm_z <- setNames(inp$zones$fsm, inp$zones$zone)
share_o <- setNames(unname(fsm_z[sub("#.*$", "", names(o_pop))]), names(o_pop))
grp <- dsg$school[inp$schools$name[inp$schools$city & !inp$schools$faith]]
single <- names(table(grp[!is.na(grp)]))[table(grp[!is.na(grp)]) == 1]
in_c <- d26$in_catch == 1
s6 <- unname(dsg$zone[d26$zone]) %in% single & !in_c
com <- d26$name %in% rl$community
st <- overflow_setup(d26$wanted, d26$orig, d26$name, d26$catchment, d26$pop, kern,
                     partner_map(inp$params$exclusive))
attribute <- function(refused, room) {
  to_p <- st$pr; to_p[st$pr] <- room[st$p_idx[st$pr]]
  part <- tibble::tibble(from = d26$name[to_p], to = d26$name[st$p_idx[to_p]],
                         n = refused[to_p], route = "The other school of the pair")
  rest <- refused; rest[to_p] <- 0
  w <- st$bw * room; ws <- rowsum(w, st$oid)[st$oid, 1]
  w2 <- st$u * room; ws2 <- rowsum(w2, st$oid)[st$oid, 1]
  sh <- ifelse(ws > 0, w / ws, ifelse(ws2 > 0, w2 / ws2, 0))
  kern_part <- tibble::tibble(oid = st$oid, from = d26$name, n = rest) |> dplyr::filter(n > 0) |>
    dplyr::inner_join(tibble::tibble(oid = st$oid, to = d26$name, s = sh) |> dplyr::filter(s > 0),
                      by = "oid", relationship = "many-to-many") |>
    dplyr::transmute(from, to, n = n * s, route = "Where their catchment's second preferences go")
  dplyr::bind_rows(part, kern_part)
}
ff <- d26$wanted * unname(share_o[d26$orig]); fn <- d26$wanted - ff
moves <- list()
for (it in 1:500) {
  a <- tier_accept(ff, fn, d26$name, cap_all, cap_all, in_c, s6, com, rl$p6_share, rl$fsm_cap_share)
  Hf <- ff * a$f; Hn <- fn * a$n; rf <- ff - Hf; rn <- fn - Hn
  if (sum(rf + rn) <= 1e-6) break
  room <- overflow_room(Hf + Hn, d26$name, cap_all)
  if (!any(room)) break
  moves[[it]] <- attribute(rf + rn, room)
  ff <- Hf + overflow_add(rf, room, st); fn <- Hn + overflow_add(rn, room, st)
}
mv_all <- dplyr::bind_rows(moves)
mv <- mv_all |>
  dplyr::mutate(row = ifelse(route == "The other school of the pair",
                             paste(short_sch(from), "→ its pair"),
                             paste(short_sch(from), "→ second preferences"))) |>
  dplyr::group_by(row, to = short_sch(to)) |>
  dplyr::summarise(n = sum(n), .groups = "drop") |>
  dplyr::filter(n > 0.5) |>
  dplyr::mutate(row = forcats::fct_reorder(row, n, sum), to = forcats::fct_reorder(to, n, sum))
ggplot(mv, aes(n, row, fill = to)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.4) +
  scale_fill_brewer(palette = "Set2", name = "Offered instead") +
  labs(x = "Children, 2026", y = "Refused at",
       title = sprintf("%.0f children are re-offered somewhere else, and the schools with room take them",
                       sum(mv_all$n)),
       subtitle = "First the other school of a pair, then wherever their catchment's second preferences point") +
  theme_slide(15)
```

## Faith schools draw from the whole city

```{r fig-faith}
#| fig-width: 11.5
#| fig-height: 5.2
faith_n <- sch_city$name[sch_city$faith]
mix <- d26 |>
  dplyr::filter(name %in% c(faith_n, "Varndean School", "Longhill High School")) |>
  dplyr::group_by(name, catchment) |> dplyr::summarise(n = sum(flow), .groups = "drop") |>
  dplyr::group_by(name) |> dplyr::mutate(share = n / sum(n)) |> dplyr::ungroup() |>
  dplyr::mutate(school = factor(short_sch(name), c("Cardinal Newman", "King's", "Varndean", "Longhill")),
                Catchment = catch_lab(catchment))
ggplot(mix, aes(share, forcats::fct_rev(school), fill = Catchment)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.4) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_fill_brewer(palette = "Set2", name = "Where the children live") +
  labs(x = NULL, y = NULL,
       title = "No catchment: their pull rests on attractiveness and distance alone",
       subtitle = "Modelled intake by home catchment, 2026. Religion is in no published data, so it sits inside attractiveness.") +
  theme_slide(15)
```

# Does the model reproduce what we know?

## It reproduces the 2026 offers, which it was not fitted to

```{r fig-valid-offers}
#| fig-width: 11
#| fig-height: 5.8
vo <- mt$runs$M5 |> dplyr::mutate(short = short_sch(name))
r2 <- cor(vo$observed, vo$modelled)^2
ggplot(vo, aes(observed, modelled)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "31") +
  geom_point(size = 4.5, colour = INK) +
  ggrepel::geom_text_repel(aes(label = short), size = 4.3, seed = 3, colour = INK, box.padding = 0.6) +
  scale_x_continuous(limits = c(0, 380)) + scale_y_continuous(limits = c(0, 380)) +
  labs(x = "Places offered, September 2026 (council factsheet)", y = "Modelled intake, 2026",
       title = sprintf("Modelled against actual offers: R² = %.2f", r2),
       subtitle = str_wrap(paste("Fitted to catchment preferences, not to offers. The four schools that filled come out exactly full;",
                                 "the two academies also take children placed there without naming them."), 110)) +
  theme_slide(14)
```

## It reproduces where each catchment's families want to go

```{r fig-valid-cells}
#| fig-width: 11.5
#| fig-height: 5.8
cells <- mt$calibrated$cells |>
  tidyr::pivot_longer(c(m4_share, m5_share), names_to = "model", values_to = "modelled") |>
  dplyr::mutate(model = ifelse(model == "m4_share", "One catchment term for the whole city", "Fitted to each catchment"))
r2c <- cells |> dplyr::group_by(model) |> dplyr::summarise(r2 = cor(obs_share, modelled)^2, .groups = "drop")
cells <- cells |> dplyr::left_join(r2c, by = "model") |>
  dplyr::mutate(model = sprintf("%s (R² = %.2f)", model, r2),
                model = factor(model, sort(unique(model), decreasing = TRUE)))
ggplot(cells, aes(obs_share, modelled)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "31") +
  geom_point(aes(colour = model), size = 3.2, alpha = 0.85, show.legend = FALSE) +
  facet_wrap(~ model) +
  scale_colour_manual(values = c(MUTED, ACCENT)) +
  scale_x_continuous(labels = scales::label_percent()) + scale_y_continuous(labels = scales::label_percent()) +
  coord_equal() +
  labs(x = "Share of the catchment's first preferences, published", y = "Modelled share",
       title = "Each dot is one catchment and one school: 60 combinations",
       subtitle = "Council's catchment-level preference table, three rounds pooled") +
  theme_slide(14)
```

## It reproduces who the council admitted, and why

```{r fig-tiers}
#| fig-width: 12
#| fig-height: 5.6
TIERS <- c("Free school meals (4, 5)", "Single-school catchment (6)",
           "In catchment, incl. siblings (3, 7)", "Everyone else (8)")
t_obs <- rl$outturn_2026 |>
  dplyr::mutate(tier = dplyr::case_when(priority %in% 4:5 ~ TIERS[1], priority == 6 ~ TIERS[2],
                                        priority %in% c(3, 7) ~ TIERS[3], priority == 8 ~ TIERS[4])) |>
  dplyr::filter(!is.na(tier)) |>
  dplyr::group_by(School = short_sch(school), tier) |>
  dplyr::summarise(Council = sum(offers), .groups = "drop")
t_mod <- r26$tiers |>
  dplyr::filter(name %in% rl$rationing) |>
  dplyr::transmute(School = short_sch(name), p45 = p45_in + p45_out, p6, p7, p8) |>
  tidyr::pivot_longer(-School, names_to = "k", values_to = "Model") |>
  dplyr::mutate(tier = TIERS[match(k, c("p45", "p6", "p7", "p8"))])
t_obs |>
  dplyr::inner_join(t_mod, by = c("School", "tier")) |>
  tidyr::pivot_longer(c(Council, Model), names_to = "src", values_to = "n") |>
  dplyr::mutate(tier = factor(tier, rev(TIERS))) |>
  ggplot(aes(n, tier, fill = src)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ School, nrow = 1) +
  scale_fill_manual(values = c(Council = INK, Model = WARM), name = NULL) +
  labs(x = "Offers, September 2026", y = NULL,
       title = "Close on catchment and priority 6; too generous on free school meals at Varndean",
       subtitle = "The three schools that rationed places. The model cannot see siblings, who are admitted before free-school-meal children.") +
  theme_slide(13) + theme(legend.position = "top", strip.text = element_text(face = "bold", size = 14))
```

## Counting disadvantage by neighbourhood got the schools in the wrong order

```{r fig-neigh}
#| fig-width: 11.5
#| fig-height: 5.6
inp$params$dis$fit |>
  dplyr::transmute(short = short_sch(name), Neighbourhood = neighbourhood, School = target) |>
  tidyr::pivot_longer(-short, names_to = "what", values_to = "share") |>
  dplyr::mutate(short = forcats::fct_reorder(short, share, max),
                what = factor(what, c("Neighbourhood", "School"),
                              c("By the neighbourhoods its children come from",
                                "What the school actually has (Year 7 offers or DfE)"))) |>
  ggplot(aes(share, short)) +
  geom_line(aes(group = short), colour = "grey75", linewidth = 1.1) +
  geom_point(aes(colour = what), size = 4.2) +
  scale_colour_manual(values = c(WARM, INK), name = NULL) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Disadvantaged share of the intake", y = NULL,
       title = "In a mixed neighbourhood, better-off and poorer families choose different schools",
       subtitle = "So the model now places disadvantaged children school by school, fitted to what each school actually admits") +
  theme_slide(15) + theme(legend.position = "top", legend.direction = "vertical",
                          legend.justification = "left")
```

# A system that moves

## The same system, year by year

```{r dyn-runs}
#| include: false
city_s <- inp$schools |> dplyr::filter(city)
YRS <- 2026:2035
p_base <- setNames(city_s$pan, city_s$name)
run_year <- function(y, pans = p_base)
  run_sim(inp, pans = pans, year = y, rules = list(rule = "priorities"))
runs_y <- lapply(YRS, run_year)
dyn <- purrr::map_dfr(seq_along(YRS), function(k)
  runs_y[[k]]$schools |> dplyr::filter(city) |> dplyr::transmute(year = YRS[k], name, intake, pan))
lh_of <- function(r) r$schools$intake[r$schools$name == "Longhill High School"]
lever <- purrr::map_dfr(seq_along(YRS), function(k) {
  y <- YRS[k]; b <- lh_of(runs_y[[k]])
  pc <- p_base; pc[c("Dorothy Stringer School", "Varndean School", "Blatchington Mill School")] <- c(300, 270, 300)
  pf <- p_base; pf[c("Cardinal Newman Catholic School", "King's School")] <- c(270, 150)
  tibble::tibble(year = y,
                 `Stringer, Varndean and Blatchington Mill trimmed` = lh_of(run_year(y, pc)) - b,
                 `Cardinal Newman 270 and King's 150` = lh_of(run_year(y, pf)) - b)
})
empty_by_year <- dyn |> dplyr::group_by(year) |> dplyr::summarise(empty = sum(pmax(pan - intake, 0)), .groups = "drop")
money_y <- purrr::map_dfr(seq_along(YRS), function(k)
  outcomes(inp, runs_y[[k]])$by_school |> dplyr::transmute(year = YRS[k], short, gap, gap_pct))
```

```{r fig-dynamic}
#| fig-width: 12
#| fig-height: 5.8
dd <- dyn |> dplyr::mutate(short = short_sch(name), lh = grepl("Longhill", name))
ggplot(dd, aes(year, intake, group = short)) +
  geom_line(aes(colour = lh, linewidth = lh)) +
  ggrepel::geom_text_repel(data = dd |> dplyr::filter(year == max(YRS)),
                           aes(label = short, colour = lh), hjust = 0, nudge_x = 0.3, direction = "y",
                           size = 4, segment.colour = "grey80", show.legend = FALSE) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = BAD), guide = "none") +
  scale_linewidth_manual(values = c(`FALSE` = 0.8, `TRUE` = 1.8), guide = "none") +
  scale_x_continuous(breaks = YRS, limits = c(2026, 2037.5)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = NULL, y = "Modelled Year 7 intake",
       title = "As the cohort falls, the full schools stay full and the others empty",
       subtitle = sprintf("Empty Year 7 places across the city rise from %s in 2026 to %s in 2035. Council priorities, today's map.",
                          fmt_n(empty_by_year$empty[empty_by_year$year == 2026]),
                          fmt_n(empty_by_year$empty[empty_by_year$year == 2035]))) +
  theme_slide(15)
```

## The same lever works in one year and not the next

```{r fig-lever-years}
#| fig-width: 11
#| fig-height: 5.6
lv <- lever |> tidyr::pivot_longer(-year, names_to = "lever", values_to = "gain")
LV_COL <- c("Stringer, Varndean and Blatchington Mill trimmed" = WARM,
            "Cardinal Newman 270 and King's 150" = INK)
ggplot(lv, aes(year, gain, colour = lever)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line(linewidth = 1.6) + geom_point(size = 2.6) +
  geom_text(data = lv |> dplyr::filter(year == 2030),
            aes(label = lever), size = 4.6, vjust = -1, hjust = 0, show.legend = FALSE) +
  scale_colour_manual(values = LV_COL, guide = "none") +
  scale_x_continuous(breaks = YRS) +
  labs(x = NULL, y = "Extra children at Longhill",
       title = "Trimming the central schools helps Longhill only while they are full",
       subtitle = "Once the cohort falls, they have room and stop overflowing east. Fewer faith-school places keeps adding a few.") +
  theme_slide(15)
```

## What the falling cohort does to budgets

```{r fig-budgets}
#| fig-width: 12
#| fig-height: 5.6
mb <- money_y |>
  dplyr::filter(year %in% c(2026, 2035)) |>
  dplyr::mutate(year = factor(year), short = forcats::fct_reorder(short, gap_pct, min))
n_def <- function(y) sum(money_y$gap[money_y$year == y] < 0)
ggplot(mb, aes(gap_pct, short)) +
  geom_vline(xintercept = 0, colour = "grey45") +
  geom_line(aes(group = short), colour = "grey78", linewidth = 1.4) +
  geom_point(aes(colour = year), size = 4.2) +
  scale_colour_manual(values = c(`2026` = INK, `2035` = WARM), name = "Intake year") +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Annual surplus or deficit, share of income", y = NULL,
       title = sprintf("%d schools in deficit on the 2026 intake, %d on the 2035 one", n_def(2026), n_def(2035)),
       subtitle = "Roll at five times the intake; 75% of lost funding cut from costs. One dot: the school fills in both years.") +
  theme_slide(15) + theme(legend.position = "top")
```

# Complex interactions, surprising results

## Pull one lever, and the effect turns up somewhere else

```{r fig-surprises}
#| fig-width: 12.5
#| fig-height: 6.2
mini <- function(df, title, sub, money = FALSE) {
  df <- df |> dplyr::mutate(stage = factor(stage, stage),
                            lab = if (money) gbp_m(value) else fmt_n(value))
  ggplot(df, aes(stage, value, fill = hi)) +
    geom_col(width = 0.6) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_text(aes(label = lab, vjust = ifelse(value < 0, 1.5, -0.5)), size = 5, colour = INK) +
    scale_fill_manual(values = c(`FALSE` = "#b9c2cc", `TRUE` = ACCENT), guide = "none") +
    scale_y_continuous(expand = expansion(mult = 0.22),
                       labels = if (money) function(x) gbp_m(x) else waiver()) +
    labs(x = NULL, y = NULL, title = title, subtitle = sub) +
    theme_slide(13) + theme(plot.title = element_text(size = 14.5),
                            plot.subtitle = element_text(size = 11.5))
}
s1 <- mini(tibble::tibble(stage = c("210 places", "150 places"),
                          value = c(co("today", "lh_intake"), co("lh150", "lh_intake")), hi = c(FALSE, TRUE)),
           "Cut Longhill's places: its intake does not move",
           "Longhill's intake, 2030. It was never full.")
s2 <- mini(tibble::tibble(stage = c("Today", "Only the council's\nschools cut", "Every school cut"),
                          value = c(co("today", "comm_gap", 2035), co("shrink", "comm_gap", 2035),
                                    co("shrink_all", "comm_gap", 2035)),
                          hi = c(FALSE, TRUE, TRUE)),
           "Cut only the council's schools: their budgets get worse",
           "Community schools' combined balance, 2035. Children move to the others.", money = TRUE)
s3 <- mini(tibble::tibble(stage = c("Today", "Central schools\ntrimmed"),
                          value = c(co("today", "displaced"), co("central", "displaced")), hi = c(FALSE, TRUE)),
           "Trim the central schools: twice as many displaced",
           sprintf("Children refused their first choice, 2030. Longhill gains %.1f.",
                   co("central", "lh_intake") - co("today", "lh_intake")))
s4 <- mini(tibble::tibble(stage = c("Council\npriorities", "Equal chance\nfor all"),
                          value = c(co("today", "displaced", 2026), co("equal", "displaced", 2026)),
                          hi = c(FALSE, TRUE)),
           sprintf("Drop the priorities: %.0f times as many displaced",
                   co("equal", "displaced", 2026) / co("today", "displaced", 2026)),
           "Children refused their first choice, 2026. Catchment families lose their edge.")
(s1 | s2) / (s3 | s4)
```

## Open admissions barely changes segregation

```{r fig-p6}
#| fig-width: 12
#| fig-height: 5.6
p6l <- p6x$sweep |> dplyr::filter(fsm, p6 <= 25) |> dplyr::mutate(yr = factor(year))
g5 <- p6l$gorard[p6l$year == 2026 & p6l$p6 == 5]
pa <- ggplot(p6l, aes(p6, gorard, colour = yr)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = g5 - 0.01, ymax = g5 + 0.01,
           fill = "grey80", alpha = 0.45) +
  annotate("text", x = 12.5, y = g5 - 0.0085, label = "within measurement noise",
           size = 4.2, colour = MUTED) +
  geom_vline(xintercept = c(5, 20), linetype = "22", colour = "grey55") +
  annotate("text", x = c(5, 20), y = 0.093, label = c("council's rule", "first proposal"),
           hjust = -0.08, size = 4, colour = MUTED) +
  geom_line(linewidth = 1.5) + geom_point(size = 2.6) +
  scale_colour_manual(values = c(`2026` = INK, `2030` = WARM), name = NULL) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  coord_cartesian(ylim = c(0.09, 0.14)) +
  labs(x = "Places open to single-school catchments (priority 6)", y = "Segregation index",
       subtitle = "More open places: the index hardly moves") +
  theme_slide(14) + theme(legend.position = "top")
wn <- p6x$winners |>
  dplyr::mutate(dir = ifelse(dep_p6 > dep_all, "Used by more disadvantaged families",
                             "Used by less disadvantaged families"),
                label = catch_lab(catchment)) |>
  tidyr::pivot_longer(c(dep_all, dep_p6), names_to = "who", values_to = "share")
pb <- ggplot(wn, aes(share, label)) +
  geom_line(aes(group = label, colour = dir), linewidth = 3, alpha = 0.5) +
  geom_point(aes(shape = who), size = 4.5, colour = INK) +
  scale_colour_manual(values = c("Used by more disadvantaged families" = ACCENT,
                                 "Used by less disadvantaged families" = WARM), guide = "none") +
  scale_shape_manual(values = c(dep_all = 1, dep_p6 = 16),
                     labels = c(dep_all = "Whole catchment",
                                dep_p6 = "Priority-6 takers"), name = NULL) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Share disadvantaged", y = NULL,
       subtitle = "Orange: taken by better-off families. Blue: by poorer ones.") +
  theme_slide(14) + theme(legend.position = "top", legend.justification = "left")
(pa | pb) + plot_layout(widths = c(1, 1.1)) + plot_annotation(
  title = "Opening places to single-school catchments neither mixes nor separates intakes",
  subtitle = "Who takes the places differs by catchment, and the effects cancel. An earlier version of the model said otherwise;\nthat finding did not survive a better measure of disadvantage.",
  theme = theme_slide(15))
```

## Measuring disadvantage better changed one answer

```{r fig-measures}
#| fig-width: 11.5
#| fig-height: 5.4
MEAS <- c("By neighbourhood" = "1  Counted by neighbourhood",
          "Whole-school shares, fitted on today's map" = "2  DfE whole-school shares, today's map",
          "Whole-school shares, fitted before the policy change" = "3  DfE whole-school shares, before the 2024 change",
          "Year 7 offers, 2026 (the model's measure)" = "4  Year 7 offers, 2026 (used now)")
dm <- whx$measures |>
  dplyr::filter(year == 2026) |>
  dplyr::select(measure, map, gorard) |>
  tidyr::pivot_wider(names_from = map, values_from = gorard) |>
  dplyr::mutate(delta = whitehawk - today,
                label = factor(unname(MEAS[measure]), rev(unname(MEAS))),
                dir = ifelse(delta > 0, "more segregated", "less segregated"))
ggplot(dm, aes(delta, label, fill = dir)) +
  annotate("rect", xmin = -0.01, xmax = 0.01, ymin = -Inf, ymax = Inf, fill = "grey88") +
  geom_col(width = 0.6) +
  geom_vline(xintercept = 0, colour = "grey40") +
  geom_text(aes(label = sprintf("%+.3f", delta), hjust = ifelse(delta > 0, -0.2, 1.2)),
            size = 5, colour = INK) +
  scale_fill_manual(values = c("more segregated" = BAD, "less segregated" = ACCENT), name = NULL) +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  labs(x = "Change in the segregation index, 2026",
       y = "How disadvantage was counted",
       title = "The Whitehawk redraw's effect on segregation flipped as the measure improved",
       subtitle = "Grey band: within measurement noise. The case for the redraw rests on journeys and Longhill's intake.") +
  theme_slide(15) + theme(legend.position = "top")
```

# The options

## How to read an option

::: {.columns}
::: {.column width="50%"}
**Compare in the same year.** The cohort falls by a fifth by 2035. Read an option against doing nothing *in that year*.

**A cap that is never reached does nothing.** Cutting Longhill to 150 changes only the count of empty places.
:::
::: {.column width="50%"}
**Small segregation differences are not findings.** Anything under about 0.01 is within what re-measuring moved it.

**Some results lean on held assumptions.** Journeys are measured directly; segregation after a move assumes a school keeps its draw on disadvantaged families at a new site.
:::
:::

29 options, each run for 2026, 2030 and 2035 on every measure at once. [How they were chosen, and seven rules for reading them.](https://adamdennett.github.io/bh-school-system/model-guide.html#sec-scenarios)

## Eight options, four priorities

```{r fig-options}
#| fig-width: 12.5
#| fig-height: 5.8
show <- c("today", "whitehawk", "central", "elm", "comart", "package_e", "package_f", "close_lh")
short <- c(today = "Today", whitehawk = "Whitehawk back to Longhill",
           central = "Longhill 150, central schools trimmed",
           elm = "Longhill at Elm Grove (150), map redrawn",
           comart = "CoMArt re-opened (not available)",
           package_e = "Package without a move",
           package_f = "Elm Grove package",
           close_lh = "Longhill closed")
MEAS_O <- c(lh_intake = "Longhill's intake", dep_gap = "Deprived children's extra journey (min)",
            gorard = "Segregation index", displaced = "Children displaced",
            comm_gap = "Community schools' balance (£m)")
ob <- cop$runs |>
  dplyr::filter(year == 2030, id %in% show) |>
  dplyr::mutate(comm_gap = comm_gap / 1e6,
                dep_gap = ifelse(abs(dep_gap) < 0.05, 0, dep_gap),
                label = factor(unname(short[id]), rev(unname(short[show]))),
                elm = id %in% c("elm", "package_f")) |>
  tidyr::pivot_longer(names(MEAS_O), names_to = "m", values_to = "v") |>
  dplyr::mutate(m = factor(unname(MEAS_O[m]), MEAS_O))
ref <- ob |> dplyr::filter(id == "today") |> dplyr::select(m, ref = v)
ggplot(ob, aes(v, label)) +
  geom_vline(data = ref, aes(xintercept = ref), colour = "grey70") +
  geom_point(aes(colour = elm), size = 4.2, na.rm = TRUE) +
  facet_wrap(~ m, nrow = 1, scales = "free_x",
             labeller = labeller(m = label_wrap_gen(13))) +
  scale_colour_manual(values = c(`FALSE` = INK, `TRUE` = ACCENT), guide = "none") +
  scale_x_continuous(n.breaks = 3) +
  labs(x = NULL, y = NULL,
       title = "Each option helps one priority and costs another",
       subtitle = "2030, council priorities. Grey line: today. Blue: Longhill at the top of Elm Grove. Longhill closed has no intake to show.") +
  theme_slide(13) +
  theme(strip.text = element_text(face = "bold", size = 12, hjust = 0),
        panel.spacing.x = unit(3, "lines"))
```

"Package without a move": Whitehawk back, Longhill 150, central trims, no priority 6, Cardinal Newman 300, Brighton Aldridge 150. "Elm Grove package": the move, with Cardinal Newman 300, King's 150, Brighton Aldridge 150.

## Only a nearer school closes the access gap

```{r fig-gap}
#| fig-width: 11
#| fig-height: 5.2
gap <- cop$runs |>
  dplyr::filter(year == 2030, id %in% show) |>
  dplyr::mutate(dep_gap = ifelse(abs(dep_gap) < 0.05, 0, dep_gap),
                label = unname(short[as.character(id)]),
                elm = id %in% c("elm", "package_f"),
                label = forcats::fct_reorder(label, dep_gap))
ggplot(gap, aes(dep_gap, label)) +
  geom_col(aes(fill = elm), width = 0.62, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f min", dep_gap)), hjust = -0.15, size = 5, colour = INK) +
  scale_fill_manual(values = c(`FALSE` = "#9aa7b4", `TRUE` = ACCENT)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Extra minutes children from the most deprived neighbourhoods travel, each way", y = NULL,
       title = "How much further deprived children travel than everyone else",
       subtitle = "2030, council priorities. Blue: Longhill at the top of Elm Grove.") +
  theme_slide(15)
```

## Access or mixing? No option does both

```{r fig-tradeoff}
#| fig-width: 11.5
#| fig-height: 5.6
tr <- cop$runs |>
  dplyr::filter(year == 2030) |>
  dplyr::mutate(dep_gap = ifelse(abs(dep_gap) < 0.05, 0, dep_gap),
                shown = id %in% show,
                label = ifelse(shown, unname(short[id]), NA_character_),
                elm = id %in% c("elm", "package_f"))
ggplot(tr, aes(dep_gap, gorard)) +
  geom_point(data = tr |> dplyr::filter(!shown), colour = "grey78", size = 2.6) +
  geom_point(data = tr |> dplyr::filter(shown), aes(colour = elm), size = 4.6) +
  ggrepel::geom_text_repel(data = tr |> dplyr::filter(shown), aes(label = label, colour = elm),
                           size = 4.2, seed = 5, box.padding = 0.6, show.legend = FALSE) +
  scale_colour_manual(values = c(`FALSE` = INK, `TRUE` = ACCENT), guide = "none") +
  labs(x = "Deprived children's extra journey, minutes  (← better)",
       y = "Segregation index  (↓ better)",
       title = "A school near Whitehawk shortens journeys and concentrates disadvantage",
       subtitle = "2030, all 29 options (grey), the eight above labelled. Nothing reaches the bottom-left corner.") +
  theme_slide(15)
```

The evidence leans towards **access**: a more mixed intake is a weak lever on results, and the journey is made by every child, twice a day. Whether shorter journeys raise attendance is what the council's own records could settle.

# What this means for Longhill

## A smaller Longhill at the top of Elm Grove

```{r tbl-elm}
elm_rows <- c("today", "elm", "package_f")
elm_lab <- c(today = "Today", elm = "Elm Grove, 150 places, map redrawn",
             package_f = "Elm Grove package")
tibble::tibble(id = elm_rows) |>
  dplyr::mutate(
    Option = unname(elm_lab[id]),
    `Longhill intake, 2026 → 2030` = sprintf("%s → %s", fmt_n(vapply(id, co, 0, col = "lh_intake", y = 2026)),
                                             fmt_n(vapply(id, co, 0, col = "lh_intake"))),
    `Deprived children's extra journey` = sprintf("%.1f min", pmax(0, vapply(id, co, 0, col = "dep_gap"))),
    `Whitehawk's disadvantaged children travel` = sprintf("%.0f min", vapply(id, whe, 0, col = "minutes_disadvantaged")),
    `Segregation index` = sprintf("%.3f", vapply(id, co, 0, col = "gorard")),
    `Longhill's annual position` = gbp_m(vapply(id, co, 0, col = "lh_gap"))) |>
  dplyr::select(-id) |>
  knitr::kable(align = "lrrrrr")
```

::: {.columns}
::: {.column width="50%"}
**What it gains**

- The biggest cut in deprived children's journeys of any option.
- **`r cop$elm_moves$whitehawk_to_longhill` of the `r cop$elm_moves$whitehawk_zones`** Whitehawk neighbourhoods back in Longhill's catchment, as part of a redesign rather than a reversal.
- More children for Longhill.
:::
::: {.column width="50%"}
**What it costs**

- A site, **not costed** here. Longhill **still does not break even**.
- About **`r fmt_n(cop$elm_moves$children)`** children a year in neighbourhoods that change catchment; longer journeys for Rottingdean, Saltdean and Woodingdean.
- Intakes more concentrated --- and that figure assumes families follow a **new map** as closely as today's.
:::
:::

# Recommendations

## What to do

1. **Lead with a smaller Longhill at the top of Elm Grove**, with catchments redrawn round it --- and test it properly, costs stated.
2. **If the site cannot be had, return Whitehawk to Longhill's catchment** on its own.
3. **Set Longhill at 150** on either site, and budget for the children it will get, not for 150.
4. **Ask Cardinal Newman (300), King's (150) and Brighton Aldridge (150)** to take fewer, and every school to share the fall in proportion by 2030.
5. **Trim Stringer, Varndean and Blatchington Mill only if Longhill stays at Ovingdean and only with a catchment redraw.**
6. **Publish where priority 6's places went.**
7. **Ask the inequalities review to settle the measure** --- journeys or mixing --- and whether journey length affects attendance.

## Next steps

- **Cost the Elm Grove site**, and re-run the move on the council's own records before deciding.
- **Draw the Elm Grove catchment with Rottingdean, Saltdean and Woodingdean in the room**, before rather than after.
- **Check the disadvantaged-intake calibration against a second year** of Year 7 offers.
- **Release six small tables** the council already holds --- applicants' home areas, full preferences, allocations and the criterion each place was given under --- which would turn most of the ranges here into estimates.
- **Link journeys to attendance** in the council's records, for the inequalities review.

## What this rests on

- A model calibrated to **published catchment totals**, not individual applications.
- Catchment attachment carried to **redrawn maps** at the strength families follow the maps they have --- the assumption every redraw result depends on.
- Finances as a **five-year steady state**; site and capital costs **not included**.
- The academies and faith schools modelled as **responding to demand**, not setting their own numbers.
- **The direction of each finding is more reliable than its size.**

[Full document](https://adamdennett.github.io/bh-school-system/) · [The model, stage by stage](https://adamdennett.github.io/bh-school-system/model-guide.html) · [Simulator](https://adamdennett-bh-school-system.share.connect.posit.cloud/) · [Code and data](https://github.com/adamdennett/bh-school-system)