---
title: "Brighton & Hove secondary schools"
subtitle: "What the evidence says, how it was built, and what to do about Longhill"
author: "Adam Dennett, UCL Centre for Advanced Spatial Analysis"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    width: 1280
    height: 720
    margin: 0.06
    smaller: true
    scrollable: true
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
source(here::here("R", "00_core.R"))
knitr::opts_chunk$set(dev = "ragg_png", dpi = 150)

fmt_n   <- function(x, d = 0) formatC(round(x, d), big.mark = ",", format = "f", digits = d)
fmt_pct <- function(x, d = 0) paste0(formatC(x, format = "f", digits = d), "%")
gbp_m   <- function(x) sprintf("%s£%.1fm", ifelse(x < 0, "−", ""), abs(x) / 1e6)

inp <- readRDS(here::here("app", "data", "sim_inputs.rds"))
cop <- bh_data("council_options.rds")
whx <- bh_data("whitehawk_explained.rds")
acc <- bh_data("accessibility.rds")

co  <- function(id, col, y = 2030) cop$runs[[col]][cop$runs$id == id & cop$runs$year == y]
whe <- function(id, col, y = 2026)
  cop$whitehawk_elm[[col]][cop$whitehawk_elm$id == id & cop$whitehawk_elm$year == y]

city   <- inp$schools[inp$schools$city, ]
places <- sum(city$pan)
kids   <- function(y) inp$demand$cohort[inp$demand$year == y]
no30   <- acc$lsoa[acc$lsoa$places_30 < 1, ]

INK <- "#1f3b57"; MUTED <- "#6b7785"; ACCENT <- "#2a78d6"
theme_slide <- function(base = 16)
  ggplot2::theme_minimal(base) +
  ggplot2::theme(panel.grid.major.y = ggplot2::element_blank(),
                 panel.grid.minor = ggplot2::element_blank(),
                 plot.title = ggplot2::element_text(face = "bold", colour = INK),
                 plot.subtitle = ggplot2::element_text(colour = MUTED),
                 axis.text = ggplot2::element_text(colour = INK))
theme_map <- function(base = 14)
  ggplot2::theme_void(base) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", colour = INK),
                 plot.subtitle = ggplot2::element_text(colour = MUTED),
                 legend.position = "right", strip.text = ggplot2::element_text(face = "bold", size = base))

library(ggplot2)
library(patchwork)

oi    <- bh_data("open_inputs.rds")
lsoa  <- bh_data("lsoa.geojson") |> sf::st_transform(4326)
catch <- bh_data("catchments_current.geojson") |> sf::st_transform(4326)
sch   <- schools_sf(oi$schools) |> dplyr::filter(name != "Peacehaven Community School")
rc    <- bh_data("reception_cohort.rds")
sfin  <- bh_data("school_finance.rds")
os    <- bh_data("open_scenarios.rds")
brt   <- bh_data("brightopia.rds")
mt    <- bh_data("model_terms.rds")
fm    <- bh_data("flow_map.rds")
elm_pt <- sf::st_as_sf(data.frame(e = oi$elm_grove$easting, n = oi$elm_grove$northing),
                       coords = c("e", "n"), crs = 27700) |> sf::st_transform(4326)
CITY_X <- c(-0.25, -0.01); CITY_Y <- c(50.80, 50.885)
```

## Before anything else

::: {.callout-warning}
**Beta.** Every number in these slides is **modelled** from published data, not a record of what happened or a forecast to plan on. The model has not been fully validated.
:::

- Built from **published data only** --- no pupil-level records.
- The full working, every source and every caveat: [The Brighton Secondary School System: a strategic view](https://adamdennett.github.io/bh-school-system/)
- Try any option yourself: [the policy simulator](https://adamdennett-bh-school-system.share.connect.posit.cloud/)

# The situation

## A system built for more children than it has

::: {.columns}
::: {.column width="55%"}
- The city publishes **`r fmt_n(places)`** Year 7 places across its ten secondary schools.
- The modelled cohort is **`r fmt_n(kids(2026))`** in 2026 and **`r fmt_n(kids(2035))`** by 2035 --- those children are already in primary school.
- By 2035 that is about **`r fmt_n(co("today", "empty", 2035))` empty Year 7 places a year**.
- Nine of the ten schools have been spending their reserves. The money is **a city problem, not only a Longhill problem**.
:::
::: {.column width="45%"}
**Longhill**

- Takes about **`r fmt_n(co("today", "lh_intake", 2026))`** children in 2026 and **`r fmt_n(co("today", "lh_intake"))`** in 2030, on 210 places.
- About **`r gbp_m(-co("today", "lh_gap"))` a year** short of break-even in 2030.
- Sits at the far edge of the city, at Ovingdean.
:::
:::

## Ten schools, six catchments, and where the children are

```{r fig-system-map}
#| fig-width: 11
#| fig-height: 6
kids <- oi$zones |> dplyr::group_by(lsoa) |> dplyr::summarise(children = sum(Oi), .groups = "drop")
kid_pts <- lsoa |> dplyr::inner_join(kids, by = c("lsoa21cd" = "lsoa")) |>
  sf::st_point_on_surface() |> suppressWarnings()
ggplot() +
  geom_sf(data = catch, aes(fill = catchment), alpha = 0.22, colour = "grey35", linewidth = 0.35) +
  geom_sf(data = kid_pts, aes(size = children), colour = "#1f4e79", alpha = 0.35) +
  geom_sf(data = sch, shape = 21, fill = "white", colour = "black", size = 3.4, stroke = 1.1) +
  ggrepel::geom_text_repel(data = sch, aes(lon, lat, label = short_sch(name)), size = 4,
                           colour = INK, seed = 2, box.padding = 0.5, min.segment.length = 0) +
  annotate("point", x = COMART$lon, y = COMART$lat, colour = "#b2182b", size = 3, shape = 4, stroke = 1.6) +
  annotate("text", x = COMART$lon, y = COMART$lat - 0.004, label = "CoMArt (closed 2005)",
           colour = "#b2182b", size = 3.4) +
  scale_fill_manual(values = CATCH_COLOURS, labels = CATCH_LABELS, name = "Catchment") +
  scale_size_area(max_size = 7, name = "Children per\nneighbourhood") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = "Most children live in the centre and west; Longhill sits at the far eastern edge",
       subtitle = "Two catchments hold a pair of schools; four hold one. King's and Cardinal Newman admit city-wide.") +
  theme_map()
```

## The cohort is already falling

```{r fig-cohort}
#| fig-width: 11
#| fig-height: 5.4
y7 <- rc$secondary_city; recp <- rc$reception_city; proj <- rc$projection
ggplot() +
  geom_ribbon(data = proj, aes(entry_year, ymin = lo, ymax = hi), fill = "#b2182b", alpha = 0.18) +
  geom_line(data = recp, aes(year, rec_offers), colour = "#2c7fb8", linewidth = 1.1) +
  geom_line(data = y7, aes(year, y7_offers), colour = INK, linewidth = 1.3) +
  geom_line(data = proj, aes(entry_year, central), colour = "#b2182b", linewidth = 1.2, linetype = "22") +
  annotate("text", x = min(recp$year) + 0.3, y = max(recp$rec_offers) * 1.03, label = "Reception offers",
           colour = "#2c7fb8", hjust = 0, size = 4.4) +
  annotate("text", x = min(y7$year) + 0.3, y = min(y7$y7_offers) * 0.93, label = "Year 7 offers",
           colour = INK, hjust = 0, size = 4.4, fontface = "bold") +
  annotate("text", x = max(proj$entry_year) - 0.2, y = max(proj$hi) * 1.06,
           label = "Projected from children\nalready in primary school", colour = "#b2182b", hjust = 1, size = 4) +
  scale_y_continuous(labels = scales::label_comma(), limits = c(0, NA)) +
  labs(x = NULL, y = "Offers",
       title = "Reception offers turned down first; Year 7 follows seven years later",
       subtitle = sprintf("Year 7 demand projected to about %s by %s, from children already counted",
                          fmt_n(proj$central[which.max(proj$entry_year)]), max(proj$entry_year))) +
  theme_slide(15)
```

## Who can reach a school within 30 minutes

```{r fig-access}
#| fig-width: 11.5
#| fig-height: 5.6
acc_sf <- lsoa |>
  dplyr::inner_join(acc$lsoa |> dplyr::select(lsoa, places_30, places_30_elm), by = c("lsoa21cd" = "lsoa")) |>
  tidyr::pivot_longer(c(places_30, places_30_elm), names_to = "site", values_to = "places") |>
  dplyr::mutate(site = factor(ifelse(site == "places_30", "Longhill at Ovingdean (today)", "Longhill at the top of Elm Grove"),
                              levels = c("Longhill at Ovingdean (today)", "Longhill at the top of Elm Grove")))
elm_lab <- sf::st_sf(site = factor("Longhill at the top of Elm Grove", levels = levels(acc_sf$site)),
                     geometry = sf::st_geometry(elm_pt))
lh_lab <- sch |> dplyr::filter(grepl("Longhill", name)) |>
  dplyr::mutate(site = factor("Longhill at Ovingdean (today)", levels = levels(acc_sf$site)))
ggplot() +
  geom_sf(data = acc_sf, aes(fill = places), colour = NA) +
  geom_sf(data = catch, fill = NA, colour = "white", linewidth = 0.25) +
  geom_sf(data = lh_lab, shape = 21, fill = "white", colour = "black", size = 3.4, stroke = 1.2) +
  geom_sf(data = elm_lab, shape = 21, fill = "white", colour = "black", size = 3.4, stroke = 1.2) +
  facet_wrap(~ site) +
  scale_fill_viridis_c(option = "viridis", direction = -1, labels = scales::label_comma(),
                       name = "School places\nwithin 30 min") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = sprintf("%s neighbourhoods can reach no school place within 30 minutes by walk and bus", nrow(no30)),
       subtitle = "Darker is more places in reach; yellow is few or none. The white dot is Longhill.") +
  theme_map(14)
```

## Poor access and child poverty in the same places

```{r fig-bivariate}
#| fig-width: 11
#| fig-height: 6
BIV <- c("1-1" = "#3b4994", "1-2" = "#8c62aa", "1-3" = "#be64ac",
         "2-1" = "#5698b9", "2-2" = "#a5add3", "2-3" = "#dfb0d6",
         "3-1" = "#5ac8c8", "3-2" = "#ace4e4", "3-3" = "#e8e8e8")
biv <- lsoa |> dplyr::inner_join(acc$bivariate |> dplyr::select(lsoa, biv_key), by = c("lsoa21cd" = "lsoa"))
main <- ggplot() +
  geom_sf(data = biv, aes(fill = biv_key), colour = "white", linewidth = 0.05) +
  geom_sf(data = catch, fill = NA, colour = "grey15", linewidth = 0.4) +
  geom_sf(data = sch, shape = 21, fill = "white", colour = "black", size = 2.4) +
  scale_fill_manual(values = BIV, guide = "none") +
  coord_sf(xlim = CITY_X, ylim = CITY_Y) +
  labs(title = "Dark blue: the least reachable school places and the most child poverty, together",
       subtitle = sprintf("Accessibility and child poverty correlate at %.2f across neighbourhoods", acc$acc_dep_spearman)) +
  theme_map(14)
key <- expand.grid(a = 1:3, d = 1:3) |>
  dplyr::mutate(k = paste0(a, "-", d)) |>
  ggplot(aes(d, a, fill = k)) + geom_tile(colour = "white") +
  scale_fill_manual(values = BIV, guide = "none") +
  labs(x = "Less deprived →", y = "More reachable →") + coord_fixed() +
  theme_minimal(10) + theme(axis.text = element_blank(), panel.grid = element_blank())
main + inset_element(key, left = 0.0, bottom = 0.02, right = 0.2, top = 0.36)
```

## Nine of the ten schools are spending their reserves

```{r fig-reserves}
#| fig-width: 12
#| fig-height: 5.6
fex <- sfin$exposure |> dplyr::mutate(short = short_sch(name))
ord <- fex |> dplyr::arrange(reserve_pct) |> dplyr::pull(name)
rp <- sfin$panel |> dplyr::mutate(school = factor(short_sch(name), short_sch(ord)), neg = pmin(reserve_pct, 0))
ggplot(rp, aes(year, reserve_pct)) +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +
  geom_ribbon(aes(ymin = neg, ymax = 0), fill = "#d03b3b", alpha = 0.3) +
  geom_line(linewidth = 0.9, colour = INK) + geom_point(size = 1.2, colour = INK) +
  facet_wrap(~ school, ncol = 5) +
  scale_x_continuous(breaks = sort(unique(rp$year)),
                     labels = function(x) paste0(substr(x, 3, 4), "/", substr(x + 1, 3, 4))) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = NULL, y = "Reserve as a share of income",
       title = "Revenue reserves, worst first; red shading is a school in deficit",
       subtitle = "DfE school income and expenditure returns") +
  theme_slide(12) + theme(axis.text.x = element_text(size = 7))
```

## Small schools get more per pupil and still run deficits

```{r fig-size-finance}
#| fig-width: 11
#| fig-height: 5.2
sz <- os$size_fin |> dplyr::mutate(size_band = factor(size_band, levels = size_band))
pa <- ggplot(sz, aes(size_band, income_pp)) + geom_col(fill = "#2166ac", width = 0.7) +
  scale_y_continuous(labels = scales::label_dollar(prefix = "£", accuracy = 1)) +
  labs(x = "Pupils on roll", y = NULL, subtitle = "Income per pupil") + theme_slide(14)
pb <- ggplot(sz, aes(size_band, pct_deficit)) + geom_col(fill = "#b2182b", width = 0.7) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(x = "Pupils on roll", y = NULL, subtitle = "Share of school-years in deficit") + theme_slide(14)
(pa | pb) + plot_annotation(title = "A shrinking school is a financial problem, not just a smaller one",
                            subtitle = "Every secondary school in England, by size",
                            theme = theme_slide(15))
```

## What the evidence says

- **This is not mainly a school quality problem.** Most of what separates schools' headline results arrives with the children; the part a school can reach runs mostly through **attendance**.
- **Families choose on the headline score**, which largely describes the intake --- so choice is self-fulfilling.
- **Admissions criteria only bind at some schools.** Several schools offer a place to every first-preference applicant.
- **Physical access is unequal, in the worst direction.** `r nrow(no30)` neighbourhoods can reach **no** school place within 30 minutes by walking and bus, and access and child poverty go together.
- Children from the most deprived neighbourhoods travel **`r sprintf("%.1f", co("today", "dep_gap", 2026))` minutes further** each way than everyone else, under the map in force.

# How the model works

## From published data to options

```{mermaid}
flowchart LR
  A["Where children live<br/>census, ONS small areas"] --> M
  B["Journey times<br/>routed walk and bus"] --> M
  C["What families ask for<br/>16 years of factsheets,<br/>catchment preferences"] --> M
  D["Catchment maps"] --> M
  M["Spatial interaction model<br/>M0 to M5"] --> R["Admission numbers and<br/>the council's priorities"]
  R --> S["Refused children<br/>re-offered by 2nd preference"]
  S --> O["Intakes, journeys,<br/>money, social mix"]
  O --> P["The simulator,<br/>and the options scored"]
```

## A spatial interaction model

::: {.columns}
::: {.column width="48%"}
$$T_{ij} = A_i\, O_i\, W_j\, c_{ij}^{-\beta}\, e^{\gamma\, \text{in catchment}}\, C_j^{\delta}$$

- $T_{ij}$ --- children from neighbourhood $i$ choosing school $j$
- $O_i$ --- children living in $i$
- $W_j$ --- how attractive school $j$ is
- $c_{ij}$ --- walk-and-bus minutes; $\beta$ how fast demand falls with them
- $\gamma$ --- how strongly families follow their catchment
- $C_j$ --- rival schools nearby
- $A_i$ --- makes every neighbourhood place all its children
:::
::: {.column width="52%"}
```{mermaid}
flowchart TB
  N["Each neighbourhood's children"] --> Sh["Shared across schools by<br/>distance, attractiveness,<br/>catchment, competition"]
  Sh --> Cap{"School full?"}
  Cap -- "no" --> In["Admitted"]
  Cap -- "yes" --> Pr["Council priorities decide<br/>who is admitted"]
  Pr --> Re["The rest re-offered at their<br/>next choice with room"]
  Re --> Cap
```
:::
:::

## The map alone: every school equally attractive

```{r fig-brightopia}
#| fig-width: 11
#| fig-height: 5.6
bt <- brt$geog_only_at_ref |> dplyr::mutate(fills = modelled >= pan2026)
ggplot(bt, aes(pan2026, modelled)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = "31") +
  geom_point(aes(colour = fills), size = 4.5) +
  ggrepel::geom_text_repel(aes(label = short_sch(name)), size = 4.2, seed = 1, colour = INK) +
  scale_colour_manual(values = c(`TRUE` = "#2166ac", `FALSE` = "#b2182b"),
                      labels = c(`TRUE` = "Geography alone would fill it", `FALSE` = "It would not"), name = NULL) +
  scale_x_continuous(limits = c(0, NA)) + scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Admission number, 2026", y = "Children drawn on geography alone",
       title = "Distance alone explains much of who goes where",
       subtitle = "Every school identical, only journey time differs. Dashed line: exactly full.") +
  theme_slide(15) + theme(legend.position = "top")
```

## Adding one thing at a time

```{r fig-ladder}
#| fig-width: 12
#| fig-height: 5.4
lad <- mt$ladder |> dplyr::mutate(model = factor(model, model))
LAD_LAB <- c(M0 = "Distance\nonly", M1 = "+ what families\nask for", M2 = "+ school\ncapacity",
             M3 = "+ one\ncatchment term", M4 = "+ competition", M5 = "Fitted to each\ncatchment")
p1 <- ggplot(lad, aes(model, mae_free)) +
  geom_col(fill = INK, width = 0.62) +
  geom_text(aes(label = round(mae_free)), vjust = -0.4, size = 4.6, colour = INK) +
  scale_x_discrete(labels = LAD_LAB) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Average error, children per school",
       subtitle = "Error on the schools not rationing places") + theme_slide(13)
gam <- tibble::tibble(catchment = names(mt$calibrated$gamma), gamma = unname(mt$calibrated$gamma)) |>
  dplyr::mutate(label = unname(CATCH_LABELS[CATCH_FROM_MODEL[catchment]]),
                label = forcats::fct_reorder(label, gamma))
p2 <- ggplot(gam, aes(gamma, label)) +
  geom_col(fill = ACCENT, width = 0.62) +
  geom_text(aes(label = sprintf("%.1f", gamma)), hjust = -0.25, size = 4.6, colour = INK) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Fitted catchment term", y = NULL,
       subtitle = "How strongly each catchment follows its map") + theme_slide(13)
(p1 | p2) + plot_annotation(title = "Capacity does most of the work; families follow their catchment strongly",
                            theme = theme_slide(15))
```

## Longhill's modelled catchment, as the model gets more realistic

```{r fig-flows}
#| fig-width: 12.5
#| fig-height: 5
FM_LABELS <- c(M0 = "Distance only", M1 = "With what families ask for", M5 = "Full model")
FM_COL <- c(Bus = "#2166ac", Walk = "#d95f02")
fl <- fm$net |> dplyr::filter(name == "Longhill High School") |>
  dplyr::mutate(model = factor(model, names(FM_LABELS), FM_LABELS))
fl_bb <- sf::st_bbox(fl)
fm_sch <- schools_sf() |> dplyr::filter(name %in% unique(fm$net$name))
ggplot() +
  geom_sf(data = catch, fill = NA, colour = "grey80", linewidth = 0.3) +
  geom_sf(data = fl |> dplyr::arrange(flow), aes(linewidth = flow, colour = leg_mode),
          alpha = 0.75, lineend = "round") +
  geom_sf(data = fm_sch, colour = "grey45", size = 1.3) +
  geom_sf(data = fm_sch |> dplyr::filter(name == "Longhill High School"),
          colour = "black", fill = "white", shape = 21, size = 3.2, stroke = 1.1) +
  facet_wrap(~ model, nrow = 1) +
  scale_linewidth_continuous(range = c(0.15, 2.8), guide = "none") +
  scale_colour_manual(values = FM_COL, name = NULL) +
  coord_sf(xlim = c(fl_bb["xmin"], fl_bb["xmax"]), ylim = c(fl_bb["ymin"], fl_bb["ymax"])) +
  labs(title = "Routed over the real network, Longhill's draw shrinks back towards its own catchment",
       subtitle = "Line width is modelled children on each street or bus segment. White dot: Longhill.") +
  theme_map(13) + theme(legend.position = "top")
```

## Checked, and corrected, along the way

The segregation results changed as the measure was tested against published figures.

```{r tbl-measures}
whx$measures |>
  dplyr::filter(year == 2026) |>
  dplyr::select(measure, map, gorard) |>
  tidyr::pivot_wider(names_from = map, values_from = gorard) |>
  dplyr::transmute(`How disadvantaged pupils were counted` = measure,
                   `Index today` = sprintf("%.3f", today),
                   `Whitehawk back to Longhill` = sprintf("%.3f", whitehawk)) |>
  knitr::kable(align = "lrr")
```

- Scored **by neighbourhood**, the schools came out in the wrong order against DfE's published shares.
- The published shares **lag the policy** by up to five years, so the model is now fitted to the **first Year 7 offers** under today's map and rules (council factsheet, 2026).
- The Whitehawk redraw's effect on segregation turned out **too small to call**. Its case rests on journeys and Longhill, not on segregation.

# The options

## Three priorities, scored at once

```{r tbl-options}
show <- c("today", "whitehawk", "central", "elm", "comart", "package_e", "package_f", "close_lh")
short <- c(today = "Today", whitehawk = "Whitehawk back to Longhill",
           central = "Longhill 150, central schools trimmed",
           elm = "Longhill at Elm Grove (150), map redrawn",
           comart = "CoMArt re-opened (not available)",
           package_e = "Package without a move",
           package_f = "Elm Grove package",
           close_lh = "Longhill closed")
cop$runs |>
  dplyr::filter(year == 2030, id %in% show) |>
  dplyr::mutate(id = factor(id, show)) |>
  dplyr::arrange(id) |>
  dplyr::transmute(Option = unname(short[as.character(id)]),
                   `Longhill intake` = ifelse(is.na(lh_intake), "---", fmt_n(lh_intake)),
                   `Deprived children's extra journey` = sprintf("%.1f min", ifelse(abs(dep_gap) < 0.05, 0, dep_gap)),
                   `Segregation index` = sprintf("%.3f", gorard),
                   Displaced = fmt_n(displaced),
                   `Community schools' balance` = gbp_m(comm_gap)) |>
  knitr::kable(align = "lrrrrr")
```

2030, council priorities. "Package without a move": Whitehawk back, Longhill 150, central trims, no priority 6, Cardinal Newman 300, Brighton Aldridge 150. "Elm Grove package": Longhill at Elm Grove at 150 with the map redrawn, Cardinal Newman 300, King's 150, Brighton Aldridge 150.

## Only a nearer school closes the access gap

```{r fig-gap}
#| fig-width: 11
#| fig-height: 5.2
gap <- cop$runs |>
  dplyr::filter(year == 2030, id %in% show) |>
  dplyr::mutate(dep_gap = ifelse(abs(dep_gap) < 0.05, 0, dep_gap),
                label = unname(short[as.character(id)]),
                elm = id %in% c("elm", "package_f"),
                label = forcats::fct_reorder(label, dep_gap))
ggplot2::ggplot(gap, ggplot2::aes(dep_gap, label)) +
  ggplot2::geom_col(ggplot2::aes(fill = elm), width = 0.62, show.legend = FALSE) +
  ggplot2::geom_text(ggplot2::aes(label = sprintf("%.1f min", dep_gap)),
                     hjust = -0.15, size = 5, colour = INK) +
  ggplot2::scale_fill_manual(values = c(`FALSE` = "#9aa7b4", `TRUE` = ACCENT)) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0, 0.18))) +
  ggplot2::labs(x = "Extra minutes children from the most deprived neighbourhoods travel, each way", y = NULL,
                title = "How much further deprived children travel than everyone else",
                subtitle = "2030, council priorities. Blue: Longhill at the top of Elm Grove.") +
  theme_slide(15)
```

# What this means for Longhill

## A smaller Longhill at the top of Elm Grove

::: {.columns}
::: {.column width="50%"}
**What it does** (150 places, catchments redrawn)

- Longhill takes **`r fmt_n(co("elm", "lh_intake", 2026))`** in 2026 and **`r fmt_n(co("elm", "lh_intake"))`** in 2030 --- against `r fmt_n(co("today", "lh_intake", 2026))` and `r fmt_n(co("today", "lh_intake"))` today.
- Deprived children's extra journey falls from **`r sprintf("%.1f", co("today", "dep_gap"))`** to **`r sprintf("%.1f", max(0, co("elm", "dep_gap")))`** minutes.
- Whitehawk's disadvantaged children travel **`r sprintf("%.0f", whe("elm", "minutes_disadvantaged"))`** minutes instead of **`r sprintf("%.0f", whe("today", "minutes_disadvantaged"))`**.
- The map drawn round it puts **`r cop$elm_moves$whitehawk_to_longhill` of the `r cop$elm_moves$whitehawk_zones`** Whitehawk neighbourhoods back in Longhill's catchment --- as part of a redesign, not a reversal of the 2024 change.
- With Cardinal Newman, King's and Brighton Aldridge taking fewer: **`r fmt_n(co("package_f", "lh_intake", 2026))`** in 2026 and **`r fmt_n(co("package_f", "lh_intake"))`** in 2030.
:::
::: {.column width="50%"}
**What it costs**

- A site --- **not costed** in this work.
- Intakes more concentrated: segregation index **`r sprintf("%.3f", co("elm", "gorard"))`** against `r sprintf("%.3f", co("today", "gorard"))`.
- About **`r fmt_n(cop$elm_moves$children)`** children a year in neighbourhoods that change catchment.
- Longer journeys for Rottingdean, Saltdean and Woodingdean, and more children leaving the city.
- Longhill **still does not break even** --- about `r gbp_m(-co("elm", "lh_gap"))` a year short in 2030.
- It needs families to follow a **new map** as closely as the one they have --- the assumption every redraw result rests on.
:::
:::

## Access or mixing? The review has to choose

- A school near Whitehawk **shortens disadvantaged children's journeys** and **concentrates them in one school**. No option does both.
- The evidence here leans towards **access**: a more mixed intake is a weak lever on results, and the journey is made by every child, twice a day.
- What this work cannot show is whether **shorter journeys raise attendance** --- the lever that matters most. The council's own records can.
- Re-opening CoMArt would have given much of the access gain; it is off the table. Closing Longhill is the cheapest system and has the longest journeys in the east.

# Recommendations

## What to do

1. **Lead with a smaller Longhill at the top of Elm Grove**, with catchments redrawn round it --- and test it properly, costs stated.
2. **If the site cannot be had, return Whitehawk to Longhill's catchment** on its own.
3. **Set Longhill at 150** on either site, and budget for the children it will get, not for 150.
4. **Ask Cardinal Newman (300), King's (150) and Brighton Aldridge (150)** to take fewer, and every school to share the fall in proportion by 2030.
5. **Trim Stringer, Varndean and Blatchington Mill only if Longhill stays at Ovingdean and only with a catchment redraw.**
6. **Publish where priority 6's places went.**
7. **Ask the inequalities review to settle the measure** --- journeys or mixing --- and whether journey length affects attendance.

## Next steps

- **Cost the Elm Grove site**, and re-run the move on the council's own records before deciding.
- **Draw the Elm Grove catchment with Rottingdean, Saltdean and Woodingdean in the room**, before rather than after.
- **Check the disadvantaged-intake calibration against a second year** of Year 7 offers.
- **Release six small tables** the council already holds --- applicants' home areas, full preferences, allocations and the criterion each place was given under --- which would turn most of the ranges here into estimates.
- **Link journeys to attendance** in the council's records, for the inequalities review.

## What this rests on

- A model calibrated to **published catchment totals**, not individual applications.
- Catchment attachment carried to **redrawn maps** at the strength families follow the maps they have --- the assumption every redraw result depends on.
- Finances as a **five-year steady state**; site and capital costs **not included**.
- The academies and faith schools modelled as **responding to demand**, not setting their own numbers.
- **The direction of each finding is more reliable than its size.**

[Full document](https://adamdennett.github.io/bh-school-system/) · [Simulator](https://adamdennett-bh-school-system.share.connect.posit.cloud/) · [Code and data](https://github.com/adamdennett/bh-school-system)
